### 2021-07-26 manu-draft.Rmd
skeleton manu draft

---
title: "Title placeholder"
output: 
  bookdown::word_document2:
    reference_docx: my_styles.docx
    number_sections: false
always_allow_html: true
# bibliography: refs.bib
author: "Marcus W. Beck (mbeck@tbep.org), Gary E. Raulerson (graulerson@tbep.org), Maya C. Burke (mburke@tbep.org), Joe Whalen (jwhalen@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T. Sherwood (esherwood@tbep.org)"
urlcolor: blue
csl: marine-pollution-bulletin.csl
link-citations: true
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F,
  fig.process = function(x) {
  x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
  if (file.rename(x, x2)) x2 else x
  })

# # extract bib entries from online
# bib_scrp('manu_draft.Rmd', 'refs.bib')
```

`r paste('Last manuscript build', Sys.time())`

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('manu_draft.Rmd')
```

# Abstract

*Key words*:

# Introduction

https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit

# Methods

# Results

# Conclusions

# Acknowledgments

# Figures

# Tables

# References

### 2021-08-04 manu-draft.Rmd
very short outline, building refs

@@
 ---
-title: "Title placeholder"
+title: "Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida"
 output: 
@@
 always_allow_html: true
-# bibliography: refs.bib
+bibliography: refs.bib
 author: "Marcus W.
@@
 
-https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit
+* Nutrient management paradigm for southwest Florida estuaries, historical progress
+* Threats/challenges to protecting water quality: nutrient inputs, climate change
+* Wastewater byproducts from mining are a global threat to the quality of surface and groundwater sources worldwide [@Hudson11].  Phosphate mining industry in Florida, then/now
+* Piney Point
+* Goals/objectives
 
@@
 
+* Historical timeline: https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit
+* Initial response, data collection
+* Synthesis and Analysis
+
 # Results
 
-# Conclusions
+* 2021 timeline
+* Historical context/baseline condition
+* effluent characteristics
+* water-column water quality changes
+* red tide impacts
+* Seagrass/macroalgae
+
+# Discussion
+
+* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16]
+* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.)
+* Potential long-term impacts TBD
+* Current challenges in TB/southwest FL - OTB, seagrass loss, red tide, climate change
+* Risk of decline (IRL ex.)/regression of past progress
 

### 2021-08-04 manu-draft.Rmd
adding some lit

@@
 * Threats/challenges to protecting water quality: nutrient inputs, climate change
-* Wastewater byproducts from mining are a global threat to the quality of surface and groundwater sources worldwide [@Hudson11].  Phosphate mining industry in Florida, then/now
+* Wastewater byproducts from mining are a global threat to the quality of surface and groundwater sources worldwide [@Hudson11]. 
+     * [@Tayibi09] reviews environmental impact of phosphogypsum production, including contaminantion of groundwater/leaching to coastal waters.  
+     * [@Burnett01] evaluate radiochemistry of phosphogypsum at Piney Point to evaluate potential for contamination of groundwater sources [pdf](https://www.osti.gov/etdeweb/servlets/purl/20243518)
+    * Citations from [@Perez16] to track down: Santos-Cubatao estuarine system (Brazil; Sanders et al., 2013), in the Lebanese coast (El Samad et al., 2014) and in the Gulf of Gabes (Tunisia; El Zrelli et al., 2016)
+    * Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
+* Phosphate mining industry in Florida, then/now
 * Piney Point
@@
 
-* Historical timeline: https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit
+* Historical timeline: https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit, https://thebradentontimes.com/piney-point-a-retrospective-p6328-158.htm
 * Initial response, data collection

### 2021-08-11 manu-draft.Rmd
initial abstract

@@
 
+From March 30th to April 9th, 2021, an estimated 215 million gallons of legacy phosphate mining wastewater from the Piney Point phosphogypsum stacks were released into Tampa Bay (Florida, USA).
+Ammonium concentrations in the wastewater were measured in excess of 200 mg/L and an estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
+In response to these events, a coordinated environmental monitoring effort consisting of multiple government, university, and private sector partners began to assess conditions of surface waters around Piney Point to understand conditions prior to, during, and after the wastewater release.
+These efforts included sampling of surface water chemistry, surveys of algal community response, assessments of wastewater contaminants, and biological surveys of seagrass, macroalgae, benthic, and nekton communities.
+Over a five month period, environmental conditions in Tampa Bay deviated from historical baseline data.
+Water quality monitoring data showed an initial phytoplankton response in April dominated by centric diatoms in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
+Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.
+Beginning in May and continuing through July, bloom concentrations of the red tide organism (*Karenia brevis*) were observed in Tampa Bay.
+Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish from coastal areas of the bay.
+Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.
+Ongoing monitoring will continue to assess potential long-term impacts to Tampa Bay to ensure decades of progress in environmental restoration are not reversed by acute, short-term events.
+
 *Key words*:
@@
 * Current challenges in TB/southwest FL - OTB, seagrass loss, red tide, climate change
-* Risk of decline (IRL ex.)/regression of past progress
+* Risk of decline (IRL ex.), regression of past progress
 

### 2021-08-11 manu-draft.Rmd
shorten abstract

--- 2021-08-04 manu-draft.Rmd
@@
 
+From March 30th to April 9th, 2021, an estimated 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
+An estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
+An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
+Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.
+Blooms of *Karenia brevis* were observed beginning in May and continuing through July.
+Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish.
+Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.
+
 *Key words*:

### 2021-08-11 manu-draft.Rmd
goals/objectives paragraph

@@
 
-From March 30th to April 9th, 2021, an estimated 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
+From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
 An estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
@@
 
-*Key words*:
+*Key words*: nitrogen, phosphate mining, Tampa Bay, wastewater, water quality
 
@@
 * Goals/objectives
+
+This paper provides an initial assessment of environmental conditions in Tampa Bay following the release of 215 million gallons of legacy phosphate mining wastewater over a ten-day period.
+The goal is to describe the results of monitoring data collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.
+We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
+A timeline of events in 2021 is also provided, which is supported by the results from response-based monitoring of conditions in and around Piney Point.
+The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.
 

### 2021-08-11 manu-draft.Rmd
intro draft

@@
 From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated ~205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
+An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
 An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
@@
 
-* Nutrient management paradigm for southwest Florida estuaries, historical progress
-* Threats/challenges to protecting water quality: nutrient inputs, climate change
-* Wastewater byproducts from mining are a global threat to the quality of surface and groundwater sources worldwide [@Hudson11]. 
-     * [@Tayibi09] reviews environmental impact of phosphogypsum production, including contaminantion of groundwater/leaching to coastal waters.  
-     * [@Burnett01] evaluate radiochemistry of phosphogypsum at Piney Point to evaluate potential for contamination of groundwater sources [pdf](https://www.osti.gov/etdeweb/servlets/purl/20243518)
-    * Citations from [@Perez16] to track down: Santos-Cubatao estuarine system (Brazil; Sanders et al., 2013), in the Lebanese coast (El Samad et al., 2014) and in the Gulf of Gabes (Tunisia; El Zrelli et al., 2016)
-    * Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
-* Phosphate mining industry in Florida, then/now
-* Piney Point
-* Goals/objectives
+Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from stormwater and wastewater sources.
+The effects of nitrogen from source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
+Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18g].
+Tampa Bay is the largest estuary in Florida located in a heavily urbanized watershed of nearly 3 million individuals.
+Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads into major bay segments [@Janicki96;@Greening16].
+Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay following the release of 215 million gallons of legacy phosphate mining wastewater over a ten-day period.
-The goal is to describe the results of monitoring data collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.
+Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.
+Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute to nutrient inputs, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
+Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].
+Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
+Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay[@Ohara17].
+The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
+
+Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources worldwide [@Tayibi09;@Hudson11].
+Fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
+The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
+Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
+The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18].
+There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
+Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18].
+
+The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].
+By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
+Currently, fourteen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.
+The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
+Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).
+Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
+Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].
+Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+
+This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
+The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.
 We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
-A timeline of events in 2021 is also provided, which is supported by the results from response-based monitoring of conditions in and around Piney Point.
+A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Piney Point.
 The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.
+We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage existing partnerships among local resource management institutions.
 
@@
 
-* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16]
-* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.)
+* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
+* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
 * Potential long-term impacts TBD

### 2021-08-13 manu-draft.Rmd
map fig added

@@
   })
+
+library(here)
 
@@
 Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.
-Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute to nutrient inputs, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
+Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute nutrients to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
 Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].
@@
 Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].
-Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
 
@@
 
-* 2021 timeline
-* Historical context/baseline condition
-* effluent characteristics
-* water-column water quality changes
-* red tide impacts
-* Seagrass/macroalgae
+* Figure \@ref(fig:timeline) timeline of events
+* Fig 2 map, response and baseline data, w/ areas inset?
+* Fig 3 Boxplots by week with +/-1sd for monthly long-term, or value median of long-term and monthly for seasonal eval
+* Fig 4 seagrass and macroalgae, correlation matrix
+* Fig 5 Red tide historic and current, include fish kill reports
+
+* Table 1 Data collected
+* Table 2 Stack/effluent characteristics, load estimates
 
@@
 
+```{r timeline, fig.cap = 'Areas of interest for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type in response to wastewater discharge from Piney Point.
+Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
+Inset shows location of Tampa Bay on the Gulf coast of Florida, USA.'}
+knitr::include_graphics(here('figs/map.jpeg'))
+```
+
+```{r map, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting Mrach 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
+knitr::include_graphics(here('figs/timeline.jpeg'))
+```
+
 # Tables

### 2021-08-13 R/figures.R

@@ -1,0 +1,185 @@
+
+# setup -------------------------------------------------------------------
+
+library(tidyverse)
+library(lubridate)
+library(sf)
+library(here)
+library(ggmap)
+library(ggspatial)
+library(patchwork)
+library(units)
+library(grid)
+library(scales)
+library(USAboundaries)
+
+data(rsallpts)
+data(ppseg)
+data(segmask)
+
+# map ---------------------------------------------------------------------
+
+# piney point loc
+pineypoint <- tibble(
+  lon = -82.52469352586753, 
+  lat = 27.629819505234703
+  ) %>% 
+  st_as_sf(coords = c('lon', 'lat'), crs = 4326)
+
+# segments
+areas <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  ) %>% 
+  st_intersection(segmask) %>% 
+  st_cast('POLYGON') %>% 
+  mutate(
+    acres = st_area(.), 
+    acres = set_units(acres, 'acres'), 
+    acres = as.numeric(acres)
+  ) %>% 
+  dplyr::filter(acres > 1e4) %>% 
+  st_buffer(dist = set_units(0.0005, degree)) %>% 
+  st_buffer(dist = set_units(-0.0005, degree)) 
+
+cols <- c("#E16A86", "#50A315", "#009ADE")
+names(cols) <- levels(areas$area)
+
+##
+# map
+
+buffdist <- 0.01
+northloc <- 'tr' 
+scaleloc <- 'bl'
+
+# layer extent as bbox plus buffer
+dat_ext <- areas %>% 
+  st_bbox %>% 
+  st_as_sfc %>% 
+  st_buffer(dist = set_units(buffdist, degree)) %>%
+  st_bbox %>% 
+  unname
+
+# reference data for ggsn, MUST have geometry named column
+ggsnref <- areas %>% 
+  st_bbox %>% 
+  st_as_sfc %>%
+  st_buffer(dist = set_units(buffdist / 2, degree)) %>% 
+  st_as_sf %>%
+  st_cast('POINT') %>% 
+  rename(geometry = x)
+
+# stamen base map
+bsmap1 <- get_stamenmap(bbox = dat_ext, maptype = 'toner-background', zoom = 10)
+
+# change opacity of basemap
[…]

### 2021-08-13 manu-draft.Rmd
wq map figure added

@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, fourteen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.
+Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.
 The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
@@
 The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.
-We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage existing partnerships among local resource management institutions.
+We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage resources from existing partnerships among local resource management institutions.
 
@@
 
+* Figure \@ref(fig:map) map
 * Figure \@ref(fig:timeline) timeline of events
-* Fig 2 map, response and baseline data, w/ areas inset?
+* Figure \@ref(fig:wqmap) map of water quality observations
 * Fig 3 Boxplots by week with +/-1sd for monthly long-term, or value median of long-term and monthly for seasonal eval
@@
 * Potential long-term impacts TBD
-* Current challenges in TB/southwest FL - OTB, seagrass loss, red tide, climate change
+* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
 * Risk of decline (IRL ex.), regression of past progress
@@
 
-```{r timeline, fig.cap = 'Areas of interest for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type in response to wastewater discharge from Piney Point.
+```{r map, fig.cap = 'Areas of interest (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to wastewater discharge from Piney Point.
 Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
@@
 
-```{r map, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting Mrach 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
+```{r timeline, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting Mrach 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
 knitr::include_graphics(here('figs/timeline.jpeg'))
+```
+
+```{r wqmap, fig.cap = 'Sampled water quality data for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+Values outside of the normal range (above for total nitrogen and chlorophyll, below for secchi) are outlined in black and those in normal range are outlined in light grey.
+Color ramps and point sizes show relative values (reversed for Secchi).
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
+Chlorophyll observations were truncated to a maximum of 30 ug/L to remove some outliers.'}
+knitr::include_graphics(here('figs/wqmap.jpeg'))
 ```

### 2021-08-13 R/figures.R

@@ -16,6 +16,8 @@ library(USAboundaries)
 data(rsallpts)
 data(ppseg)
 data(segmask)
+data(rswqdat)
+data(rsstatloc)
 
 # map ---------------------------------------------------------------------
 
@@ -182,4 +184,132 @@ pout <- p1 + p2
 
 jpeg(here('figs/map.jpeg'), height = 4.2, width = 9, family = 'serif', units = 'in', res = 500)
 print(pout)
+dev.off()
+
+# nutrients, chloropyll map -----------------------------------------------
+
+# nonbay stations
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+# colors 
+vrscols <- rev(RColorBrewer::brewer.pal(n = 9, name = 'RdYlBu'))
+vrscols[7:9] <- 'red'
+
+# ggplot base size
+bssz <- 13
+
+thm <-  theme(
+    axis.title = element_blank(),
+    legend.position = 'bottom', 
+    legend.box= 'vertical'
+  ) 
+
+# combine water quality data with locations
+wqdat <- rswqdat %>% 
+  filter(var %in% c('tn', 'chla', 'secchi')) %>% 
+  filter(!station %in% nonbay) %>% 
+  inner_join(rsstatloc, ., by = 'station') %>% 
+  mutate(
+    lng = st_coordinates(.)[, 1], 
+    lat = st_coordinates(.)[, 2],
+    dategrp = floor_date(date, unit = 'week'), 
+    inrng = case_when(
+      inrng %in% c('below', 'in range') & var %in% c('chla', 'tn') ~ 'in range', 
+      inrng %in% c('above') & var %in% c('chla', 'tn') ~ 'above', 
+      inrng %in% c('above', 'in range') & var %in% c('secchi') ~ 'in range',
+      inrng %in% c('below') & var %in% c('secchi') ~ 'below'
+    )
+  ) %>% 
+  select(station, date, dategrp, var, val, lat, lng, inrng)
+
+# reference data for ggsn, MUST have geometry named column
+dat_ext <- wqdat %>% 
+  st_bbox %>% 
+  st_as_sfc %>% 
+  st_buffer(dist = set_units(0.01, degree)) %>%
+  st_bbox %>% 
+  unname
+
+# stamen base map
[…]

### 2021-08-13 manu-draft.Rmd
wqtrnds figure added

@@
 * Figure \@ref(fig:wqmap) map of water quality observations
-* Fig 3 Boxplots by week with +/-1sd for monthly long-term, or value median of long-term and monthly for seasonal eval
-* Fig 4 seagrass and macroalgae, correlation matrix
-* Fig 5 Red tide historic and current, include fish kill reports
+* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
+* Fig x seagrass and macroalgae, correlation matrix
+* Fig x Red tide historic and current, include fish kill reports
 
@@
 
+```{r wqtrnds, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+Observations are aggregated by week and within assessment areas shown in Figure \\@ref(fig:map)a.
+Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area.'}
+knitr::include_graphics(here('figs/wqtrnds.jpeg'))
+```
+
 # Tables

### 2021-08-13 R/figures.R

@@ -14,6 +14,8 @@ library(scales)
 library(USAboundaries)
 
 data(rsallpts)
+data(bswqdat)
+data(bsstatloc)
 data(ppseg)
 data(segmask)
 data(rswqdat)
@@ -311,5 +313,128 @@ p3 <- bsmap +
 p <- p1 + p2 + p3
 
 jpeg(here('figs/wqmap.jpeg'), height = 7, width = 12, units = 'in', res = 500, family = 'serif')
+print(p)
+dev.off()
+
+# weekly plots ------------------------------------------------------------
+
+# water quality plot fun
+wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet = FALSE, ttl, ylb){
+  
+  nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+  
+  ##
+  # wq data
+  
+  # monitoring data
+  rswqtmp <- rswqdat %>% 
+    filter(var == vr) %>% 
+    filter(!station %in% nonbay) %>% 
+    inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
+    mutate(
+      date = floor_date(date, unit = 'week'), 
+      mo = month(date), 
+      fillcl = factor(area, levels = levels(area), labels = cols), 
+      fillcl = as.character(fillcl)
+    ) 
+  
+  # baseline data
+  bswqtmp <- bswqdat %>% 
+    select(-source, -uni) %>% 
+    filter(var == vr) %>% 
+    filter(yr > 2005) %>% 
+    inner_join(bsstatloc, ., by = 'station') %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    group_by(mo, var, area) %>% 
+    summarise(   
+      avev = mean(val, na.rm = T), 
+      stdv = sd(val, na.rm = T), 
+      .groups = 'drop'
+    ) %>%
+    left_join(parms, by = 'var') %>% 
+    mutate(
+      avev = round(avev, sigdig), 
[…]

### 2021-08-16 manu-draft.Rmd
pca and correlation figure added

@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.
+Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to prevent impacts to the environment.
 The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
@@
 * Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
+* Fig x pca of all wq variables, by area? 
 * Fig x seagrass and macroalgae, correlation matrix
@@
 
+```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge form Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).
+All variables were standardized to zero mean and unit variance prior to PCA.
+Variables with log-normal distribution were log-transformed prior to analysis.
+Pearson correlations indicate the linear strength of assocation between pairs of variables.
+** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
+knitr::include_graphics(here('figs/pcacors.jpeg'))
+```
+
 # Tables

### 2021-08-16 R/figures.R

@@ -12,6 +12,15 @@ library(units)
 library(grid)
 library(scales)
 library(USAboundaries)
+library(ggord)
+library(lubridate)
+library(patchwork)
+box::use(
+  scales = scales[muted], 
+  units = units[set_units], 
+  vegan = vegan[decostand], 
+  FactoMineR = FactoMineR[PCA]
+)
 
 data(rsallpts)
 data(bswqdat)
@@ -20,6 +29,8 @@ data(ppseg)
 data(segmask)
 data(rswqdat)
 data(rsstatloc)
+
+source(here('R/funcs.R'))
 
 # map ---------------------------------------------------------------------
 
@@ -438,3 +449,184 @@ p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Week
 jpeg(here('figs/wqtrnds.jpeg'), height = 6, width = 8, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
+
+# PCA and correlations ----------------------------------------------------
+
+vrs <- c('chla', 'dosat', 'nh34', 'ph', 'secchi', 'temp', 'tn', 'no23', 'tp', 'sal')
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+cols <- c("#E16A86", "#50A315", "#009ADE")
+names(cols) <- c('Area 1', 'Area 2', 'Area 3')
+
+ppsegbf <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+rswqsub <- rswqdat %>% 
+  filter(var %in% vrs) %>% 
+  filter(source == 'fldep') %>%
+  filter(!station %in% nonbay) %>% 
+  inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+  st_intersection(ppsegbf) %>% 
+  st_set_geometry(NULL) %>% 
+  select(date, var, val, station, area) %>% 
+  mutate(
+    var = case_when(
+      var == 'chla' ~ 'Chl-a', 
[…]

### 2021-08-16 R/funcs.R

@@ -1,0 +1,12 @@
+# function for formatting p-values in tables
+p_ast <- function(x){
+  
+  sig_cats <- c('**', '*', '')
+  sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
+  out <- as.character(out)
+  
+  return(out)
+  
+}

### 2021-08-16 manu-draft.Rmd
seagrass macroalgae figures added

@@
 library(here)
+
+data(rstrnpts)
 
@@
 * Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
-* Fig x pca of all wq variables, by area? 
-* Fig x seagrass and macroalgae, correlation matrix
+* Figure \@ref(fig:pcacors) pca and correlations of all wq variables, by area 
+* Figure \@ref(fig:trnex) seagrass and macroalgae example transect
+* Figure \@ref(fig:trnfrq) seagrass, macroalgae frequency occurrence estimates
+* Figure \@ref(fig:allpcacors) seagrass, macralgae, wq pca and correlation matrix
 * Fig x Red tide historic and current, include fish kill reports
@@
 * Table 2 Stack/effluent characteristics, load estimates
+* Table 3 Tukey multiple comparisons of water quality by week/area, with median/min/max
 
@@
 
-```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge form Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).
+```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).
 All variables were standardized to zero mean and unit variance prior to PCA.
@@
 
+```{r trnex, fig.cap = 'Results for (a) seagrass and (b) macroalgae rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.
+Sample dates in 2021 are shown in rows with transect meter results shown in columns.
+Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Lyngbya sp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva sp.*) increasing in July.
+Abundances are Braun-Blanquet coverage estimates.'}
+knitr::include_graphics(here('figs/trnex.jpeg'))
+```
+
+```{r trnfrq, fig.cap = 'Frquency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.
+Estimates are grouped by sample months in 2021.
+Frequency occurrences are absolute for each taxa based on presence/absence.'}
+knitr::include_graphics(here('figs/trnfrq.jpeg'))
+```
+
+```{r allpcacors, fig.cap = 'Principal components analysis (PCA) and correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
+Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
+All variables were standardized to zero mean and unit variance prior to PCA.
+Variables with log-normal distribution were log-transformed prior to analysis.
+Thick black lines in the correlation matrix separate water quality, macroalgae, and seagrass variables.
+Pearson correlations indicate the linear strength of assocation between pairs of variables.
+** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
+knitr::include_graphics(here('figs/allpcacors.jpeg'))
+```
+
 # Tables

### 2021-08-16 R/figures.R

@@ -29,6 +29,7 @@ data(ppseg)
 data(segmask)
 data(rswqdat)
 data(rsstatloc)
+data(rstrndat)
 
 source(here('R/funcs.R'))
 
@@ -329,101 +330,6 @@ dev.off()
 
 # weekly plots ------------------------------------------------------------
 
-# water quality plot fun
-wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet = FALSE, ttl, ylb){
-  
-  nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
-  
-  ##
-  # wq data
-  
-  # monitoring data
-  rswqtmp <- rswqdat %>% 
-    filter(var == vr) %>% 
-    filter(!station %in% nonbay) %>% 
-    inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
-    st_intersection(ppseg) %>% 
-    st_set_geometry(NULL) %>% 
-    select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
-    mutate(
-      date = floor_date(date, unit = 'week'), 
-      mo = month(date), 
-      fillcl = factor(area, levels = levels(area), labels = cols), 
-      fillcl = as.character(fillcl)
-    ) 
-  
-  # baseline data
-  bswqtmp <- bswqdat %>% 
-    select(-source, -uni) %>% 
-    filter(var == vr) %>% 
-    filter(yr > 2005) %>% 
-    inner_join(bsstatloc, ., by = 'station') %>% 
-    st_intersection(ppseg) %>% 
-    st_set_geometry(NULL) %>% 
-    group_by(mo, var, area) %>% 
-    summarise(   
-      avev = mean(val, na.rm = T), 
-      stdv = sd(val, na.rm = T), 
-      .groups = 'drop'
-    ) %>%
-    left_join(parms, by = 'var') %>% 
-    mutate(
-      avev = round(avev, sigdig), 
-      stdv = round(stdv, sigdig), 
-      minv = avev - stdv, 
-      minv = pmax(0, minv),
-      maxv = avev + stdv,
-      lbunis = gsub('^.*\\s(\\(.*\\))$', '\\1', lbs), 
[…]

### 2021-08-16 R/funcs.R

@@ -10,3 +10,302 @@ p_ast <- function(x){
   return(out)
   
 }
+
+# water quality plot fun
+wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet = FALSE, ttl, ylb){
+  
+  nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+  
+  ##
+  # wq data
+  
+  # monitoring data
+  rswqtmp <- rswqdat %>% 
+    filter(var == vr) %>% 
+    filter(!station %in% nonbay) %>% 
+    inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    select(-qual, -bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
+    mutate(
+      date = floor_date(date, unit = 'week'), 
+      mo = month(date), 
+      fillcl = factor(area, levels = levels(area), labels = cols), 
+      fillcl = as.character(fillcl)
+    ) 
+  
+  # baseline data
+  bswqtmp <- bswqdat %>% 
+    select(-source, -uni) %>% 
+    filter(var == vr) %>% 
+    filter(yr > 2005) %>% 
+    inner_join(bsstatloc, ., by = 'station') %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    group_by(mo, var, area) %>% 
+    summarise(   
+      avev = mean(val, na.rm = T), 
+      stdv = sd(val, na.rm = T), 
+      .groups = 'drop'
+    ) %>%
+    left_join(parms, by = 'var') %>% 
+    mutate(
+      avev = round(avev, sigdig), 
+      stdv = round(stdv, sigdig), 
+      minv = avev - stdv, 
+      minv = pmax(0, minv),
+      maxv = avev + stdv,
+      lbunis = gsub('^.*\\s(\\(.*\\))$', '\\1', lbs), 
+      lbunis = gsub('pH', '', lbunis), 
+      datestr= paste0('2021-', mo, '-01'), 
+      datestr = ymd(datestr), 
+      dateend = ceiling_date(datestr, unit = 'month')
+    )
+  
+  # boxplot colors
[…]

### 2021-08-16 manu-draft.Rmd
redtide and fish kills figure added

@@
 * Figure \@ref(fig:allpcacors) seagrass, macralgae, wq pca and correlation matrix
-* Fig x Red tide historic and current, include fish kill reports
+* Figure \@ref(fig:redtide) red tide historic and current, fish kill reports historic and current
 
@@
 
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) and number of fish kill reports for the contiguous record showing cell concentrations (a) by year and (b) by week in 2021 and reported fish kills by city (Tampa, St.
+Petersburg) (c) by year and (d) by week in 2021.
+Red tide concentrations show minimum, tenth percentile, median, 90th percentile, and maximum for each year or week for middle and lower Tampa Bay.
+*K.
+brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Wildlife Research Insitute Fish Kill Database, attributed to *K.
+brevis* (https://public.myfwc.com/FWRI/FishKillReport/).'}
+knitr::include_graphics('figs/redtide.jpeg')
+```
+
 # Tables

### 2021-08-16 R/figures.R

@@ -871,3 +871,211 @@ jpeg(here('figs/allpcacors.jpeg'), height = 4, width = 10, units = 'in', res = 5
 print(p)
 dev.off()
 
+# red tide and fish kills -------------------------------------------------
+
+# data from https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx
+# requested hillsborough, pinellas, manatee 1/1/95 to 8/13/21
+fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
+  select(
+    date = textBox6, 
+    county = tEMPDataTextBox,
+    city = cOUNTYDataTextBox, 
+    waterbody = lOCATIONDataTextBox,
+    species = textBox18
+  ) %>% 
+  mutate(
+    date = mdy(date),
+    yr = year(date),
+    week = floor_date(date, unit = 'week'), 
+    week = factor(format(week, '%b %d')), 
+    week = factor(week, levels = as.character(unique(week))), 
+    county = case_when(
+      county %in% c('Pinellas ', 'Pinellas', 'pinellas') ~ 'Pinellas', 
+      county %in% c('Hillsborough', 'Hillsborough ') ~ 'Hillsborough', 
+      T ~ county
+    ),
+    city = gsub('\\s+$', '', city),
+    city = gsub('^St\\s', 'St. ', city),
+    city = case_when(
+      city %in% c('St. pete Beach', 'St. Pete Beach', 'St. Petersburg Beach') ~ 'St. Petersburg Beach', 
+      city %in% 'Tierra Ceia' ~ 'Terra Ceia', 
+      city %in% 'dunedin' ~ 'Dunedin',
+      T ~ city
+    )
+  )
+
+
+# levels for week, starts on first of week from jan through july
+weeklv <- seq.Date(from = as.Date('2021-01-01'), to = Sys.Date(), by = 'days') %>% 
+  lubridate::floor_date(unit = 'week') %>% 
+  unique %>% 
+  tibble(
+    dt = ., 
+    yr = year(dt), 
+    mo = month(dt), 
+    lb = format(dt, '%b %d')
+  ) %>%
+  filter(yr > 2020) %>% 
+  filter(mo <= 8) %>% 
+  pull(lb)
+
+# MTB subset
+toplo <- kbrdat %>%
+  .[tbseg[tbseg$bay_segment %in% c('MTB', 'LTB'), ], ] %>%
[…]

### 2021-08-16 manu-draft.Rmd
water quality summary table added

@@
 data(rstrnpts)
+load(file = here('tables/wqsumtab.RData'))
 
@@
 
-* Table 1 Data collected
+* Table \@ref(tab:wqsumtab) water quality data collected and summarized
 * Table 2 Stack/effluent characteristics, load estimates
@@
 
+```{r wqsumtab}
+# table stuff
+cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point.
+Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
+Summaries are median, maximum, and minimum values.
+Total observations (N obs.) and values in rnage, above, or below normal ranges are also shown.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
+tab <- wqsumtab
+# table
+knitr::kable(tab, booktabs = T, caption = cap.val)
+```
+
 # References

### 2021-08-16 R/tables.R

@@ -1,0 +1,80 @@
+
+# setup -------------------------------------------------------------------
+
+library(tidyverse)
+library(sf)
+box::use(
+  units = units[set_units], 
+)
+
+data(rswqdat)
+data(ppseg)
+data(rsstatloc)
+data(parms)
+
+# water quality summary table ---------------------------------------------
+
+# segments
+ppsegbf <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+vrs <- c('chla', 'dosat', 'nh34', 'ph', 'secchi', 'temp', 'tn', 'no23', 'tp', 'sal')
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+sigs <- parms %>% 
+  select(lbs, sigdig)
+
+totab <- rswqdat %>% 
+  left_join(rsstatloc, ., by = c('station', 'source')) %>% 
+  filter(var %in% vrs) %>% 
+  filter(!station %in% nonbay) %>% 
+  select(station, date, lbs, val, inrng) %>% 
+  st_intersection(ppsegbf) %>% 
+  st_set_geometry(NULL)
+
+totab1 <- totab %>% 
+  group_by(area, lbs, inrng) %>% 
+  summarise(
+    cnt = n(), 
+    .groups = 'drop'
+  ) %>% 
+  complete(area, lbs, inrng, fill = list(cnt = 0)) %>% 
+  spread(inrng, cnt)
+
+totab2 <- totab %>% 
+  group_by(area, lbs) %>% 
+  left_join(sigs, by = 'lbs') %>% 
+  summarise(
+    medv = round(median(val, na.rm = T), unique(sigdig)), 
+    minv = round(min(val, na.rm = T), unique(sigdig)), 
+    maxv = round(max(val, na.rm = T), unique(sigdig)), 
+    cnt = n(), 
+    .groups = 'drop'
+  ) %>% 
+  mutate(
+    minv = paste0(' (', minv, ', '), 
+    maxv = paste0(maxv, ')')
+  ) %>% 
[…]

### 2021-08-17 manu-draft.Rmd
stack table added

--- 2021-08-16 manu-draft.Rmd
@@
 data(rstrnpts)
+load(file = here('tables/wqsumtab.RData'))
+load(file = here('tables/stktab.RData'))
 
@@
 
-* Table 1 Data collected
-* Table 2 Stack/effluent characteristics, load estimates
+* Table \@ref(tab:stktab) Stack characteristics
+* Table \@ref(tab:wqsumtab) water quality data collected and summarized
 * Table 3 Tukey multiple comparisons of water quality by week/area, with median/min/max
+* Table 4 Seagrass/macroalgae freq. occurrence summaries
 
@@
 *K.
-brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Wildlife Research Insitute Fish Kill Database, attributed to *K.
+brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
 brevis* (https://public.myfwc.com/FWRI/FishKillReport/).'}
@@
 
+```{r stktab}
+cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point on April 9th, 2021 for relevant water quality variables compared to normal annual means (min, max) for concentrations in lower Tampa Bay.
+Normal means are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by Hillsborough County Environmental Protection Commission.'
+knitr::kable(stktab, booktabs = T, caption = cap.val)
+```
+
+```{r wqsumtab}
+cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point.
+Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
+Summaries are median, maximum, and minimum values.
+Total observations (N obs.) and values in rnage, above, or below normal ranges are also shown.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
+knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
+```
+
 # References

### 2021-08-17 R/tables.R

@@ -1,6 +1,8 @@
 
 # setup -------------------------------------------------------------------
 
+library(here)
+library(tbeptools)
 library(tidyverse)
 library(sf)
 box::use(
@@ -11,6 +13,71 @@ data(rswqdat)
 data(ppseg)
 data(rsstatloc)
 data(parms)
+data(bsstatloc)
+data(bswqdat)
+
+# stack characteristics ---------------------------------------------------
+
+# ltb seg
+ltb <- tbseg %>% 
+  filter(bay_segment %in% 'LTB')
+
+# relevant stack measurements form ngs-s, FDEP measurements
+# original data from https://docs.google.com/spreadsheets/d/1AgRPWUv8TLVcLf6KmDCSIcTOVZmQL7ij/edit#gid=1685851796
+stkraw <- read.csv(here('data-raw/stack_samples.csv')) %>% 
+  select(var = Parameter, val = NGS.S, Units) %>% 
+  filter(var %in% c('NO2NO3-N', 'Ammonia N', 'Total N', 'Total-P', 'Ortho Phosphate as P', 'Dissolved Oxygen', 'pH*')) %>% 
+  filter(!(var %in% 'Dissolved Oxygen' & Units %in% 'mg/L')) %>% 
+  mutate(
+    var = case_when(
+      var == 'NO2NO3-N' ~ 'no23', 
+      var == 'Ammonia N' ~ 'nh34', 
+      var == 'Total N' ~ 'tn', 
+      var == 'Total-P' ~ 'tp', 
+      var == 'Ortho Phosphate as P' ~ 'orthop', 
+      var == 'Dissolved Oxygen' ~ 'dosat', 
+      var == 'pH*' ~ 'ph'
+    )
+  ) %>% 
+  left_join(parms, by = 'var') %>% 
+  select(var, lbs, val)
+
+bssum <- bswqdat %>% 
+  filter(yr > 2005 & yr < 2021) %>% 
+  filter(source == 'epchc') %>% 
+  filter(var %in% unique(stkraw$var)) %>% 
+  inner_join(bsstatloc, ., by = c('station')) %>% 
+  .[ltb, ] %>% 
+  st_set_geometry(NULL) %>% 
+  group_by(yr, var) %>% 
+  summarise(
+    val = median(val, na.rm = T),
+    .groups = 'drop'
+  ) %>%
+  left_join(parms, by = 'var') %>% 
+  group_by(var) %>% 
+  summarise(
+    minv = round(min(val), unique(sigdig)),
[…]

### 2021-08-17 manu-draft.Rmd
water quality comparison table added

@@
 load(file = here('tables/stktab.RData'))
+load(file = here('tables/wqcmptab.RData'))
 
@@
 
+For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
+An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
+
 # Results
@@
 * Table \@ref(tab:wqsumtab) water quality data collected and summarized
-* Table 3 Tukey multiple comparisons of water quality by week/area, with median/min/max
+* Table \@ref(tab:wqcmptab) Multiple comparisons of water quality by month/area, with median/min/max
 * Table 4 Seagrass/macroalgae freq. occurrence summaries
@@
 cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point on April 9th, 2021 for relevant water quality variables compared to normal annual means (min, max) for concentrations in lower Tampa Bay.
-Normal means are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by Hillsborough County Environmental Protection Commission.'
+Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by Hillsborough County Environmental Protection Commission.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)
@@
 
+```{r wqcmps}
+cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
+Multiple comparisons with Mann-Whitney U tests were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
+Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different at $\\alpha$ = 0.05.
+Probability values were adjusted using the Bonferroni method in @Holm79.'
+knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
+```
+
 # References

### 2021-08-17 R/tables.R

@@ -5,8 +5,11 @@ library(here)
 library(tbeptools)
 library(tidyverse)
 library(sf)
+library(lubridate)
 box::use(
   units = units[set_units], 
+  forcats = forcats[fct_drop], 
+  multcompView = multcompView[multcompLetters]
 )
 
 data(rswqdat)
@@ -59,7 +62,7 @@ bssum <- bswqdat %>%
   summarise(
     minv = round(min(val), unique(sigdig)),
     maxv = round(max(val), unique(sigdig)),
-    avev = round(mean(val), unique(sigdig)), 
+    avev = round(median(val), unique(sigdig)), 
     .groups = 'drop'
   ) %>% 
   mutate(
@@ -72,7 +75,7 @@ tab <- full_join(stkraw, bssum, by = 'var') %>%
   select(
     `Water quality variable` = lbs, 
     `Stack value` = val, 
-    `Normal mean (min, max)` = sumv
+    `Normal median (min, max)` = sumv
   )
 
 stktab <- tab
@@ -145,3 +148,105 @@ totab2 <- totab %>%
 
 wqsumtab <- totab2
 save(wqsumtab, file = here('tables/wqsumtab.RData'))
+
+# water quality comparison trnds ------------------------------------------
+
+vrs <- c('chla', 'secchi', 'tn')
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+ppsegbf <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+rswqsub <- rswqdat %>% 
+  filter(var %in% vrs) %>% 
+  filter(source == 'fldep') %>%
+  filter(!station %in% nonbay) %>% 
+  inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+  st_intersection(ppsegbf) %>% 
+  st_set_geometry(NULL) %>% 
+  left_join(parms, by = c('var', 'lbs')) %>% 
+  select(date, var, lbs, val, sigdig, station, area) %>% 
+  mutate(
[…]

### 2021-08-17 R/figures.R
spearman rank cor for all

@@ -15,6 +15,7 @@ library(USAboundaries)
 library(ggord)
 library(lubridate)
 library(patchwork)
+library(tbeptools)
 box::use(
   scales = scales[muted], 
   units = units[set_units], 
@@ -30,6 +31,9 @@ data(segmask)
 data(rswqdat)
 data(rsstatloc)
 data(rstrndat)
+data(parms)
+data(kbrdat)
+data(rstrnpts)
 
 source(here('R/funcs.R'))
 
@@ -179,7 +183,6 @@ lbs2 <- tibble(
   st_as_sf(coords = c('lon', 'lat'), crs = 4326)
 
 p2b <- ggplot() + 
-  geom_sf(data = continent, fill = 'grey', colour = 'grey') +
   geom_sf(data = states, fill = 'grey', colour = 'grey') +
   geom_sf(data = insetbb, fill = NA, color = 'blue', size = 1.5) +
   geom_sf_text(data = lbs1, aes(label = label), size = 5) + 
@@ -485,7 +488,7 @@ rswqtmp <- rswqsub %>%
         var1 <- out[[i, 'var1']]
         var2 <- out[[i, 'var2']]
         
-        tst <- cor.test(x[[var1]], x[[var2]])
+        tst <- cor.test(x[[var1]], x[[var2]], method = 'spearman')
         
         corv <- tst$estimate
         pval <- tst$p.value
@@ -701,7 +704,7 @@ trncrs <- crossing(var1 = names(trncors), var2 = names(trncors)) %>%
       vr2 <- trncors[[x$var2]]
       
       # pearson
-      pr_ts <- cor.test(vr1, vr2, method = 'pearson')
+      pr_ts <- cor.test(vr1, vr2, method = 'spearman')
       pr_cr <- round(pr_ts$estimate, 2)
       pr_pv <- p_ast(pr_ts$p.value)
       pr <- paste(pr_cr, pr_pv)
@@ -793,7 +796,7 @@ crs <- crossing(var1 = c(wqlab, mcrsel, savsel), var2 = c(wqlab, mcrsel, savsel)
       vr2 <- tocor[[x$var2]]
       
       # pearson
-      pr_ts <- try(cor.test(vr1, vr2, method = 'pearson'))
+      pr_ts <- try(cor.test(vr1, vr2, method = 'spearman'))
       pr_cr <- round(pr_ts$estimate, 2)
       pr_pv <- p_ast(pr_ts$p.value)
       pr <- paste(pr_cr, pr_pv)

### 2021-08-17 manu-draft.Rmd
kruskal test to wq comp table

@@
 Pearson correlations indicate the linear strength of assocation between pairs of variables.
-** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
 knitr::include_graphics(here('figs/allpcacors.jpeg'))
@@
 cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
-Multiple comparisons with Mann-Whitney U tests were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
-Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different at $\\alpha$ = 0.05.
-Probability values were adjusted using the Bonferroni method in @Holm79.'
+Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
+Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
 knitr::kable(wqcmptab, booktabs = T, caption = cap.val)

### 2021-08-17 R/funcs.R

@@ -1,8 +1,11 @@
 # function for formatting p-values in tables
-p_ast <- function(x){
+p_ast <- function(x, usens = F){
   
   sig_cats <- c('**', '*', '')
   sig_vals <- c(-Inf, 0.005, 0.05, Inf)
+  
+  if(usens)
+    sig_cats[3] <- 'ns'
   
   out <- cut(x, breaks = sig_vals, labels = sig_cats, right = FALSE)
   out <- as.character(out)

### 2021-08-17 R/tables.R

@@ -12,6 +12,8 @@ box::use(
   multcompView = multcompView[multcompLetters]
 )
 
+source(here('R/funcs.R'))
+
 data(rswqdat)
 data(ppseg)
 data(rsstatloc)
@@ -208,6 +210,9 @@ cmps <- rswqsub %>%
       # group membership based on multiple comparisons
       lets <- multcompLetters(vecs)$Letters
       
+      # kruskal test
+      tst <- kruskal.test(val ~ mo, x)
+      
       # standard summary stats
       sums <- group_by(x, mo) %>%
         summarise(
@@ -219,9 +224,12 @@ cmps <- rswqsub %>%
         ) %>% 
         mutate(
           minval = paste0(' (', minval, ', '),
-          maxval = paste0(maxval, ')')
+          maxval = paste0(maxval, ')'), 
+          krusk_pval = p_ast(tst$p.value),
+          krusk_chis = round(tst$statistic, 2)
         ) %>% 
-        unite('sumv', medval, minval, maxval, sep = '')
+        unite('sumv', medval, minval, maxval, sep = '') %>% 
+        unite('krusk', krusk_chis, krusk_pval, sep = '')
       
       data.frame(lets, sums, stringsAsFactors = FALSE)
       
@@ -237,11 +245,14 @@ cmps <- rswqsub %>%
   group_by(area) %>%  
   mutate(
     area = ifelse(duplicated(area), '', area), 
-    lbs = ifelse(duplicated(lbs), '', as.character(lbs))
+    lbs = ifelse(duplicated(lbs), '', as.character(lbs)), 
+    krusk = ifelse(duplicated(krusk), '', krusk)
   ) %>% 
+  ungroup() %>% 
   select(
     Area = area, 
     `Water quality variable` = lbs, 
+    `Chi-Sq.` = krusk,
     `Comp.` = lets, 
     Month = mo, 
     `N obs.` = length, 
@@ -250,3 +261,4 @@ cmps <- rswqsub %>%
 
 wqcmptab <- cmps
 save(wqcmptab, file = here('tables/wqcmptab.RData'))
+

### 2021-08-17 manu-draft.Rmd
seagrass and macroalgae fo tables

@@
 load(file = here('tables/wqcmptab.RData'))
+load(file = here('tables/savtab.RData'))
+load(file = here('tables/mcrtab.RData'))
 
@@
 
+Need to note that K. brevis cell data are from event-based sampling, i.e., not appropriate to combine with other data for stats analysis.
+
+All stats based on FLDEP monitoring, fixed sample design, etc.
+
 For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 * Table \@ref(tab:wqcmptab) Multiple comparisons of water quality by month/area, with median/min/max
-* Table 4 Seagrass/macroalgae freq. occurrence summaries
+* Table \@ref(tab:mcrtab) Macroalgae freq. occurrence summaries by month/area, with median/min/max
+* Table \@ref(tab:savtab) Seagrass freq. occurrence summaries by month/area, with median/min/max
+
+Seagrass: note decline of HW and correlations in the matrix, these make sense.
+
+Seagrass interp from DT:
+
+First off, Marcus please correct me if any confusion on my part about how to read the chart, but it’s my take that the asterisks reflect levels of significance, and that a box without an asterisk – I didn’t pay much attention to that.
+Also, a negative number would be a “cool” color, reflecting an inverse relationship, while a positive number (warm color) would reflect a positive correlation?
+
+If I have this correct, then what stood out to me was some of the following – seagrass distributions are highly inversely correlated between most of the species, which reflects seagrass being clonal organisms that tend not to grow interspersed with each other in a random manner.
+S.
+filiforme does not grow in shallow waters, because it’s blades can’t “lay down” during low tides, and it often is the deepest growing species in some of our cleaner areas, while H.
+wrightii can be both the shallowest and deepest growing species in areas like Hillsborough Bay.
+
+More relevant here is that there is no significant correlation between the abundance of cyanobacteria and either chlorophyll-a or TN or TP.
+To me, that makes sense, since macroalgae like cyanos seem to reflect an alternative nutrient destination, compared to phytoplankton.
+The strongest correlations out there are the well expected ones between TN and Chl-a, but also between TP and NH3/4, which suggests overall nutrient availability tracks together, and phytoplankton continue to be a major destination of nitrogen availability.
+
+But what about the positive correlation between cyanobacteria and S.
+filiforme?
+From my experience in Sarasota Bay, that makes sense, since the highest levels of cyanobacteria were found entangled within the manatee grass meadows in the fairly deeper waters farther offshore.
+Cyanos are strongly negatively correlated with red algae, since they occurred at different times, basically.
+
+Finally, keep in mind that “Lyngbya” life cycles are such that most of the biomass produced does not stay where it was produced.
+Those clumps of Lyngbya growing on the bottom get lifted up off the bottom by the oxygen bubbles you can see in their blooms, and they then are carried along the shoreline to decompose on whatever shoreline that they’ve been blown to.
+Our photos (and others) clearly show that there is MUCH more Lyngbya in our bays than what we’ve sampled along the bottom of the bay – that’s a bit different than “traditional” red algae biomass.
 
@@
 
+```{r mcrtab}
+cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
+Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
+knitr::kable(mcrtab, booktabs = T, caption = cap.val)
+```
+
+```{r savtab}
+cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
+Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
+knitr::kable(savtab, booktabs = T, caption = cap.val)
+```
+
 # References

### 2021-08-17 R/figures.R

@@ -598,6 +598,23 @@ areas <- ppseg %>%
     area = factor(area)
   )
 
+# # view sample effort by transect, area, month
+# smpeff <- rstrndat %>% 
+#   left_join(rstrnpts, ., by = 'station') %>% 
+#   select(-source, -type, -lng, -lat) %>% 
+#   st_intersection(areas) %>% 
+#   st_set_geometry(NULL) %>% 
+#   mutate(
+#     mo = month(date)
+#   ) %>% 
+#   group_by(station, area, mo) %>% 
+#   summarise(
+#     obs = (any(bb > 0)),
+#     .groups = 'drop'
+#   ) %>% 
+#   spread(mo, obs)
+# View(smpeff)
+
 # add area
 trnsum <- rstrndat %>%
   inner_join(rstrnpts, ., by = 'station') %>% 
@@ -689,7 +706,6 @@ trncors <- trnsum %>%
   spread(taxa, foest) %>% 
   # filter(area %in% 'Area 3') %>% 
   select(-station, -date)
-
 
 trncrs <- crossing(var1 = names(trncors), var2 = names(trncors)) %>% 
   filter(var1 != var2) %>% 

### 2021-08-17 R/tables.R

@@ -262,3 +262,161 @@ cmps <- rswqsub %>%
 wqcmptab <- cmps
 save(wqcmptab, file = here('tables/wqcmptab.RData'))
 
+# sav and mcr trends by month ---------------------------------------------
+
+mcrsel <- c("Red", "Green", "Cyanobacteria")
+savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
+
+# segments
+areas <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+# # view sample effort by transect, area, month
+# smpeff <- rstrndat %>%
+#   mutate(
+#     mo = month(date)
+#   ) %>%
+#   group_by(station, mo) %>%
+#   summarise(
+#     obs = (any(bb > 0)),
+#     .groups = 'drop'
+#   ) %>% 
+#   group_by(station) %>% 
+#   summarise(
+#     cnt = n(), 
+#     .groups = 'drop'
+#   ) %>% 
+#   filter(cnt >= 4) %>% 
+#   pull(station)
+
+# add area
+trnsf <- rstrndat %>%
+  inner_join(rstrnpts, ., by = 'station') %>% 
+  filter(taxa %in% c(mcrsel, savsel)) %>% 
+  # filter(station %in% smpeff) %>% 
+  st_intersection(areas) %>% 
+  st_set_geometry(NULL) %>% 
+  select(station, date, location, typ, taxa, bb, area)
+
+trnsum <- trnsf %>% 
+  mutate(
+    mo = month(date, label = T),
+    pa = bb > 0
+  ) %>%
+  dplyr::group_by(area, typ, mo, station, taxa) %>%
+  dplyr::summarise(
+    fo = sum(pa) / length(pa), 
+    .groups = 'drop'
+  ) %>% 
+  group_by(area, taxa) %>% 
+  nest %>% 
+  mutate(
+    ests = map(data, function(x){
+      
+      # pairwise comparisons with mann-whitney (wilcox)
+      grps <- unique(x$mo)
+      grps <- combn(grps, 2)
+      pval <- rep(NA, ncol(grps))
+      for(col in 1:ncol(grps)){
+        grp <- x$mo %in% grps[, col, drop = TRUE]
[…]

### 2021-08-19 manu-draft.Rmd
stack and effluent characteristics to table 1

@@
 ```{r stktab}
-cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point on April 9th, 2021 for relevant water quality variables compared to normal annual means (min, max) for concentrations in lower Tampa Bay.
-Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by Hillsborough County Environmental Protection Commission.'
+cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.
+Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.
+Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by the Environmental Protection Commission of Hillsborough County.
+Effluent concentrations were taken from two samples on April 6th and 13th at the outflow at Port Manatee.
+Averages were taken when two measured values were available from both effluent sample dates.
+Missing values were not measured in the effluent.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)

### 2021-08-19 R/tables.R

@@ -45,7 +45,22 @@ stkraw <- read.csv(here('data-raw/stack_samples.csv')) %>%
     )
   ) %>% 
   left_join(parms, by = 'var') %>% 
-  select(var, lbs, val)
+  select(var, lbs, stkval = val) %>% 
+  bind_rows(
+    tibble(
+      var = 'chla', 
+      lbs = c('Chl-a (ug/L)'), 
+      stkval = NA
+    )
+  )
+
+# effluent characteristics (end of pipe)
+effraw <- tibble(
+  var = c('no23', 'nh34', 'tn', 'tp', 'orthop', 'dosat', 'ph', 'chla'),
+  lbs = c("Nitrate/Nitrite (mg/L)", "NH3, NH4+ (mg/L)", "TN (mg/L)", 
+          "TP (mg/L)", "Ortho-P (mg/L)", "DO (% sat.)", "pH", "Chl-a (ug/L)"),
+  effval = c(0.292, 210, 220, mean(c(140, 161)), mean(c(140, 155)), NA, NA, 105)
+)
 
 bssum <- bswqdat %>% 
   filter(yr > 2005 & yr < 2021) %>% 
@@ -74,10 +89,12 @@ bssum <- bswqdat %>%
   unite(sumv, avev, minv, maxv, sep = '')
 
 tab <- full_join(stkraw, bssum, by = 'var') %>% 
+  full_join(effraw, by = c('var', 'lbs')) %>% 
   select(
     `Water quality variable` = lbs, 
-    `Stack value` = val, 
-    `Normal median (min, max)` = sumv
+    `2019 stack value` = stkval, 
+    `2021 end-of-pipe value` = effval,
+    `Bay median (min, max)` = sumv
   )
 
 stktab <- tab

### 2021-08-20 manu-draft.Rmd
history and timeline text added

@@
 An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed in May, with biomass estimated at 0.5 kg/m2 at some locations.
+Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass estimated at 0.5 kg/m2 at some locations.
 Blooms of *Karenia brevis* were observed beginning in May and continuing through July.
@@
 Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
-Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay[@Ohara17].
+Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay [@Ohara17].
 The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
@@
 Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to prevent impacts to the environment.
-The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).
-Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
-Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].
-Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
+Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FDEP).
+Decreasing holding capacity of the ponds with seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
+Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay [@Switzer11;@Garrett11].
+Recently, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
 
@@
 # Methods
+
+## History of Piney Point
+
+The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
+Port operations were primarily for export of phosphate production by the plant.
+Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
+Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
+In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
+
+The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
+Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
+Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
+Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater from the stack to be released into Bishop Harbor, with an estimated 15.4 tons fo nitrogen (pers.
+comm.
+D.
+Eckenrod, USEPA).
+Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
+From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to reduce the likelihood of an uncontrolled spill.
+@Switzer11 reported minimal impacts to nekton communities, although an increase in macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* was observed as a potential indication of nutrient eutrophication.
+Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on holding capacity of storage ponds [@Hu03].
+Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
+
+HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.
+This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
+In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
+This material was added to the existing holding pond at Piney Point, further reducing holding capacity leading to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to Bishop Harbor.
+The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
+HRK maintains ownership and responsibility of the site to present day, with oversight by FDEP.
+
+Discharges from Piney Point did not occur again until 2021.
+Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
+The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.
+Water quality measured in NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
+Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.
+This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.
+From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
+Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].
+
+## Response to the emergency discharge
+
+Monitoring of the natural resources of Tampa Bay in response to the discharge release of Piney Point began in April, 2021 and continued over the following months.
 
@@
 * Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
+* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
 * Potential long-term impacts TBD

### 2021-08-20 R/tables.R

@@ -90,12 +90,16 @@ bssum <- bswqdat %>%
 
 tab <- full_join(stkraw, bssum, by = 'var') %>% 
   full_join(effraw, by = c('var', 'lbs')) %>% 
+  mutate(
+    stkval = ifelse(is.na(stkval), '-', as.character(stkval)),
+    effval = ifelse(is.na(effval), '-', as.character(effval))
+  ) %>% 
   select(
     `Water quality variable` = lbs, 
     `2019 stack value` = stkval, 
     `2021 end-of-pipe value` = effval,
     `Bay median (min, max)` = sumv
-  )
+  ) 
 
 stktab <- tab
 

### 2021-08-20 manu-draft.Rmd
working on methods, data sources

@@
 This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
-The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, seagrass, benthic, and nekton datasets available for the region.
+The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
 We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
@@
 The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.
-Water quality measured in NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
 Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
@@
 Monitoring of the natural resources of Tampa Bay in response to the discharge release of Piney Point began in April, 2021 and continued over the following months.
+These data were collected through a coordinated effort, facilitated in part by the TBEP.
+Monitoring agencies and local partners that collected included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Florida Fish and Wildlife Conservation Commission Fish and Wildlife Research Institute, City of St.
+Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing and *in situ* measurements of discrete samples), phytoplankton cell counts, benthic sediment samples, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted but they are not reported here.
+
+Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
+Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
+Discrete water quality samples were taken from surface grabs by boat and processed by the respective laboratories of each participating agency.
+
+## Long-term monitoring data
+
+## Data Analysis
 

### 2021-08-22 manu-draft.Rmd
working on methods data description

@@
 
-## Response to the emergency discharge
+## Monitoring response to the emergency discharge
 
-Monitoring of the natural resources of Tampa Bay in response to the discharge release of Piney Point began in April, 2021 and continued over the following months.
+Monitoring of the natural resources of Tampa Bay in response to the wastewater release at Piney Point began in April, 2021 and continued over the following months.
 These data were collected through a coordinated effort, facilitated in part by the TBEP.
-Monitoring agencies and local partners that collected included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Florida Fish and Wildlife Conservation Commission Fish and Wildlife Research Institute, City of St.
+Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
 Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing and *in situ* measurements of discrete samples), phytoplankton cell counts, benthic sediment samples, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
 Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted but they are not reported here.
@@
 Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
+To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
 Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
-Discrete water quality samples were taken from surface grabs by boat and processed by the respective laboratories of each participating agency.
+Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
+For this paper, we focus on parameters related to the nutrient management paradigm for the bay and the expected phytoplankton response form inorganic nitrogen entering the day.
+This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orhophosphate (PO$_4^{3-}$), and chlorophyll-a (ug/L) concentrations.
+Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) are also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
+Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
+As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts of the discharge relative to Piney Point.
+
+Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
+Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa) as common primary producers observed throughout the growing season and species associated with harmful algal blooms (HABs) as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
+Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* and *Pyrodinium bahamense* that can occur in the bay depending on salinity and temperature conditions during the growing season.
+Occurrence of both these species has historically been spatially distinct, with *K.
+brevis* originating in the Gulf oc Mexico and occuring in higher salinity portions of the bay, whereas *P.
+bahamense* has been observed consistently each year since 2008 in Old Tampa Bay (northwest segment) during the summer.
+Data for *K.
+brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
+Because of the increased occurrence of red tide samples in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
+Petersberg) impacted by the event.
+Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
+
+Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point starting in April.
+Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay.
+Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid" design to sample seagrasses and macroalgae along a fifty meter transect at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet coverage within a 50 cm$^2$ quadrat at 10m distances along each transect.
+Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
+Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
+Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
+Seagrasses and macroalgae abundances were converted to frequency occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure \@ref(fig:map)a) depending on the analysis described below.
 

### 2021-08-23 manu-draft.Rmd
methods draft done

@@
 
+Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the Environmental Protection Commission of Hillsborough County.
+These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.
+Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection.
+Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).
+The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.
+Additionally, the period of record for monitoring data from Manatee County began in 1996.
+
+Long-term water quality monitoring data from Hillsborough and Manatee counties were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+Station data from Hillsborough County were obtained for the middle and lower segments of Tampa Bay.
+Station data from Manatee County were obtained for areas in lower Tampa Bay, Terra Ceia Bay, Anna Maria Sound, and northern Sarasota Bay.
+For the same water quality parameters noted above (i.e., nitrogen, phosphorus, pH, temperature, salinity), observations at each monitoring station were averaged for each month across years from 2006 to 2020.
+This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
+For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference concentrations relative to results at the closest monitoring station that was sampled in response to Piney Point.
+This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in August).
+Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].
+In some cases, the nearest long-term station did not include data for a every monitoring parameter at a response location and the next closest station was used as reference.
+Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
+
 ## Data Analysis
 
-* Historical timeline: https://docs.google.com/document/d/1KqaEAYEG7pdcGpWbUjONfVKMHHB6_kOkKCyncp4JYKA/edit, https://thebradentontimes.com/piney-point-a-retrospective-p6328-158.htm
-* Initial response, data collection
-* Synthesis and Analysis
+The R statistical programming language (v4.0.2) was used to import, synthesize, and analyze all datasets provided by multiple partners [@RCT20].
+Partner data were uploaded or entered manually as Google spreadsheets, where they were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.
+The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.
+The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.
 
-Need to note that K. brevis cell data are from event-based sampling, i.e., not appropriate to combine with other data for stats analysis.
+Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparisons tests to assess trends over time.
+Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general assessment of potential patterns in eutrophication following wastewater release.
+Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
+Data describing *K.
+brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing.
+Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.
+Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a.
+For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.
+Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.
 
-All stats based on FLDEP monitoring, fixed sample design, etc.
-
+The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
 For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
+For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
 
@@
 
-```{r trnfrq, fig.cap = 'Frquency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.
+```{r trnfrq, fig.cap = 'Frequency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.
 Estimates are grouped by sample months in 2021.
@@
 
-```{r wqcmps}
+```{r wqcmptab}
 cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.

### 2021-08-23 manu-draft.Rmd
starting results

@@
 The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.
-Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table (\@ref(tab:stktab), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
 Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
@@
 
-Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparisons tests to assess trends over time.
-Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general assessment of potential patterns in eutrophication following wastewater release.
+Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
+Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
 Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
@@
 
-* Figure \@ref(fig:map) map
+A general narrative of events in Tampa Bay following release of wastewater from Piney Point on March 30th is shown in Figure \@ref(fig:timeline).
+These events are supported by data summaries and results of statistical tests described below.
+After the the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
+Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
+The inition diatom bloom did not perist past April.
+On April 20th, *K.
+brevis* was first observed near Anna Maria Sound at the mouth of Tampa Bay and reached bloom concentrations by May 23rd, although observations were limited to lower Tampa Bay.
+Also during May, increased observations of *Lyngbya sp.* (a macroalgal cyanobacteria)
+
 * Figure \@ref(fig:timeline) timeline of events

### 2021-08-24 manu-draft.Rmd
water quality results draft done

@@
 
-library(here)
+box::use(
+  here[here],
+  dplyr[...]
+)
 
@@
 
-A general narrative of events in Tampa Bay following release of wastewater from Piney Point on March 30th is shown in Figure \@ref(fig:timeline).
-These events are supported by data summaries and results of statistical tests described below.
-After the the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
+## Timeline of events from April 2021
+
+A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
+After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
-The inition diatom bloom did not perist past April.
+The initial diatom bloom did not persist past April.
 On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the mouth of Tampa Bay and reached bloom concentrations by May 23rd, although observations were limited to lower Tampa Bay.
-Also during May, increased observations of *Lyngbya sp.* (a macroalgal cyanobacteria)
+brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
+Also during May, *Lyngbya sp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
+*Lyngbya* were observed in large floating mats on the surface and covering benthic and seagrass habitat below the water column at these locationa.
+By June 27th, fish kill reports attributed to red tide were increasing as *K.
+brevis* increased in abundance in lower and middle Tampa Bay.
+The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.
+This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically for the cities of St.
+Petersburg and Tampa.
+Concentrations of *K.
+brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay by August.
+The remainder of this section quantifies these events with monitoring data collected in response to Piney Point.
+
+## Water quality response
+
+```{r}
+nwq <- sum(wqsumtab[['N obs.']])
+outrng1 <- wqsumtab %>% 
+  filter(`Water quality variable` == 'Secchi (m)') %>% 
+  pull(below) %>%  
+  sum
+outrng2 <- wqsumtab %>% 
+  filter(`Water quality variable` != 'Secchi (m)') %>% 
+  pull(above) %>%  
+  sum
+outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
+```
+
+From April to August 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, ammonia, nitrate/nitrite, pH, salinity, and secchi depth (Table \@ref(tab:wqsumtab)).
+Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
+The percent of observations outside of the normal range varied by location and parameter.
+For chl-a, over half of the observations were above the normal range for area 1, whereas only 9.1\% and 24\% were above for areas 2 and 3, respectively.
+Similarly, total nitrogen concentrations were above the normal range for 39.9\% of observations in area 1, whereas concentrations were above for only 12.2\% and  23.4\% of observations in areas 2 and 3.
+Secchi observations for the period of observation were below the normal range for 46.4\% of observations in area1 and for   24.2\% and 43.5\% of observations in areas 2 and 3.
+Notable differences were also observed for dissolved oxygen (e.g., 49.7\% were above in area 1, 47.6\% in area 2).
+Physical parameters (e.g., salinity, temperature) were generally within range over the five month period.
+Inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations in April near area 1.
+Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi) for many locations concentrated near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
+
+Boxplots of total nitrogen, chl-a, and secchi depth show the temporal progression of observations by area relative to the normal ranges from long-term monitoring data (Figure \@ref(fig:wqtrnds)).
+For area 1, total nitrogen and chlorophyll concentrations were frequenlty above normal ranges beginning the week of April 4th and lasting through the month.
+Concentrations remained similar to baseline conditions until June and July when median values were often the baseline range.
+Secchi observations in area 1 were below baseline ranges in April and June.
+Observations in areas 2 and 3 were more often within the normal range, wiht some exceptions for total nitrogen and chl-a in area 3 during some weeks of May, June, and July.
+Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter and were significant for total nitrogen, chl-a, and secchi depth for area 1, total nitrogen and chl-a for area 2, and chl-a and secchi depth for area 3 (Table \@ref(tab:wqcmptab)).
+Results of multiple comparison tests that evaluated differneces between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
+Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
+brevis* abundances.
+
+Multivariate assessment of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)).
+The first principal component explained 41\%, 74\%, and 47\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.
+These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.
+The second and third principal components explained a minorit of the variation of among water quality parameters such that 36\%, 20\%, and 34\% was explained by both axes for areas 1, 2, and 3, respectively.
+The patterns explained by these axes were less clear, although physical drivers were partially explained in some cases (e.g., component 2 for area 2, Figure \@ref(fig:pcacors)b).
+Spearman rank correlations between the pairs of variables used for each of the PCAs showed significant assocations.
+Total nitrogen was positively associated with chlorophyll and negatively associated with secchi depth for areas 1 and 2 ($p < 0.005$ and $p < 0.05$, respectively).
+Dissolved oxygen was negatively associated with chl-a only in area 3, although the temporal progression likely changed over time in the other areas (e.g., increase followed by decrease).
+Temperature generally had a positive association with chlorophyll ($p < 0.05$ for area 2, $p < 0.005$ for area 3).
+pH was positively associated with chl-a in areas 1 and 2 ($p < 0.05$ for both).
+
+## Seagrass and macroalgae trends
+
+## Red tide and fish kill reports
 
@@
 
+* Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
 * Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
@@
 Summaries are median, maximum, and minimum values.
-Total observations (N obs.) and values in rnage, above, or below normal ranges are also shown.
+Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
 Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'

### 2021-08-24 R/tables.R

@@ -159,6 +159,11 @@ totab2 <- totab %>%
   unite('sumv', medv, rng, sep = '') %>% 
   left_join(totab1, by = c('area', 'lbs')) %>% 
   mutate(area = ifelse(duplicated(area), '', area)) %>% 
+  mutate(
+    `in range` = round(100 * `in range` / cnt, 1),
+    `above` = round(100 * `above` / cnt, 1),
+    `below` = round(100 * `below` / cnt, 1)
+  ) %>% 
   select(
     Area = area, 
     `Water quality variable` = lbs, 

### 2021-08-25 manu-draft.Rmd
minor edits to wq results, wqmap scale updated for tn

--- 2021-08-23 manu-draft.Rmd
@@
 Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater from the stack to be released into Bishop Harbor, with an estimated 15.4 tons fo nitrogen (pers.
+Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
 comm.
@@
 Occurrence of both these species has historically been spatially distinct, with *K.
-brevis* originating in the Gulf oc Mexico and occuring in higher salinity portions of the bay, whereas *P.
+brevis* originating in the Gulf of Mexico and occurring in higher salinity portions of the bay, whereas *P.
 bahamense* has been observed consistently each year since 2008 in Old Tampa Bay (northwest segment) during the summer.
@@
 
-A general narrative of events in Tampa Bay following release of wastewater from Piney Point on March 30th is shown in Figure \@ref(fig:timeline).
-These events are supported by data summaries and results of statistical tests described below.
-After the the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
+## Timeline of events from April 2021
+
+A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
+After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
-The inition diatom bloom did not perist past April.
+The initial diatom bloom did not persist past April.
 On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the mouth of Tampa Bay and reached bloom concentrations by May 23rd, although observations were limited to lower Tampa Bay.
-Also during May, increased observations of *Lyngbya sp.* (a macroalgal cyanobacteria)
+brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
+Also during May, *Lyngbya sp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
+*Lyngbya* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
+By June 27th, fish kill reports attributed to red tide increased with *K.
+brevis* cell concentrations in lower and middle Tampa Bay.
+The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.
+This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St.
+Petersburg and Tampa.
+Concentrations of *K.
+brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay by August.
+The remainder of this section is quantitative description these events.
+
+## Water quality response
+
+```{r}
+nwq <- sum(wqsumtab[['N obs.']])
+outrng1 <- wqsumtab %>% 
+  filter(`Water quality variable` == 'Secchi (m)') %>% 
+  pull(below) %>%  
+  sum
+outrng2 <- wqsumtab %>% 
+  filter(`Water quality variable` != 'Secchi (m)') %>% 
+  pull(above) %>%  
+  sum
+outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
+```
+
+From April to August 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
+Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
+The percent of observations outside of the normal range varied by location and parameter.
+For chl-a, 51\% of the observations were above the normal range for area 1, whereas only 9\% and 24\% were above for areas 2 and 3, respectively.
+Similarly, total nitrogen concentrations were above the normal range for 40\% of observations in area 1, whereas concentrations were above for only 12\% and  23\% of observations in areas 2 and 3.
+Secchi observations for the period of observation were below the normal range for 46\% of observations in area 1 and for 24\% and 44\% of observations in areas 2 and 3.
+Notable differences were also observed for dissolved oxygen (e.g., 50\% were above in area 1, 48\% in area 2).
+Physical parameters (salinity, temperature) were generally within range over the five month period.
+Inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
+Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
+
+Boxplots of total nitrogen, chl-a, and secchi depth show the temporal progression of observations by week and area relative to the normal ranges from long-term monitoring data (Figure \@ref(fig:wqtrnds)).
+For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges beginning the week of April 4th and lasting through the month.
+Concentrations remained similar to baseline conditions until June and July when median values were often the baseline range.
+Secchi observations in area 1 were below baseline ranges in April and June.
+Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 during weeks of May, June, and July.
+Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1, total nitrogen and chl-a for area 2, and chl-a and secchi depth for area 3 (Table \@ref(tab:wqcmptab)).
+Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
+Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
+brevis* abundances.
+
+Multivariate assessments of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)).
+The first principal component explained 41\%, 74\%, and 47\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.
+These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.
+The second and third principal components explained less of the variation among water quality parameters, such that 36\%, 20\%, and 34\% was explained by both axes for areas 1, 2, and 3, respectively.
+The patterns explained by these axes were less clear, although physical drivers were partially explained in some cases (e.g., component 2 for area 2 explained a positive association for salinity and temperature, Figure \@ref(fig:pcacors)b).
+Spearman rank correlations between the pairs of variables used for each of the PCAs showed significant associations.
+Total nitrogen was positively correlated with chlorophyll and negatively associated with secchi depth for areas 1 and 2 ($p < 0.005$ and $p < 0.05$, respectively).
+Dissolved oxygen was negatively associated with chl-a only in area 3 ($p < 0.005$), although the temporal progression likely changed over time in the other areas (e.g., increase followed by decrease).
+Temperature generally had a positive association with chlorophyll ($p < 0.05$ for area 2, $p < 0.005$ for area 3).
+pH was positively associated with chl-a in areas 1 and 2 ($p < 0.05$ for both).
+
+## Seagrass and macroalgae trends
+
+## Red tide and fish kill reports
 
@@
 Averages were taken when two measured values were available from both effluent sample dates.
-Missing values were not measured in the effluent.'
+Missing values were not measured in the effluent or stack water.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)

### 2021-08-25 R/figures.R

@@ -203,7 +203,7 @@ jpeg(here('figs/map.jpeg'), height = 4.2, width = 9, family = 'serif', units = '
 print(pout)
 dev.off()
 
-# nutrients, chloropyll map -----------------------------------------------
+# nutrients, chloropyll, secchi map ---------------------------------------
 
 # nonbay stations
 nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
@@ -261,14 +261,17 @@ attributes(bsmap1_transparent) <- mapatt
 bsmap <- ggmap(bsmap1_transparent)
 
 toplo1 <- wqdat %>% 
-  filter(var == 'tn')
+  filter(var == 'tn') %>% 
+  mutate(val = pmin(2, val))
 
 # static plot
+brks <- c(0.1, 0.25, 0.5, 2)
+lbs <- c('0.1', '0.25', '0.5', '>2')
 p1 <- bsmap +
   geom_point(data = toplo1, aes(x = lng, y = lat, size = val, fill = val, group = dategrp, color = inrng), pch = 21, alpha = 0.8) +
-  scale_fill_gradientn('mg/L', trans = 'log10', colours = vrscols) +
+  scale_fill_gradientn('mg/L', trans = 'log10', colours = vrscols, breaks = brks, labels = lbs) +
   scale_color_manual('In normal range?', values = c('black', 'lightgrey')) +
-  scale_size('mg/L', range = c(0.5, 5), trans = 'log10') + 
+  scale_size('mg/L', range = c(0.5, 5), trans = 'log10', breaks = brks, labels = lbs) + 
   coord_map() + 
   guides(
     fill = guide_legend(order = 1), 

### 2021-08-25 manu-draft.Rmd
working on transect results

@@
   here[here],
-  dplyr[...]
+  dplyr[...], 
+  english[english], 
+  lubridate[month], 
+  tibble[deframe]
 )
@@
 data(rstrnpts)
+data(rstrndat)
+data(trnsum)
 load(file = here('tables/wqsumtab.RData'))
@@
 A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnd)b).
+After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnds)b).
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
@@
 
-## Water quality response
+## Water quality trends
 
@@
 ## Seagrass and macroalgae trends
+
+```{r}
+ntrns <- length(unique(rstrndat$station))
+ntrns <- english(ntrns)
+visdat <- rstrndat %>% 
+  mutate(
+    mo = month(date)
+  ) %>% 
+  group_by(station, mo) %>% 
+  summarise(
+    cnt = length(unique(date)), 
+    .groups = 'drop'
+  ) %>% 
+  group_by(mo) %>% 
+  summarise(
+    cnt = mean(cnt)
+  )
+avevis <- round(mean(visdat$cnt), 1)
+
+# all fo
+foall <- rstrndat %>% 
+  mutate(
+    pa = bb > 0
+  ) %>% 
+  mutate(
+    taxa = as.character(taxa), 
+    taxa = case_when(
+      grepl('^Halophila', taxa) ~ 'Halophila sp', 
+      T ~ taxa
+    )
+  ) %>% 
+  group_by(taxa) %>% 
+  summarise(
+    fo = sum(pa) / length(pa), 
+    .groups = 'drop'
+  ) %>% 
+  mutate(
+    fo = round(100 * fo, 1)
+  ) %>% 
+  deframe
+
+# fo by month, area
+fomo <- trnsum %>% 
+  mutate(
+    foest = round(100 * foest, 1), 
+    mo = month(date)
+  )
+fomoarea1 <- fomo %>% 
+  filter(area == '(a) Area 1')
+fomoarea3 <- fomo %>% 
+  filter(area == '(b) Area 3')
+fomoarea1apr <- fomoarea1 %>% 
+  filter(mo %in% 4) %>% 
+  select(taxa, foest) %>% 
+  deframe
+fomoarea1aug <- fomoarea1 %>% 
+  filter(mo %in% 8) %>% 
+  select(taxa,foest) %>% 
+  deframe
+fomoarea3apr <- fomoarea3 %>% 
+  filter(mo %in% 4) %>% 
+  select(taxa, foest) %>% 
+  deframe
+fomoarea3jul <- fomoarea3 %>% 
+  filter(mo %in% 7) %>% 
+  select(taxa,foest) %>% 
+  deframe
+fomoarea1jun <- fomoarea1 %>% 
+  filter(mo %in% 6) %>% 
+  select(taxa, foest) %>% 
+  deframe
+fomoarea3jun <- fomoarea3 %>% 
+  filter(mo %in% 6) %>% 
+  select(taxa, foest) %>% 
+  deframe
+```
+
+A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April to August, each visited on average `r avevis` times per month.
+Turtle grass (*T.
+testudinum*) was the dominant seagrass species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
+Manatee grass (*S.
+filiforme*) and shoal grass (*H.
+wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
+Widgeon grass (*R.
+maritima*) and star grass (*H.
+engelmanni* and *H.
+decipiens*) were uncommon,  with frequency occurrences of `r foall['Ruppia maritima']`\% and `r foall['Halophila sp']`\%.
+The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].
+At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.
+Macroalgae observed along the transects also varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
+Common taxa in the red group included *Gracilaria sp.* and *Acanthophora sp.*.
+Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
+Common taxa in the green group included *Ulva sp.* and *Caulerpa sp.*, whereas *Lyngbya sp.* was the only cyanobacteria macroalgae observed.
+Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
+
+Temporal progression of seagrassess and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
+Transect S3T6 is located less than one kilometer to the north of Port Manatee.
+The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
+Overall abundunce of seagrass did not change dramatically from April 22nd to August 9th.
+However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
+Red macroalgae were present in high abundances from April to May.
+*Lyngbya sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
+*Lyngbya sp.* peristed through June and July, but not was observed after July 20th.
+Green macroalgae taxa were first observed in July, although at generally low abundances.
+By August, macroalgae abundances were low.
+Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae.
+However, overall macroalgae trends, including June dominance of cyanobacteria was observed at nearly all transects that suggests a broader impact from wastewater discharge at Piney Point.
+
+Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnex)) provided an indication of seagrass and macroalgae trends across all transects.
+No transects were sampled in area 2 to the north of Piney Point.
+Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in *H.
+wrightii* was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, August frequency occurrence of `r fomoarea1aug['Halodule wrightii']`\%).
+A similar decline in coverage of *H wrightii* was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
+Frequency occurrence of *S.
+syringodium* also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, August frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
+Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
+Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3).
+Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
+Green macroalgae had the second lowest frequency occurrence, with the highest coverage of `r fomoarea1aug['Green']`\% observed in August in area 1.
+Brown macroalgae was only observed at one transect.
 

### 2021-08-25 R/figures.R

@@ -636,7 +636,8 @@ trnsum <- rstrndat %>%
   ) %>% 
   group_by(area, typ, date, taxa) %>% 
   summarize(
-    foest = sum(pa) / length(pa)
+    foest = sum(pa) / length(pa), 
+    .groups = 'drop'
   ) %>% 
   filter(taxa %in% c(mcrsel, savsel))
 
@@ -666,6 +667,8 @@ p <- ggplot(trnsum, aes(x =date, y = foest, fill = taxa)) +
 jpeg(here('figs/trnfrq.jpeg'), height = 6, width = 8, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
+
+save(trnsum, file = here('data/trnsum.RData'))
 
 # all PCA and correlations ------------------------------------------------
 

### 2021-08-25 manu-draft.Rmd
Suggested title edits for your consideration.

@@
 ---
-title: "Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida"
+title: "Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
 output: 

### 2021-08-25 manu-draft.Rmd
Suggested abstract tweaks for your consideration.

@@
 From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual nitrogen load estimates in a matter of days.
-An initial phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
+An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
+An immediate phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
 Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass estimated at 0.5 kg/m2 at some locations.
-Blooms of *Karenia brevis* were observed beginning in May and continuing through July.
-Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1700 tons of dead fish.
+Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* began to be measured in May and continued through July within Tampa Bay proper.
+Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
 Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.

### 2021-08-26 manu-draft.Rmd
More abstract and intro edits for your consideration.

@@
 Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
-Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region.
+Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years.
 
@@
 
-Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from stormwater and wastewater sources.
-The effects of nitrogen from source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
+Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.
+The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
 Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18g].
 Tampa Bay is the largest estuary in Florida located in a heavily urbanized watershed of nearly 3 million individuals.
-Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads into major bay segments [@Janicki96;@Greening16].
-Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
+Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].
+Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
 
 Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.
-Although point-source inputs of nutrient loads from wastewater treatment plants into Tampa Bay have been reduced, non-point sources from wastewater and stormwater runoff contribute nutrients to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
-Atmospheric deposition of nutrients from coal-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs to the bay [@Poor13].
+While point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been much reduced, non-point source loads from wastewater sources and stormwater runoff are estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
+Atmospheric deposition of nutrients from fossil fuel based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
 Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
-Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program to implement a science-based resource management plan for the Bay [@Ohara17].
-The Tampa Bay Estuary Program has been instrumental in coordinating efforts among local and regional partners to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
+Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program through implementation of a science-based resource management plan for the Bay [@Ohara17].
+The Tampa Bay Estuary Program and its partners has been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
 
 Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources worldwide [@Tayibi09;@Hudson11].
-Fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
+Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
 The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to prevent impacts to the environment.
-The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].
+Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste to prevent unforeseen impacts to the environment.
+The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack located less than two miles from the shore of Tampa Bay and near two Florida aquatic preserves [@Henderson04].
 Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FDEP).
@@
 Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay [@Switzer11;@Garrett11].
+Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
 Recently, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+At that time, approximately 480 million gallons of legacy phosphate mining wastewater was being held in the failing stack.
 
 This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
-The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess deviation of current conditions relative to long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
+The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
 We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
-A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Piney Point.
+A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
 The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.

### 2021-08-26 manu-draft.Rmd
Methods section comments. I see you're also pulling in the lobo data. I think reporting the observed DO anomalies in the dataset would complement some of the analyses you've presented from the discrete datasets here too.

@@
 D.
-Eckenrod, USEPA).
+Eckenrod to USEPA, Nov.
+28, 2001).
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
@@
 In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to the existing holding pond at Piney Point, further reducing holding capacity leading to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to Bishop Harbor.
+This material was added to an existing, gypsumstack holding pond at Piney Point.
+Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
 The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
@@
 
-Discharges from Piney Point did not occur again until 2021.
+Legacy wastewater discharges from Piney Point did not occur again until 2021, although onsite stormwater management and discharges have occurred throughout its history.
 Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
@@
 Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release wastewater through a stormwater management system (under NPDES permit) that emptied at Port Manatee.
-This was done with the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent from the site.
+Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged at Berth 12 in Port Manatee.
+This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
 From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
-Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding the annual assimilative nutrient capacity of lower Tampa Bay in a matter of days [@tbep03a10].
+Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep03a10].
 
@@
 Monitoring of the natural resources of Tampa Bay in response to the wastewater release at Piney Point began in April, 2021 and continued over the following months.
-These data were collected through a coordinated effort, facilitated in part by the TBEP.
+These data were collected through a coordinated effort, facilitated in part by the FDEP and TBEP.
 Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
@@
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted but they are not reported here.
+Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted, but they are not reported here in anticipation of future monitoring events.
 
@@
 Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
-For this paper, we focus on parameters related to the nutrient management paradigm for the bay and the expected phytoplankton response form inorganic nitrogen entering the day.
+For this paper, we focus on parameters related to the nutrient management paradigm for the bay and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
 This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orhophosphate (PO$_4^{3-}$), and chlorophyll-a (ug/L) concentrations.
@@
 Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
-Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa) as common primary producers observed throughout the growing season and species associated with harmful algal blooms (HABs) as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
+Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
 Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* and *Pyrodinium bahamense* that can occur in the bay depending on salinity and temperature conditions during the growing season.
@@
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay.
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid" design to sample seagrasses and macroalgae along a fifty meter transect at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
-Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet coverage within a 50 cm$^2$ quadrat at 10m distances along each transect.
+Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along a fifty meter transect at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 50 cm$^2$ quadrat at 10m distances along each transect.
 Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
@@
 Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
-Seagrasses and macroalgae abundances were converted to frequency occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure \@ref(fig:map)a) depending on the analysis described below.
+Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure \@ref(fig:map)a) depending on the analysis described below.
 
@@
 Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].
-In some cases, the nearest long-term station did not include data for a every monitoring parameter at a response location and the next closest station was used as reference.
-Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
+In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
+These long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
 

### 2021-08-27 manu-draft.Rmd
results draft done

@@
 author: "Marcus W.
-Beck (mbeck@tbep.org), Gary E.
-Raulerson (graulerson@tbep.org), Maya C.
-Burke (mburke@tbep.org), Joe Whalen (jwhalen@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T.
-Sherwood (esherwood@tbep.org)"
+Beck (mbeck@tbep.org), Maya C.
+Burke (mburke@tbep.org), Emma E.
+Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
+Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T.
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue
@@
   dplyr[...],
+  tidyr[fill],
   english[english], 
-  lubridate[month], 
+  lubridate[month,year,mdy,floor_date], 
   tibble[deframe]
@@
 Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass estimated at 0.5 kg/m2 at some locations.
-Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* began to be measured in May and continued through July within Tampa Bay proper.
+Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
 Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
@@
 While point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been much reduced, non-point source loads from wastewater sources and stormwater runoff are estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
-Atmospheric deposition of nutrients from fossil fuel based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
+Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
 Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
@@
 Decreasing holding capacity of the ponds with seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
-Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay [@Switzer11;@Garrett11].
+Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.
 Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
@@
   deframe
+
+# kw pvals
+cyanoarea3 <- mcrtab %>% 
+  mutate(
+    Area = ifelse(Area == '', NA, Area)
+  ) %>% 
+  fill(Area) %>% 
+  filter(Area == 3 & `Macroalgae group` == 'Cyanobacteria') %>% 
+  pull(pval) %>% 
+  as.numeric %>% 
+  round(3)
 ```
@@
 
-Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnex)) provided an indication of seagrass and macroalgae trends across all transects.
+Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of seagrass and macroalgae trends across all transects.
 No transects were sampled in area 2 to the north of Piney Point.
-Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in *H.
-wrightii* was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, August frequency occurrence of `r fomoarea1aug['Halodule wrightii']`\%).
-A similar decline in coverage of *H wrightii* was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
-Frequency occurrence of *S.
-syringodium* also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, August frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
+Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in shoal grass (*H.
+wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, August frequency occurrence of `r fomoarea1aug['Halodule wrightii']`\%).
+A similar decline in coverage of shoal grass was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
+Frequency occurrence of manatee grass (*S.
+syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, August frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
 Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
@@
 
+Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrtab) and \@ref(tab:savtab) further suggested that seagrass trends were minor compared to those of macroalgae.
+Results by month showed that seagrass frequency occurrence, by species and area, had insignificant changes ($p > 0.05$) over the study period.
+Although trends were not significant, estimated medians in each month in Table \@ref(tab:savtab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.
+These changes were more pronounced in area 3 as compared to area 1.
+For macroalgae (Table \@ref(tab:mcrtab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).
+Coverage of cyanobacteria in area 1 during June was significantly higher than coverage in April.
+No significant differences between months were observed for the green macroalgae group.
+
+## Water quality compared to seagrass and macrolagae
+
+A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpcacors)).
+The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.
+For the water quality varialbes, the first principal component axis had positive loadings for secchi depth and dissolved oyxgen and negatative assocations for total nitrogen, chl-a, and temperature.
+For macroalgae, the first axis was positively associated with frequency occurrence of red and green macroalgae and negatively associated with cyanobacteria.
+Loading vectors for the seagrasses were minor, although turtle grass (*T.
+testudinum*) had a weak positive association with axis 1.
+The second principal component axis had a strong postive assocation with total phosphorus and negative association with salinity.
+The third axis showed a strong positive association with manatee grass (*S.
+filiforme*) and negative association with shoal grass (*H.
+wrightii*).
+
+Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables.
+Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).
+Chlorophyll-a was also negatively correlated red and green macroalgae ($p < 0.05$) and turtle grass ($p < 0.005$).
+Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).
+For water quality and seagrass, chl-a, total nitrogen, and total phosphorus was negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).
+For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$ for both), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
+Among the seagrasses, manatee grass was negatively correlated with turtle grass and shoal grass ($p < 0.005$ for both).
+
 ## Red tide and fish kill reports
 
-* Figure \@ref(fig:timeline) timeline of events
-* Figure \@ref(fig:wqmap) map of water quality observations
-* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
-* Figure \@ref(fig:pcacors) pca and correlations of all wq variables, by area 
-* Figure \@ref(fig:trnex) seagrass and macroalgae example transect
-* Figure \@ref(fig:trnfrq) seagrass, macroalgae frequency occurrence estimates
-* Figure \@ref(fig:allpcacors) seagrass, macralgae, wq pca and correlation matrix
-* Figure \@ref(fig:redtide) red tide historic and current, fish kill reports historic and current
+```{r}
+fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
+  select(
+    date = textBox6, 
+    city = cOUNTYDataTextBox, 
+    waterbody = lOCATIONDataTextBox,
+    species = textBox18
+  ) %>% 
+  mutate(
+    date = mdy(date),
+    yr = year(date),
+    week = floor_date(date, unit = 'week'), 
+    week = factor(format(week, '%b %d')), 
+    week = factor(week, levels = as.character(unique(week))), 
+    city = gsub('\\s+$', '', city),
+    city = gsub('^St\\s', 'St. ', city)
+  ) %>% 
+  filter(city %in% c('Tampa', 'St. Petersburg'))
+fishann <- fishdat %>% 
+  group_by(yr, city) %>% 
+  summarise(cnt = n(), .groups = 'drop') %>% 
+  group_by(city) %>% 
+  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
+  mutate(per = round(100 * cnt / sum(cnt), 1))
+annstpper <- fishann %>% 
+  filter(city == 'St. Petersburg') %>% 
+  pull(per)
+anntmpper <- fishann %>% 
+  filter(city == 'Tampa') %>% 
+  pull(per)
+fishwk <- fishdat %>% 
+  filter(yr == 2021) %>% 
+  group_by(week, city) %>% 
+  summarise(cnt = n(), .groups = 'drop') %>% 
+  group_by(city) %>% 
+  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
+  mutate(per = round(100 * cnt / sum(cnt), 1))
+wkstpcnt <- fishwk %>% 
+  filter(city == 'St. Petersburg') %>% 
+  pull(cnt)
+wktmpcnt <- fishwk %>% 
+  filter(city == 'Tampa') %>% 
+  pull(cnt)
+```
 
-* Table \@ref(tab:stktab) Stack characteristics
-* Table \@ref(tab:wqsumtab) water quality data collected and summarized
-* Table \@ref(tab:wqcmptab) Multiple comparisons of water quality by month/area, with median/min/max
-* Table \@ref(tab:mcrtab) Macroalgae freq. occurrence summaries by month/area, with median/min/max
-* Table \@ref(tab:savtab) Seagrass freq. occurrence summaries by month/area, with median/min/max
+The increase in *K.
+brevis* from April to July to bloom concentrations exceeding 10k cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10k cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels.
+Median cell concentrations for most years were well below 1k cells/L.
+The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L.
+Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined through August (Figure \@ref(fig:redtide)b).
 
-Seagrass: note decline of HW and correlations in the matrix, these make sense.
+Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K.
+brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c).
+Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.
+In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` reports were made in Tampa (Figure \@ref(fig:redtide)d).
+The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K.
+brevis* cell concentrations (Figure \@ref(fig:redtide)b).
+Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed over nearly five months.
+Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.
+As a result, the city of St.
+Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
+Hammer Levy, City of St.
+Petersburg, pers.
+comm.
+Aug.
+2021).
 
-Seagrass interp from DT:
+# Discussion
 
-First off, Marcus please correct me if any confusion on my part about how to read the chart, but it’s my take that the asterisks reflect levels of significance, and that a box without an asterisk – I didn’t pay much attention to that.
+<!-- Seagrass: note decline of HW and correlations in the matrix, these make sense.  -->
+
+<!-- Seagrass interp from DT:  -->
+
+<!-- First off, Marcus please correct me if any confusion on my part about how to read the chart, but it’s my take that the asterisks reflect levels of significance, and that a box without an asterisk – I didn’t pay much attention to that.
 Also, a negative number would be a “cool” color, reflecting an inverse relationship, while a positive number (warm color) would reflect a positive correlation?
+-->
 
-If I have this correct, then what stood out to me was some of the following – seagrass distributions are highly inversely correlated between most of the species, which reflects seagrass being clonal organisms that tend not to grow interspersed with each other in a random manner.
+<!-- If I have this correct, then what stood out to me was some of the following – seagrass distributions are highly inversely correlated between most of the species, which reflects seagrass being clonal organisms that tend not to grow interspersed with each other in a random manner.
 S.
@@
 wrightii can be both the shallowest and deepest growing species in areas like Hillsborough Bay.
+-->
 
-More relevant here is that there is no significant correlation between the abundance of cyanobacteria and either chlorophyll-a or TN or TP.
+<!-- More relevant here is that there is no significant correlation between the abundance of cyanobacteria and either chlorophyll-a or TN or TP.
 To me, that makes sense, since macroalgae like cyanos seem to reflect an alternative nutrient destination, compared to phytoplankton.
 The strongest correlations out there are the well expected ones between TN and Chl-a, but also between TP and NH3/4, which suggests overall nutrient availability tracks together, and phytoplankton continue to be a major destination of nitrogen availability.
+-->
 
-But what about the positive correlation between cyanobacteria and S.
+<!-- But what about the positive correlation between cyanobacteria and S.
 filiforme?
@@
 Cyanos are strongly negatively correlated with red algae, since they occurred at different times, basically.
+-->
 
-Finally, keep in mind that “Lyngbya” life cycles are such that most of the biomass produced does not stay where it was produced.
+<!-- Finally, keep in mind that “Lyngbya” life cycles are such that most of the biomass produced does not stay where it was produced.
 Those clumps of Lyngbya growing on the bottom get lifted up off the bottom by the oxygen bubbles you can see in their blooms, and they then are carried along the shoreline to decompose on whatever shoreline that they’ve been blown to.
 Our photos (and others) clearly show that there is MUCH more Lyngbya in our bays than what we’ve sampled along the bottom of the bay – that’s a bit different than “traditional” red algae biomass.
-
-# Discussion
+-->
 
@@
 
-```{r trnfrq, fig.cap = 'Frequency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = `r nrow(rstrnpts)`) near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = ', ntrns, ') near Piney Point.
 Estimates are grouped by sample months in 2021.
-Frequency occurrences are absolute for each taxa based on presence/absence.'}
+Frequency occurrences are absolute for each taxa based on presence/absence.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))
@@
 ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
-knitr::kable(mcrtab, booktabs = T, caption = cap.val)
+totab <- mcrtab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
 ```
@@
 ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
-knitr::kable(savtab, booktabs = T, caption = cap.val)
+totab <- savtab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
 ```

### 2021-08-27 R/tables.R

@@ -17,6 +17,8 @@ source(here('R/funcs.R'))
 data(rswqdat)
 data(ppseg)
 data(rsstatloc)
+data(rstrnpts)
+data(rstrndat)
 data(parms)
 data(bsstatloc)
 data(bswqdat)
@@ -383,6 +385,7 @@ trnsum <- trnsf %>%
         mutate(
           minval = paste0(' (', minval, ', '),
           maxval = paste0(maxval, ')'), 
+          pval = tst$p.value,
           krusk_pval = p_ast(tst$p.value),
           krusk_chis = round(tst$statistic, 2)
         ) %>% 
@@ -407,13 +410,15 @@ savests <- ests %>%
   mutate(
     area = ifelse(duplicated(area), '', area), 
     taxa = ifelse(duplicated(taxa), '', as.character(taxa)), 
-    krusk = ifelse(duplicated(krusk), '', krusk)
+    krusk = ifelse(duplicated(krusk), '', krusk),
+    pval = ifelse(duplicated(pval), '', pval)
   ) %>% 
   ungroup() %>% 
   select(
     Area = area, 
     `Seagrass species` = taxa, 
     `Chi-Sq.` = krusk,
+    pval,
     `Comp.` = lets, 
     Month = mo, 
     `N obs.` = length, 
@@ -428,13 +433,15 @@ mcrests <- ests %>%
   mutate(
     area = ifelse(duplicated(area), '', area), 
     taxa = ifelse(duplicated(taxa), '', as.character(taxa)), 
-    krusk = ifelse(duplicated(krusk), '', krusk)
+    krusk = ifelse(duplicated(krusk), '', krusk),
+    pval = ifelse(duplicated(pval), '', pval)
   ) %>% 
   ungroup() %>% 
   select(
     Area = area, 
     `Macroalgae group` = taxa, 
     `Chi-Sq.` = krusk,
+    pval,
     `Comp.` = lets, 
     Month = mo, 
     `N obs.` = length, 

### 2021-08-29 manu-draft.Rmd
adding legend items for map figure

@@
 Additionally, the period of record for monitoring data from Manatee County began in 1996.
+Long-term monitoring stations for Hillsborough and Manatee Counties are shown in Figure \@ref(fig:map)a.
 
@@
 
-```{r map, fig.cap = 'Areas of interest (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to wastewater discharge from Piney Point.
+```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to wastewater discharge from Piney Point.
 Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).

### 2021-08-29 R/figures.R

@@ -20,7 +20,9 @@ box::use(
   scales = scales[muted], 
   units = units[set_units], 
   vegan = vegan[decostand], 
-  FactoMineR = FactoMineR[PCA]
+  FactoMineR = FactoMineR[PCA], 
+  ggnewscale[new_scale_fill], 
+  grid[unit]
 )
 
 data(rsallpts)
@@ -104,17 +106,17 @@ bsmap1_transparent <- matrix(adjustcolor(bsmap1,
                              nrow = nrow(bsmap1))
 attributes(bsmap1_transparent) <- mapatt
 
-#For reference, add long-term station locations to Figure 1a or create another panel w/ stations?
-#Added stations, but now how to combine with existing 1b legend or add a new legend here?
 # plot
 p1 <- ggmap(bsmap1_transparent) +
   geom_sf(data = areas, aes(fill = area), color = NA, inherit.aes = F, alpha = 0.8) +
-    scale_fill_manual(values = cols, drop = F) +
-  geom_sf(data = bsstatloc, aes(fill = 'ambient'), pch =21, color = 'black', fill = 'white', inherit.aes= F) +
-  geom_sf(data = pineypoint, aes(fill = 'Piney Point'), pch = 24, color = 'black', fill = 'black', size = 3, inherit.aes= F) + 
+  scale_fill_manual(values = cols, drop = F, guide = 'none') +
+  new_scale_fill() + 
+  geom_sf(data = bsstatloc, aes(fill = 'Long-term monitoring', shape = 'Long-term monitoring'), color = 'black', size = 2, inherit.aes= F) +
+  geom_sf(data = pineypoint, aes(fill = 'Piney Point', shape = 'Piney Point'), color = 'black', size = 3, inherit.aes= F) + 
   geom_sf_text(data = areas, aes(label = area), color = 'black', inherit.aes = F, alpha = 0.8, size = 6) +
-  
-    theme_bw() + 
+  scale_fill_manual('test', values = c('white', 'black')) +
+  scale_shape_manual('test', values= c(21, 24)) +
+  theme_bw() + 
   theme(
     legend.title = element_blank(), 
     panel.grid = element_blank(), 
@@ -125,7 +127,7 @@ p1 <- ggmap(bsmap1_transparent) +
     panel.background = element_rect(fill = 'white'),
     axis.ticks = element_line(colour = 'grey'),
[…]

### 2021-08-29 manu-draft.Rmd
working on censored data #1 with NADA methods

@@
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts of the discharge relative to Piney Point.
+Values below laboratory detection limits (or Secchi values on the bottom) were assessed prior to analysis.
+Appropriate analysis methods to include parameters with observations below detection are described below.
 
@@
 
-Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
+Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), correlations between pairs of variables, and multiple comparison tests to assess trends between months.
 Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
@@
 The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
-For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+Water quality values below detection were removed prior to PCA.
+For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated by testing differences among empirical cumulative distribution functions using the "cendiff" function from the NADA R package [@Lee20] to accommodate observations below detection.
+Secchi observations that were visually identified on the bottom (12 observations of 518 total) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with the above methods.
+Evaluations were made for each parameter in each area using month as a group variable, followed by pairwise comparisons between months.
 Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
+Similar comparisons were made for seagrass and macroalgae using the Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+All correlation analyses used the Kendall tau correlation to accommodate censored data in the water quality results ("cenken" function from the NADA R package).
+For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partnerps.
 
@@
 Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
-Chlorophyll observations were truncated to a maximum of 30 ug/L to remove some outliers.'}
+Values below detection limits (or secchi on bottom) are not shown.'}
 knitr::include_graphics(here('figs/wqmap.jpeg'))
@@
 Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area.'}
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area.
+Values below detection limits (or secchi on bottom) are not shown.'}
 knitr::include_graphics(here('figs/wqtrnds.jpeg'))
@@
 cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point.
-Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
+Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
 Summaries are median, maximum, and minimum values.
 Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
+The final column shows the percentage of total observations that were below detection limits (including Secchi on bottom).'
 knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
@@
 cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
+Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown based on differences between empirical cumulative distribution functions with adjustments for observations below detection.
+Multiple comparisons (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
 Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different.

### 2021-08-29 R/figures.R

@@ -236,6 +236,7 @@ thm <-  theme(
 wqdat <- rswqdat %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!station %in% nonbay) %>% 
+  filter(!qual %in% c('(VOB)', 'U')) %>% 
   inner_join(rsstatloc, ., by = 'station') %>% 
   mutate(
     lng = st_coordinates(.)[, 1], 
@@ -319,14 +320,16 @@ p2 <- bsmap +
   )
 
 toplo3 <- wqdat %>% 
-  filter(var %in% 'secchi')
+  filter(var %in% 'secchi') %>% 
+  mutate(val = pmin(5, val))
 
-brks <- c(1, 2, 3, 4)
+brks <- c(0.5, 2, 3.5, 5)
+lbs <- c('0.5', '2', '3.5', '>5')
 p3 <- bsmap +
   geom_point(data = toplo3, aes(x = lng, y = lat, size = val, fill = val, group = dategrp, color = inrng), pch = 21, alpha = 0.8) +
-  scale_fill_gradientn('meters',colours = rev(vrscols), breaks = brks) +
+  scale_fill_gradientn('meters',colours = rev(vrscols), breaks = brks, labels = lbs) +
   scale_color_manual('In normal range?', values = c('black', 'grey'), guide = T) +
-  scale_size('meters', range = c(6, 0.5), breaks = brks) + 
+  scale_size('meters', range = c(6, 0.5), breaks = brks, labels = lbs) + 
   coord_map() + 
   guides(
     fill = guide_legend(order = 1), 
@@ -361,9 +364,13 @@ ppsegbf <- ppseg %>%
 cols <- c("#E16A86", "#50A315", "#009ADE")
 names(cols) <- levels(ppseg$area)
 
-p1 <- wqplo_fun(rswqdat, bswqdat, ppsegbf, vr = 'tn', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
-p2 <- wqplo_fun(rswqdat, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
[…]

### 2021-08-29 R/funcs.R

@@ -22,6 +22,19 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
   ##
   # wq data
   
+  # unique weeks
+  wklv <- rswqdat %>%
+    pull(date) %>% 
+    range
+  wklv <- seq.Date(wklv[1], wklv[2],by = 'days') %>% 
+    floor_date(unit = 'week') %>% 
+    unique %>% 
+    crossing(date = ., area = c('Area 1', 'Area 2', 'Area 3')) %>% 
+    mutate(
+      fillcl = factor(area, levels = unique(area), labels = cols), 
+      fillcl = as.character(fillcl)
+    )
+  
   # monitoring data
   rswqtmp <- rswqdat %>% 
     filter(var == vr) %>% 
@@ -35,7 +48,8 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
       mo = month(date), 
       fillcl = factor(area, levels = levels(area), labels = cols), 
       fillcl = as.character(fillcl)
-    ) 
+    ) %>% 
+    left_join(wklv, ., by = c('date', 'area', 'fillcl'))
   
   # baseline data
   bswqtmp <- bswqdat %>% 
@@ -67,12 +81,14 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
   
   # boxplot colors
   bxcls <- rswqtmp %>% 
+    na.omit() %>% 
     select(area, date, fillcl) %>% 
-    unique
+    unique %>% 
+    arrange(area)
   
   p1 <- ggplot() + 
     geom_rect(data = bswqtmp, aes(xmin = datestr, xmax = dateend, ymin = minv, ymax = maxv, group = mo, fill = 'Monthly baseline (mean +/- 1 sd)'), alpha = 0.2) +
-    geom_boxplot(data = rswqtmp, aes(x = date, y = val, group = date), fill= bxcls$fillcl, outlier.colour = NA, lwd = 0.5, alpha = 0.8, show.legend = F) + 
+    geom_boxplot(data = rswqtmp, aes(x = date, y = val, group = date), fill = bxcls$fillcl, outlier.colour = NA, lwd = 0.5, alpha = 0.8, show.legend = F) + 
     geom_jitter(data = rswqtmp, aes(x = date, y = val, group = date), alpha = 0.4, size = 0.5) + 
[…]

### 2021-08-29 R/tables.R

@@ -6,6 +6,7 @@ library(tbeptools)
 library(tidyverse)
 library(sf)
 library(lubridate)
+library(NADA)
 box::use(
   units = units[set_units], 
   forcats = forcats[fct_drop], 
@@ -130,8 +131,12 @@ totab <- rswqdat %>%
   left_join(rsstatloc, ., by = c('station', 'source')) %>% 
   filter(var %in% vrs) %>% 
   filter(!station %in% nonbay) %>% 
-  select(station, date, lbs, val, inrng) %>% 
+  select(station, date, lbs, val, inrng, qual) %>% 
   st_intersection(ppsegbf) %>% 
+  filter(!qual %in% '(VOB)') %>% 
+  mutate(
+    qual = grepl('^U|^LOD', qual) # censored data
+  ) %>% 
   st_set_geometry(NULL)
 
 totab1 <- totab %>% 
@@ -147,13 +152,15 @@ totab2 <- totab %>%
   group_by(area, lbs) %>% 
   left_join(sigs, by = 'lbs') %>% 
   summarise(
-    medv = round(median(val, na.rm = T), unique(sigdig)), 
+    nondetect = sum(qual),
+    medv = round(median(ros(val, censored = qual, forwardT = NULL, reverseT = NULL), na.rm = T), unique(sigdig)),
     minv = round(min(val, na.rm = T), unique(sigdig)), 
     maxv = round(max(val, na.rm = T), unique(sigdig)), 
     cnt = n(), 
     .groups = 'drop'
   ) %>% 
   mutate(
+    medv = ifelse(medv < 0, minv, medv),
     minv = paste0(' (', minv, ', '), 
     maxv = paste0(maxv, ')')
   ) %>% 
@@ -164,7 +171,8 @@ totab2 <- totab %>%
   mutate(
     `in range` = round(100 * `in range` / cnt, 1),
     `above` = round(100 * `above` / cnt, 1),
-    `below` = round(100 * `below` / cnt, 1)
+    `below` = round(100 * `below` / cnt, 1), 
+    `Below detection` = round(100 * nondetect / cnt, 1)
   ) %>% 
   select(
     Area = area, 
@@ -173,7 +181,8 @@ totab2 <- totab %>%
     `N obs.` = cnt, 
     `in range`, 
     above, 
-    below
+    below, 
+    `Below detection`
   )
 
 wqsumtab <- totab2
@@ -229,7 +238,7 @@ cmps <- rswqsub %>%
       
[…]

### 2021-08-30 manu-draft.Rmd
all analyses now use appropriate summary estimates for non-detects #1, need to update text results

--- 2021-08-29 manu-draft.Rmd
@@
 The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
-For total nitrogen, chlorophyll-a, and secchi depth, differences in concentrations between months within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+Water quality values below detection were removed prior to PCA.
+For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
+Secchi observations that were visually identified on the bottom (12 observations of 518 total) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with methods described below.
+Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
 Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
+When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.
+These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from the NADA R package.
+For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.
+Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.
+PCA and correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.
 
@@
 cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point.
-Variables are grouped by major areas interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
+Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
 Summaries are median, maximum, and minimum values.
 Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.'
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
+The final column shows the percentage of total observations that were below detection limits (including Secchi on bottom).
+Medians denoted by "-" could not be calculated due to insufficient values above detection.'
 knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
@@
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
 knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
@@
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
 totab <- mcrtab %>% 
@@
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
 totab <- savtab %>% 

### 2021-08-30 R/figures.R

@@ -16,6 +16,7 @@ library(ggord)
 library(lubridate)
 library(patchwork)
 library(tbeptools)
+library(NADA)
 box::use(
   scales = scales[muted], 
   units = units[set_units], 
@@ -382,7 +383,7 @@ dev.off()
 
 # PCA and correlations ----------------------------------------------------
 
-vrs <- c('chla', 'dosat', 'nh34', 'ph', 'secchi', 'temp', 'tn', 'no23', 'tp', 'sal')
+vrs <- c('chla', 'dosat', 'nh34', 'ph', 'secchi', 'temp', 'tn', 'tp', 'sal')
 nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
 
 cols <- c("#E16A86", "#50A315", "#009ADE")
@@ -402,10 +403,14 @@ rswqsub <- rswqdat %>%
   filter(var %in% vrs) %>% 
   filter(source == 'fldep') %>%
   filter(!station %in% nonbay) %>% 
+  filter(!grepl('VOB', qual)) %>% 
   inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
   st_intersection(ppsegbf) %>% 
   st_set_geometry(NULL) %>% 
-  select(date, var, val, station, area) %>% 
+  mutate(
+    cens = grepl('U', qual)
+  ) %>% 
+  select(date, var, val, cens, station, area) %>% 
   mutate(
     var = case_when(
       var == 'chla' ~ 'Chl-a', 
@@ -415,7 +420,6 @@ rswqsub <- rswqdat %>%
       var == 'secchi' ~ 'Secchi', 
       var == 'temp' ~ 'Temp', 
       var == 'tn' ~ 'TN', 
-      var == 'no23' ~ 'NOx',
       var == 'tp' ~ 'TP', 
       var == 'sal' ~ 'Sal'
     ),
@@ -423,9 +427,10 @@ rswqsub <- rswqdat %>%
   ) %>% 
   group_by(date, var, area) %>% 
   summarise(
-    val = median(val, na.rm = T), 
+    val = ifelse(any(cens), median(cenfit(val, cens), na.rm = T), median(val, na.rm = T)),
     .groups = 'drop'
-  ) 
+  ) %>% 
+  na.omit()
 
 rswqtmp <- rswqsub %>% 
[…]

### 2021-08-30 R/tables.R

@@ -69,20 +69,24 @@ bssum <- bswqdat %>%
   filter(yr > 2005 & yr < 2021) %>% 
   filter(source == 'epchc') %>% 
   filter(var %in% unique(stkraw$var)) %>% 
+  filter(!is.na(val)) %>% 
   inner_join(bsstatloc, ., by = c('station')) %>% 
   .[ltb, ] %>% 
   st_set_geometry(NULL) %>% 
+  mutate(
+    cens = grepl('U', qual)
+  ) %>% 
   group_by(yr, var) %>% 
   summarise(
-    val = median(val, na.rm = T),
+    val = median(cenfit(val, censored = cens), na.rm = T),
     .groups = 'drop'
   ) %>%
   left_join(parms, by = 'var') %>% 
   group_by(var) %>% 
   summarise(
-    minv = round(min(val), unique(sigdig)),
-    maxv = round(max(val), unique(sigdig)),
-    avev = round(median(val), unique(sigdig)), 
+    minv = round(min(val, na.rm = T), unique(sigdig)),
+    maxv = round(max(val, na.rm = T), unique(sigdig)),
+    avev = round(median(val, na.rm = T), unique(sigdig)), 
     .groups = 'drop'
   ) %>% 
   mutate(
@@ -153,14 +157,14 @@ totab2 <- totab %>%
   left_join(sigs, by = 'lbs') %>% 
   summarise(
     nondetect = sum(qual),
-    medv = round(median(ros(val, censored = qual, forwardT = NULL, reverseT = NULL), na.rm = T), unique(sigdig)),
+    medv = round(median(cenfit(val, censored = qual), na.rm = T), unique(sigdig)),
     minv = round(min(val, na.rm = T), unique(sigdig)), 
     maxv = round(max(val, na.rm = T), unique(sigdig)), 
     cnt = n(), 
     .groups = 'drop'
   ) %>% 
   mutate(
-    medv = ifelse(medv < 0, minv, medv),
+    medv = ifelse(is.na(medv), '-', medv),
     minv = paste0(' (', minv, ', '), 
     maxv = paste0(maxv, ')')
   ) %>% 
@@ -207,11 +211,12 @@ rswqsub <- rswqdat %>%
   filter(var %in% vrs) %>% 
   filter(source == 'fldep') %>%
   filter(!station %in% nonbay) %>% 
+  filter(!qual %in% '(VOB)') %>% # no U quals in fldep tn, chla, secchi data
[…]

### 2021-08-31 manu-draft.Rmd
slight text edit to analysis methods
scrap separate wq pca by area, added groups to pca plots, working on results edits with better repro

@@
 
+data(rswqdat)
 data(rstrnpts)
@@
 For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference concentrations relative to results at the closest monitoring station that was sampled in response to Piney Point.
+Methods in the NADA R package [@Lee20] to account for observations below detection limits were used for all summary statistics.
 This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in August).
@@
 ## Data Analysis
+
+```{r}
+seconb <- rswqdat %>% 
+  filter(var == 'secchi') %>% 
+  filter(source == 'fldep') %>% 
+  pull(qual) %>% 
+  grepl('S', .)
+sumb <- sum(seconb)
+totb <- length(seconb) 
+```
 
@@
 The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
-Water quality values below detection were removed prior to PCA.
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-Secchi observations that were visually identified on the bottom (12 observations of 518 total) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with methods described below.
+Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with methods described below.
 Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
+
 When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.
-These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from the NADA R package.
+These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from NADA.
 For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.
@@
 outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
+
+# percentages out of range
+wqsumfl <- wqsumtab %>% 
+  mutate(
+    Area = ifelse(Area == '', NA, Area)
+  ) %>% 
+  fill(Area) 
+chlabv <- wqsumfl %>% 
+  filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
+  select(Area, above) %>% 
+  deframe %>% 
+  round(0)
+tnabv <- wqsumfl %>% 
+  filter(`Water quality variable` == 'TN (mg/L)') %>% 
+  select(Area, above) %>% 
+  deframe %>% 
+  round(0)
+secblw <- wqsumfl %>% 
+  filter(`Water quality variable` == 'Secchi (m)') %>% 
+  select(Area, below) %>% 
+  deframe %>% 
+  round(0)
+dosabv <- wqsumfl %>% 
+  filter(`Water quality variable` == 'DO (% sat.)') %>% 
+  select(Area, above) %>% 
+  deframe %>% 
+  round(0)
 ```
@@
 The percent of observations outside of the normal range varied by location and parameter.
-For chl-a, 51\% of the observations were above the normal range for area 1, whereas only 9\% and 24\% were above for areas 2 and 3, respectively.
-Similarly, total nitrogen concentrations were above the normal range for 40\% of observations in area 1, whereas concentrations were above for only 12\% and  23\% of observations in areas 2 and 3.
-Secchi observations for the period of observation were below the normal range for 46\% of observations in area 1 and for 24\% and 44\% of observations in areas 2 and 3.
-Notable differences were also observed for dissolved oxygen (e.g., 50\% were above in area 1, 48\% in area 2).
+For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
+Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.
+Secchi observations for the period of observation were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.
+Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).
 Physical parameters (salinity, temperature) were generally within range over the five month period.
-Inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
-Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
+Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
+Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
 
@@
 For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges beginning the week of April 4th and lasting through the month.
-Concentrations remained similar to baseline conditions until June and July when median values were often the baseline range.
+Concentrations remained similar to baseline conditions until June and July when median values were often above the baseline range.
 Secchi observations in area 1 were below baseline ranges in April and June.
@@
 Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds).
-Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1, total nitrogen and chl-a for area 2, and chl-a and secchi depth for area 3 (Table \@ref(tab:wqcmptab)).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for areas 1 and 3 and total nitrogen and chl-a for area 2 (Table \@ref(tab:wqcmptab)).
 Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
@@
 Multivariate assessments of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)).
-The first principal component explained 41\%, 74\%, and 47\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.
+The first principal component explained 36\%, 49\%, and 42\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.
 These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.

### 2021-08-31 R/figures.R

@@ -232,12 +232,12 @@ thm <-  theme(
     legend.box= 'vertical'
   ) 
 
-#Secchi VOB and other qualified values should be filtered prior to use below, correct?
 # combine water quality data with locations
+# removed secchi on bottom and non-detect for tn, chla
 wqdat <- rswqdat %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!station %in% nonbay) %>% 
-  filter(!qual %in% c('(VOB)', 'U')) %>% 
+  filter(!qual %in% c('S', 'U')) %>% 
   inner_join(rsstatloc, ., by = 'station') %>% 
   mutate(
     lng = st_coordinates(.)[, 1], 
@@ -278,8 +278,8 @@ toplo1 <- wqdat %>%
   mutate(val = pmin(2, val))
 
 # static plot
-brks <- c(0.1, 0.25, 0.5, 2)
-lbs <- c('0.1', '0.25', '0.5', '>2')
+brks <- c(0.25, 0.5, 1, 2)
+lbs <- c('0.25', '0.5', '1', '>2')
 p1 <- bsmap +
   geom_point(data = toplo1, aes(x = lng, y = lat, size = val, fill = val, group = dategrp, color = inrng), pch = 21, alpha = 0.8) +
   scale_fill_gradientn('mg/L', trans = 'log10', colours = vrscols, breaks = brks, labels = lbs) +
@@ -367,7 +367,7 @@ names(cols) <- levels(ppseg$area)
 
 datin <- rswqdat %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
-  filter(!qual %in% c('(VOB)', 'U'))
+  filter(!qual %in% c('S', 'U')) # remove secchi on bottom, nondetect for chla, tn
 
 p1 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
 p2 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
@@ -377,7 +377,7 @@ p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Week
[…]

### 2021-08-31 R/funcs.R

@@ -50,19 +50,28 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
       fillcl = as.character(fillcl)
     ) %>% 
     left_join(wklv, ., by = c('date', 'area', 'fillcl'))
-  
+
   # baseline data
   bswqtmp <- bswqdat %>% 
     select(-source, -uni) %>% 
-    filter(var == vr) %>% 
+    filter(var == vr) %>%
+    filter(!(var == 'secchi' & grepl('S', qual))) %>% # remove secchi on bottom
     filter(yr > 2005) %>% 
+    filter(!is.na(val)) %>% 
     inner_join(bsstatloc, ., by = 'station') %>% 
     st_intersection(ppseg) %>% 
     st_set_geometry(NULL) %>% 
+    mutate(cens = grepl('U', qual)) %>% 
     group_by(mo, var, area) %>% 
     summarise(   
-      avev = mean(val, na.rm = T), 
-      stdv = sd(val, na.rm = T), 
+      avev = ifelse(
+        any(cens), mean(cenfit(val, cens), na.rm = T),
+        mean(val, na.rm = T)
+      ),
+      stdv = ifelse(
+        any(cens), sd(cenfit(val, cens), na.rm = T),
+        sd(val, na.rm = T)
+      ),
       .groups = 'drop'
     ) %>%
     left_join(parms, by = 'var') %>% 
@@ -106,7 +115,7 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
       axis.title.x = element_blank(),
       panel.grid.minor = element_blank(),
       strip.text = element_text(size = 14), 
-      axis.text.x = element_text(size = 6.5, angle = 45, hjust = 1)
+      axis.text.x = element_text(size = 6.25, angle = 45, hjust = 1)
     )
   
   if(logtr)

### 2021-08-31 R/tables.R

@@ -137,9 +137,9 @@ totab <- rswqdat %>%
   filter(!station %in% nonbay) %>% 
   select(station, date, lbs, val, inrng, qual) %>% 
   st_intersection(ppsegbf) %>% 
-  filter(!qual %in% '(VOB)') %>% 
+  filter(!qual %in% 'S') %>%  # remove secchi on bottom
   mutate(
-    qual = grepl('^U|^LOD', qual) # censored data
+    qual = grepl('U', qual) # censored data
   ) %>% 
   st_set_geometry(NULL)
 
@@ -211,7 +211,7 @@ rswqsub <- rswqdat %>%
   filter(var %in% vrs) %>% 
   filter(source == 'fldep') %>%
   filter(!station %in% nonbay) %>% 
-  filter(!qual %in% '(VOB)') %>% # no U quals in fldep tn, chla, secchi data
+  filter(!qual %in% 'S') %>% # remove secchi on bottom
   inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
   st_intersection(ppsegbf) %>% 
   st_set_geometry(NULL) %>% 

### 2021-08-31 manu-draft.Rmd
no separate PCA and corr analysis for water quality

@@
 Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
-Further analyses were conducted to compare all data types together, including the entire suite of water quality data, seagrass frequency occurrences, and macroalgae frequency occurrences to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
+Further analyses were conducted to compare all data types together using PCA and correlation analyses, to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
 Data describing *K.
@@
 brevis* abundances.
-
-Multivariate assessments of all water quality parameters that combined results across weeks by area showed significant associations between groups of variables (Figure \@ref(fig:pcacors)).
-The first principal component explained 36\%, 49\%, and 42\% of the variation among water quality parameters for areas 1, 2, and 3, respectively.
-These components generally described a positive association between nutrients and chlorophyll concentrations and a negative association with secchi depth and dissolved oxygen.
-The second and third principal components explained less of the variation among water quality parameters, such that 36\%, 20\%, and 34\% was explained by both axes for areas 1, 2, and 3, respectively.
-The patterns explained by these axes were less clear, although physical drivers were partially explained in some cases (e.g., component 2 for area 2 explained a positive association for salinity and temperature, Figure \@ref(fig:pcacors)b).
-Spearman rank correlations between the pairs of variables used for each of the PCAs showed significant associations.
-Total nitrogen was positively correlated with chlorophyll and negatively associated with secchi depth for areas 1 and 2 ($p < 0.005$ and $p < 0.05$, respectively).
-Dissolved oxygen was negatively associated with chl-a only in area 3 ($p < 0.005$), although the temporal progression likely changed over time in the other areas (e.g., increase followed by decrease).
-Temperature generally had a positive association with chlorophyll ($p < 0.05$ for area 2, $p < 0.005$ for area 3).
-pH was positively associated with chl-a in areas 1 and 2 ($p < 0.05$ for both).
 
@@
 
-```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).
-All variables were standardized to zero mean and unit variance prior to PCA.
-Variables with log-normal distribution were log-transformed prior to analysis.
-Pearson correlations indicate the linear strength of assocation between pairs of variables.
-** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
-knitr::include_graphics(here('figs/pcacors.jpeg'))
-```
-
 ```{r trnex, fig.cap = 'Results for (a) seagrass and (b) macroalgae rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.

### 2021-08-31 R/figures.R

@@ -381,146 +381,6 @@ jpeg(here('figs/wqtrnds.jpeg'), height = 6, width = 8.5, units = 'in', res = 500
 print(p)
 dev.off()
 
-# PCA and correlations ----------------------------------------------------
-
-vrs <- c('chla', 'dosat', 'nh34', 'ph', 'secchi', 'temp', 'tn', 'tp', 'sal')
-nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
-
-cols <- c("#E16A86", "#50A315", "#009ADE")
-names(cols) <- c('Area 1', 'Area 2', 'Area 3')
-
-ppsegbf <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
-  mutate(
-    area = factor(area)
-  )
-
-rswqsub <- rswqdat %>% 
-  filter(var %in% vrs) %>% 
-  filter(source == 'fldep') %>%
-  filter(!station %in% nonbay) %>% 
-  filter(!grepl('S', qual)) %>% # remove secchi on bottom
-  inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
-  st_intersection(ppsegbf) %>% 
-  st_set_geometry(NULL) %>% 
-  mutate(
-    cens = grepl('U', qual)
-  ) %>% 
-  select(date, var, val, cens, station, area) %>% 
-  mutate(
-    var = case_when(
-      var == 'chla' ~ 'Chl-a', 
-      var == 'dosat' ~ 'DOsat', 
-      var == 'nh34' ~ 'NH3, NH4+', 
-      var == 'ph' ~ 'pH', 
-      var == 'secchi' ~ 'Secchi', 
-      var == 'temp' ~ 'Temp', 
-      var == 'tn' ~ 'TN', 
-      var == 'tp' ~ 'TP', 
-      var == 'sal' ~ 'Sal'
-    ),
-    date = floor_date(date, unit = 'week')
-  ) %>% 
-  group_by(date, var, area) %>% 
-  summarise(
-    val = ifelse(any(cens), median(cenfit(val, cens), na.rm = T), median(val, na.rm = T)),
-    .groups = 'drop'
-  ) %>% 
-  na.omit() %>% 
[…]

### 2021-08-31 manu-draft.Rmd
separate pca and cor matrix for all variables, need to update text results

@@
 
-A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpcacors)).
+A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpca)).
 The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.
@@
 
-Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables.
+Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
 Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).
@@
 
-```{r allpcacors, fig.cap = 'Principal components analysis (PCA) and correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
+```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
 Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
 All variables were standardized to zero mean and unit variance prior to PCA.
+Variables with log-normal distribution were log-transformed prior to analysis.'}
+knitr::include_graphics(here('figs/allpca.jpeg'))
+```
+
+```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
+Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
 Variables with log-normal distribution were log-transformed prior to analysis.
-Thick black lines in the correlation matrix separate water quality, macroalgae, and seagrass variables.
-Pearson correlations indicate the linear strength of assocation between pairs of variables.
+Thick black lines separate water quality, macroalgae, and seagrass variables.
+Spearman correlations indicate the linear strength of assocation between pairs of variables.
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
-knitr::include_graphics(here('figs/allpcacors.jpeg'))
+knitr::include_graphics(here('figs/allcor.jpeg'))
 ```

### 2021-08-31 R/figures.R

@@ -509,11 +509,143 @@ dev.off()
 
 save(trnsum, file = here('data/trnsum.RData'))
 
-# all PCA and correlations ------------------------------------------------
+# all PCA -----------------------------------------------------------------
 
 vrs <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
 mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
 savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
+wqsel <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
+wqlab <- c('Chl-a', 'DOsat', 'NH3, NH4+', 'pH', 'Sal', 'Secchi', 'Temp', 'TN', 'TP')
+
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+# segments
+areas <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+# combine water quality with transect data at aggregated scale
+
+# add area
+trnsum <- rstrndat %>% 
+  mutate(
+    date = floor_date(date, unit = 'week')
+  ) %>% 
+  inner_join(rstrnpts, ., by = 'station') %>% 
+  st_intersection(areas) %>% 
+  st_set_geometry(NULL) %>% 
+  dplyr::group_by(area, date, station, location, taxa) %>%
+  dplyr::summarise(
+    pa = as.numeric(any(bb > 0))
+  ) %>%
+  group_by(area, date, taxa) %>% 
+  summarize(
+    val = sum(pa) / length(pa) # freq occ.
+  ) %>% 
+  ungroup %>% 
+  filter(taxa %in% c(savsel, mcrsel)) %>% 
+  rename(var = taxa) %>%
+  spread(var, val) %>% 
+  rename(
+    Cyano = Cyanobacteria, 
+    `H. wrightii` = `Halodule wrightii`, 
+    `T. testudinum` = `Thalassia testudinum`, 
[…]

### 2021-09-01 manu-draft.Rmd
results text updated with better handling of nondetects/secchi on bottom, fix #1

@@
 data(trnsum)
+data(allpcadat)
+
 load(file = here('tables/wqsumtab.RData'))
@@
 
-A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates shows significant assocations between groups of variables (Figure \@ref(fig:allpca)).
-The first principal component axis explained 31\% of the variation among all variables, whereas the second and third axes explained 33\% of the remaining variation.
-For the water quality varialbes, the first principal component axis had positive loadings for secchi depth and dissolved oyxgen and negatative assocations for total nitrogen, chl-a, and temperature.
-For macroalgae, the first axis was positively associated with frequency occurrence of red and green macroalgae and negatively associated with cyanobacteria.
-Loading vectors for the seagrasses were minor, although turtle grass (*T.
-testudinum*) had a weak positive association with axis 1.
-The second principal component axis had a strong postive assocation with total phosphorus and negative association with salinity.
-The third axis showed a strong positive association with manatee grass (*S.
-filiforme*) and negative association with shoal grass (*H.
-wrightii*).
+```{r}
+expvr <- allpcadat$eig[, 2] %>% 
+  round(0)
+```
+
+A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).
+The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]` of the remaining variation.
+For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and strong negative loadings for total nitrogen, chl-a, and temperature.
+For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae.
+For seagrasses, turtle grass (*T.
+testudinum*) had a strong positive loading with axis 1.
+The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.
+The third axis showed a strong positive loading with shoal grass (*H.
+wrightii*) and negative association with manatee grass (*S.
+filiforme*).
+Observations by area showed that samples in area 3 were grouped by positive loadings on on the first PCA axis and negative loadings on the second PCA axis and were distinct from points in area 1.
 
 Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
-Chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH (all $p < 0.005$) and negatively correlated with secchi depth ($p < 0.005$).
-Chlorophyll-a was also negatively correlated red and green macroalgae ($p < 0.05$) and turtle grass ($p < 0.005$).
+Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all, except $p < 0.05$ for pH) and negatively correlated with secchi depth ($p < 0.05$).
+Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$) and negatively correlated with salinity and secchi ($p < 0.05$).
 Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).
-For water quality and seagrass, chl-a, total nitrogen, and total phosphorus was negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).
-For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$ for both), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
-Among the seagrasses, manatee grass was negatively correlated with turtle grass and shoal grass ($p < 0.005$ for both).
+Among the seagrasses, manatee grass was negatively correlated wtih shoal grass and turtle grass ($p < 0.005$).
+Several notable correlations were also observed between water quality, macroalgae, and seagrass.
+Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$).
+For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).
+For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
 
@@
 * Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
+* Explain correlations/PCA - area 3 distinct from area 1 (high salinity/turtle grass dominated, likely just a natural grouping)
 * Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)

### 2021-09-01 R/figures.R

@@ -629,15 +629,17 @@ exp <- 0.1
 parse <- F
 ellipse <- F
 
-ppp <- PCA(toord, scale.unit = F, graph = F) 
-p1 <- ggord(ppp, axes = c('1', '2'), grp_in = grps, ellipse = ellipse, cols = cols, parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, exp = exp)
-p2 <- ggord(ppp, axes = c('2', '3'), grp_in = grps, ellipse = ellipse, cols = cols, parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, exp = exp)
+allpcadat <- PCA(toord, scale.unit = F, graph = F) 
+p1 <- ggord(allpcadat, axes = c('1', '2'), grp_in = grps, ellipse = ellipse, cols = cols, parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, exp = exp)
+p2 <- ggord(allpcadat, axes = c('2', '3'), grp_in = grps, ellipse = ellipse, cols = cols, parse = parse, vec_ext = vec_ext, coord_fix = coord_fix, size = size, repel = repel, arrow = arrow, txt = txt, alpha = alpha, ext = ext, exp = exp)
 
 p <- p1 + p2 + plot_layout(ncol = 2, guides = 'collect') & theme(legend.position = 'top', legend.title = element_blank())
 
 jpeg(here('figs/allpca.jpeg'), height = 4, width = 7, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
+
+save(allpcadat, file = here('data/allpcadat.RData'))
 
 # all correlations --------------------------------------------------------
 
@@ -826,7 +828,7 @@ pbase <- theme(
   panel.grid.minor = element_blank(), 
   axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1, size = 7), 
   axis.text.y = element_text(size = 7),
-  legend.position = c(0.5, 1.1),
+  legend.position = c(0.5, 1.07),
   legend.direction = 'horizontal',
   plot.margin = unit(c(4,4,0,0), "lines"),
[…]

### 2021-09-01 manu-draft.Rmd
started discussion text

@@
 
-<!-- Seagrass: note decline of HW and correlations in the matrix, these make sense.  -->
-
 <!-- Seagrass interp from DT:  -->
@@
 
-* Explain results - what's up with inorganic nutrients?  Likely driving phyto response but concentrations are not a good indicator.  Uptake is rapid...
-* Explain correlations/PCA - area 3 distinct from area 1 (high salinity/turtle grass dominated, likely just a natural grouping)
+Several of the correlations in Figure \@ref(fig:allcor) can be explained through mechanistic relationshps among water quality, macroalgae, and seagrasses.
+Conversely, many of the correlations may be spurious and due to collinearity among variables that is likely not causative.
+Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca).
+An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.
+Temporally, these changes were observed at different times and by different species of phytoplankton.
+The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K.
+brevis* in June and July.
+Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
+Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.
+Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.
+Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.
+
+From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability that may negatively affect seagrass growth.
+However, changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous.
+Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.
+Spatially, area 3 closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
+This reflects a natural partitioning of the species along a salinity gradient (as shown in the PCA in Figure \@ref(fig:allpca)) and not any association with effluent from Piney Point.
+Correlations among the macroalgae and seagrasses can also be explained by spatial differences among species along a depth gradient.
+Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.
+Seagrasses are clonal organisms that grow together and species are usually not randomly dispersed in the same patch.
+Shoal grass and turtle grass typically grow at shallower depths than manatee grass and are often exposed during low tides.
+As an adaptive mechanism to varying water levels, blades of shoal grass and turtle grass can "lay down" during low tide to minimize dessication stress.
+Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species is better suited at deeper depths that are minimally affected by varying tidal height.
+As such, the correlation analysis can be explained by spatial partitioning of seagrasses along depth gradients due to adaptive characteristics of each species.
+These associations may also reflect natural variation in water clarity along the longitidinal axis of Tampa Bay.
+Shoal grass can occur in more turbid waters that increase towards the northern portion of Tampa Bay, which are also shallower.
+
+Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.
+This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.
+The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.
+However past spill events were associated with an increase in both red and green macroalgae in smaller subembayments of Tampa Bay, prompting manual removal to prevent conditions associated with hypoxia from decomposition of macroalgal biomass [@Switzer11].
+Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites as a result of nutrient cycling stimulated by wastewater release.
+A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., Lyngbya sp.).
+Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021.
+The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
+*Lyngbya sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
+
+The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
+Although the seagrass community did not change dramatically across the study period, long-term impacts may be observed.
+Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that peristed over a year and half.
+These changes may have resulted from a reduction in water clarity from heavy concentrations of *K.
+brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.
+Annually, seagrass coverage in Tampa Bay has declined...
+what other factors?
+PP is an affront...
+
 * Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)

### 2021-09-01 manu-draft.Rmd
remove placeholder text from DT

@@
 
-<!-- Seagrass interp from DT:  -->
-
-<!-- First off, Marcus please correct me if any confusion on my part about how to read the chart, but it’s my take that the asterisks reflect levels of significance, and that a box without an asterisk – I didn’t pay much attention to that.
-Also, a negative number would be a “cool” color, reflecting an inverse relationship, while a positive number (warm color) would reflect a positive correlation?
--->
-
-<!-- If I have this correct, then what stood out to me was some of the following – seagrass distributions are highly inversely correlated between most of the species, which reflects seagrass being clonal organisms that tend not to grow interspersed with each other in a random manner.
-S.
-filiforme does not grow in shallow waters, because it’s blades can’t “lay down” during low tides, and it often is the deepest growing species in some of our cleaner areas, while H.
-wrightii can be both the shallowest and deepest growing species in areas like Hillsborough Bay.
--->
-
-<!-- More relevant here is that there is no significant correlation between the abundance of cyanobacteria and either chlorophyll-a or TN or TP.
-To me, that makes sense, since macroalgae like cyanos seem to reflect an alternative nutrient destination, compared to phytoplankton.
-The strongest correlations out there are the well expected ones between TN and Chl-a, but also between TP and NH3/4, which suggests overall nutrient availability tracks together, and phytoplankton continue to be a major destination of nitrogen availability.
--->
-
-<!-- But what about the positive correlation between cyanobacteria and S.
-filiforme?
-From my experience in Sarasota Bay, that makes sense, since the highest levels of cyanobacteria were found entangled within the manatee grass meadows in the fairly deeper waters farther offshore.
-Cyanos are strongly negatively correlated with red algae, since they occurred at different times, basically.
--->
-
-<!-- Finally, keep in mind that “Lyngbya” life cycles are such that most of the biomass produced does not stay where it was produced.
-Those clumps of Lyngbya growing on the bottom get lifted up off the bottom by the oxygen bubbles you can see in their blooms, and they then are carried along the shoreline to decompose on whatever shoreline that they’ve been blown to.
-Our photos (and others) clearly show that there is MUCH more Lyngbya in our bays than what we’ve sampled along the bottom of the bay – that’s a bit different than “traditional” red algae biomass.
--->
+Make inventory of anomalous events..
 

### 2021-09-02 manu-draft.Rmd
more draft discussion

@@
 
-Make inventory of anomalous events..
+The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
+These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
+The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
+Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.
+In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.
+Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
+Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.
+From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years.
+These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.
 
-Several of the correlations in Figure \@ref(fig:allcor) can be explained through mechanistic relationshps among water quality, macroalgae, and seagrasses.
-Conversely, many of the correlations may be spurious and due to collinearity among variables that is likely not causative.
+Several of the correlations in Figure \@ref(fig:allcor) support known responses of water quality to nutrient loads and can be explained through mechanistic relationships among water quality, macroalgae, and seagrasses.
 Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca).
@@
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
-Although the seagrass community did not change dramatically across the study period, long-term impacts may be observed.
-Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that peristed over a year and half.
+Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.
+Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.
 These changes may have resulted from a reduction in water clarity from heavy concentrations of *K.
 brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.
-Annually, seagrass coverage in Tampa Bay has declined...
-what other factors?
-PP is an affront...
+Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
+In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
+The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.
+General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.
+Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.
+From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
+These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
+Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
 

### 2021-09-02 manu-draft.Rmd
more discussion draft

@@
 Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.
-Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.
+Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year in southwest Florida.
 These changes may have resulted from a reduction in water clarity from heavy concentrations of *K.
@@
 The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.
-General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.
-Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.
+General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities.
+Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Management District.
 From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
-These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
+These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
 Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+
+Potential short and long-term impacts to Tampa Bay resources beyond water quality, macroalgae, and seagrasses may also have occurred.
+Fisheries resources have undoubtedly been impacted by *K.
+brevis* blooms in June and July as documented by the fish kill reports (Figure \@ref(fig:redtide)).
+Quantitative data on changes in abundance and diversity of fish species in Tampa Bay following the red tide event are forthcoming.
+Routine sampling by the Fisheries Independent Monitoring Program of the FWC occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay.
+@Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay and demonstrated a sensitivity of the index to a significant red tide event in lower Tampa Bay in 2005.
+Given the observed red tide concentration in 2021 and the magnitude of reported fish kills, a similar response in the fish community as reflected by the TBNI is expected.
+Of note is that fisheries resources improved quickly following surveys after 2005, although temporary recreational or commercial closures may need to be implemented to allow the fish community to recover.
+As of writing, catch and release only limits have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).
+@SWitzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.
+Fish community structure and species composition did not differ compared to the pre-impact period.
+From 2003 to 2007, an estimated 1.9 billion gallons was discharged to Bishop Harbor, where various treatment methods were used to mitigate potential impacts of introducing acidic and nutrient-rich process water to surface waters.
+@Garrett11 noted bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
+brevis* and dense diatom blooms, in Bishop Harbor during this time.
+
+Additional monitoring was conducted for benthic communities to sample sediment quality at Port Manatee and surrounding waters in April 2021.
+Biological surveys were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthic community based on the quality of the sediments they inhabit [@tbep0106].
+Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions.
+Results from initial sampling suggest a mix of conditions with dominated by "fair" and "good" samples near Port Manatee reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.
+Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earlist sampline in 1993.
+Additional sampling is planned at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources.
+Detailed baseline information and data relative to 2021 are lacking for these species relative to the primary data reported herein.
+At least one [report](https://myfwc.com/research/manatee/rescue-mortality-response/statistics/mortality/2021/) to FWC of a manatee mortality in Tampa Bay at the peak of the red tide bloom in July was recorded, although cause of death was undetermined.
+Regardless, it is reasonable to assume toxicity of *K.
+brevis* can have a negative effect on regional populations of large mammals.
+This is of particular concern given the recent report of manatee mass die-offs on the east coast of Florida related to seagrass losses.
 
@@
 * Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
-* Potential long-term impacts TBD
 * Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change

### 2021-09-03 manu-draft.Rmd
draft discussion done, need to add refs

@@
 Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.
+
 In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.
-Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
+Spill events reported to FDEP (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
+An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
 Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.
-From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years.
-These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.
+From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in 2021 is an outlier compared to past years.
+These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
 
@@
 Fisheries resources have undoubtedly been impacted by *K.
-brevis* blooms in June and July as documented by the fish kill reports (Figure \@ref(fig:redtide)).
+brevis* blooms in June and July as documented by the fish kill reports to FWC (Figure \@ref(fig:redtide)).
 Quantitative data on changes in abundance and diversity of fish species in Tampa Bay following the red tide event are forthcoming.
-Routine sampling by the Fisheries Independent Monitoring Program of the FWC occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay.
-@Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay and demonstrated a sensitivity of the index to a significant red tide event in lower Tampa Bay in 2005.
+Routine sampling by the FWC Fisheries Independent Monitoring Program occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information on species richness and abundance for the major segments of the bay.
+@Schrandt21 developed a Tampa Bay Nekton Index to assist with fisheries management strategies for the bay that demonstrated a sensitivity of fisheres resrouces to a significant red tide event in lower Tampa Bay in 2005.
 Given the observed red tide concentration in 2021 and the magnitude of reported fish kills, a similar response in the fish community as reflected by the TBNI is expected.
 Of note is that fisheries resources improved quickly following surveys after 2005, although temporary recreational or commercial closures may need to be implemented to allow the fish community to recover.
-As of writing, catch and release only limits have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).
-@SWitzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.
+As of writing, catch and release only limits for 2021 have been extended through the fall for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*).
+For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following dishcarge to the subembayment and compared the results to fisheries data for years prior.
 Fish community structure and species composition did not differ compared to the pre-impact period.
@@
 
-Additional monitoring was conducted for benthic communities to sample sediment quality at Port Manatee and surrounding waters in April 2021.
+Additional monitoring and evaluation of results may also be required to fully understand the long-term impacts to bay resources beyond water quality.
+Sampling was conducted for benthic communities to evaluate sediment quality at Port Manatee and surrounding waters in April 2021.
 Biological surveys were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthic community based on the quality of the sediments they inhabit [@tbep0106].
 Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions.
-Results from initial sampling suggest a mix of conditions with dominated by "fair" and "good" samples near Port Manatee reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.
-Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earlist sampline in 1993.
-Additional sampling is planned at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
-Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources.
-Detailed baseline information and data relative to 2021 are lacking for these species relative to the primary data reported herein.
-At least one [report](https://myfwc.com/research/manatee/rescue-mortality-response/statistics/mortality/2021/) to FWC of a manatee mortality in Tampa Bay at the peak of the red tide bloom in July was recorded, although cause of death was undetermined.
-Regardless, it is reasonable to assume toxicity of *K.
-brevis* can have a negative effect on regional populations of large mammals.
-This is of particular concern given the recent report of manatee mass die-offs on the east coast of Florida related to seagrass losses.
+Results from initial sampling suggested a mix of conditions with dominated by "fair" and "good" samples near Port Manatee, reflecting the high spatial variability of macroinvertebrate communities in estuarine sediments.
+Comparison of the initial 2021 samples to historical benthic conditions using TBBI scores for comparison suggested relatively consistent sediment quality back to earliest sampling in 1993.
+Additional sampling is planned later in 2021 at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their abundances in the bay and use of bay resources within the broader spatial distribution of these species.
+Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) within the known red tide boundary of the impacted counties of Tampa Bay through August 2021.
+This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is suspected to be strongly linked to seagrass losses on the east coast (e.g., Indian River Lagoon) and current trends of seagrass loss for southwest Florida.
 
-* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
-* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
-* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
-* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
-* Risk of decline (IRL ex.), regression of past progress
+Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly at the port instead of through channels to back-bay habitats (i.e., Bishop Harbor).
+The comparison to past years provided a stark contrast to the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impact on different habitats that vary spatially and throughout the water column.
+The Tampa Bay Coastal Ocean Model [@Chen19] was used to create a three-dimensional picture of simulated daily tracer particle movement and expected dilution of nutrient concentrations following release at the point of discharge.
+Although these efforts initially informed monitoring to track potential plume impacts, the results suggested that the spread of the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
+Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient rich wastewater from Piney Point.
+Further, estimates of tracer concentrations suggested that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a matter of days.
+However, diluted concentrations were considered biologically relevant given the initial concentrations of inorganic nitrogen measured in stack effluent (Table \@ref(tab:stktab)) and the background concentrations for the bay from baseline monitoring (Table \@ref(tab:wqsumtab)).
+
+In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to human health and the environment.
+This study adds to the growing body of literature on the impacts of effluent from gypstack release to coastal waters that can inform both event-based monitoring and anticipated environmental response.
+In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.
+Grand Bay is a 7500 hectare protected area in southern Mississippi that has historically been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18].
+The first spill event in 2005 occurred following failure of the retaining walls for the gypstack after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
+Following both of these events, massive fish kills were observed likely related to acute toxicity with low pH exposure with discharge of wastewater.
+Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concern of the long-term impacts focused primarily on the effects of heavy loads of orthophosphate to the system [@Dillon15].
+Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
+A general trend is that legacy wastewater from past production of fertilizer is poorly maintained and long-term plans ae insufficient to dispose of pollutants that can significantly affect coastal resources.
+Piney Point and Grand Bay are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities in an environmentally responsible manner are needed at the national and international level.
+
+Limitations of our analysis are also important to note to inform future event-based monitoring and additional research to better understand acute, short-term threats to the sustained recovery of Tampa Bay.
+All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.
+However, the interpretations are supported by well-established linkages between drivers of primary production in coastal waters.
+The outstanding weight of evidence that 2021 is an anomaly relative to past years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
+Additional data to support these results could include explicit load-based estimates for all sources entering the bay in 2021 and these estimates are forthcoming as additional context to the events of this year.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more definite description of the recycling and fate of nitrogen in the system.
+There may likely never be an irrefutable link between the observations in 2021 and the wastewater from Piney Point, but this is not necessary to support future actions that ensure the continued threats of these facilities to the health of coastal resources are eliminated through collaborative efforts by local, regional, and state partners.
+These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
 

### 2021-09-04 manu-draft.Rmd
editing discussion

--- 2021-09-02 manu-draft.Rmd
@@
 From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
-Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep03a10].
+Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
 
@@
 
-The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point together represent an adverse environmental response to a large pulse of inorganic nitrogen into the system.
-Collectively, these observations provide a weight of evidence that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
-These anomalous events included 1) a large, spring diatom bloom in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for an unusual amount of observations supported by the decades of long-term monitoring data as a baseline for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
-Although these out-of-range results had specific temporal and spatial patterns during the period of observation, they were seen throughout the bay suggesting a system-wide effect of effluent from Piney Point.
+The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay in June and July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
+The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
+Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
+
 In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.
-Spill events reported to FDEP have been documented for years (e.g., industrial spills, stormwater pump failures, sanitary sewer overflows) and spill volumes for these for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
-Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point discharged over the ten day period exceeded current annual estimates of the external load into lower Tampa Bay.
-From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in the Bay in 2021 is an outlier compared to past years.
-These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator and asset for environmental quality.
+Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
+An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
+Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of the external load into lower Tampa Bay [@tbep0417].
+From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in 2021 is an outlier compared to past years.
+These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
 
-Several of the correlations in Figure \@ref(fig:allcor) support known responses of water quality to nutrient loads and can be explained through mechanistic relationships among water quality, macroalgae, and seagrasses.
-Regarding the nutrient management paradigm for Tampa Bay, the linear association between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations and loadings from the PCA in Figure \@ref(fig:allpca).
-An increase in nitrogen was associated with an increase in chlorophyll, which in turn was associated with a decrease in water clarity.
-Temporally, these changes were observed at different times and by different species of phytoplankton.
+Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
+Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)) and loadings from the PCA results (Figure \@ref(fig:allpca)).
+Temporally, these changes were observed at different times and for different species of phytoplankton.
 The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K.
@@
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
-Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa.
+Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
 Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.
-Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen already utilized by different taxa, including macroalgae.
+Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
 
@@
 Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.
-Spatially, area 3 closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
-This reflects a natural partitioning of the species along a salinity gradient (as shown in the PCA in Figure \@ref(fig:allpca)) and not any association with effluent from Piney Point.
-Correlations among the macroalgae and seagrasses can also be explained by spatial differences among species along a depth gradient.
+Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
+This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].
+Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient.
 Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.
-Seagrasses are clonal organisms that grow together and species are usually not randomly dispersed in the same patch.
-Shoal grass and turtle grass typically grow at shallower depths than manatee grass and are often exposed during low tides.
-As an adaptive mechanism to varying water levels, blades of shoal grass and turtle grass can "lay down" during low tide to minimize dessication stress.
-Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species is better suited at deeper depths that are minimally affected by varying tidal height.
-As such, the correlation analysis can be explained by spatial partitioning of seagrasses along depth gradients due to adaptive characteristics of each species.
-These associations may also reflect natural variation in water clarity along the longitidinal axis of Tampa Bay.
-Shoal grass can occur in more turbid waters that increase towards the northern portion of Tampa Bay, which are also shallower.
+Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.
+As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.
+Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper.
+These associations may also reflect natural variation in water clarity along the main axis of Tampa Bay.
+Shoal grass can occur in more turbid waters of oligohaline areas.
 
@@
 The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.
-However past spill events were associated with an increase in both red and green macroalgae in smaller subembayments of Tampa Bay, prompting manual removal to prevent conditions associated with hypoxia from decomposition of macroalgal biomass [@Switzer11].
-Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites as a result of nutrient cycling stimulated by wastewater release.
-A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., Lyngbya sp.).
+However past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
+Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
+A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., *Lyngbya sp.*).
 Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021.
@@
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
-Although the seagrass community did not change significantly across the study period, long-term impacts may be observed.
-Reductions in seagrass coverage in Sarasota Bay were observed as a result of the 2018 red tide event that persisted over a year and half in southwest Florida.
-These changes may have resulted from a reduction in water clarity from heavy concentrations of *K.
-brevis*, similar to what was observed in June and July in middle and lower Tampa Bay.
-Species shifts from macroalgal  to seagrass dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
+Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.
+Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern.
+Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
 In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
-The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality of *Caulerpa* dominance relative to seagrasses.
-General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communites.
-Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Managagement District.
+The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.
+General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
+Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Management District.
 From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
-These broader trends suggest drivers acting at a regional scale are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
+These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
 Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
 
-* Comparison to other locations/past events - Grand Bay, Bishop Harbor, Huelva estuary [@Perez10;@Perez16], Dillon report about Grand Bay [link](https://www.wrri.msstate.edu/pdf/2016dillon_finalreport.pdf)
-* Analysis limitations: no smoking gun but 2021 is an anomaly, additional info (benthic diversity TBD, nekton diversity TBD, large mammals, etc.), response-based monitoring may be biased
-* Long-term closure plan?  NGS-N is treated with spray evaporation system, but concentrations of TN/TP higher than NGS-S (see EO)/
-* Potential long-term impacts TBD
-* Current challenges in TB/southwest FL - OTB, seagrass loss (possibly linked to 2018 red tide, effects of Hurricanes [@Tomasko20]), red tide, climate change
-* Risk of decline (IRL ex.), regression of past progress
+Fisheries resources have undoubtedly been impacted by *K.
+brevis* blooms in June and July (Figure \@ref(fig:redtide)).
+Quantitative data on changes in abundance and diversity of fish species in Tampa Bay in 2021 are forthcoming.
+Routine sampling by the FWC Fisheries Independent Monitoring Program occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
+Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Schrandt21].
+Given the observed *K.
+brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
+For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
+Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
+brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
+
+Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.
+For benthic communities, sediments were sampled in April during the discharge near Port Manatee and surrounding waters.
+These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
+Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions.
+Results suggested a mix of conditions dominated by "fair" and "good" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
+Additional sampling is planned later at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
+Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
+This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass loss for southwest Florida.
+
+Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels to back-bay habitats (i.e., Bishop Harbor).
+The comparison to past years provided a stark contrast of the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
+The Tampa Bay Coastal Ocean Model [@Chen19] was used to simulate daily tracer particle movement and expected dilution of nutrient concentrations from the point of discharge.
+Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
+Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient for phytoplankton production.
+Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of inorganic nitrogen from stack effluent (Table \@ref(tab:stktab)) and the background concentrations for the bay (Table \@ref(tab:wqsumtab)).
+
+In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to environmental resources.
+In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.
+Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18].
+The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
+Massive fish kills were observed likely related to low pH of the effluent.
+Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
+Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
+A common theme is that legacy wastewater from fertilizer production is poorly maintained and long-term plans are insufficient to dispose of pollutants that can significantly affect coastal resources.
+These are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities in an environmentally responsible manner are needed at the local, national and international level.
+
+Limitations of our analysis are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.
+All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.
+However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
+The weight of evidence that 2021 is an anomaly relative to past years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
+Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more definite description of the recycling and fate of nitrogen from Piney Point.
+There may likely never be a definitive link, but this is not necessary to support future actions that ensure the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.
+These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
 

### 2021-09-12 manu-draft.Rmd
edit all to keep data from april to july only

@@
 load(file = here('tables/mcrtab.RData'))
+
+maxdt <- as.Date('2021-08-01')
 
@@
 Methods in the NADA R package [@Lee20] to account for observations below detection limits were used for all summary statistics.
-This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in August).
+This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
 Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].
@@
 seconb <- rswqdat %>% 
+  filter(date < maxdt) %>% 
   filter(var == 'secchi') %>% 
@@
 
-From April to August 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
+From April through July 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
 Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
@@
 Concentrations remained similar to baseline conditions until June and July when median values were often above the baseline range.
-Secchi observations in area 1 were below baseline ranges in April and June.
+Secchi observations in area 1 were below baseline ranges in April and July.
 Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 during weeks of May, June, and July.
@@
 visdat <- rstrndat %>% 
+  filter(date < maxdt) %>% 
   mutate(
@@
 foall <- rstrndat %>% 
+  filter(date < maxdt) %>% 
   mutate(
@@
   deframe
-fomoarea1aug <- fomoarea1 %>% 
-  filter(mo %in% 8) %>% 
+fomoarea1jul <- fomoarea1 %>% 
+  filter(mo %in% 7) %>% 
   select(taxa,foest) %>% 
@@
 
-A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April to August, each visited on average `r avevis` times per month.
+A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through July, each visited on average `r avevis` times per month.
 Turtle grass (*T.
@@
 The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
-Overall abundunce of seagrass did not change dramatically from April 22nd to August 9th.
+Overall abundunce of seagrass did not change dramatically from April 22nd through July.
 However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
@@
 *Lyngbya sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
-*Lyngbya sp.* peristed through June and July, but not was observed after July 20th.
+*Lyngbya sp.* persisted through June and July, but not was observed after July 20th.
 Green macroalgae taxa were first observed in July, although at generally low abundances.
-By August, macroalgae abundances were low.
 Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae.
@@
 Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in shoal grass (*H.
-wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, August frequency occurrence of `r fomoarea1aug['Halodule wrightii']`\%).
+wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea1jul['Halodule wrightii']`\%).
 A similar decline in coverage of shoal grass was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
 Frequency occurrence of manatee grass (*S.
-syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, August frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
+syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, July frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
 Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
@@
 Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
-Green macroalgae had the second lowest frequency occurrence, with the highest coverage of `r fomoarea1aug['Green']`\% observed in August in area 1.
-Brown macroalgae was only observed at one transect.
+Green macroalgae had the second lowest frequency occurrence, followed by Brown macroalgae that was only observed at one transect.
 
@@
 The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]` of the remaining variation.
-For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and strong negative loadings for total nitrogen, chl-a, and temperature.
+For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and secchi depth and strong negative loadings for total nitrogen, chl-a, and temperature.
 For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae.
@@
 The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.
-The third axis showed a strong positive loading with shoal grass (*H.
-wrightii*) and negative association with manatee grass (*S.
+The third axis showed a strong negative loading with shoal grass (*H.
+wrightii*) and positive loading with manatee grass (*S.
 filiforme*).
@@
 Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
-Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all, except $p < 0.05$ for pH) and negatively correlated with secchi depth ($p < 0.05$).
-Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$) and negatively correlated with salinity and secchi ($p < 0.05$).
-Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas both were negatively correlated with cyanobacteria ($p < 0.005$ for red, $p < 0.05$ for green).
-Among the seagrasses, manatee grass was negatively correlated wtih shoal grass and turtle grass ($p < 0.005$).
+Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$).
+Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$, except $p < 0.05$ for temperature) and negatively correlated with secchi ($p < 0.005$).
+Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas red macroalgae was negatively correlated with cyanobacteria ($p < 0.005$).
+Among the seagrasses, manatee grass was negatively correlated with shoal grass and turtle grass ($p < 0.005$).
 Several notable correlations were also observed between water quality, macroalgae, and seagrass.
-Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$).
-For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively associated with shoal grass ($p < 0.05$).
+Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).
+For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).
 For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
@@
   ) %>% 
-  filter(city %in% c('Tampa', 'St. Petersburg'))
+  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
+  filter(date < maxdt)
 fishann <- fishdat %>% 
@@
 The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L.
-Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined through August (Figure \@ref(fig:redtide)b).
+Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 

### 2021-09-12 R/figures.R

@@ -235,6 +235,7 @@ thm <-  theme(
 # combine water quality data with locations
 # removed secchi on bottom and non-detect for tn, chla
 wqdat <- rswqdat %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!station %in% nonbay) %>% 
   filter(!qual %in% c('S', 'U')) %>% 
@@ -366,6 +367,7 @@ cols <- c("#E16A86", "#50A315", "#009ADE")
 names(cols) <- levels(ppseg$area)
 
 datin <- rswqdat %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!qual %in% c('S', 'U')) # remove secchi on bottom, nondetect for chla, tn
 
@@ -387,11 +389,13 @@ trn <- 'S3T6b'
 rmdt <- as.Date('2021-04-07')
 
 mcrdat <- rstrndat %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(station %in% trn) %>% 
   filter(typ == 'mcr') %>% 
   filter(date != rmdt) %>% 
   mutate(taxa = fct_drop(taxa))
 savdat <- rstrndat %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(station %in% trn) %>% 
   filter(typ == 'sav') %>% 
   filter(date != rmdt) %>% 
@@ -459,6 +463,7 @@ areas <- ppseg %>%
 
 # add area
 trnsum <- rstrndat %>%
+  filter(date < as.Date('2021-08-01')) %>% 
   inner_join(rstrnpts, ., by = 'station') %>% 
   st_intersection(areas) %>% 
   st_set_geometry(NULL) %>%
@@ -534,6 +539,7 @@ areas <- ppseg %>%
 
 # add area
 trnsum <- rstrndat %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   mutate(
     date = floor_date(date, unit = 'week')
   ) %>% 
@@ -563,6 +569,7 @@ trnsum <- rstrndat %>%
 rswqsum <- rswqdat %>% 
   filter(var %in% vrs) %>% 
   filter(source == 'fldep') %>%
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(!station %in% nonbay) %>% 
[…]

### 2021-09-12 R/funcs.R

@@ -115,7 +115,7 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
       axis.title.x = element_blank(),
       panel.grid.minor = element_blank(),
       strip.text = element_text(size = 14), 
-      axis.text.x = element_text(size = 6.25, angle = 45, hjust = 1)
+      axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
     )
   
   if(logtr)

### 2021-09-12 R/tables.R

@@ -134,6 +134,7 @@ sigs <- parms %>%
 totab <- rswqdat %>% 
   left_join(rsstatloc, ., by = c('station', 'source')) %>% 
   filter(var %in% vrs) %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(!station %in% nonbay) %>% 
   select(station, date, lbs, val, inrng, qual) %>% 
   st_intersection(ppsegbf) %>% 
@@ -209,6 +210,7 @@ ppsegbf <- ppseg %>%
 
 rswqsub <- rswqdat %>% 
   filter(var %in% vrs) %>% 
+  filter(date < as.Date('2021-08-01')) %>% 
   filter(source == 'fldep') %>%
   filter(!station %in% nonbay) %>% 
   filter(!qual %in% 'S') %>% # remove secchi on bottom
@@ -340,6 +342,7 @@ areas <- ppseg %>%
 
 # add area
 trnsf <- rstrndat %>%
+  filter(date < as.Date('2021-08-01')) %>% 
   inner_join(rstrnpts, ., by = 'station') %>% 
   filter(taxa %in% c(mcrsel, savsel)) %>% 
   # filter(station %in% smpeff) %>% 

### 2021-09-13 manu-draft.Rmd
personal edits to whole draft

@@
 An immediate phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass estimated at 0.5 kg/m2 at some locations.
+Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass peaking in June at 48\% frequency occurrence for some locations.
 Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
@@
 
-*Key words*: nitrogen, phosphate mining, Tampa Bay, wastewater, water quality
+*Key words*: nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality
 
@@
 The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18g].
+Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
 Tampa Bay is the largest estuary in Florida located in a heavily urbanized watershed of nearly 3 million individuals.
@@
 Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program through implementation of a science-based resource management plan for the Bay [@Ohara17].
-The Tampa Bay Estuary Program and its partners has been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
+The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
 
@@
 Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
-The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18].
+The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
 There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
-Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18].
+Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste to prevent unforeseen impacts to the environment.
+Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack located less than two miles from the shore of Tampa Bay and near two Florida aquatic preserves [@Henderson04].
@@
 Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
-Recently, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
-At that time, approximately 480 million gallons of legacy phosphate mining wastewater was being held in the failing stack.
+Recently, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+At that time, approximately 480 million gallons of legacy phosphate mining wastewater was being held in the failing gypstack.
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay over five months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
+This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
 The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
 We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
-A timeline of events in 2021 is also provided, which is supported by the results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
+A timeline of events in 2021 is also provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
 The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.
-We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of long-term management goals that leverage resources from existing partnerships among local resource management institutions.
+We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
 
@@
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to reduce the likelihood of an uncontrolled spill.
+From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
 @Switzer11 reported minimal impacts to nekton communities, although an increase in macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* was observed as a potential indication of nutrient eutrophication.
-Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on holding capacity of storage ponds [@Hu03].
+Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
 Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
@@
 In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing, gypsumstack holding pond at Piney Point.
+This material was added to an existing gypstack at Piney Point.
 Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
@@
 Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged at Berth 12 in Port Manatee.
+Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
 This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
@@
 Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, which included water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
 Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted, but they are not reported here in anticipation of future monitoring events.
@@
 For this paper, we focus on parameters related to the nutrient management paradigm for the bay and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orhophosphate (PO$_4^{3-}$), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) are also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
+This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
+Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
 Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts of the discharge relative to Piney Point.
-Values below laboratory detection limits (or Secchi values on the bottom) were assessed prior to analysis.
-Appropriate analysis methods to include parameters with observations below detection are described below.
+These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
 
@@
 Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
-Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* and *Pyrodinium bahamense* that can occur in the bay depending on salinity and temperature conditions during the growing season.
-Occurrence of both these species has historically been spatially distinct, with *K.
-brevis* originating in the Gulf of Mexico and occurring in higher salinity portions of the bay, whereas *P.
-bahamense* has been observed consistently each year since 2008 in Old Tampa Bay (northwest segment) during the summer.
+Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
+Occurrence of this species has historically been spatially distinct, with blooms originating in the Gulf of Mexico and occasionally occurring at bloom concentrations in lower Tampa Bay.
 Data for *K.
@@
 This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
-For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference concentrations relative to results at the closest monitoring station that was sampled in response to Piney Point.
-Methods in the NADA R package [@Lee20] to account for observations below detection limits were used for all summary statistics.
+For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
 This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
@@
 In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
-These long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
+Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
 
@@
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and can be evaluated with methods described below.
+Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.
 Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 Concentrations of *K.
-brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay by August.
+brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.
 The remainder of this section is quantitative description these events.
@@
 Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
-The percent of observations outside of the normal range varied by location and parameter.
+The percentage of observations outside of the normal range varied by location and parameter.
 For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
 Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.
-Secchi observations for the period of observation were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.
+Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.
 Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).
-Physical parameters (salinity, temperature) were generally within range over the five month period.
+Physical parameters (salinity, temperature) were generally within range over the study period.
 Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
@@
 ```{r}
-ntrns <- length(unique(rstrndat$station))
-ntrns <- english(ntrns)
+ntrnscnt <- length(unique(rstrndat$station))
+ntrns <- english(ntrnscnt)
 visdat <- rstrndat %>% 
@@
 wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
-Widgeon grass (*R.
-maritima*) and star grass (*H.
-engelmanni* and *H.
-decipiens*) were uncommon,  with frequency occurrences of `r foall['Ruppia maritima']`\% and `r foall['Halophila sp']`\%.
 The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].
@@
 
-Temporal progression of seagrassess and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
+Temporal progression of seagrasses and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
 Transect S3T6 is located less than one kilometer to the north of Port Manatee.
 The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
-Overall abundunce of seagrass did not change dramatically from April 22nd through July.
+Overall abundance of seagrass did not change dramatically from April 22nd through July.
 However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
@@
 Green macroalgae taxa were first observed in July, although at generally low abundances.
-Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae.
-However, overall macroalgae trends, including June dominance of cyanobacteria was observed at nearly all transects that suggests a broader impact from wastewater discharge at Piney Point.
+Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae by mid-summer.
 
@@
 
-## Water quality compared to seagrass and macrolagae
+## Water quality compared to seagrass and macroalgae
 
@@
 A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).
-The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]` of the remaining variation.
+The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]`\% of the remaining variation.
 For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and secchi depth and strong negative loadings for total nitrogen, chl-a, and temperature.
@@
 filiforme*).
-Observations by area showed that samples in area 3 were grouped by positive loadings on on the first PCA axis and negative loadings on the second PCA axis and were distinct from points in area 1.
+Observations by area showed that samples in area 3 were distinct from points in area 1 and grouped by positive loadings on the first PCA axis and negative loadings on the second PCA axis.
 
@@
 Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).
-For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated to turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).
+For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated with turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).
 For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
@@
 Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.
-In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` reports were made in Tampa (Figure \@ref(fig:redtide)d).
+In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).
 The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K.
 brevis* cell concentrations (Figure \@ref(fig:redtide)b).
-Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed over nearly five months.
+Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.
 Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.
@@
 Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper.
-These associations may also reflect natural variation in water clarity along the main axis of Tampa Bay.
+In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.
 Shoal grass can occur in more turbid waters of oligohaline areas.
@@
 In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.
-The dominance of the red groups early in the period of record followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.
-However past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
+The dominance of the red groups early in the summer followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.
+Past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
 Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
@@
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
+Regionally, seagrasses have declined in coverage as estimated from biennial maps provided by the Southwest Florida Water Management District.
+From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
+These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
 Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.
@@
 Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
-In particular, increasing abundance of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
+In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
 The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.
 General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Regionally, seagrasses have declined in coverage as estimated from biennal maps provided by the Southwest Florida Water Management District.
-From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
-These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
 Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
@@
 Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
-This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass loss for southwest Florida.
+This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
 
@@
 In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.
-Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18].
+Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18b].
 The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
@@
 A common theme is that legacy wastewater from fertilizer production is poorly maintained and long-term plans are insufficient to dispose of pollutants that can significantly affect coastal resources.
-These are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities in an environmentally responsible manner are needed at the local, national and international level.
+These are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities are needed at the local, national and international level.
 
@@
 
+We thank the many TBEP partners and collaborators for their continuing efforts to restore and monitor Tampa Bay.
+We thank the administrative staff, field crews, and laboratory staff from the Florida Department of Environmental Protection, Environmental Protection Commission of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission, City of St.
+Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
+Our partners’ willingness to adapt and implement innovative monitoring and management actions in response to Piney Point and the ever evolving challenges threatening Tampa Bay is greatly appreciated.
+
 # Figures
@@
 
-```{r timeline, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting Mrach 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
+```{r timeline, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting March 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
 knitr::include_graphics(here('figs/timeline.jpeg'))
@@
 Color ramps and point sizes show relative values (reversed for Secchi).
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location (Figure \\@ref(fig:map)a).
 Values below detection limits (or secchi on bottom) are not shown.'}
@@
 
-```{r wqtrnds, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure \\@ref(fig:map)a.
 Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure \\@ref(fig:map)a).
 Values below detection limits (or secchi on bottom) are not shown.'}
@@
 
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 2 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = ', ntrns, ') near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 3 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
 Estimates are grouped by sample months in 2021.
@@
 
-```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
+```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
 Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
@@
 
-```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrasses by week for April to July 2021 in response to wastewater discharge from Piney Point.
+```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
 Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
@@
 Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.
-Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay collected monthly by the Environmental Protection Commission of Hillsborough County.
-Effluent concentrations were taken from two samples on April 6th and 13th at the outflow at Port Manatee.
-Averages were taken when two measured values were available from both effluent sample dates.
+Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay (Figure \\@ref(fig:map)a).
+End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
 Missing values were not measured in the effluent or stack water.'
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected from March to July 2021 in response to wastewater discharge from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from March to July 2021 in response to wastewater discharge from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
-Summaries are median, maximum, and minimum values.
+Summaries are median, minimum, and maximum values.
 Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
@@
 Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
-Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
+Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different between month pairs.
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
@@
 Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
-Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
+Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs.
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
@@
 Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
-Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
+Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs.
 Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.

### 2021-09-17 manu-draft.Rmd
adding co-authors

@@
 author: "Marcus W.
-Beck (mbeck@tbep.org), Maya C.
-Burke (mburke@tbep.org), Emma E.
+Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Maya C.
+Burke (mburke@tbep.org), Diana W.
+Chin (diwchin@gmail.com), Emma E.
 Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T.
-Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Joe Whalen (jwhalen@tbep.org)"
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue

### 2021-09-19 R/figures.R
fix figure 9a/9b to 25th/75th

@@ -930,9 +930,9 @@ toplo <- kbrdat %>%
   summarise(
     cnt = n(),
     y0 = min(val, na.rm = T), 
-    y10 = quantile(val, prob = 0.25, na.rm = T),
+    y25 = quantile(val, prob = 0.25, na.rm = T),
     y50 = quantile(val, prob = 0.5, na.rm = T),
-    y90 = quantile(val, prob = 0.75, na.rm = T),
+    y75 = quantile(val, prob = 0.75, na.rm = T),
     y100 = max(val, na.rm = T),
     .groups = 'drop'
   ) %>%
@@ -943,7 +943,7 @@ toplo <- kbrdat %>%
 # plot
 p1 <- ggplot(toplo, aes(x = yr)) +
   geom_boxplot(
-    aes(ymin = y0, lower = y10, middle = y50, upper = y90, ymax = y100),
+    aes(ymin = y0, lower = y25, middle = y50, upper = y75, ymax = y100),
     stat = "identity", width = 0.75, fill = '#00806E'
   ) +
   scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
@@ -975,9 +975,9 @@ toplo <- kbrdat %>%
   summarise(
     cnt = n(),
     y0 = min(val, na.rm = T), 
-    y10 = quantile(val, prob = 0.1, na.rm = T),
+    y25 = quantile(val, prob = 0.25, na.rm = T),
     y50 = quantile(val, prob = 0.5, na.rm = T),
-    y90 = quantile(val, prob = 0.9, na.rm = T),
+    y75 = quantile(val, prob = 0.75, na.rm = T),
     y100 = max(val, na.rm = T),
     .groups = 'drop'
   ) %>%
@@ -986,7 +986,7 @@ toplo <- kbrdat %>%
 # plot
 p2 <- ggplot(toplo, aes(x = week)) +
   geom_boxplot(
-    aes(ymin = y0, lower = y10, middle = y50, upper = y90, ymax = y100),
+    aes(ymin = y0, lower = y25, middle = y50, upper = y75, ymax = y100),
     stat = "identity", width = 0.75, fill = '#00806E'
   ) +
   scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE))) +

### 2021-09-21 manu-draft.Rmd
fix figure 9a/9b to 25th/75th
added CL co-author

@@
 Chin (diwchin@gmail.com), Emma E.
-Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
+Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T.
@@
 Petersburg) (c) by year and (d) by week in 2021.
-Red tide concentrations show minimum, tenth percentile, median, 90th percentile, and maximum for each year or week for middle and lower Tampa Bay.
+Red tide concentrations show minimum, 25th percentile, median, 75th percentile, and maximum for each year or week for middle and lower Tampa Bay.
 *K.

### 2021-09-28 manu-draft.Rmd
Update manu-draft.Rmd

@@
 An immediate phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass peaking in June at 48\% frequency occurrence for some locations.
+Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
 Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
@@
 
-*Key words*: nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality
+*Key words*: algae, nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality
 
@@
 Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Tampa Bay is the largest estuary in Florida located in a heavily urbanized watershed of nearly 3 million individuals.
+Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3 million individuals.
 Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].
@@
 Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.
-While point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been much reduced, non-point source loads from wastewater sources and stormwater runoff are estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
+Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
 Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
@@
 
-Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources worldwide [@Tayibi09;@Hudson11].
+Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
 Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
@@
 Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
-The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack located less than two miles from the shore of Tampa Bay and near two Florida aquatic preserves [@Henderson04].
+The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
 Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FDEP).
@@
 A timeline of events in 2021 is also provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
-The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region.
+The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region as well as a potential template for future emergency responses.
 We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
@@
 
-Legacy wastewater discharges from Piney Point did not occur again until 2021, although onsite stormwater management and discharges have occurred throughout its history.
+Legacy wastewater discharges from Piney Point were not reported again until 2021, although onsite stormwater management and discharges have occurred throughout its history.
 Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
@@
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-Additional samples for contaminants (e.g., heavy metals), benthic sediment, and nekton surveys were also conducted, but they are not reported here in anticipation of future monitoring events.
+Additional sampling for contaminants (e.g., heavy metals), benthic invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
 
@@
 brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
-Because of the increased occurrence of red tide samples in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
+Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
 Petersberg) impacted by the event.
@@
 Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point starting in April.
-Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay.
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along a fifty meter transect at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
-Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 50 cm$^2$ quadrat at 10m distances along each transect.
+Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@sherwood17].
+Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.
 Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.

### 2021-10-04 manu-draft.Rmd
Update manu-draft.Rmd
Some additional edits.

@@
 Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
-Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Ed T.
+Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T.
 Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
@@
 An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
-An immediate phytoplankton response was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
+An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
 Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
@@
 Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3 million individuals.
+Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3.1 million individuals.
 Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].
@@
 
-Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite historical gains in environmental recovery.
+Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
 Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
@@
 From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
-@Switzer11 reported minimal impacts to nekton communities, although an increase in macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* was observed as a potential indication of nutrient eutrophication.
+@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
 Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
@@
 The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
-HRK maintains ownership and responsibility of the site to present day, with oversight by FDEP.
+HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
 
@@
 Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
-For this paper, we focus on parameters related to the nutrient management paradigm for the bay and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
+For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
 This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
+Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
 Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts of the discharge relative to Piney Point.
+As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.
 These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
@@
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
-The initial diatom bloom did not persist past April.
+The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.
 On April 20th, *K.
@@
 brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.
-The remainder of this section is quantitative description these events.
+A quantitative description of these events follows.
 

### 2021-10-04 manu-draft.Rmd
Update manu-draft.Rmd
Some additional edits to the results and discussion.

@@
 brevis* from April to July to bloom concentrations exceeding 10k cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10k cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels.
-Median cell concentrations for most years were well below 1k cells/L.
-The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L.
+The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels.
+Median cell concentrations for most years were well below 1,000 cells/L.
+The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L [@Flaherty11].
 Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
@@
 Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
-Hammer Levy, City of St.
-Petersburg, pers.
+Hammer Levy, Pinellas County, pers.
 comm.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay in June and July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
 The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
@@
 
-In addition to the changing environmental conditions observed during the summer, the Piney Point event is also anomalous in the volume of effluent entering the system and estimated nutrient load to lower Tampa Bay.
+In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen effluent directly into lower Tampa Bay.
 Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
 An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
-Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of the external load into lower Tampa Bay [@tbep0417].
-From both a monitoring perspective and the nature of the effluent into Tampa Bay, the observed response of conditions in 2021 is an outlier compared to past years.
+Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
+From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years.
 These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
@@
 
-From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability that may negatively affect seagrass growth.
-However, changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous.
+From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the Summer 2021 in comparison to prior years.
+Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.
+Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apaprent over the initial 4 month sampling period.
 Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.
@@
 Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
-A clear difference in the macroalgae community in 2021 compared to past years was a high abundance of cyanobacteria macroalgae (i.e., *Lyngbya sp.*).
-Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee was observed in 2021.
+A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Lyngbya sp.*).
+Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee were observed in 2021.
 The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
@@
 General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health and restoration activities.
 
-Fisheries resources have undoubtedly been impacted by *K.
+Fisheries resources were undoubtedly impacted by *K.
 brevis* blooms in June and July (Figure \@ref(fig:redtide)).
-Quantitative data on changes in abundance and diversity of fish species in Tampa Bay in 2021 are forthcoming.
-Routine sampling by the FWC Fisheries Independent Monitoring Program occurs annually in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
-Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Schrandt21].
+Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
+Routine sampling by the FWC Fisheries Independent Monitoring Program occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
+Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
 Given the observed *K.
@@
 
-Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels to back-bay habitats (i.e., Bishop Harbor).
+Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels, rivers or other back-bay habitats (e.g., Alafia River and Bishop Harbor).
 The comparison to past years provided a stark contrast of the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
@@
 Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
-Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient for phytoplankton production.
-Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of inorganic nitrogen from stack effluent (Table \@ref(tab:stktab)) and the background concentrations for the bay (Table \@ref(tab:wqsumtab)).
+Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
+Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of labile inorganic nitrogen forms from stack effluent (Table \@ref(tab:stktab)) relative to background concentrations for the bay (Table \@ref(tab:wqsumtab)).
 
@@
 The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
-Massive fish kills were observed likely related to low pH of the effluent.
+Massive fish kills were observed and likely related to low pH of the effluent.
 Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
 Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
-A common theme is that legacy wastewater from fertilizer production is poorly maintained and long-term plans are insufficient to dispose of pollutants that can significantly affect coastal resources.
-These are not isolated examples and sustained attention and regulatory oversight on efforts to close these facilities are needed at the local, national and international level.
+Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.
+These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.
 
-Limitations of our analysis are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.
+Limitations of our analyses are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.
 All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.
 However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-The weight of evidence that 2021 is an anomaly relative to past years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
+The weight of evidence that 2021 is an anomaly relative to recent prior years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more definite description of the recycling and fate of nitrogen from Piney Point.
-There may likely never be a definitive link, but this is not necessary to support future actions that ensure the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.
+There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.
 These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.

### 2021-10-05 manu-draft.Rmd
Initial edits and comments

@@
 An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Macroalgae blooms of cyanobacteria (*Lyngbya* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
-Though present south of Tampa Bay prior to the event, blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
+Cyanobacteria blooms (*Dapis* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
+Blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
 Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
@@
 Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
-Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
-Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
-Many of these challenges are addressed by ongoing efforts of the US EPA National Estuary Program through implementation of a science-based resource management plan for the Bay [@Ohara17].
-The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources [@Greening14;@Greening16].
+The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources through implementation of a science-based resource management plan for the Bay [@Greening14;@Greening16].
 
@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
-The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FDEP).
-Decreasing holding capacity of the ponds with seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
+Currently, seventeen phosphogypsum stacks, or gypstacks, (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
+The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
+Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC, with oversight by the Florida Departmental of Environmental Protection (FDEP).
+Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
 Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.
@@
 
-## History of Piney Point
-
-The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
-Port operations were primarily for export of phosphate production by the plant.
-Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
-Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
-In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
-
-The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
-Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
-Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
-comm.
-D.
-Eckenrod to USEPA, Nov.
-28, 2001).
-Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
-@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
-Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
-
-HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.
-This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
-In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing gypstack at Piney Point.
-Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
-The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
-HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
+## Background
 

### 2021-10-05 manu-draft.Rmd
Initial edits part 2

@@
 By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks, or gypstacks, (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
+Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
@@
 Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
-Recently, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
-At that time, approximately 480 million gallons of legacy phosphate mining wastewater was being held in the failing gypstack.
+
+Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
+In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
+At that time, approximately 480 million gallons of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.
+Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
+Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
+This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
+From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
+Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
 
@@
 # Methods
-
-## Background
-
-Legacy wastewater discharges from Piney Point were not reported again until 2021, although onsite stormwater management and discharges have occurred throughout its history.
-Leakages from a tear in the plastic liner of the southern holding pond (NGS-S) were suspected when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
-The NGS-S holding pond held 480 million gallons of wastewater, as a mixture of remnant process water from phosphate production and seawater from port dredging operations.
-Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
-Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
-This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
-From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
-Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
 
@@
 brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
-Also during May, *Lyngbya sp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
-*Lyngbya* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
+Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
+*Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
 By June 27th, fish kill reports attributed to red tide increased with *K.
@@
 Macroalgae observed along the transects also varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
-Common taxa in the red group included *Gracilaria sp.* and *Acanthophora sp.*.
+Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.
 Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
-Common taxa in the green group included *Ulva sp.* and *Caulerpa sp.*, whereas *Lyngbya sp.* was the only cyanobacteria macroalgae observed.
+Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
 Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
@@
 Red macroalgae were present in high abundances from April to May.
-*Lyngbya sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
-*Lyngbya sp.* persisted through June and July, but not was observed after July 20th.
+*Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
+*Dapis sp.* persisted through June and July, but not was observed after July 20th.
 Green macroalgae taxa were first observed in July, although at generally low abundances.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of the macroalgae *Lyngbya sp.* in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
 The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
@@
 This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria sp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva sp.*, *Caulerpa sp.*) and seasonally occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in *Ulva sp.* likely reflects a natural phenology in Tampa Bay.
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.
+The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* likely reflects a natural phenology in Tampa Bay.
 Past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
 Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
-A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Lyngbya sp.*).
-Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Lyngbya sp.* in Anna Maria Sound and near Port Manatee were observed in 2021.
+A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Dapis sp.*).
+Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Dapis sp.* in Anna Maria Sound and near Port Manatee were observed in 2021.
 The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
-*Lyngbya sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
+*Dapis sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
 
@@
 Sample dates in 2021 are shown in rows with transect meter results shown in columns.
-Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Lyngbya sp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva sp.*) increasing in July.
+Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis sp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
 Abundances are Braun-Blanquet coverage estimates.'}

### 2021-10-06 manu-draft.Rmd
minor edits to history

@@
 Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
+Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
+Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
 The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources through implementation of a science-based resource management plan for the Bay [@Greening14;@Greening16].
@@
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of 215 million gallons of legacy phosphate mining wastewater in April, 2021.
+This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of legacy phosphate mining wastewater in April, 2021.
 The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
@@
 # Methods
+
+## History of Piney Point
+
+The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
+Port operations were primarily for export of phosphate production by the plant.
+Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
+Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
+In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
+
+The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
+Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
+Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
+Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
+comm.
+D.
+Eckenrod to USEPA, Nov.
+28, 2001).
+Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
+From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
+Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
+Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
+
+HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.
+This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
+In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
+This material was added to an existing gypstack at Piney Point.
+Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
+The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
+HRK maintains ownership and responsibility of the site to present day with oversight by FDEP
 
@@
 Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point starting in April.
-Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@sherwood17].
+Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
 Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.

### 2021-10-07 manu-draft.Rmd
update seagrass macro fo plot

@@
 Estimates are grouped by sample months in 2021.
-Frequency occurrences are absolute for each taxa based on presence/absence.')}
+Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))

### 2021-10-07 R/figures.R

@@ -431,7 +431,8 @@ savcol <- savcol[savsel]
 mcrcol <- c('tomato1', 'lightgreen', 'burlywood3', 'lightblue')
 names(mcrcol) <- mcrsel
 mcrcol <- mcrcol[mcrsel]
-cols <- c(savcol, mcrcol)
+cols <- c(mcrcol, savcol)
+cols <- c(cols, Total = 'white')
 
 # segments
 areas <- ppseg %>% 
@@ -463,7 +464,6 @@ areas <- ppseg %>%
 
 # add area
 trnsum <- rstrndat %>%
-  filter(date < as.Date('2021-08-01')) %>% 
   inner_join(rstrnpts, ., by = 'station') %>% 
   st_intersection(areas) %>% 
   st_set_geometry(NULL) %>%
@@ -473,28 +473,46 @@ trnsum <- rstrndat %>%
   ) %>%  
   mutate(
     date = floor_date(date, unit = 'month'), 
-    typ = factor(typ, levels = c('sav', 'mcr'), labels = c('Seagrasses', 'Macroalgae')), 
-    taxa = factor(taxa, levels = c(savsel, mcrsel)), 
-    area = forcats::fct_drop(area), 
-    area = factor(area, levels = c('Area 1', 'Area 3'), labels = c('(a) Area 1', '(b) Area 3'))
+    typ = factor(typ, levels = c('mcr', 'sav'), labels = c('Macroalgae', 'Seagrass'))
   ) %>% 
   group_by(area, typ, date, taxa) %>% 
   summarize(
-    foest = sum(pa) / length(pa), 
-    .groups = 'drop'
+    foest = sum(pa) / length(pa)
   ) %>% 
   filter(taxa %in% c(mcrsel, savsel))
 
-p <- ggplot(trnsum, aes(x =date, y = foest, fill = taxa)) + 
-  geom_bar(pch = 16, stat = 'identity', color = 'grey', alpha = 0.8) +
-  facet_grid(typ ~ area, scales = 'free') +
+trnsumtots <- rstrndat %>%
+  inner_join(rstrnpts, ., by = 'station') %>% 
+  st_intersection(areas) %>% 
+  st_set_geometry(NULL) %>%
+  dplyr::group_by(area, typ, date, station, location) %>%
+  dplyr::summarise(
+    pa = as.numeric(any(bb > 0))
+  ) %>%  
+  mutate(
+    date = floor_date(date, unit = 'month'), 
[…]

### 2021-10-08 manu-draft.Rmd
adding co-authors

@@
 author: "Marcus W.
-Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Maya C.
+Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C.
 Burke (mburke@tbep.org), Diana W.
 Chin (diwchin@gmail.com), Emma E.
-Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
+Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
+Phlips (phlips@ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T.
-Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue

### 2021-10-08 manu-draft.Rmd
line 72 revised to untreated wastewater

@@
 Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
+Unlike past discharges from the site, HRK was authorized to release untreated wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
 This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.

### 2021-10-08 manu-draft.Rmd
scriveners corrections to lines 88 & 90

@@
 The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
-HRK maintains ownership and responsibility of the site to present day with oversight by FDEP
+HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
 
@@
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.
-These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
+These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
 

### 2021-10-08 manu-draft.Rmd
add Pinellas County to line 335
Pin Co removed the majority of the fish, even in St. Pete
scriveners edits line 345

@@
 Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.
-As a result, the city of St.
+As a result, Pinellas County and the city of St.
 Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
@@
 
-From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the Summer 2021 in comparison to prior years.
+From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years.
 Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.
-Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apaprent over the initial 4 month sampling period.
+Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.
 Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.

### 2021-10-11 manu-draft.Rmd
adding co-authors
edits to benthic monitoring

@@
 Beck (mbeck@tbep.org), Andrew Altieri (andrew.altieri@essie.ufl.edu), Christine Angelini (christine.angelini@essie.ufl.edu), Maya C.
-Burke (mburke@tbep.org), Diana W.
+Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W.
 Chin (diwchin@gmail.com), Emma E.
@@
 These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
-Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions.
-Results suggested a mix of conditions dominated by "fair" and "good" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions.
+Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
 Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Additional sampling is planned later at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Additional sampling occurred in September at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
 Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.

### 2021-10-11 manu-draft.Rmd
Edits to benthic section (line 353)

@@
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-Additional sampling for contaminants (e.g., heavy metals), benthic invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
+Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
 
@@
 Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments were sampled in April during the discharge near Port Manatee and surrounding waters.
+For benthic communities, sediments were sampled near Port Manatee and surrounding waters in April 2021 during the discharge and in September 2021 to further assess potential impacts of Piney Point on sediment quality.
 These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
@@
 Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
-Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Additional sampling occurred in September at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Comparison of the April 2021 samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
+Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples.
 Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.

### 2021-10-11 manu-draft.Rmd
Address some abbreviations.
Edited some instances of abbreviations. The Environmental Protection Commission of Hillsborough County typically prefers EPCHC, as their abbreviation. But, I didn't modify from what you have included.

@@
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC, with oversight by the Florida Departmental of Environmental Protection (FDEP).
+Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).
 Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
@@
 
-HRK Holdings, LLC (hereafter, HRK) acquired Piney Point in August 2006 through an administrative agreement with FDEP.
+HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.
 This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
@@
 
-Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the Environmental Protection Commission of Hillsborough County.
+Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the EPC.
 These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.

### 2021-10-11 manu-draft.Rmd
nit picky time frame for seagrass/macroalgae monitoring
adding co-authors

--- 2021-10-11 manu-draft.Rmd
@@
 Chin (diwchin@gmail.com), Emma E.
-Dontis (edontis@co.pinellas.fl.us), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
+Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
 Phlips (phlips@ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
@@
 These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
-Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions.
-Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "good", "fair", and "poor" as a qualitative summary of enumerated taxonomic data that can support management decisions.
+Results suggested a mix of conditions dominated by "fair" and "good" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
 Comparison of the initial 2021 samples to historical benthic conditions suggested relatively consistent sediment quality from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Additional sampling occurred in September at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
+Additional sampling is planned later at the same sites surveyed in April to further assess potential impacts of Piney Point on sediment quality.
 Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.

### 2021-10-11 manu-draft.Rmd

--- 2021-10-11 manu-draft.Rmd
@@
 
-Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point starting in April.
+Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].

### 2021-10-13 manu-draft.Rmd
precip to redtide figure, update figure caption

@@
 
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) and number of fish kill reports for the contiguous record showing cell concentrations (a) by year and (b) by week in 2021 and reported fish kills by city (Tampa, St.
-Petersburg) (c) by year and (d) by week in 2021.
-Red tide concentrations show minimum, 25th percentile, median, 75th percentile, and maximum for each year or week for middle and lower Tampa Bay.
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, and (e) wind rose plots for 2021 before, passage (two days pre/post), and after tropical storm Elsa.
+Red tide concentrations show minimum, 25th percentile, median, 75th percentile, and maximum for each year or week for middle Tampa Bay.
 *K.
 brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
-brevis* (https://public.myfwc.com/FWRI/FishKillReport/).'}
+brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, and wind rose plots are from St.
+Petersburg, Florida.
+Wind roses show relative counts of six minute observations in directional (30 degree bins) and speed (m/s) categories.'}
 knitr::include_graphics('figs/redtide.jpeg')

### 2021-10-13 R/figures.R

@@ -39,6 +39,7 @@ data(parms)
 data(rstrnpts)
 data(kbrdat)
 data(winddat)
+data(raindat)
 
 source(here('R/funcs.R'))
 
@@ -915,17 +916,17 @@ fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>%
       T ~ city
     )
   ) %>% 
-  filter(date < as.Date('2021-08-01'))
+  filter(date < as.Date('2021-10-01'))
 
 # k brevis data from CL
 habdat <- read.csv(here('data-raw/KB_LowMid_1995-2021.csv')) %>% 
-  # filter(Segment %in% c('MTB', 'LTB')) %>% 
+  filter(Segment %in% c('MTB')) %>%
   select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
   mutate(date = mdy(date))
 
 # add k brevis data from HABSOS that are collected after last date in habdat
 kbrdat <- kbrdat %>% 
-  .[tbseg[tbseg$bay_segment %in% c('MTB', 'LTB'), ], ] %>% 
+  .[tbseg[tbseg$bay_segment %in% c('MTB'), ], ] %>% 
   filter(var == 'kb') %>% 
   filter(date > max(habdat$date)) %>% 
   mutate(
@@ -948,12 +949,12 @@ weeklv <- seq.Date(from = as.Date('2021-01-01'), to = Sys.Date(), by = 'days') %
     lb = format(dt, '%b %d')
   ) %>%
   filter(yr > 2020) %>% 
-  filter(mo < 8) %>% 
+  filter(mo < 10) %>% 
   pull(lb)
 
 # habdat p1
 toplo <- habdat %>%
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   mutate(
     dtgrp = quarter(date),
     yr = year(date)
@@ -990,13 +991,13 @@ p1 <- ggplot(toplo, aes(x = yr)) +
   labs(
     x = 'Year',
     y = 'Cells (100k / L)',
-    title = expression(paste('(a) ', italic('K. brevis'), ' concentrations by year, middle/lower Tampa Bay'))
+    title = expression(paste('(a) ', italic('K. brevis'), ' concentrations by year, middle Tampa Bay'))
   )
 
 # habdat p2
 toplo <- habdat %>%
[…]

### 2021-10-13 R/funcs.R

@@ -446,7 +446,7 @@ plot.windrose <- function(data,
   p.windrose <- ggplot(data = data,
                        aes(x = dir.binned,
                            fill = spd.binned)) +
-    geom_bar() + 
+    geom_bar(color = 'lightgrey') + 
     scale_x_discrete(drop = FALSE,
                      labels = waiver()) +
     coord_polar(start = -((dirres/2)/360) * 2*pi) +

### 2021-10-13 manu-draft.Rmd
started supplement figures

@@
 Petersburg, Florida.
-Wind roses show relative counts of six minute observations in directional (30 degree bins) and speed (m/s) categories.'}
+Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
 knitr::include_graphics('figs/redtide.jpeg')

### 2021-10-13 supplement.Rmd
started supplement figures

--- 2021-10-11 manu-draft.Rmd
@@
 ---
-title: "Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
+title: "Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
 output: 
@@
   })
-
-box::use(
-  here[here],
-  dplyr[...],
-  tidyr[fill],
-  english[english], 
-  lubridate[month,year,mdy,floor_date], 
-  tibble[deframe]
-)
-
-data(rswqdat)
-data(rstrnpts)
-data(rstrndat)
-data(trnsum)
-data(allpcadat)
-
-load(file = here('tables/wqsumtab.RData'))
-load(file = here('tables/stktab.RData'))
-load(file = here('tables/wqcmptab.RData'))
-load(file = here('tables/savtab.RData'))
-load(file = here('tables/mcrtab.RData'))
-
-maxdt <- as.Date('2021-08-01')
-
-# # extract bib entries from online
-# bib_scrp('manu_draft.Rmd', 'refs.bib')
 ```
 
-`r paste('Last manuscript build', Sys.time())`
-
-```{r echo = F, cache = F, eval = F}
-spelling::spell_check_files('manu_draft.Rmd')
+```{r windroses, fig.cap = 'Wind rose plots for 2021 by month.
+Data are from St.
+Petersburg, Florida.
+Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
+knitr::include_graphics('figs/windroses.jpeg')
 ```
-
-# Abstract
-
-From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
-An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Cyanobacteria blooms (*Dapis* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
-Blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
-Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
-Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years.
-
-*Key words*: algae, nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality
-
-# Introduction
-
-Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.
-The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3.1 million individuals.
-Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].
-Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
-
-Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
-Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
-Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
-Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
-The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources through implementation of a science-based resource management plan for the Bay [@Greening14;@Greening16].
-
-Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
-Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
-Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
-The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
-There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
-Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
-
-The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].
-By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
-The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).
-Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
-Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.
-Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
-
-Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
-In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
-At that time, approximately 480 million gallons of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.
-Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
-Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release untreated wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
-This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
-From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
-Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
-
-This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of legacy phosphate mining wastewater in April, 2021.
-The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
-We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
-A timeline of events in 2021 is also provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
-The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region as well as a potential template for future emergency responses.
-We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
-
-# Methods
-
-## History of Piney Point
-
-The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
-Port operations were primarily for export of phosphate production by the plant.
-Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
-Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
-In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
-
-The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
-Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
-Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
-comm.
-D.
-Eckenrod to USEPA, Nov.
-28, 2001).
-Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
-@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
-Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
-
-HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.
-This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
-In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing gypstack at Piney Point.
-Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
-The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
-HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
-
-## Monitoring response to the emergency discharge
-
-Monitoring of the natural resources of Tampa Bay in response to the wastewater release at Piney Point began in April, 2021 and continued over the following months.
-These data were collected through a coordinated effort, facilitated in part by the FDEP and TBEP.
-Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
-Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
-
-Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
-To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
-Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
-Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
-For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
-Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.
-These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
-
-Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
-Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
-Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
-Occurrence of this species has historically been spatially distinct, with blooms originating in the Gulf of Mexico and occasionally occurring at bloom concentrations in lower Tampa Bay.
-Data for *K.
-brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
-Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
-Petersberg) impacted by the event.
-Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
-
-Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.
-Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
-Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.
-Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
-Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
-Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
-Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure \@ref(fig:map)a) depending on the analysis described below.
-
-## Long-term monitoring data
-
-Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the EPC.
-These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.
-Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection.
-Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).
-The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.
-Additionally, the period of record for monitoring data from Manatee County began in 1996.
-Long-term monitoring stations for Hillsborough and Manatee Counties are shown in Figure \@ref(fig:map)a.
-
-Long-term water quality monitoring data from Hillsborough and Manatee counties were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
-Station data from Hillsborough County were obtained for the middle and lower segments of Tampa Bay.
-Station data from Manatee County were obtained for areas in lower Tampa Bay, Terra Ceia Bay, Anna Maria Sound, and northern Sarasota Bay.
-For the same water quality parameters noted above (i.e., nitrogen, phosphorus, pH, temperature, salinity), observations at each monitoring station were averaged for each month across years from 2006 to 2020.
-This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
-For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
-This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
-Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].
-In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
-Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
-
-## Data Analysis
-
-```{r}
-seconb <- rswqdat %>% 
-  filter(date < maxdt) %>% 
-  filter(var == 'secchi') %>% 
-  filter(source == 'fldep') %>% 
-  pull(qual) %>% 
-  grepl('S', .)
-sumb <- sum(seconb)
-totb <- length(seconb) 
-```
-
-The R statistical programming language (v4.0.2) was used to import, synthesize, and analyze all datasets provided by multiple partners [@RCT20].
-Partner data were uploaded or entered manually as Google spreadsheets, where they were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.
-The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.
-The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.
-
-Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
-Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
-Further analyses were conducted to compare all data types together using PCA and correlation analyses, to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
-Data describing *K.
-brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing.
-Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.
-Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a.
-For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.
-Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.
-
-The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
-For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.
-Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
-Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
-An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-
-When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.
-These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from NADA.
-For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.
-Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.
-PCA and correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.
-
-# Results
-
-## Timeline of events from April 2021
-
-A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnds)b).
-Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
-The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.
-On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
-Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
-*Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
-By June 27th, fish kill reports attributed to red tide increased with *K.
-brevis* cell concentrations in lower and middle Tampa Bay.
-The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.
-This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St.
-Petersburg and Tampa.
-Concentrations of *K.
-brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.
-A quantitative description of these events follows.
-
-## Water quality trends
-
-```{r}
-nwq <- sum(wqsumtab[['N obs.']])
-outrng1 <- wqsumtab %>% 
-  filter(`Water quality variable` == 'Secchi (m)') %>% 
-  pull(below) %>%  
-  sum
-outrng2 <- wqsumtab %>% 
-  filter(`Water quality variable` != 'Secchi (m)') %>% 
-  pull(above) %>%  
-  sum
-outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
-
-# percentages out of range
-wqsumfl <- wqsumtab %>% 
-  mutate(
-    Area = ifelse(Area == '', NA, Area)
-  ) %>% 
-  fill(Area) 
-chlabv <- wqsumfl %>% 
-  filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
-  select(Area, above) %>% 
-  deframe %>% 
-  round(0)
-tnabv <- wqsumfl %>% 
-  filter(`Water quality variable` == 'TN (mg/L)') %>% 
-  select(Area, above) %>% 
-  deframe %>% 
-  round(0)
-secblw <- wqsumfl %>% 
-  filter(`Water quality variable` == 'Secchi (m)') %>% 
-  select(Area, below) %>% 
-  deframe %>% 
-  round(0)
-dosabv <- wqsumfl %>% 
-  filter(`Water quality variable` == 'DO (% sat.)') %>% 
-  select(Area, above) %>% 
-  deframe %>% 
-  round(0)
-```
-
-From April through July 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
-Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
-The percentage of observations outside of the normal range varied by location and parameter.
-For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
-Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.
-Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.
-Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).
-Physical parameters (salinity, temperature) were generally within range over the study period.
-Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
-Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
-
-Boxplots of total nitrogen, chl-a, and secchi depth show the temporal progression of observations by week and area relative to the normal ranges from long-term monitoring data (Figure \@ref(fig:wqtrnds)).
-For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges beginning the week of April 4th and lasting through the month.
-Concentrations remained similar to baseline conditions until June and July when median values were often above the baseline range.
-Secchi observations in area 1 were below baseline ranges in April and July.
-Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 during weeks of May, June, and July.
-Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds).
-Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for areas 1 and 3 and total nitrogen and chl-a for area 2 (Table \@ref(tab:wqcmptab)).
-Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
-Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
-brevis* abundances.
-
-## Seagrass and macroalgae trends
-
-```{r}
-ntrnscnt <- length(unique(rstrndat$station))
-ntrns <- english(ntrnscnt)
-visdat <- rstrndat %>% 
-  filter(date < maxdt) %>% 
-  mutate(
-    mo = month(date)
-  ) %>% 
-  group_by(station, mo) %>% 
-  summarise(
-    cnt = length(unique(date)), 
-    .groups = 'drop'
-  ) %>% 
-  group_by(mo) %>% 
-  summarise(
-    cnt = mean(cnt)
-  )
-avevis <- round(mean(visdat$cnt), 1)
-
-# all fo
-foall <- rstrndat %>% 
-  filter(date < maxdt) %>% 
-  mutate(
-    pa = bb > 0
-  ) %>% 
-  mutate(
-    taxa = as.character(taxa), 
-    taxa = case_when(
-      grepl('^Halophila', taxa) ~ 'Halophila sp', 
-      T ~ taxa
-    )
-  ) %>% 
-  group_by(taxa) %>% 
-  summarise(
-    fo = sum(pa) / length(pa), 
-    .groups = 'drop'
-  ) %>% 
-  mutate(
-    fo = round(100 * fo, 1)
-  ) %>% 
-  deframe
-
-# fo by month, area
-fomo <- trnsum %>% 
-  mutate(
-    foest = round(100 * foest, 1), 
-    mo = month(date)
-  )
-fomoarea1 <- fomo %>% 
-  filter(area == '(a) Area 1')
-fomoarea3 <- fomo %>% 
-  filter(area == '(b) Area 3')
-fomoarea1apr <- fomoarea1 %>% 
-  filter(mo %in% 4) %>% 
-  select(taxa, foest) %>% 
-  deframe
-fomoarea1jul <- fomoarea1 %>% 
-  filter(mo %in% 7) %>% 
-  select(taxa,foest) %>% 
-  deframe
-fomoarea3apr <- fomoarea3 %>% 
-  filter(mo %in% 4) %>% 
-  select(taxa, foest) %>% 
-  deframe
-fomoarea3jul <- fomoarea3 %>% 
-  filter(mo %in% 7) %>% 
-  select(taxa,foest) %>% 
-  deframe
-fomoarea1jun <- fomoarea1 %>% 
-  filter(mo %in% 6) %>% 
-  select(taxa, foest) %>% 
-  deframe
-fomoarea3jun <- fomoarea3 %>% 
-  filter(mo %in% 6) %>% 
-  select(taxa, foest) %>% 
-  deframe
-
-# kw pvals
-cyanoarea3 <- mcrtab %>% 
-  mutate(
-    Area = ifelse(Area == '', NA, Area)
-  ) %>% 
-  fill(Area) %>% 
-  filter(Area == 3 & `Macroalgae group` == 'Cyanobacteria') %>% 
-  pull(pval) %>% 
-  as.numeric %>% 
-  round(3)
-```
-
-A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through July, each visited on average `r avevis` times per month.
-Turtle grass (*T.
-testudinum*) was the dominant seagrass species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
-Manatee grass (*S.
-filiforme*) and shoal grass (*H.
-wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
-The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].
-At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.
-Macroalgae observed along the transects also varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
-Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.
-Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
-Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
-Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
-
-Temporal progression of seagrasses and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
-Transect S3T6 is located less than one kilometer to the north of Port Manatee.
-The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
-Overall abundance of seagrass did not change dramatically from April 22nd through July.
-However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
-Red macroalgae were present in high abundances from April to May.
-*Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
-*Dapis sp.* persisted through June and July, but not was observed after July 20th.
-Green macroalgae taxa were first observed in July, although at generally low abundances.
-Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae by mid-summer.
-
-Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of seagrass and macroalgae trends across all transects.
-No transects were sampled in area 2 to the north of Piney Point.
-Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in shoal grass (*H.
-wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea1jul['Halodule wrightii']`\%).
-A similar decline in coverage of shoal grass was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
-Frequency occurrence of manatee grass (*S.
-syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, July frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
-Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
-Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3).
-Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
-Green macroalgae had the second lowest frequency occurrence, followed by Brown macroalgae that was only observed at one transect.
-
-Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrtab) and \@ref(tab:savtab) further suggested that seagrass trends were minor compared to those of macroalgae.
-Results by month showed that seagrass frequency occurrence, by species and area, had insignificant changes ($p > 0.05$) over the study period.
-Although trends were not significant, estimated medians in each month in Table \@ref(tab:savtab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.
-These changes were more pronounced in area 3 as compared to area 1.
-For macroalgae (Table \@ref(tab:mcrtab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).
-Coverage of cyanobacteria in area 1 during June was significantly higher than coverage in April.
-No significant differences between months were observed for the green macroalgae group.
-
-## Water quality compared to seagrass and macroalgae
-
-```{r}
-expvr <- allpcadat$eig[, 2] %>% 
-  round(0)
-```
-
-A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).
-The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]`\% of the remaining variation.
-For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and secchi depth and strong negative loadings for total nitrogen, chl-a, and temperature.
-For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae.
-For seagrasses, turtle grass (*T.
-testudinum*) had a strong positive loading with axis 1.
-The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.
-The third axis showed a strong negative loading with shoal grass (*H.
-wrightii*) and positive loading with manatee grass (*S.
-filiforme*).
-Observations by area showed that samples in area 3 were distinct from points in area 1 and grouped by positive loadings on the first PCA axis and negative loadings on the second PCA axis.
-
-Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
-Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$).
-Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$, except $p < 0.05$ for temperature) and negatively correlated with secchi ($p < 0.005$).
-Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas red macroalgae was negatively correlated with cyanobacteria ($p < 0.005$).
-Among the seagrasses, manatee grass was negatively correlated with shoal grass and turtle grass ($p < 0.005$).
-Several notable correlations were also observed between water quality, macroalgae, and seagrass.
-Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).
-For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated with turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).
-For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
-
-## Red tide and fish kill reports
-
-```{r}
-fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
-  select(
-    date = textBox6, 
-    city = cOUNTYDataTextBox, 
-    waterbody = lOCATIONDataTextBox,
-    species = textBox18
-  ) %>% 
-  mutate(
-    date = mdy(date),
-    yr = year(date),
-    week = floor_date(date, unit = 'week'), 
-    week = factor(format(week, '%b %d')), 
-    week = factor(week, levels = as.character(unique(week))), 
-    city = gsub('\\s+$', '', city),
-    city = gsub('^St\\s', 'St. ', city)
-  ) %>% 
-  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
-  filter(date < maxdt)
-fishann <- fishdat %>% 
-  group_by(yr, city) %>% 
-  summarise(cnt = n(), .groups = 'drop') %>% 
-  group_by(city) %>% 
-  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
-  mutate(per = round(100 * cnt / sum(cnt), 1))
-annstpper <- fishann %>% 
-  filter(city == 'St. Petersburg') %>% 
-  pull(per)
-anntmpper <- fishann %>% 
-  filter(city == 'Tampa') %>% 
-  pull(per)
-fishwk <- fishdat %>% 
-  filter(yr == 2021) %>% 
-  group_by(week, city) %>% 
-  summarise(cnt = n(), .groups = 'drop') %>% 
-  group_by(city) %>% 
-  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
-  mutate(per = round(100 * cnt / sum(cnt), 1))
-wkstpcnt <- fishwk %>% 
-  filter(city == 'St. Petersburg') %>% 
-  pull(cnt)
-wktmpcnt <- fishwk %>% 
-  filter(city == 'Tampa') %>% 
-  pull(cnt)
-```
-
-The increase in *K.
-brevis* from April to July to bloom concentrations exceeding 10k cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels.
-Median cell concentrations for most years were well below 1,000 cells/L.
-The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L [@Flaherty11].
-Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
-
-Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K.
-brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c).
-Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.
-In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).
-The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K.
-brevis* cell concentrations (Figure \@ref(fig:redtide)b).
-Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.
-Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.
-As a result, Pinellas County and the city of St.
-Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
-Hammer Levy, Pinellas County, pers.
-comm.
-Aug.
-2021).
-
-# Discussion
-
-The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
-Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
-Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
-
-In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen effluent directly into lower Tampa Bay.
-Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
-An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
-Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
-From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years.
-These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
-
-Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
-Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)) and loadings from the PCA results (Figure \@ref(fig:allpca)).
-Temporally, these changes were observed at different times and for different species of phytoplankton.
-The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K.
-brevis* in June and July.
-Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
-Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
-Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.
-Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
-
-From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years.
-Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.
-Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.
-Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.
-Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
-This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].
-Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient.
-Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.
-Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.
-As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.
-Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper.
-In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.
-Shoal grass can occur in more turbid waters of oligohaline areas.
-
-Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.
-This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* likely reflects a natural phenology in Tampa Bay.
-Past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
-Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
-A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Dapis sp.*).
-Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Dapis sp.* in Anna Maria Sound and near Port Manatee were observed in 2021.
-The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
-*Dapis sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
-
-The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
-Regionally, seagrasses have declined in coverage as estimated from biennial maps provided by the Southwest Florida Water Management District.
-From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
-These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
-Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.
-Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern.
-Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
-In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
-The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.
-General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health and restoration activities.
-
-Fisheries resources were undoubtedly impacted by *K.
-brevis* blooms in June and July (Figure \@ref(fig:redtide)).
-Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
-Routine sampling by the FWC Fisheries Independent Monitoring Program occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
-Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
-Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
-For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
-Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
-brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-
-Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments were sampled near Port Manatee and surrounding waters in April 2021 during the discharge and in September 2021 to further assess potential impacts of Piney Point on sediment quality.
-These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
-Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions.
-Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
-Comparison of the April 2021 samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples.
-Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
-Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
-This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
-
-Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels, rivers or other back-bay habitats (e.g., Alafia River and Bishop Harbor).
-The comparison to past years provided a stark contrast of the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
-The Tampa Bay Coastal Ocean Model [@Chen19] was used to simulate daily tracer particle movement and expected dilution of nutrient concentrations from the point of discharge.
-Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
-Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
-Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of labile inorganic nitrogen forms from stack effluent (Table \@ref(tab:stktab)) relative to background concentrations for the bay (Table \@ref(tab:wqsumtab)).
-
-In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to environmental resources.
-In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.
-Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18b].
-The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
-Massive fish kills were observed and likely related to low pH of the effluent.
-Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
-Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.
-These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.
-
-Limitations of our analyses are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.
-All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.
-However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-The weight of evidence that 2021 is an anomaly relative to recent prior years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
-Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.
-There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.
-These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
-
-# Acknowledgments
-
-We thank the many TBEP partners and collaborators for their continuing efforts to restore and monitor Tampa Bay.
-We thank the administrative staff, field crews, and laboratory staff from the Florida Department of Environmental Protection, Environmental Protection Commission of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission, City of St.
-Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
-Our partners’ willingness to adapt and implement innovative monitoring and management actions in response to Piney Point and the ever evolving challenges threatening Tampa Bay is greatly appreciated.
-
-# Figures
-
-```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to wastewater discharge from Piney Point.
-Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
-Inset shows location of Tampa Bay on the Gulf coast of Florida, USA.'}
-knitr::include_graphics(here('figs/map.jpeg'))
-```
-
-```{r timeline, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting March 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
-knitr::include_graphics(here('figs/timeline.jpeg'))
-```
-
-```{r wqmap, fig.cap = 'Sampled water quality data for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
-Values outside of the normal range (above for total nitrogen and chlorophyll, below for secchi) are outlined in black and those in normal range are outlined in light grey.
-Color ramps and point sizes show relative values (reversed for Secchi).
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location (Figure \\@ref(fig:map)a).
-Values below detection limits (or secchi on bottom) are not shown.'}
-knitr::include_graphics(here('figs/wqmap.jpeg'))
-```
-
-```{r wqtrnds, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
-Observations are aggregated by week and within assessment areas shown in Figure \\@ref(fig:map)a.
-Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure \\@ref(fig:map)a).
-Values below detection limits (or secchi on bottom) are not shown.'}
-knitr::include_graphics(here('figs/wqtrnds.jpeg'))
-```
-
-```{r trnex, fig.cap = 'Results for (a) seagrass and (b) macroalgae rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.
-Sample dates in 2021 are shown in rows with transect meter results shown in columns.
-Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis sp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
-Abundances are Braun-Blanquet coverage estimates.'}
-knitr::include_graphics(here('figs/trnex.jpeg'))
-```
-
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 3 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
-Estimates are grouped by sample months in 2021.
-Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.')}
-knitr::include_graphics(here('figs/trnfrq.jpeg'))
-```
-
-```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
-Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
-All variables were standardized to zero mean and unit variance prior to PCA.
-Variables with log-normal distribution were log-transformed prior to analysis.'}
-knitr::include_graphics(here('figs/allpca.jpeg'))
-```
-
-```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
-Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
-Variables with log-normal distribution were log-transformed prior to analysis.
-Thick black lines separate water quality, macroalgae, and seagrass variables.
-Spearman correlations indicate the linear strength of assocation between pairs of variables.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
-knitr::include_graphics(here('figs/allcor.jpeg'))
-```
-
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) and number of fish kill reports for the contiguous record showing cell concentrations (a) by year and (b) by week in 2021 and reported fish kills by city (Tampa, St.
-Petersburg) (c) by year and (d) by week in 2021.
-Red tide concentrations show minimum, 25th percentile, median, 75th percentile, and maximum for each year or week for middle and lower Tampa Bay.
-*K.
-brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
-brevis* (https://public.myfwc.com/FWRI/FishKillReport/).'}
-knitr::include_graphics('figs/redtide.jpeg')
-```
-
-# Tables
-
-```{r stktab}
-cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.
-Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.
-Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay (Figure \\@ref(fig:map)a).
-End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
-Missing values were not measured in the effluent or stack water.'
-knitr::kable(stktab, booktabs = T, caption = cap.val)
-```
-
-```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from March to July 2021 in response to wastewater discharge from Piney Point.
-Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
-Summaries are median, minimum, and maximum values.
-Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
-The final column shows the percentage of total observations that were below detection limits (including Secchi on bottom).
-Medians denoted by "-" could not be calculated due to insufficient values above detection.'
-knitr::kable(wqsumtab, booktabs = T, caption = cap.val)
-```
-
-```{r wqcmptab}
-cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp.
-column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
-Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-knitr::kable(wqcmptab, booktabs = T, caption = cap.val)
-```
-
-```{r mcrtab}
-cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp.
-column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
-Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- mcrtab %>% 
-  select(-pval)
-knitr::kable(totab, booktabs = T, caption = cap.val)
-```
-
-```{r savtab}
-cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp.
-column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
-Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- savtab %>% 
-  select(-pval)
-knitr::kable(totab, booktabs = T, caption = cap.val)
-```
-
-# References

### 2021-10-13 R/figures.R

@@ -1110,17 +1110,17 @@ p4 <- ggplot(toplo, aes(x = xvals, y = precip_cm, group = yr, color = flvl, size
   )
 
 # wind roses
-winddat <- na.omit(winddat)
+toplo <- na.omit(winddat)
 
 dts <- as.Date(c('2021-01-01', '2021-07-03', '2021-07-05', '2021-07-07', '2021-10-01'))
 
-toplo1 <- winddat %>% 
+toplo1 <- toplo %>% 
   filter(as.Date(datetime) >= dts[1] & as.Date(datetime) < dts[2])
-toplo2 <- winddat %>% 
+toplo2 <- toplo %>% 
   filter(as.Date(datetime) >= dts[2] & as.Date(datetime) < dts[3])
-toplo3 <- winddat %>% 
+toplo3 <- toplo %>% 
   filter(as.Date(datetime) >= dts[3] & as.Date(datetime) < dts[4])
-toplo4 <- winddat %>% 
+toplo4 <- toplo %>% 
   filter(as.Date(datetime) >= dts[4] & as.Date(datetime) < dts[5])
 
 spdmin <- 0
@@ -1161,5 +1161,33 @@ p <- p1 + p2 + p3 + p4 + p5 +
 
 
 jpeg(here('figs/redtide.jpeg'), height = 11, width = 9, units = 'in', res = 500, family = 'serif')
+print(p)
+dev.off()
+
+# Supplement figures ------------------------------------------------------
+
+# wind roses
+toplo <- na.omit(winddat) %>% 
+  mutate(
+    mo = month(datetime, label = T)
+  ) %>% 
+  filter(mo != 'Oct')
+
+spdmin <- 0
+spdmax <- 14
+spdres = 3
+thm <- theme_minimal()
+
+p <- plot.windrose(data = winddat, spd = 'wind_ms', dir = 'wind_dir', spdres = spdres, spdmin = spdmin, spdmax = spdmax) + 
+  facet_wrap(~mo, ncol = 3) + 
+  theme_minimal() + 
+  theme(
+    axis.text = element_blank(), 
+    axis.title = element_blank(), 
+    axis.ticks.y = element_blank(), 
+    legend.position = 'right'
+  )
+
+jpeg(here('figs/windroses.jpeg'), height = 6, width = 7, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()

### 2021-10-14 R/figures.R
red tide plot apr sep only, precip as percentile, five wind roses

@@ -1090,21 +1090,33 @@ toplo <- raindat %>%
     xvals = ymd(paste('2021', month(date), day(date), sep = '-'))
   )
 
-p4 <- ggplot(toplo, aes(x = xvals, y = precip_cm, group = yr, color = flvl, size = flvl)) + 
-  geom_line() + 
-  scale_size_manual(values = c(0.5, 1.5)) +
-  scale_color_manual(values = c('grey', '#00806E')) +
+toplo1 <- toplo %>% 
+  filter(yr == 2021)
+toplo2 <- toplo %>% 
+  filter(yr < 2021) %>% 
+  group_by(xvals) %>% 
+  summarise(
+    lov = quantile(precip_cm, 0.25, na.rm = T),
+    hiv = quantile(precip_cm, 0.75, na.rm = T),
+  ) %>% 
+  ungroup()
+p4 <- ggplot() + 
+  geom_ribbon(data = toplo2, aes(x = xvals, ymin = lov, ymax = hiv, fill = '1995 - 2020\n25th - 75th %tile'), alpha = 0.7) +
+  geom_line(data = toplo1, aes(x = xvals, y = precip_cm, color = '2021'), size = 1.2) +
+  scale_color_manual(values = '#00806E') +
+  scale_fill_manual(values = 'grey') + 
   scale_x_date(date_breaks = 'month', date_labels = '%b %d', expand = c(0, 0)) +
   theme_minimal() +
   theme(
     axis.text.x = element_text(angle = 45, size = 8, hjust = 1),
     panel.grid.minor.x = element_blank(),
-    panel.grid.major.x = element_blank()
+    panel.grid.major.x = element_blank(), 
+    legend.title = element_blank()
   ) +
   labs(
     x = 'Day of year',
     y = 'Cumulative precip. (cm)',
-    title = '(d) Cumulative precipitation by year for Tampa International Airport',
+    title = '(d) Cumulative precipitation in 2021 at Tampa International Airport',
     color = 'Year group', 
     size = 'Year group'
   )
@@ -1112,7 +1124,7 @@ p4 <- ggplot(toplo, aes(x = xvals, y = precip_cm, group = yr, color = flvl, size
 # wind roses
 toplo <- na.omit(winddat)
 
[…]

### 2021-10-14 supplement.Rmd
supplement wqtrnds by week added

@@
 
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+Observations are aggregated by week and within assessment areas shown in Figure 1a.
+Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).
+Values below detection limits are not shown.'}
+knitr::include_graphics('figs/wqtrnds-supp.jpeg')
+```
+
 ```{r windroses, fig.cap = 'Wind rose plots for 2021 by month.

### 2021-10-14 R/figures.R

@@ -920,13 +920,13 @@ fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>%
 
 # k brevis data from CL
 habdat <- read.csv(here('data-raw/KB_LowMid_1995-2021.csv')) %>% 
-  filter(Segment %in% c('MTB')) %>%
+  filter(Segment %in% c('MTB','LTB')) %>%
   select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
   mutate(date = mdy(date))
 
 # add k brevis data from HABSOS that are collected after last date in habdat
 kbrdat <- kbrdat %>% 
-  .[tbseg[tbseg$bay_segment %in% c('MTB'), ], ] %>% 
+  .[tbseg[tbseg$bay_segment %in% c('MTB', 'LTB'), ], ] %>% 
   filter(var == 'kb') %>% 
   filter(date > max(habdat$date)) %>% 
   mutate(
@@ -955,33 +955,37 @@ weeklv <- seq.Date(from = as.Date('2021-01-01'), to = Sys.Date(), by = 'days') %
 # habdat p1
 toplo <- habdat %>%
   filter(date < as.Date('2021-10-01')) %>% 
+  filter(month(date) > 3 & month(date) < 10) %>% 
   mutate(
     dtgrp = quarter(date),
     yr = year(date)
   ) %>%
   mutate(
     yr = factor(yr, levels = seq(min(yr), max(yr)))
-  ) %>%
-  group_by(yr) %>%
-  summarise(
-    cnt = n(),
-    y0 = min(val, na.rm = T), 
-    y25 = quantile(val, prob = 0.25, na.rm = T),
-    y50 = quantile(val, prob = 0.5, na.rm = T),
-    y75 = quantile(val, prob = 0.75, na.rm = T),
-    y100 = max(val, na.rm = T),
-    .groups = 'drop'
-  ) %>%
-  complete(yr) %>% 
-  filter(as.numeric(as.character(yr)) >= 1995)
+  ) #%>%
+  # group_by(yr) %>%
+  # summarise(
+  #   cnt = n(),
+  #   y0 = min(val, na.rm = T), 
+  #   y25 = quantile(val, prob = 0.25, na.rm = T),
+  #   y50 = quantile(val, prob = 0.5, na.rm = T),
+  #   y75 = quantile(val, prob = 0.75, na.rm = T),
[…]

### 2021-10-14 R/figures.R
reproc all figs up to Oct

@@ -238,7 +238,7 @@ thm <-  theme(
 # combine water quality data with locations
 # removed secchi on bottom and non-detect for tn, chla
 wqdat <- rswqdat %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!station %in% nonbay) %>% 
   filter(!qual %in% c('S', 'U')) %>% 
@@ -370,7 +370,7 @@ cols <- c("#E16A86", "#50A315", "#009ADE")
 names(cols) <- levels(ppseg$area)
 
 datin <- rswqdat %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
   filter(!qual %in% c('S', 'U')) # remove secchi on bottom, nondetect for chla, tn
 
@@ -560,7 +560,7 @@ areas <- ppseg %>%
 
 # add area
 trnsum <- rstrndat %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   mutate(
     date = floor_date(date, unit = 'week')
   ) %>% 
@@ -590,7 +590,7 @@ trnsum <- rstrndat %>%
 rswqsum <- rswqdat %>% 
   filter(var %in% vrs) %>% 
   filter(source == 'fldep') %>%
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   filter(!station %in% nonbay) %>% 
   inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
   st_intersection(areas) %>% 
@@ -696,7 +696,7 @@ areas <- ppseg %>%
 
 # add area
 trnsum <- rstrndat %>%
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   mutate(
     date = floor_date(date, unit = 'week')
   ) %>%
@@ -747,7 +747,7 @@ trncrs <- crossing(var1 = names(trncors), var2 = names(trncors)) %>%
 
 # add area
 trnsum <- rstrndat %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
[…]

### 2021-10-14 manu-draft.Rmd
reprocess all tables up to Oct, need to update manu text

@@
 
-maxdt <- as.Date('2021-08-01')
+maxdt <- as.Date('2021-10-01')
 
@@
 
-From April through July 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
+From April through September 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
 Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).

### 2021-10-14 R/figures.R

@@ -1177,7 +1177,8 @@ p5 <- pa + pb + pc + pd + pe + plot_layout(ncol = 5, guides = 'collect') &
     axis.text = element_blank(), 
     axis.title = element_blank(), 
     axis.ticks.y = element_blank(), 
-    legend.position = 'right'
+    legend.position = 'right',
+    plot.subtitle = element_text(size = 9)
   )
 
 # all plots together

### 2021-10-14 R/funcs.R

@@ -115,7 +115,7 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
       axis.title.x = element_blank(),
       panel.grid.minor = element_blank(),
       strip.text = element_text(size = 14), 
-      axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
+      axis.text.x = element_text(size = 4.5, angle = 45, hjust = 1)
     )
   
   if(logtr)
@@ -446,7 +446,7 @@ plot.windrose <- function(data,
   p.windrose <- ggplot(data = data,
                        aes(x = dir.binned,
                            fill = spd.binned)) +
-    geom_bar(color = 'lightgrey') + 
+    geom_bar(color = 'grey') + 
     scale_x_discrete(drop = FALSE,
                      labels = waiver()) +
     coord_polar(start = -((dirres/2)/360) * 2*pi) +

### 2021-10-14 R/tables.R

@@ -134,7 +134,7 @@ sigs <- parms %>%
 totab <- rswqdat %>% 
   left_join(rsstatloc, ., by = c('station', 'source')) %>% 
   filter(var %in% vrs) %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   filter(!station %in% nonbay) %>% 
   select(station, date, lbs, val, inrng, qual) %>% 
   st_intersection(ppsegbf) %>% 
@@ -210,7 +210,7 @@ ppsegbf <- ppseg %>%
 
 rswqsub <- rswqdat %>% 
   filter(var %in% vrs) %>% 
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   filter(source == 'fldep') %>%
   filter(!station %in% nonbay) %>% 
   filter(!qual %in% 'S') %>% # remove secchi on bottom
@@ -222,7 +222,7 @@ rswqsub <- rswqdat %>%
   mutate(
     mo = month(date, label = T)
   ) %>% 
-  filter(mo %in% c('Apr', 'May', 'Jun', 'Jul')) %>% 
+  filter(mo %in% c('Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep')) %>% 
   mutate(mo = fct_drop(mo))
 
 # multiple comparisons of time frames within sites, all nutrients
@@ -342,7 +342,7 @@ areas <- ppseg %>%
 
 # add area
 trnsf <- rstrndat %>%
-  filter(date < as.Date('2021-08-01')) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
   inner_join(rstrnpts, ., by = 'station') %>% 
   filter(taxa %in% c(mcrsel, savsel)) %>% 
   # filter(station %in% smpeff) %>% 

### 2021-10-14 manu-draft.Rmd
inflow to red tide figure

@@
 
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, and (e) wind rose plots for 2021 before, passage (two days pre/post), and after tropical storm Elsa.
-Red tide concentrations show minimum, 25th percentile, median, 75th percentile, and maximum for each year or week for middle Tampa Bay.
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, (e) cumulative inflow in 2021 compared to past years, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa.
 *K.
 brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
-brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, and wind rose plots are from St.
+brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for all of Tampa Bay from USGS gaged sites, and wind rose plots are from St.
 Petersburg, Florida.

### 2021-10-14 R/dat_copy.R

@@ -35,7 +35,7 @@
 
 # sync data files from piney-point
 
-fls <- c('bsstatloc', 'bstransect', 'bstransectocc', 'bswqdat', 'bswqdatsub', 'bswqrngs', 'kbrdat', 'lobodat', 'loboeco', 
+fls <- c('bsstatloc', 'bstransect', 'bstransectocc', 'bswqdat', 'bswqdatsub', 'bswqrngs', 'hydrodat', 'kbrdat', 'lobodat', 'loboeco', 
          'mcrtmp', 'parms', 'ppseg', 'raindat', 'rsallpts', 'rsbntdat', 'rsbntpts', 'rscntdat', 'rscntthr', 'rsphydat', 'rsphypts', 
          'rsstatloc', 'rstrndat', 'rstrnlns', 'rstrnpts', 'rstrnwts', 'rswqdat', 'sbshed', 'segmask', 'tbhished', 'trnlns', 
          'trnpts', 'winddat', 'wqrefmap')

### 2021-10-14 R/figures.R

@@ -40,6 +40,7 @@ data(rstrnpts)
 data(kbrdat)
 data(winddat)
 data(raindat)
+data(hydrodat)
 
 source(here('R/funcs.R'))
 
@@ -884,7 +885,8 @@ jpeg(here('figs/allcor.jpeg'), height = 4.4, width = 4.5, units = 'in', res = 50
 print(p)
 dev.off()
 
-# red tide, fish kills, wind, precip --------------------------------------
+# red tide, fish kills, wind, flow, precip --------------------------------
+
 
 # data from https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx
 # requested hillsborough, pinellas, manatee 1/1/95 to early Oct
@@ -963,18 +965,18 @@ toplo <- habdat %>%
   mutate(
     yr = factor(yr, levels = seq(min(yr), max(yr)))
   ) #%>%
-  # group_by(yr) %>%
-  # summarise(
-  #   cnt = n(),
-  #   y0 = min(val, na.rm = T), 
-  #   y25 = quantile(val, prob = 0.25, na.rm = T),
-  #   y50 = quantile(val, prob = 0.5, na.rm = T),
-  #   y75 = quantile(val, prob = 0.75, na.rm = T),
-  #   y100 = max(val, na.rm = T),
-  #   .groups = 'drop'
-  # ) %>%
-  # complete(yr) %>% 
-  # filter(as.numeric(as.character(yr)) >= 1995)
+# group_by(yr) %>%
+# summarise(
+#   cnt = n(),
+#   y0 = min(val, na.rm = T), 
+#   y25 = quantile(val, prob = 0.25, na.rm = T),
+#   y50 = quantile(val, prob = 0.5, na.rm = T),
+#   y75 = quantile(val, prob = 0.75, na.rm = T),
+#   y100 = max(val, na.rm = T),
+#   .groups = 'drop'
+# ) %>%
+# complete(yr) %>% 
+# filter(as.numeric(as.character(yr)) >= 1995)
 
 # plot
 p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
@@ -1022,7 +1024,7 @@ toplo <- habdat %>%
 # plot
 p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
   geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + 
-    # geom_boxplot(
+  # geom_boxplot(
[…]

### 2021-10-17 manu-draft.Rmd
nutrient flow figure and supp fo figure done

@@
 
+```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) *Karenia brevis* cell concentrations, and (e) cyanobacteria macroalgal abundances.
+Values are summarized for all samples within each week.
+The values suggest nutrient cycling between water column phytoplankton in the initial April diatom bloom, then to cyanobacteria macroalgae in May to June, and then to *K.
+brevis* peaking in early July.
+Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurence estimates across all dates.
+Diatom concentrations are based on combined cell counts from *Asterionellopsis sp.* and *Skeletonema sp*.'}
+knitr::include_graphics('figs/nutrientflow.jpeg')
+```
+
 # Tables

### 2021-10-17 supplement.Rmd
nutrient flow figure and supp fo figure done

@@
 ```
+
+```{r phymcrfoest, fig.cap = 'Frequency occurrence estimates for (a) phytoplankton (diatoms and *K.
+brevis*) and (b) macroalgae groups.
+Frequency occurrence estimates are aggregated by week of observation based on all sample locations where a phytoplankton or macroalgal taxa was observed divided by all sample locations in a week.
+Estimates are not additive and are specific to each taxa.
+Sample dates are noted by vertical grey lines in each plot.
+Diatoms are based on presence/absense of *Asterionellopsis sp*.
+and *Skeletonema sp*.
+Note that frequency occurrence is not a precise measure of cell concentrations.'}
+knitr::include_graphics('figs/phymcrfoest.jpeg')
+```

### 2021-10-17 R/figures.R

@@ -41,6 +41,9 @@ data(kbrdat)
 data(winddat)
 data(raindat)
 data(hydrodat)
+data(rstrnwts)
+data(rsphydat)
+data(bstransect)
 
 source(here('R/funcs.R'))
 
@@ -887,7 +890,6 @@ dev.off()
 
 # red tide, fish kills, wind, flow, precip --------------------------------
 
-
 # data from https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx
 # requested hillsborough, pinellas, manatee 1/1/95 to early Oct
 fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>% 
@@ -964,30 +966,13 @@ toplo <- habdat %>%
   ) %>%
   mutate(
     yr = factor(yr, levels = seq(min(yr), max(yr)))
-  ) #%>%
-# group_by(yr) %>%
-# summarise(
-#   cnt = n(),
-#   y0 = min(val, na.rm = T), 
-#   y25 = quantile(val, prob = 0.25, na.rm = T),
-#   y50 = quantile(val, prob = 0.5, na.rm = T),
-#   y75 = quantile(val, prob = 0.75, na.rm = T),
-#   y100 = max(val, na.rm = T),
-#   .groups = 'drop'
-# ) %>%
-# complete(yr) %>% 
-# filter(as.numeric(as.character(yr)) >= 1995)
+  )
 
 # plot
 p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
   geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + 
-  # geom_boxplot(
-  #   aes(ymin = y0, lower = y25, middle = y50, upper = y75, ymax = y100),
-  #   stat = "identity", width = 0.75, fill = '#00806E'
-  # ) +
-  # scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
   scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
-  # geom_hline(aes(yintercept = 1e5, color = 'Bloom\nconcentration')) +
+
   theme_minimal() +
   theme(
     axis.text.x = element_text(angle = 45, size = 8, hjust = 1),
@@ -1246,6 +1231,295 @@ jpeg(here('figs/redtide.jpeg'), height = 11, width = 9, units = 'in', res = 500,
 print(p)
 dev.off()
 
[…]

### 2021-10-17 R/nutrient_flow_fig.R

@@ -1,235 +1,0 @@
-
-# setup -------------------------------------------------------------------
-
-library(tidyverse)
-library(lubridate)
-library(sf)
-library(here)
-library(ggmap)
-library(ggspatial)
-library(patchwork)
-library(units)
-library(grid)
-library(scales)
-library(USAboundaries)
-library(ggord)
-library(lubridate)
-library(patchwork)
-library(tbeptools)
-library(NADA)
-library(RColorBrewer)
-box::use(
-  scales = scales[muted], 
-  units = units[set_units], 
-  vegan = vegan[decostand], 
-  FactoMineR = FactoMineR[PCA], 
-  ggnewscale[new_scale_fill], 
-  grid[unit]
-)
-
-data(rsallpts)
-data(bswqdat)
-data(bsstatloc)
-data(ppseg)
-data(segmask)
-data(rswqdat)
-data(rsstatloc)
-data(rstrndat)
-data(parms)
-data(rstrnpts)
-data(kbrdat)
-data(winddat)
-data(raindat)
-data(hydrodat)
-data(rsphydat)
-data(rsphypts)
-
-source(here('R/funcs.R'))
-
-
-vrs <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
-mcrlab <- c("Red", "Green", "Brown", "Cyano")
-savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
-savlab <- c('T. testudinum', 'H. wrightii', 'S. filiforme')
-wqsel <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-wqlab <- c('Chl-a', 'DOsat', 'NH3, NH4+', 'pH', 'Sal', 'Secchi', 'Temp', 'TN', 'TP')
-
-nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
-
-# segments
-areas <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
[…]

### 2021-10-18 manu-draft.Rmd
nutrient flow ordination added

@@
 
+```{r nutrientfloword, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.
+Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.
+Observations are grouped by month periods based on phytoplankton (diatoms or *K.
+brevis*) or macroalgal dominance.'}
+knitr::include_graphics('figs/nutrientfloword.jpeg')
+```
+
 # Tables

### 2021-10-18 R/figures.R

@@ -21,7 +21,7 @@ library(RColorBrewer)
 box::use(
   scales = scales[muted], 
   units = units[set_units], 
-  vegan = vegan[decostand], 
+  vegan = vegan[decostand, metaMDS], 
   FactoMineR = FactoMineR[PCA], 
   ggnewscale[new_scale_fill], 
   grid[unit]
@@ -1517,6 +1517,259 @@ p5 <- ggplot(wtsest, aes(x = week, y = medv)) +
 p <- p1 + p2 + p3 + p4 + p5 + plot_layout(ncol = 1)
 
 jpeg(here('figs/nutrientflow.jpeg'), height = 8, width = 6, units = 'in', res = 500, family = 'serif')
+print(p)
+dev.off()
+
+# nutrient flow ordination ------------------------------------------------
+
+vrs <- c('chla', 'tn')
+nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+
+# water quality summarized
+wqsum <- rswqdat %>% 
+  filter(var %in% vrs) %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
+  filter(source == 'fldep') %>%
+  filter(!station %in% nonbay) %>% 
+  # inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+  # st_intersection(areas) %>% 
+  # st_set_geometry(NULL) %>% 
+  mutate(cens = grepl('U', qual)) %>% 
+  select(date, var, val, cens, station) %>% 
+  mutate(
+    date = floor_date(date, unit = 'week')
+  ) %>% 
+  group_by(date, var) %>% 
+  summarise(
+    medv = ifelse(
+      any(cens), median(cenfit(val, cens), na.rm = T),
+      median(val, na.rm = T)
+    ),
+    hiv = ifelse(
+      any(cens), quantile(cenfit(val, cens), prob = 0.975, na.rm = T),
+      quantile(val, prob = 0.975, na.rm = T)
+    ),
+    lov = ifelse(
+      any(cens), quantile(cenfit(val, cens), prob = 0.025, na.rm = T),
+      quantile(val, prob = 0.025, na.rm = T)
+    ),
+    .groups = 'drop'
+  )
+
+# phytoplankton summarized
[…]

### 2021-10-18 manu-draft.Rmd
gam seasonal estimate plot added

@@
 
+```{r wqgamtrnds, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
+Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
+Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.
+Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
+knitr::include_graphics(here('figs/wqgamtrnds.jpeg'))
+```
+
 ```{r trnex, fig.cap = 'Results for (a) seagrass and (b) macroalgae rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.

### 2021-10-18 R/figures.R

@@ -18,6 +18,7 @@ library(patchwork)
 library(tbeptools)
 library(NADA)
 library(RColorBrewer)
+library(mgcv)
 box::use(
   scales = scales[muted], 
   units = units[set_units], 
@@ -387,6 +388,37 @@ p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Week
   theme(legend.position = 'top')
 
 jpeg(here('figs/wqtrnds.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
+print(p)
+dev.off()
+
+# seasonal trend plots ----------------------------------------------------
+
+# segments
+ppsegbf <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+cols <- c("#E16A86", "#50A315", "#009ADE")
+names(cols) <- c('Area 1', 'Area 2', 'Area 3')
+
+datin <- rswqdat %>% 
+  filter(!(var == 'secchi' & val >= 9.5)) # outlier secchi
+
+p1 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = T, rmfacet = T, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
+p2 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = T, rmfacet = T, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
+p3 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'secchi', cols, logtr = F, rmfacet = F, ttl = '(c) Secchi', ylb = 'meters')
+
+p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Day of year', gp = gpar(fontsize=14))) + 
+  plot_layout(ncol = 1, guides = 'collect', height = c(1, 0.05)) & 
+  theme(legend.position = 'top')
+
+jpeg(here('figs/wqgamtrnds.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 

### 2021-10-18 R/funcs.R

@@ -134,6 +134,178 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
   
 }
 
+# water quality plot fun using GAM predictions
+gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet = FALSE, ttl, ylb){
+  
+  nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
+  mindt <- ymd('2021-03-28')
+  dtrng <- ymd(c('2006-01-01', '2021-10-01'))
+  
+  # monitoring data
+  rswqtmp <- rswqdat %>% 
+    filter(var %in% vr) %>% 
+    filter(!station %in% nonbay) %>% 
+    inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    select(-bswqstation, -nrmrng, -source, -source_lng, -uni, -lbunis) %>% 
+    mutate(
+      date = floor_date(date, unit = 'week'),
+      cens = grepl('U', qual)
+    ) %>% 
+    select(station, date, var, val, area)
+  
+  # baseline data
+  bswqtmp <- bswqdat %>% 
+    select(-source, -uni) %>% 
+    filter(var %in% vr) %>%
+    filter(!(var == 'secchi' & grepl('S', qual))) %>% # remove secchi on bottom
+    filter(date >= dtrng[1]) %>% 
+    filter(!is.na(val)) %>% 
+    inner_join(bsstatloc, ., by = 'station') %>% 
+    st_intersection(ppseg) %>% 
+    st_set_geometry(NULL) %>% 
+    mutate(
+      date = floor_date(date, unit = 'month'),
+      cens = grepl('U', qual)
+      ) %>% 
+    select(station, date, var, val, area)
+  
+  tomod <- bind_rows(bswqtmp, rswqtmp) %>% 
+    mutate(
+      yr = lubridate::year(date), 
+      doy = lubridate::yday(date),
+      cont_year = lubridate::decimal_date(date)
+    ) %>% 
+    arrange(date, var, area) %>% 
+    group_by(area) %>% 
+    nest()
+  
+  if(logtr)
+    mods <- tomod %>% 
+      mutate(
+        mod = purrr::map(data, function(x){
[…]

### 2021-10-18 R/wqbyarea.R

@@ -69,7 +69,7 @@ bswqtmp <- bswqdat %>%
     cens = grepl('U', qual)) %>% 
   group_by(date, var, area) %>% 
   summarise(   
-    avev = ifelse(
+    val = ifelse(
       any(cens), mean(cenfit(val, cens), na.rm = T),
       mean(val, na.rm = T)
     ),
@@ -102,7 +102,7 @@ rswqtmp <- rswqdat %>%
     cens = grepl('U', qual)
   ) %>% 
   left_join(parms, by = 'var') %>% 
-  select(station, date, var, val, area, uni)
+  select(station, date, var, val, area, lbs)
 
 # baseline data
 bswqtmp <- bswqdat %>% 
@@ -117,8 +117,8 @@ bswqtmp <- bswqdat %>%
   mutate(
     date = floor_date(date, unit = 'month'),
     cens = grepl('U', qual)) %>% 
-  left_join(parms, by = 'var') %>% 
-  select(station, date, var, val, area, uni)
+  left_join(parms, by = 'var')%>% 
+  select(station, date, var, val, area, lbs)
 
 wqbyarea <- bind_rows(bswqtmp, rswqtmp) %>% 
   arrange(date, var, area)

### 2021-10-18 manu-draft.Rmd
gam seasonal estimate supplement plot added

@@
 Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure \\@ref(fig:map)a).
-Values below detection limits (or secchi on bottom) are not shown.'}
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure \\@ref(fig:map)a).'}
 knitr::include_graphics(here('figs/wqtrnds.jpeg'))

### 2021-10-18 supplement.Rmd
gam seasonal estimate supplement plot added

@@
 Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).
-Values below detection limits are not shown.'}
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
 knitr::include_graphics('figs/wqtrnds-supp.jpeg')
+```
+
+```{r wqgamtrnds-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
+Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
+Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.
+Results are grouped by assessment areas shown in Figure 1a.'}
+knitr::include_graphics('figs/wqgamtrnds-supp.jpeg')
 ```

### 2021-10-18 R/figures.R

@@ -1872,6 +1872,43 @@ jpeg(here('figs/wqtrnds-supp.jpeg'), height = 9, width = 10.5, units = 'in', res
 print(p)
 dev.off()
 
+## seasonal trend plots - additional variables ----------------------------
+
+# segments
+ppsegbf <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+cols <- c("#E16A86", "#50A315", "#009ADE")
+names(cols) <- c('Area 1', 'Area 2', 'Area 3')
+
+datin <- rswqdat %>% 
+  filter(date < as.Date('2021-10-01')) %>% 
+  filter(var %in% c('nh34', 'orthop', 'tp', 'dosat', 'turb', 'sal')) %>% 
+  filter(!qual %in% c('S', 'U')) %>% # remove secchi on bottom, nondetect for chla, tn
+  filter(!(var == 'sal' & source == 'ncf'))# there are some low salinity values in July in MR that skew the plots
+
+p1 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'nh34', cols, logtr = T, rmfacet = T, ttl = '(a) NH3, NH4+', ylb = 'mg/L (log-scale)')
+p2 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'tp', cols, logtr = T, rmfacet = T, ttl = '(b) Total phosphorus', ylb = 'mg/L (log-scale)')
+p3 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'orthop', cols, logtr = T, ttl = '(c) Ortho-phosphate', ylb = 'mg/L (log-scale)')
+p4 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'dosat', cols, logtr = F, rmfacet = T, ttl = '(d) Dissolved oxygen sat.', ylb = '%')
+p5 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'turb', cols, logtr = T, rmfacet = T, ttl = '(e) Turbidity', ylb = 'NTU (log-scale)')
+p6 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'sal', cols, logtr = F, ttl = '(f) Salinity', ylb = 'ppt')
[…]

### 2021-10-19 manu-draft.Rmd
removed pca analysis

@@
 data(trnsum)
-data(allpcadat)
 
@@
 
+The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.
+This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.
+GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
+The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].
+The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
+Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.
+The mgcv R package was used to create the models [@Wood17].
+
 ## Data Analysis
@@
 
-Quantitative assessments of trends included boxplot summaries, principal components analysis (PCA), Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
+Quantitative assessments of trends included descriptive summaries, Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
 Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
-Further analyses were conducted to compare all data types together using PCA and correlation analyses, to identify potential mechanisms of change using seagrasses as an endpoint for evaluating potential impacts.
-Data describing *K.
-brevis* cell concentrations were only evaluated qualitatively because the data are from event-based sampling and generally do not represent a random sample appropriate for statistical testing.
 Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.
 Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a.
-For PCA, all variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.
+All variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.
 Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.
 
-The "PCA" function from the FactoMineR R package [@Le08] was used for PCA and the "ggord" function from the ggord R package [@Beck21b] was used to plot the results.
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
@@
 Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.
-PCA and correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.
+Correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.
 
@@
 
-```{r}
-expvr <- allpcadat$eig[, 2] %>% 
-  round(0)
-```
-
-A PCA comparison of all water quality variables with seagrass and macroalgae frequency occurrence estimates showed significant associations between groups of variables (Figure \@ref(fig:allpca)).
-The first principal component axis explained `r expvr[1]`\% of the variation among all variables, whereas the second and third axes explained `r expvr[2]`\% and `r expvr[3]`\% of the remaining variation.
-For the water quality variables, the first principal component axis had strong, positive loadings for dissolved oxygen and secchi depth and strong negative loadings for total nitrogen, chl-a, and temperature.
-For macroalgae, the first axis had positive loadings with frequency occurrence of red and green macroalgae.
-For seagrasses, turtle grass (*T.
-testudinum*) had a strong positive loading with axis 1.
-The second principal component axis had strong positive loadings for secchi depth, total phosphorus, and ammonia and negative loadings for salinity and cyanobacteria macroalgae.
-The third axis showed a strong negative loading with shoal grass (*H.
-wrightii*) and positive loading with manatee grass (*S.
-filiforme*).
-Observations by area showed that samples in area 3 were distinct from points in area 1 and grouped by positive loadings on the first PCA axis and negative loadings on the second PCA axis.
-
-Pairwise Spearman rank correlations for the same variables used in the PCA showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
+Pairwise Spearman rank correlations showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
 Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$).
@@
 Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
-Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)) and loadings from the PCA results (Figure \@ref(fig:allpca)).
+Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)).
 Temporally, these changes were observed at different times and for different species of phytoplankton.
@@
 Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
-This reflects a natural partitioning of the species along a salinity gradient [Figure \@ref(fig:allpca), @Sherwood17].
+This reflects a natural partitioning of the species along a salinity gradient [@Sherwood17].
 Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient.
@@
 
-```{r allpca, fig.cap = 'Principal components analysis (PCA) for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
-Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
-All variables were standardized to zero mean and unit variance prior to PCA.
-Variables with log-normal distribution were log-transformed prior to analysis.'}
-knitr::include_graphics(here('figs/allpca.jpeg'))
-```
-
 ```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.

### 2021-10-19 R/figures.R

@@ -377,7 +377,8 @@ names(cols) <- levels(ppseg$area)
 datin <- rswqdat %>% 
   filter(date < as.Date('2021-10-01')) %>% 
   filter(var %in% c('tn', 'chla', 'secchi')) %>% 
-  filter(!qual %in% c('S', 'U')) # remove secchi on bottom, nondetect for chla, tn
+  filter(!qual %in% c('S', 'U')) %>% # remove secchi on bottom, nondetect for chla, tn
+  filter(!(var == 'secchi' & val >= 9.5)) # outlier secchi
 
 p1 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
 p2 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
@@ -570,140 +571,6 @@ print(p)
 dev.off()
 
 save(trnsum, file = here('data/trnsum.RData'))
-
-# all PCA -----------------------------------------------------------------
-
-vrs <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
-savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
-wqsel <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-wqlab <- c('Chl-a', 'DOsat', 'NH3, NH4+', 'pH', 'Sal', 'Secchi', 'Temp', 'TN', 'TP')
-
-nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
-
-# segments
-areas <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
-  mutate(
-    area = factor(area)
-  )
-
-# combine water quality with transect data at aggregated scale
-
-# add area
[…]

### 2021-10-19 manu-draft.Rmd
revised methods with new stuff, removed correlation matrix

@@
 
-## Long-term monitoring data
-
-Water quality data in Tampa Bay have been collected at fixed sampling sites since 1974 by the EPC.
-These data include samples at 45 stations used by the TBEP to assess annual progress towards programmatic goals and regulatory thresholds applicable for each of the major segments in Tampa Bay.
-Discrete water samples are collected monthly at mid-depth and processed in the laboratory immediately after collection.
-Similarly, the Parks and Natural Resources Department of Manatee County have been collecting data in Tampa Bay at locations south of Piney Point and south of the main dredged channel of Tampa Bay (approximate longitudinal axis of the bay).
-The sampling design is similar to the EPC data, with the exception that sites are sampled approximately every three months but at a higher spatial density per unit area.
-Additionally, the period of record for monitoring data from Manatee County began in 1996.
-Long-term monitoring stations for Hillsborough and Manatee Counties are shown in Figure \@ref(fig:map)a.
-
-Long-term water quality monitoring data from Hillsborough and Manatee counties were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
-Station data from Hillsborough County were obtained for the middle and lower segments of Tampa Bay.
-Station data from Manatee County were obtained for areas in lower Tampa Bay, Terra Ceia Bay, Anna Maria Sound, and northern Sarasota Bay.
-For the same water quality parameters noted above (i.e., nitrogen, phosphorus, pH, temperature, salinity), observations at each monitoring station were averaged for each month across years from 2006 to 2020.
-This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
-For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
-This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
-Spatial matching of each response monitoring station relative to the long-term monitoring stations was accomplished using the "st_nearest_feature()" function from the sf R package [@Pebesma18].
-In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
-Long-term water quality data are available from the University of South Florida Water Atlas (https://wateratlas.usf.edu/).
-
-The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.
-This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.
-GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
-The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].
-The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
-Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.
-The mgcv R package was used to create the models [@Wood17].
-
 ## Data Analysis
@@
 
-The R statistical programming language (v4.0.2) was used to import, synthesize, and analyze all datasets provided by multiple partners [@RCT20].
-Partner data were uploaded or entered manually as Google spreadsheets, where they were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.
-The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.
-The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.
+Long-term water quality monitoring data from Hillsborough and Manatee counties (Figure \@ref(fig:map)a, accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
+This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
+For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
+This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
+In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
 
-Quantitative assessments of trends included descriptive summaries, Spearman rank correlations between pairs of variables, and multiple comparison tests to assess trends between months.
+The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.
+This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.
+GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
+The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].
+The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
+Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.
+
 Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
@@
 Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a.
-All variables were standardized to zero mean and unit variance so that the central tendencies and ranges of all variables were similar.
 Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.
-
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
 Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.
+
 Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-
-When appropriate, methods were used from the NADA R package [@Lee20] to accommodate measured concentrations in water quality variables that were below detection.
-These included estimates of median, mean, and standard deviation of parameters using the "cenfit" function from NADA.
+When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.
+These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
 For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.
 Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.
-Correlation analyses were based on weekly aggregations by area, using appropriate methods for median estimates with non-detects in the water quality data.
+
+The R statistical programming language (v4.0.2) was used for all analyses [@RCT20].
+Partner data uploaded or entered manually as Google spreadsheets were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.
+The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.
+The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.
+The NADA R package [@Lee20] was used for analysis of censored data.
+All spatial analyses were done using the simple features (sf) R package [@Pebesma18].
+The mgcv R package [@Wood17] was used to create the the GAMs for water quality parameters.
 
@@
 fomo <- trnsum %>% 
+  ungroup() %>% 
   mutate(
@@
 fomoarea1 <- fomo %>% 
-  filter(area == '(a) Area 1')
+  filter(area == 'Area 1')
 fomoarea3 <- fomo %>% 
-  filter(area == '(b) Area 3')
+  filter(area == 'Area 3')
 fomoarea1apr <- fomoarea1 %>% 
@@
 No significant differences between months were observed for the green macroalgae group.
-
-## Water quality compared to seagrass and macroalgae
-
-Pairwise Spearman rank correlations showed significant associations among several of the water quality, seagrass, and macroalgae variables (Figure \@ref(fig:allcor)).
-Among the water quality variables, chlorophyll-a was positively correlated with total phosphorus, total nitrogen, temperature, and pH ($p < 0.005$ for all) and negatively correlated with secchi depth ($p < 0.05$).
-Total nitrogen was positively correlated with total phosphorus, temperature, and ammonia ($p < 0.005$, except $p < 0.05$ for temperature) and negatively correlated with secchi ($p < 0.005$).
-Among the macroalgae, green and red groups were positively correlated ($p < 0.05$), whereas red macroalgae was negatively correlated with cyanobacteria ($p < 0.005$).
-Among the seagrasses, manatee grass was negatively correlated with shoal grass and turtle grass ($p < 0.005$).
-Several notable correlations were also observed between water quality, macroalgae, and seagrass.
-Chlorophyll-a was negatively correlated with red and green macroalgae ($p < 0.05$), whereas secchi depth was positively correlated with both ($p < 0.05$) and negatively correlated with cyanobacteria ($p < 0.05$).
-For water quality and seagrass, chl-a, total nitrogen, and total phosphorus were negatively correlated with turtle grass ($p < 0.005$ for all, except $p < 0.05$ for total nitrogen), whereas secchi was positively correlated with shoal grass ($p < 0.05$).
-For macroalgae and seagrass, red and cyanobacteria macroalgae were positively correlated with manatee grass ($p < 0.005$), whereas cyanobacteria was negatively correlated with shoal grass ($p < 0.005$) and turtle grass ($p< 0.05$).
 
@@
 Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
-Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency is supported by the significant correlations (Figure \@ref(fig:allcor)).
+Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
 Temporally, these changes were observed at different times and for different species of phytoplankton.
@@
 The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
-*Dapis sp* were commonly observed at depth and entangled in blades of manatee grass (positive correlation in Figure \@ref(fig:allcor)), whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
+*Dapis sp* were commonly observed at depth and entangled in blades of manatee grass, whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
 
@@
 
-```{r allcor, fig.cap = 'Correlation matrix for water quality variables, macroalgae, and seagrass results aggregated by week from April to July 2021 in response to wastewater discharge from Piney Point.
-Only areas 1 and 3 (Figure \\@ref(fig:map)a) are included where seagrass transects were surveyed.
-Variables with log-normal distribution were log-transformed prior to analysis.
-Thick black lines separate water quality, macroalgae, and seagrass variables.
-Spearman correlations indicate the linear strength of assocation between pairs of variables.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'}
-knitr::include_graphics(here('figs/allcor.jpeg'))
-```
-
 ```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, (e) cumulative inflow in 2021 compared to past years, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa.

### 2021-10-19 R/figures.R

@@ -572,221 +572,6 @@ dev.off()
 
 save(trnsum, file = here('data/trnsum.RData'))
 
-# all correlations --------------------------------------------------------
-
-vrs <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
-mcrlab <- c("Red", "Green", "Brown", "Cyano")
-savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
-savlab <- c('T. testudinum', 'H. wrightii', 'S. filiforme')
-wqsel <- c('chla', 'dosat', 'nh34', 'ph', 'sal', 'secchi', 'temp', 'tn', 'tp')
-wqlab <- c('Chl-a', 'DOsat', 'NH3, NH4+', 'pH', 'Sal', 'Secchi', 'Temp', 'TN', 'TP')
-
-nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
-
-# segments
-areas <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
-  mutate(
-    area = factor(area)
-  )
-
-# correlate transect data at transect scale
-
-# add area
-trnsum <- rstrndat %>%
-  filter(date < as.Date('2021-10-01')) %>% 
-  mutate(
-    date = floor_date(date, unit = 'week')
-  ) %>%
-  dplyr::group_by(date, station, location, taxa) %>%
-  dplyr::summarise(
-    pa = as.numeric(any(bb > 0))
-  ) %>%
-  group_by(date, station, taxa) %>% 
-  summarize(
-    foest = sum(pa) / length(pa)
-  )
-
-trncors <- trnsum %>% 
-  ungroup %>% 
-  filter(taxa %in% c(savsel, mcrsel)) %>% 
-  spread(taxa, foest) %>% 
-  # filter(area %in% 'Area 3') %>% 
-  select(-station, -date)
-
-trncrs <- crossing(var1 = names(trncors), var2 = names(trncors)) %>% 
-  filter(var1 != var2) %>% 
-  rownames_to_column() %>% 
[…]

### 2021-10-19 manu-draft.Rmd
removed weekly wq trends plot

@@
 A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqtrnds)b).
+After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgamtrnds)b).
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
@@
 
-Boxplots of total nitrogen, chl-a, and secchi depth show the temporal progression of observations by week and area relative to the normal ranges from long-term monitoring data (Figure \@ref(fig:wqtrnds)).
-For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges beginning the week of April 4th and lasting through the month.
-Concentrations remained similar to baseline conditions until June and July when median values were often above the baseline range.
-Secchi observations in area 1 were below baseline ranges in April and July.
-Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 during weeks of May, June, and July.
-Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqtrnds).
+Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgamtrnds)).
+For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges during April.
+Concentrations decreased slightly until June and July when values increased again above the seasonal expectation.
+Many secchi observations in area 1 were lower than normal in April and July.
+Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 in  May, June, and July.
+Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgamtrnds).
 Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for areas 1 and 3 and total nitrogen and chl-a for area 2 (Table \@ref(tab:wqcmptab)).
@@
 These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap) and \@ref(fig:wqtrnds), Table \@ref(tab:wqsumtab)).
+The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
 Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
@@
 
-```{r wqtrnds, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
-Observations are aggregated by week and within assessment areas shown in Figure \\@ref(fig:map)a.
-Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure \\@ref(fig:map)a).'}
-knitr::include_graphics(here('figs/wqtrnds.jpeg'))
-```
-
 ```{r wqgamtrnds, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.

### 2021-10-19 R/figures.R

@@ -358,40 +358,6 @@ jpeg(here('figs/wqmap.jpeg'), height = 7, width = 12, units = 'in', res = 500, f
 print(p)
 dev.off()
 
-# weekly plots ------------------------------------------------------------
-
-# segments
-ppsegbf <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
-  mutate(
-    area = factor(area)
-  )
-
-cols <- c("#E16A86", "#50A315", "#009ADE")
-names(cols) <- levels(ppseg$area)
-
-datin <- rswqdat %>% 
-  filter(date < as.Date('2021-10-01')) %>% 
-  filter(var %in% c('tn', 'chla', 'secchi')) %>% 
-  filter(!qual %in% c('S', 'U')) %>% # remove secchi on bottom, nondetect for chla, tn
-  filter(!(var == 'secchi' & val >= 9.5)) # outlier secchi
-
-p1 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
-p2 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
-p3 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'secchi', cols, logtr = FALSE, ttl = '(c) Secchi', ylb = 'meters')
-
-p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Week of', gp = gpar(fontsize=14))) + 
-  plot_layout(ncol = 1, guides = 'collect', height = c(1, 0.05)) & 
-  theme(legend.position = 'top')
-
-jpeg(here('figs/wqtrnds.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
-print(p)
-dev.off()
-
 # seasonal trend plots ----------------------------------------------------
 
 # segments

### 2021-10-20 manu-draft.Rmd
seasonally corrected mult comp for wq table done

@@
 
-Long-term water quality monitoring data from Hillsborough and Manatee counties (Figure \@ref(fig:map)a, accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
 Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
@@
 
-The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.
-This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.
+The historical monitoring data were also used to model an expected seasonal pattern for water quality parameters from April to October in 2021.
+This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).
 GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
@@
 The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
-Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.
+Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs through 2021 provided an assessment of how the current data may have deviated from historical and normal seasonal variation.
 
-Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release.
+Statistical assessments were evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
 Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
+For water quality variables, long-term monthly averages by area were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.
+Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
 When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.
 These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
-For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.
-Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.
 

### 2021-10-20 R/tables.R

@@ -208,6 +208,32 @@ ppsegbf <- ppseg %>%
     area = factor(area)
   )
 
+# monthly baseline averages by station
+bswqtmp <- bswqdat %>% 
+  select(-source, -uni) %>% 
+  filter(var %in% vrs) %>%
+  filter(!(var == 'secchi' & grepl('S', qual))) %>% # remove secchi on bottom
+  filter(yr > 2005) %>% 
+  filter(!is.na(val)) %>% 
+  inner_join(bsstatloc, ., by = 'station') %>% 
+  st_intersection(ppsegbf) %>% 
+  st_set_geometry(NULL) %>% 
+  mutate(
+    cens = grepl('U', qual), 
+    mo = month(date, label = T)
+  ) %>% 
+  group_by(mo, var, area) %>% 
+  summarise(   
+    avev = ifelse(
+      any(cens), mean(cenfit(val, cens), na.rm = T),
+      mean(val, na.rm = T)
+    ),
+    .groups = 'drop'
+  ) %>% 
+  filter(mo %in% c('Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep')) %>% 
+  mutate(mo = fct_drop(mo))
+
+# prep response data for mult comp
 rswqsub <- rswqdat %>% 
   filter(var %in% vrs) %>% 
   filter(date < as.Date('2021-10-01')) %>% 
@@ -217,10 +243,15 @@ rswqsub <- rswqdat %>%
   inner_join(rsstatloc, ., by = c('station', 'source')) %>% 
   st_intersection(ppsegbf) %>% 
   st_set_geometry(NULL) %>% 
-  left_join(parms, by = c('var', 'lbs')) %>% 
-  select(date, var, lbs, val, sigdig, station, area, qual) %>% 
+  left_join(parms, by = c('var', 'lbs', 'uni')) %>% 
+  select(date, var, lbs, val, sigdig, station, area, qual, nrmrng) %>% 
+  separate(nrmrng, into = c('minv', 'maxv'), sep = '-') %>% 
   mutate(
-    mo = month(date, label = T)
+    mo = month(date, label = T), 
+    minv = as.numeric(minv), 
+    maxv = as.numeric(maxv),
+    avev = minv + (maxv - minv) / 2, 
+    valcr = val - avev
   ) %>% 
   filter(mo %in% c('Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep')) %>% 
[…]

### 2021-10-20 manu-draft.Rmd
added sg and macro abundance tables and figures

@@
 load(file = here('tables/wqcmptab.RData'))
-load(file = here('tables/savtab.RData'))
-load(file = here('tables/mcrtab.RData'))
+load(file = here('tables/savfotab.RData'))
+load(file = here('tables/mcrfotab.RData'))
 
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-For water quality variables, long-term monthly averages by area were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.
+For water quality variables, long-term monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.
 Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
@@
 Physical parameters (salinity, temperature) were generally within range over the study period.
-Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the effluent measurements in Table \@ref(tab:stktab).
+Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the discharge measurements in Table \@ref(tab:stktab).
 Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
@@
 Many secchi observations in area 1 were lower than normal in April and July.
-Observations in areas 2 and 3 were more often within the normal range, with some exceptions for total nitrogen and chl-a in area 3 in  May, June, and July.
+Observations in areas 2 and 3 were more often within the normal seasonal range, with some exceptions for total nitrogen and chl-a in area 3 in  April, May, and July.
 Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgamtrnds).
-Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for areas 1 and 3 and total nitrogen and chl-a for area 2 (Table \@ref(tab:wqcmptab)).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1 and for total nitrogen and chl-a for areas 2 and 3 (Table \@ref(tab:wqcmptab)).
 Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
@@
   filter(area == 'Area 3')
+fomoarea1sg <- fomoarea1 %>% 
+  group_by(taxa) %>% 
+  summarise(foest = mean(foest)) %>% 
+  deframe
+
 fomoarea1apr <- fomoarea1 %>% 
@@
 # kw pvals
-cyanoarea3 <- mcrtab %>% 
+cyanoarea3 <- mcrfotab %>% 
   mutate(
@@
 
-A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through July, each visited on average `r avevis` times per month.
+A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through September, each visited on average `r avevis` times per month.
 Turtle grass (*T.
@@
 The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
-Overall abundance of seagrass did not change dramatically from April 22nd through July.
+Overall abundance of seagrass did not change dramatically from April 22nd through September.
 However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
@@
 Green macroalgae taxa were first observed in July, although at generally low abundances.
-Trends in macroalgae somewhat followed seasonal patterns observed baywide with red macroalgae transitioning to green macroalgae by mid-summer.
+Red macroalgae were the dominant taxa at the end of September.
 
 Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of seagrass and macroalgae trends across all transects.
-No transects were sampled in area 2 to the north of Piney Point.
-Area 1 had generally balanced frequency occurrence of the three primary seagrass species observed in the bay, although a slight decrease over time in shoal grass (*H.
+No transects were sampled in area 2 to the north of Piney Point and no transects were sampled past September in Area 1.
+Area 1 had generally stable frequency occurrence over time of the three primary seagrass species observed in the bay, with turtle grass (*T.
+testudinum*) occurring slightly more often (average frequency occurrence across months `r fomoarea1sg['Thalassia testudinum']`\%) than shoal grass (*H.
+wrightii*, `r fomoarea1sg['Halodule wrightii']`\%)) or manatee grass (*S.
+filiforme*, `r fomoarea1sg['Syringodium filiforme']`\%)).
+
+, although a slight decrease over time in shoal grass (*H.
 wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea1jul['Halodule wrightii']`\%).
@@
 syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, July frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
+
 Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
@@
 
-Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrtab) and \@ref(tab:savtab) further suggested that seagrass trends were minor compared to those of macroalgae.
+Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrfotab) and \@ref(tab:savfotab) further suggested that seagrass trends were minor compared to those of macroalgae.
 Results by month showed that seagrass frequency occurrence, by species and area, had insignificant changes ($p > 0.05$) over the study period.
-Although trends were not significant, estimated medians in each month in Table \@ref(tab:savtab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.
+Although trends were not significant, estimated medians in each month in Table \@ref(tab:savfotab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.
 These changes were more pronounced in area 3 as compared to area 1.
-For macroalgae (Table \@ref(tab:mcrtab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).
+For macroalgae (Table \@ref(tab:mcrfotab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).
 Coverage of cyanobacteria in area 1 during June was significantly higher than coverage in April.
@@
 
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 3 (Figure \\@ref(fig:map)a) for seagrass (top) and macroalgae (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 3 (Figure \\@ref(fig:map)a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
 Estimates are grouped by sample months in 2021.
-Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.')}
+Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.
+Points are offset slightly for readability.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))
@@
 
-```{r mcrtab}
+```{r mcrfotab}
 cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- mcrtab %>% 
+totab <- mcrfotab %>% 
   select(-pval)
@@
 
-```{r savtab}
+```{r savfotab}
 cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- savtab %>% 
+totab <- savfotab %>% 
   select(-pval)

### 2021-10-20 supplement.Rmd
added sg and macro abundance tables and figures

@@
   })
+
+box::use(
+  here[here],
+  dplyr[...]
+)
+
+load(file = here('tables/mcrabutab.RData'))
+load(file = here('tables/savabutab.RData'))
 ```
+
+# Figures
 
@@
 knitr::include_graphics('figs/wqgamtrnds-supp.jpeg')
+```
+
+```{r trnabu, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
+Estimates are grouped by sample months in 2021.
+Points are offset slightly for readability.'}
+knitr::include_graphics(here('figs/trnabu.jpeg'))
 ```
@@
 ```
+
+# Tables
+
+```{r mcrabutab}
+cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly abundances for each macroalgae group in each area.
+Rows that share a letter within each area and macroalgae group combination have abundances that are not significantly different between month pairs.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
+totab <- mcrabutab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
+```
+
+```{r savabutab}
+cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly abudances for each seagrass species in each area.
+Rows that share a letter within each area and seagrass species combination have abundances that are not significantly different between month pairs.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
+totab <- savabutab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
+```

### 2021-10-20 R/figures.R

@@ -452,31 +452,41 @@ areas <- ppseg %>%
   )
 
 # # view sample effort by transect, area, month
-# smpeff <- rstrndat %>% 
-#   left_join(rstrnpts, ., by = 'station') %>% 
-#   select(-source, -type, -lng, -lat) %>% 
-#   st_intersection(areas) %>% 
-#   st_set_geometry(NULL) %>% 
+# smpeff <- rstrndat %>%
+#   inner_join(rstrnpts, ., by = 'station') %>%
+#   select(-source, -type, -lng, -lat) %>%
+#   st_intersection(areas) %>%
+#   st_set_geometry(NULL) %>%
 #   mutate(
 #     mo = month(date)
-#   ) %>% 
-#   group_by(station, area, mo) %>% 
+#   ) %>%
+#   group_by(station, area, mo) %>%
 #   summarise(
 #     obs = (any(bb > 0)),
 #     .groups = 'drop'
-#   ) %>% 
-#   spread(mo, obs)
+#   ) %>%
+#   spread(mo, obs) %>% 
+#   arrange(area, station)
 # View(smpeff)
+
+# tokp <- smpeff %>%
+#   gather('var', 'val', -station, -area) %>%
+#   na.omit() %>%
+#   group_by(station, area) %>%
+#   summarise(cnt = sum(val), .groups= 'drop') %>%
+#   filter(cnt >= 4) %>%
+#   pull(station)
 
 # add area
 trnsum <- rstrndat %>%
+  # filter(station %in% tokp) %>%
   inner_join(rstrnpts, ., by = 'station') %>% 
   st_intersection(areas) %>% 
   st_set_geometry(NULL) %>%
   dplyr::group_by(area, typ, date, station, taxa, location) %>%
   dplyr::summarise(
     pa = as.numeric(any(bb > 0))
-  ) %>%  
+  ) %>% 
   mutate(
     date = floor_date(date, unit = 'month'), 
     typ = factor(typ, levels = c('mcr', 'sav'), labels = c('Macroalgae', 'Seagrass'))
@@ -484,10 +494,11 @@ trnsum <- rstrndat %>%
   group_by(area, typ, date, taxa) %>% 
   summarize(
     foest = sum(pa) / length(pa)
-  ) %>% 
+  ) %>%
   filter(taxa %in% c(mcrsel, savsel))
 
 trnsumtots <- rstrndat %>%
+  # filter(station %in% tokp) %>% 
   inner_join(rstrnpts, ., by = 'station') %>% 
[…]

### 2021-10-20 R/tables.R

@@ -346,11 +346,11 @@ cmps <- rswqsub %>%
 wqcmptab <- cmps
 save(wqcmptab, file = here('tables/wqcmptab.RData'))
 
-# sav and mcr trends by month ---------------------------------------------
+# fo sav and mcr trends by month ------------------------------------------
 
 mcrsel <- c("Red", "Green", "Cyanobacteria")
 savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
-
+  
 # segments
 areas <- ppseg %>% 
   rename(area = Name) %>% 
@@ -505,8 +505,173 @@ mcrests <- ests %>%
     `Med. (Min., Max.)` = sumv
   )
 
-mcrtab <- mcrests
-savtab <- savests
+mcrfotab <- mcrests
+savfotab <- savests
 
-save(mcrtab, file = here('tables/mcrtab.RData'))
-save(savtab, file = here('tables/savtab.RData'))
+save(mcrfotab, file = here('tables/mcrfotab.RData'))
+save(savfotab, file = here('tables/savfotab.RData'))
+
+# abundance sav and mcr trends by month -----------------------------------
+
+mcrsel <- c("Red", "Green", "Cyanobacteria")
+savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
+
+# segments
+areas <- ppseg %>% 
+  rename(area = Name) %>% 
+  group_by(area) %>% 
+  summarise() %>% 
+  st_buffer(dist = set_units(0.0001, degree)) %>% 
+  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  mutate(
+    area = factor(area)
+  )
+
+# # view sample effort by transect, area, month
+# smpeff <- rstrndat %>%
+#   mutate(
+#     mo = month(date)
+#   ) %>%
+#   group_by(station, mo) %>%
+#   summarise(
+#     obs = (any(bb > 0)),
+#     .groups = 'drop'
+#   ) %>% 
+#   group_by(station) %>% 
+#   summarise(
+#     cnt = n(), 
+#     .groups = 'drop'
+#   ) %>% 
+#   filter(cnt >= 4) %>% 
+#   pull(station)
+
+# add area
+trnsf <- rstrndat %>%
+  filter(date < as.Date('2021-10-01')) %>% 
+  inner_join(rstrnpts, ., by = 'station') %>% 
[…]

### 2021-10-20 manu-draft.Rmd
transect example picture dates descending, x axis labels changed

@@
 
-```{r trnex, fig.cap = 'Results for (a) seagrass and (b) macroalgae rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.
+```{r trnex, fig.cap = 'Results for (a) macroalgae and (b) seagrass rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.
 Sample dates in 2021 are shown in rows with transect meter results shown in columns.

### 2021-10-20 R/figures.R

@@ -418,7 +418,7 @@ savsel <- savdat %>%
   unique %>% 
   as.character()
 
-p <- show_rstransect(savdat, mcrdat, savsel, mcrsel)
+p <- show_rstransect(savdat, mcrdat, savsel, mcrsel, rev = T, sclloc = T)
 
 jpeg(here('figs/trnex.jpeg'), height = 7, width = 8, units = 'in', res = 500, family = 'serif')
 print(p)

### 2021-10-20 R/funcs.R

@@ -308,7 +308,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
 
 # function for plotting rapid response transect data
 # modified from show_transect in tbpetools
-show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12){
+show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12, rev = F, sclloc = T){
   
   savlevs <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme', 'Ruppia maritima', 'Halophila engelmannii', 'Halophila decipiens')
   grplevs <- c('Red', 'Green', 'Brown', 'Cyanobacteria')
@@ -332,6 +332,14 @@ show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12){
   
   xlms <- range(savxlms, mcrxlms)
   
+  # x axis text labels, scaling location to zero if sclloc
+  savxlbs <- savxlms
+  mcrxlbs <- mcrxlms
+  if(sclloc){
+    savxlbs <- savxlbs - min(savxlbs)
+    mcrxlbs <- mcrxlbs - min(mcrxlbs)
+  }
+  
   # get dates for factor levels
   # this makes sure that y values on plots are shared
   dts1 <- savdat %>% 
@@ -342,6 +350,9 @@ show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12){
     unique %>%
     sort %>% 
     format('%b %d')
+  
+  if(rev)
+    dts <- rev(dts)
   
   # prep sav plot data
   savdatfrm <- savdat %>%
@@ -401,10 +412,10 @@ show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12){
     ggplot2::geom_point(data = toplo2a, alpha = 1, colour = 'black', size = 2) +
     ggplot2::geom_point(aes(size = bb, fill = taxa), alpha = 0.8, pch = 21) +
     ggplot2::scale_fill_manual(values = savcol) +
-    ggplot2::scale_radius(limits = range(abubrks), labels = abulabs, breaks = abubrks, range = szrng) +
+    ggplot2::scale_radius(limits = range(abubrks), labels = abulabs, breaks = abubrks, range = szrng, guide = 'none') +
     ggplot2::theme_minimal(base_size = base_size) +
[…]

### 2021-10-20 manu-draft.Rmd
working on methods revision, mult comp tests for sav, mcr to supp

@@
 load(file = here('tables/wqcmptab.RData'))
-load(file = here('tables/savfotab.RData'))
-load(file = here('tables/mcrfotab.RData'))
 
@@
 
-## Seagrass and macroalgae trends
+## Macroalgae and seagrass trends
 
@@
   filter(area == 'Area 3')
-fomoarea1sg <- fomoarea1 %>% 
+fomoarea1ave <- fomoarea1 %>% 
   group_by(taxa) %>% 
-  summarise(foest = mean(foest)) %>% 
+  summarise(foest = round(mean(foest), 1)) %>% 
+  deframe
+fomoarea3ave <- fomoarea3 %>% 
+  group_by(taxa) %>% 
+  summarise(foest = round(mean(foest), 1)) %>% 
   deframe
@@
   deframe
-
-# kw pvals
-cyanoarea3 <- mcrfotab %>% 
-  mutate(
-    Area = ifelse(Area == '', NA, Area)
-  ) %>% 
-  fill(Area) %>% 
-  filter(Area == 3 & `Macroalgae group` == 'Cyanobacteria') %>% 
-  pull(pval) %>% 
-  as.numeric %>% 
-  round(3)
+fomoarea1sep <- fomoarea1 %>% 
+  filter(mo %in% 9) %>% 
+  select(taxa, foest) %>% 
+  deframe
+fomoarea3oct <- fomoarea3 %>% 
+  filter(mo %in% 10) %>% 
+  select(taxa, foest) %>% 
+  deframe
 ```
 
-A total of `r ntrns` transects were sampled for seagrasses and macroalgae from April through September, each visited on average `r avevis` times per month.
-Turtle grass (*T.
-testudinum*) was the dominant seagrass species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
+A total of `r ntrns` transects were sampled for macroalgae and seagrass from April through September, each visited on average `r avevis` times per month.
+Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
+Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.
+Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
+Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
+Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
+For seagrasses, turtle grass (*T.
+testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
 Manatee grass (*S.
@@
 At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.
-Macroalgae observed along the transects also varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
-Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.
-Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
-Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
-Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
 
-Temporal progression of seagrasses and macroalgae varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
+Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
 Transect S3T6 is located less than one kilometer to the north of Port Manatee.
-The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
-Overall abundance of seagrass did not change dramatically from April 22nd through September.
-However, macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
+Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
 Red macroalgae were present in high abundances from April to May.
@@
 Green macroalgae taxa were first observed in July, although at generally low abundances.
-Red macroalgae were the dominant taxa at the end of September.
+Red macroalgae were the dominant taxa by the end of September.
+Overall abundance of seagrass did not change dramatically from April 22nd through September.
+The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
 
-Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of seagrass and macroalgae trends across all transects.
+Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.
 No transects were sampled in area 2 to the north of Piney Point and no transects were sampled past September in Area 1.
-Area 1 had generally stable frequency occurrence over time of the three primary seagrass species observed in the bay, with turtle grass (*T.
-testudinum*) occurring slightly more often (average frequency occurrence across months `r fomoarea1sg['Thalassia testudinum']`\%) than shoal grass (*H.
-wrightii*, `r fomoarea1sg['Halodule wrightii']`\%)) or manatee grass (*S.
-filiforme*, `r fomoarea1sg['Syringodium filiforme']`\%)).
-
-, although a slight decrease over time in shoal grass (*H.
-wrightii*) was observed (April frequency occurrence of `r fomoarea1apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea1jul['Halodule wrightii']`\%).
-A similar decline in coverage of shoal grass was also observed in area 3 (April frequency occurrence of `r fomoarea3apr['Halodule wrightii']`\%, July frequency occurrence of `r fomoarea3jul['Halodule wrightii']`\%).
-Frequency occurrence of manatee grass (*S.
-syringodium*) also declined slightly from April to July in area 3 (April frequency occurrence of `r fomoarea3apr['Syringodium filiforme']`\%, July frequency occurrence of `r fomoarea3jul['Syringodium filiforme']`\%).
-
-Changes in macroalgae frequency occurrence were much more pronounced than for seagrasses across months and by area.
+Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).
 Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3).
 Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
-Green macroalgae had the second lowest frequency occurrence, followed by Brown macroalgae that was only observed at one transect.
-
-Statistical comparisons of results in Figure \@ref(fig:trnfrq) shown in Tables \@ref(tab:mcrfotab) and \@ref(tab:savfotab) further suggested that seagrass trends were minor compared to those of macroalgae.
-Results by month showed that seagrass frequency occurrence, by species and area, had insignificant changes ($p > 0.05$) over the study period.
-Although trends were not significant, estimated medians in each month in Table \@ref(tab:savfotab) showed similar changes for seagrasses as in Figure \@ref(fig:trnex), with generally higher coverage estimates in April.
-These changes were more pronounced in area 3 as compared to area 1.
-For macroalgae (Table \@ref(tab:mcrfotab)), significant differences between months were observed for red and cyanobacteria macroalgae in both areas 1 and 3, although cyanobacteria differences for area 3 were only marginally significant ($p = `r cyanoarea3`$).
-Coverage of cyanobacteria in area 1 during June was significantly higher than coverage in April.
-No significant differences between months were observed for the green macroalgae group.
+Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in area 1, `r fomoarea3oct['Green']`\% in October in area 3).
+Brown macroalgae was only observed at one transect.
+For seagrass, both areas had generally stable total frequency occurrence.
+Turtle grass (*T.
+testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall area 3), compared to shoal grass (*H.
+wrightii*, `r fomoarea1ave['Halodule wrightii']`\% area 1, `r fomoarea3ave['Halodule wrightii']`\% area 3) and manatee grass (*S.
+filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% area 1, `r fomoarea3ave['Syringodium filiforme']`\% area 3).
+Slight changes in frequency occurrence in area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
+Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).
 
@@
 
-```{r mcrfotab}
-cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp.
-column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
-Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- mcrfotab %>% 
-  select(-pval)
-knitr::kable(totab, booktabs = T, caption = cap.val)
-```
-
-```{r savfotab}
-cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
-Multiple comparisons with Mann-Whitney U tests (Comp.
-column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
-Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
-** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- savfotab %>% 
-  select(-pval)
-knitr::kable(totab, booktabs = T, caption = cap.val)
-```
-
 # References

### 2021-10-20 supplement.Rmd
working on methods revision, mult comp tests for sav, mcr to supp

@@
 
+load(file = here('tables/mcrfotab.RData'))
+load(file = here('tables/savfotab.RData'))
 load(file = here('tables/mcrabutab.RData'))
@@
 
+```{r mcrfotab}
+cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly frequency occurrences for each macroalgae group in each area.
+Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
+totab <- mcrfotab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
+```
+
+```{r savfotab}
+cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Multiple comparisons with Mann-Whitney U tests (Comp.
+column) were used to evaluate pairwise monthly frequency occurrences for each seagrass species in each area.
+Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs.
+Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
+totab <- savfotab %>% 
+  select(-pval)
+knitr::kable(totab, booktabs = T, caption = cap.val)
+```
+
 ```{r mcrabutab}

### 2021-10-21 manu-draft.Rmd
week trend plots to supp, changing file suffix for supp

@@
 A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgamtrnds)b).
+After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgam)b).
 Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
@@
 
-Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgamtrnds)).
+Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).
 For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges during April.
@@
 Observations in areas 2 and 3 were more often within the normal seasonal range, with some exceptions for total nitrogen and chl-a in area 3 in  April, May, and July.
-Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgamtrnds).
+Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
 Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1 and for total nitrogen and chl-a for areas 2 and 3 (Table \@ref(tab:wqcmptab)).
@@
 
-```{r wqgamtrnds, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
+```{r wqgam, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
 Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
@@
 Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
-knitr::include_graphics(here('figs/wqgamtrnds.jpeg'))
+knitr::include_graphics(here('figs/wqgam.jpeg'))
 ```

### 2021-10-21 supplement.Rmd
week trend plots to supp, changing file suffix for supp

@@
 
-load(file = here('tables/mcrfotab.RData'))
-load(file = here('tables/savfotab.RData'))
-load(file = here('tables/mcrabutab.RData'))
-load(file = here('tables/savabutab.RData'))
+load(file = here('tables/mcrfotabsupp.RData'))
+load(file = here('tables/savfotabsupp.RData'))
+load(file = here('tables/mcrabutabsupp.RData'))
+load(file = here('tables/savabutabsupp.RData'))
 ```
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+Observations are aggregated by week and within assessment areas shown in Figure 1a.
+Normal ranges for the month of observation (monthly baseline) and area are shown by the blue shaded areas.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at long-term monitoring sites within each area (Figure 1a).'}
+knitr::include_graphics(here('figs/wqtrnds-supp.jpeg'))
+```
+
+```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at long-term monitoring sites within each area (Figure 1a).'}
-knitr::include_graphics('figs/wqtrnds-supp.jpeg')
+knitr::include_graphics('figs/wqtrndsadd-supp.jpeg')
 ```
 
-```{r wqgamtrnds-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
+```{r wqgamadd-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
 Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
@@
 Results are grouped by assessment areas shown in Figure 1a.'}
-knitr::include_graphics('figs/wqgamtrnds-supp.jpeg')
+knitr::include_graphics('figs/wqgamadd-supp.jpeg')
 ```
 
-```{r trnabu, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
+```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
 Estimates are grouped by sample months in 2021.
 Points are offset slightly for readability.'}
-knitr::include_graphics(here('figs/trnabu.jpeg'))
+knitr::include_graphics(here('figs/trnabu-supp.jpeg'))
 ```
 
-```{r windroses, fig.cap = 'Wind rose plots for 2021 by month.
+```{r windroses-supp, fig.cap = 'Wind rose plots for 2021 by month.
 Data are from St.
@@
 Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
-knitr::include_graphics('figs/windroses.jpeg')
+knitr::include_graphics('figs/windroses-supp.jpeg')
 ```
 
-```{r phymcrfoest, fig.cap = 'Frequency occurrence estimates for (a) phytoplankton (diatoms and *K.
+```{r phymcrfoest-supp, fig.cap = 'Frequency occurrence estimates for (a) phytoplankton (diatoms and *K.
 brevis*) and (b) macroalgae groups.
@@
 Note that frequency occurrence is not a precise measure of cell concentrations.'}
-knitr::include_graphics('figs/phymcrfoest.jpeg')
+knitr::include_graphics('figs/phymcrfoest-supp.jpeg')
 ```
@@
 
-```{r mcrfotab}
-cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+```{r mcrfotabsupp}
+cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure 1a) and month.
 Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- mcrfotab %>% 
+totab <- mcrfotabsupp %>% 
   select(-pval)
@@
 
-```{r savfotab}
-cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure \\@ref(fig:map)a) and month.
+```{r savfotabsupp}
+cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure 1a) and month.
 Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- savfotab %>% 
+totab <- savfotabsupp %>% 
   select(-pval)
@@
 
-```{r mcrabutab}
-cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.
+```{r mcrabutabsupp}
+cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure 1a) and month.
 Overall signifance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- mcrabutab %>% 
+totab <- mcrabutabsupp %>% 
   select(-pval)
@@
 
-```{r savabutab}
-cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure \\@ref(fig:map)a) and month.
+```{r savabutabsupp}
+cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure 1a) and month.
 Overall signifance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
@@
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
-totab <- savabutab %>% 
+totab <- savabutabsupp %>% 
   select(-pval)
@@
 ```
+
+# References

### 2021-10-21 R/figures.R

@@ -385,7 +385,7 @@ p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Day 
   plot_layout(ncol = 1, guides = 'collect', height = c(1, 0.05)) & 
   theme(legend.position = 'top')
 
-jpeg(here('figs/wqgamtrnds.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
+jpeg(here('figs/wqgam.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -548,142 +548,6 @@ print(p)
 dev.off()
 
 save(trnsum, file = here('data/trnsum.RData'))
-
-# transect abundance ------------------------------------------------------
-
-mcrsel <- c("Red", "Green", "Brown", "Cyanobacteria")
-savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
-
-colpal <- colorRampPalette(RColorBrewer::brewer.pal(n = 8, name = 'Dark2'))
-savlevs <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme', 'Ruppia maritima', 'Halophila engelmannii', 'Halophila decipiens')
-savcol <- colpal(length(savlevs))
-names(savcol) <- savlevs
-savcol <- savcol[savsel]
-mcrcol <- c('tomato1', 'lightgreen', 'burlywood3', 'lightblue')
-names(mcrcol) <- mcrsel
-mcrcol <- mcrcol[mcrsel]
-cols <- c(mcrcol, savcol)
-cols <- c(cols, Total = 'white')
-
-# segments
-areas <- ppseg %>% 
-  rename(area = Name) %>% 
-  group_by(area) %>% 
-  summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
-  mutate(
-    area = factor(area)
-  )
-
-# # view sample effort by transect, area, month
-# smpeff <- rstrndat %>%
-#   inner_join(rstrnpts, ., by = 'station') %>%
-#   select(-source, -type, -lng, -lat) %>%
-#   st_intersection(areas) %>%
-#   st_set_geometry(NULL) %>%
-#   mutate(
-#     mo = month(date)
[…]

### 2021-10-21 R/tables.R

@@ -346,7 +346,9 @@ cmps <- rswqsub %>%
 wqcmptab <- cmps
 save(wqcmptab, file = here('tables/wqcmptab.RData'))
 
-# fo sav and mcr trends by month ------------------------------------------
+# Supplement tables -------------------------------------------------------
+
+## fo sav and mcr trends by month -----------------------------------------
 
 mcrsel <- c("Red", "Green", "Cyanobacteria")
 savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
@@ -505,13 +507,13 @@ mcrests <- ests %>%
     `Med. (Min., Max.)` = sumv
   )
 
-mcrfotab <- mcrests
-savfotab <- savests
+mcrfotabsupp <- mcrests
+savfotabsupp <- savests
 
-save(mcrfotab, file = here('tables/mcrfotab.RData'))
-save(savfotab, file = here('tables/savfotab.RData'))
+save(mcrfotabsupp, file = here('tables/mcrfotabsupp.RData'))
+save(savfotabsupp, file = here('tables/savfotabsupp.RData'))
 
-# abundance sav and mcr trends by month -----------------------------------
+## abundance sav and mcr trends by month ----------------------------------
 
 mcrsel <- c("Red", "Green", "Cyanobacteria")
 savsel <- c('Thalassia testudinum', 'Halodule wrightii', 'Syringodium filiforme')
@@ -669,9 +671,9 @@ mcrests <- ests %>%
     `Med. (Min., Max.)` = sumv
   )
 
-mcrabutab <- mcrests
-savabutab <- savests
+mcrabutabsupp <- mcrests
+savabutabsupp <- savests
 
-save(mcrabutab, file = here('tables/mcrabutab.RData'))
-save(savabutab, file = here('tables/savabutab.RData'))
+save(mcrabutabsupp, file = here('tables/mcrabutabsupp.RData'))
+save(savabutabsupp, file = here('tables/savabutabsupp.RData'))
 

### 2021-10-21 manu-draft.Rmd
results revision done with updated figs, ord plot to supp

@@
 
-## Red tide and fish kill reports
+## Red tide impacts
 
@@
   filter(date < maxdt)
-fishann <- fishdat %>% 
-  group_by(yr, city) %>% 
-  summarise(cnt = n(), .groups = 'drop') %>% 
-  group_by(city) %>% 
-  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
-  mutate(per = round(100 * cnt / sum(cnt), 1))
-annstpper <- fishann %>% 
-  filter(city == 'St. Petersburg') %>% 
-  pull(per)
-anntmpper <- fishann %>% 
-  filter(city == 'Tampa') %>% 
-  pull(per)
 fishwk <- fishdat %>% 
@@
 The increase in *K.
-brevis* from April to July to bloom concentrations exceeding 10k cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2001 and 2012, although several years had concentrations above the median that were at bloom levels.
+brevis* from April to July to bloom concentrations exceeding 10,000 cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2005 and 2018.
 Median cell concentrations for most years were well below 1,000 cells/L.
@@
 Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical baseline values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
+This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K.
+brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.
 
-Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL somewhat tracked the annual *K.
-brevis* cell concentrations, with a notable exception in 2021 which had the most reports among all years (Figure \@ref(fig:redtide)c).
-Historically, more fish kills were reported in Saint Petersburg (`r annstpper`%) as compared to Tampa (`r anntmpper`%), where the former is closer to the mouth of Tampa Bay.
-In 2021, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa (Figure \@ref(fig:redtide)d).
+Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations (Figure \@ref(fig:redtide)e).
+In total, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa.
 The combined weekly reports in 2021 for Tampa and Saint Petersburg peaked the week of July 4th, the same week as the peak of *K.
@@
 Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.
-Increased reports in early July coincided with a shift in winds from Tropical Storm Elsa which moved dead fish closer to populated nearshore areas, as noted earlier.
-As a result, Pinellas County and the city of St.
+Increased fish kill reports in early July may coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f), which moved dead fish closer to populated nearshore areas, as noted earlier.
+By August, Pinellas County and the city of St.
 Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
@@
 2021).
+
+## Nutrient cycling
+
+The above results can be considered together to develop a concise narrative of the temporal shift of nutrient pools between different ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point.
+Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).
+This initial peak in water quality parameters indicated a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).
+Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.
+However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
+During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
+brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).
+The co-occurring decline in macroalgae and increase in *K.
+brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients were likely present (as noted by modelling efforts).
+Conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
+These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).
 
@@
 
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) fish kill reports in 2021, (d) cumulative precipitation in 2021 compared to past years, (e) cumulative inflow in 2021 compared to past years, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa.
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa.
 *K.
@@
 
-```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) *Karenia brevis* cell concentrations, and (e) cyanobacteria macroalgal abundances.
+```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) cyanobacteria macroalgal abundances, and (e) *Karenia brevis* cell concentrations.
 Values are summarized for all samples within each week.
@@
 knitr::include_graphics('figs/nutrientflow.jpeg')
-```
-
-```{r nutrientfloword, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.
-Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.
-Observations are grouped by month periods based on phytoplankton (diatoms or *K.
-brevis*) or macroalgal dominance.'}
-knitr::include_graphics('figs/nutrientfloword.jpeg')
 ```

### 2021-10-21 supplement.Rmd
results revision done with updated figs, ord plot to supp

@@
 
+```{r nutrientfloword-supp, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.
+Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.
+Observations are grouped by month periods based on phytoplankton (diatoms or *K.
+brevis*) or macroalgal dominance.'}
+knitr::include_graphics('figs/nutrientfloword-supp.jpeg')
+```
+
 # Tables

### 2021-10-21 R/figures.R

@@ -689,39 +689,6 @@ p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
     title = expression(paste('(b) ', italic('K. brevis'), ' concentrations in 2021 by week, middle/lower Tampa Bay'))
   )
 
-toplo1 <- fishdat %>% 
-  filter(city %in% c('Tampa', 'St. Petersburg')) %>% 
-  filter(yr >= 2021) %>%  
-  group_by(week, city) %>% 
-  summarise(
-    cnt = n(), 
-    .groups = 'drop'
-  ) %>% 
-  mutate(
-    week = factor(week, levels = weeklv)
-  ) %>% 
-  complete(week)
-
-p3 <- ggplot(toplo1, aes(x = week, fill = city, y = cnt)) + 
-  geom_bar(stat = 'identity', colour = 'darkgrey') + 
-  labs(
-    x = 'Week of',
-    y = 'No. of reports', 
-    title = '(c) Fish kill reports for red tide in 2021 by week'
-  ) +
-  scale_y_continuous(expand = c(0, 0)) +
-  scale_fill_brewer('City', palette = 'Pastel1') + 
-  theme_minimal() + 
-  theme(
-    # axis.ticks.x = element_line(),
-    # axis.title.x = element_blank(), 
-    axis.text.x = element_text(size = 8, angle = 45, hjust = 1),
-    legend.position = 'right', 
-    panel.grid.minor.y = element_blank(), 
-    panel.grid.minor.x = element_blank(), 
-    panel.grid.major.x = element_blank()
-  )
-
 # precip 
 toplo <- raindat %>% 
   filter(grepl('^Tampa', station)) %>% 
@@ -753,7 +720,7 @@ toplo2 <- toplo %>%
     hiv = quantile(precip_cm, 0.75, na.rm = T),
   ) %>% 
   ungroup()
-p4 <- ggplot() + 
+p3 <- ggplot() + 
   geom_ribbon(data = toplo2, aes(x = xvals, ymin = lov, ymax = hiv, fill = '1995 - 2020\n25th - 75th %tile'), alpha = 0.7) +
   geom_line(data = toplo1, aes(x = xvals, y = precip_cm, color = '2021'), size = 1.2) +
   scale_color_manual(values = '#00806E') +
@@ -769,7 +736,7 @@ p4 <- ggplot() +
   labs(
     x = 'Day of year',
     y = 'Precip. (cm)',
[…]

### 2021-10-22 manu-draft.Rmd
editing discussion, moving history to supp, minor edits to intro

@@
 Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Tampa Bay is the largest open-water estuary in Florida, located in a heavily urbanized watershed of nearly 3.1 million individuals.
-Historical gains in seagrass coverage in Tampa Bay have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to major bay segments [@Janicki96;@Greening16].
-Together, these efforts have resulted in the long-term recovery of Tampa Bay through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
+Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
+Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
 
@@
 Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
-The Tampa Bay Estuary Program and its partners have been instrumental in coordinating efforts among local and regional stakeholders to address legacy pollutants and current threats to the long-term protection of bay resources through implementation of a science-based resource management plan for the Bay [@Greening14;@Greening16].
+The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.
 
@@
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay over four months following the recent release of legacy phosphate mining wastewater in April, 2021.
+This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.
 The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
-We provide a brief overview of the history of the Piney Point facility, including past wastewater releases and impacts observed in Tampa Bay.
-A timeline of events in 2021 is also provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
-The results of this study provide an initial documentation of impacts to the natural resources of Tampa Bay that can be used to inform long-term assessments of acute wastewater discharge events on the environmental quality of the region as well as a potential template for future emergency responses.
+A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
 We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
@@
 # Methods
-
-## History of Piney Point
-
-The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
-Port operations were primarily for export of phosphate production by the plant.
-Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
-Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
-In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
-
-The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
-Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
-Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
-comm.
-D.
-Eckenrod to USEPA, Nov.
-28, 2001).
-Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
-@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
-Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
-
-HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.
-This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
-In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing gypstack at Piney Point.
-Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
-The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
-HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
 
@@
 
-The above results can be considered together to develop a concise narrative of the temporal shift of nutrient pools between different ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point.
+The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point.
 Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters indicated a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).
+This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.
@@
 The co-occurring decline in macroalgae and increase in *K.
-brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients were likely present (as noted by modelling efforts).
-Conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
+brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).
+Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
 These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).
@@
 
-The observed conditions in Tampa Bay from April to July, 2021 following effluent release from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
-Collectively, these observations show that conditions in 2021 were anomalous when compared to the long-term record of water quality and habitat data for Tampa Bay.
+The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
 These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal ranges for many observations supported by the decades of long-term monitoring data for the region (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
+The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
 Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
@@
 
-From a seagrass perspective, both major bloom events contributed to a reduction in water clarity and less light availability during the summer 2021 in comparison to prior years.
-Future effects on seagrass persistence in this region of Tampa Bay are unknown at this time, but monitoring continues.
-Changes in frequency occurrence of seagrasses and associations with water quality and macroalgae were ambiguous and not readily apparent over the initial 4 month sampling period.
-Evaluations of frequency occurrence across months showed no changes, although correlations with water quality and macroalgae can be explained with some certainty.
-Spatially, transects closer to the mouth of Tampa Bay and extending to the northern portion of Sarasota Bay (area 3) had higher frequency occurrence of turtle grass and lower frequency of shoal grass.
-This reflects a natural partitioning of the species along a salinity gradient [@Sherwood17].
-Correlations between macroalgae and seagrasses can also be explained by spatial differences along a depth gradient.
-Frequency occurrences of both shoal grass and turtle grass were negatively associated with manatee grass.
-Shoal grass and turtle grass generally grow at more shallow depths than manatee grass and are often exposed during low tides.
-As an adaptive mechanism to varying water levels at shallower depths, blades of shoal grass and turtle grass can "lay down" during low tide to minimize desiccation stress.
-Manatee grass has thicker, cylindrical blades that are more rigid and, as a result, the species grows deeper.
-In addition to species depth partitioning, these associations may also reflect natural variation in seagrass response to water clarity gradients along the main axis of Tampa Bay.
-Shoal grass can occur in more turbid waters of oligohaline areas.
+Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.
+This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.
+The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* may reflect a natural phenology in Tampa Bay.
+The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis sp.*) in May and June.
+Unusually high abundances of *Dapis sp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
+Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge.
+Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
+However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07] and it is not unreasonable to expect the conditions observed in Tampa Bay to be related to nutrients from Piney Point.
 
-Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.
-This was expected given both the documented changes from past releases from Piney Point and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* likely reflects a natural phenology in Tampa Bay.
-Past spill events were associated with an increase in both red and green macroalgae in a small embayment of Tampa Bay [@Switzer11].
-Thus, it is unclear if the trend from red to green macroalgae reflects a seasonal change or a shift to more eutrophic conditions at nearshore sites stimulated by wastewater release.
-A clear difference in the macroalgae community in 2021 compared to recent prior years was a high abundance of cyanobacteria macroalgae (i.e., *Dapis sp.*).
-Although cyanobacteria are observed in routine monitoring of seagrass transects in Tampa Bay, unusually high abundances of *Dapis sp.* in Anna Maria Sound and near Port Manatee were observed in 2021.
-The long-term effects of these blooms are uncertain, whereas the correlation of cyanobacteria with the different macroalgae groups and seagrass species may be partially explained by depth variation and a seasonal change in species dominance.
-*Dapis sp* were commonly observed at depth and entangled in blades of manatee grass, whereas floating mats closer to shore were also observed and likely driven by currents or winds moving the macroalgae away from the point of origin.
-
+For seagrasses, both major bloom events produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
-Regionally, seagrasses have declined in coverage as estimated from biennial maps provided by the Southwest Florida Water Management District.
 From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
@@
 In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
-The causes of these shifts have yet to be identified, nor is the ecological value known for habitat and water quality.
-General consensus among local managers is that *Caulerpa* may provide some ecological value over bare substrate, but it may be indicative of broader ecosystem changes concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Acute stressors from short-term events, such as wastewater release from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health and restoration activities.
+These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
+Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
 
@@
 Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
-Routine sampling by the FWC Fisheries Independent Monitoring Program occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
+Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
 Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
@@
 Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments were sampled near Port Manatee and surrounding waters in April 2021 during the discharge and in September 2021 to further assess potential impacts of Piney Point on sediment quality.
-These samples were used to calculate scores for the Tampa Bay Benthic Index (TBBI) as a measure of the ecological health of the benthos [@tbep0106].
-Similar to other reporting tools for Tampa Bay, the TBBI categorizes sediment samples into "healthy", "intermediate", and "degraded" as a qualitative summary of enumerated taxonomic data that can support management decisions.
-Results suggested a mix of conditions dominated by "intermediate" and "healthy" samples, possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
-Comparison of the April 2021 samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
+For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+Comparison of the April samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
 Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples.
@@
 
-In the broader context of mining impacts to surface waters, the results herein reinforce the understanding that legacy pollutants from phosphate mining represent a serious threat to environmental resources.
-In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have also been affected by gypstack release.
-Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic wastewater release from a defunct gypstack [@Dillon15;@Beck18b].
+In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.
+In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.
+Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].
 The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
@@
 
-Limitations of our analyses are also important to note to inform future event-based monitoring and additional research to fully understand acute, short-term threats to the sustained recovery of Tampa Bay.
-All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and, as a result, may not represent explicit cause and effect mechanisms.
+Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.
+All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.
 However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-The weight of evidence that 2021 is an anomaly relative to recent prior years provides further justification that our interpretation of results is within the limits of what would be expected by more mechanistic descriptions of the effects of large nutrient pulses to Tampa Bay.
+The weight of evidence that 2021 is an anomaly relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.

### 2021-10-22 supplement.Rmd
editing discussion, moving history to supp, minor edits to intro

@@
 ```
+
+# History of Piney Point
+
+The Piney Point facility in Palmetto, Florida was established in 1966 by the now defunct Borden Chemicals company near Port Manatee on the southeast shore of lower Tampa Bay.
+Port operations were primarily for export of phosphate production by the plant.
+Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
+Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
+In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
+
+The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
+Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
+Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
+Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
+comm.
+D.
+Eckenrod to USEPA, Nov.
+28, 2001).
+Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
+From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
+Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
+Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
+
+HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.
+This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
+In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
+This material was added to an existing gypstack at Piney Point.
+Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
+The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
+HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
 

### 2021-10-24 manu-draft.Rmd
minor edits to gam trend plots and fig caption

@@
 
-```{r wqgam, fig.cap = 'Predicted (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
-Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
-Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.
+```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
+Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021.
 Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}

### 2021-10-24 supplement.Rmd
minor edits to gam trend plots and fig caption

@@
 
-```{r wqgamadd-supp, fig.cap = 'Predicted (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
-Estimated trends are based on Generalized Additve Models fit to historical basline data from 2006 to early 2021.
-Predicted trends from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots.
+```{r wqgamadd-supp, fig.cap = 'Forecasted 2021 (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
+Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021.
 Results are grouped by assessment areas shown in Figure 1a.'}

### 2021-10-24 R/funcs.R

@@ -260,9 +260,9 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
 
   p1 <- ggplot(prds1, aes(x = xvals, group = yr)) + 
     geom_line(aes(y = val, linetype = 'Baseline modelled'), size = 0.5, color = 'lightgrey') +
-    geom_line(data = prds2, aes(y = val, color = area, linetype = '2021 predicted'), size = 2) + 
+    geom_line(data = prds2, aes(y = val, color = area, linetype = '2021 forecasted by area'), size = 2) + 
     geom_ribbon(data = prds2, aes(ymin = lov, ymax = hiv, fill = area), alpha = 0.3, color = NA) +
-    geom_point(data = obs, aes(x = date, y = val, color = area, shape = '2021 samples'), alpha = 0.8, size = 0.5, position = position_jitter()) +
+    geom_point(data = obs, aes(x = date, y = val, shape = '2021 samples'), color = 'black', alpha = 0.8, size = 0.5, position = position_jitter()) +
     scale_x_date(date_breaks = 'month', date_labels = '%b %d', expand = c(0, 0)) +
     facet_grid(area ~ ., scales = 'free_y') + 
     scale_color_manual(values = cols, guide = 'none') +

### 2021-10-24 manu-draft.Rmd
RW, YL, JC comments added

@@
 Monitoring of the natural resources of Tampa Bay in response to the wastewater release at Piney Point began in April, 2021 and continued over the following months.
-These data were collected through a coordinated effort, facilitated in part by the FDEP and TBEP.
+These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by the University of South Florida (USF) Ocean Circulation Lab, facilitated in part by the FDEP and TBEP.
 Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
@@
 Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
+
+In an immediate response to the wastewater discharge, the Ocean Circulation Lab at the University of South Florida, College of Marine Science simulated the Piney Point effluent evolution using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
+Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of effluent release and subsequent Tampa Bay distribution.
+The modeled plume evolution web product (Y.
+Liu, R.H.
+Weisberg, J.
+Chen, Y.
+Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.
 
@@
 Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site.
+As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
 These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
@@
 Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
-Occurrence of this species has historically been spatially distinct, with blooms originating in the Gulf of Mexico and occasionally occurring at bloom concentrations in lower Tampa Bay.
+Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay.
 Data for *K.
-brevis* were also obtained from event-based monitoring samples collected by FWC and available from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
+brevis* were also obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
 Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
@@
 
-Compared to past events, the 2021 event differed from previous releases to the bay in that effluent was discharged directly to lower Tampa Bay instead of through channels, rivers or other back-bay habitats (e.g., Alafia River and Bishop Harbor).
-The comparison to past years provided a stark contrast of the effect of wastewater discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
-The Tampa Bay Coastal Ocean Model [@Chen19] was used to simulate daily tracer particle movement and expected dilution of nutrient concentrations from the point of discharge.
+Compared to past events, the 2021 event differed from previous releases to the bay in that discharge was direct to lower Tampa Bay instead of through channels, rivers or other back-bay habitats (e.g., Alafia River and Bishop Harbor).
+The comparison to past years provided a stark contrast of the effect of discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
+The Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] was used to simulate Eulerian, passive trace concentration [@Chen03;@Chen08] and expected dilution of nutrients from the point of discharge.
 Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
 Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
-Although simulated particle tracking showed that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of labile inorganic nitrogen forms from stack effluent (Table \@ref(tab:stktab)) relative to background concentrations for the bay (Table \@ref(tab:wqsumtab)).
+This is due to the relatively long residence time of the water in Tampa Bay [@Weisberg06;@Zhu15].
+Although model simulations suggested that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of labile inorganic nitrogen forms from stack effluent (Table \@ref(tab:stktab)) relative to background concentrations for the bay (Table \@ref(tab:wqsumtab)).
 

### 2021-10-25 R/figures.R
gam fig shade historical lines

@@ -24,7 +24,7 @@ box::use(
   units = units[set_units], 
   vegan = vegan[decostand, metaMDS], 
   FactoMineR = FactoMineR[PCA], 
-  ggnewscale[new_scale_fill], 
+  ggnewscale[new_scale_fill, new_scale_color], 
   grid[unit]
 )
 

### 2021-10-25 R/funcs.R

@@ -258,8 +258,12 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
     unnest('data') %>% 
     filter(date >= mindt)
 
+  colyrs <- c(rep('lightgrey', 3), brewer.pal(9, 'Greys')[1:7])
+  
   p1 <- ggplot(prds1, aes(x = xvals, group = yr)) + 
-    geom_line(aes(y = val, linetype = 'Baseline modelled'), size = 0.5, color = 'lightgrey') +
+    geom_line(aes(y = val, linetype = '2006 - 2020 Baseline modelled (darker more recent)', color = yr), size = 0.5) +
+    scale_color_gradientn(colors = colyrs, guide = F) + 
+    new_scale_color() + 
     geom_line(data = prds2, aes(y = val, color = area, linetype = '2021 forecasted by area'), size = 2) + 
     geom_ribbon(data = prds2, aes(ymin = lov, ymax = hiv, fill = area), alpha = 0.3, color = NA) +
     geom_point(data = obs, aes(x = date, y = val, shape = '2021 samples'), color = 'black', alpha = 0.8, size = 0.5, position = position_jitter()) +
@@ -269,7 +273,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
     scale_fill_manual(values = cols, guide = 'none') +
     scale_linetype_manual(values = c(1, 1)) + 
     guides(
-      linetype = guide_legend(override.aes = list(size = c(2, 0.5), color = c('black', 'grey')))
+      linetype = guide_legend(override.aes = list(size = c(0.5, 2), color = c('grey', 'black')))
     ) +
     labs(
       color = 'Year', 

### 2021-10-25 manu-draft.Rmd
gam fig shade historical lines
DT comments included

@@
 
-From March 30th to April 9th, 2021, 215 million gallons of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 205 tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
+From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
+An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
 An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
@@
 Blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
-Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1800 tons of dead fish.
+Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
 Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years.
@@
 The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].
-By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
+By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
 Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
-The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located less than two miles from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
+The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
 Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).
@@
 In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
-At that time, approximately 480 million gallons of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.
+At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.
 Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
@@
 This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
-From March 30th to April 9th, approximately 215 million gallons of wastewater were released to lower Tampa Bay.
-Over this ten day period, an estimated 205 tons of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
+From March 30th to April 9th, approximately 814 million liters (215 million gallons) of wastewater were released to lower Tampa Bay.
+Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
 
@@
 
-Monitoring of the natural resources of Tampa Bay in response to the wastewater release at Piney Point began in April, 2021 and continued over the following months.
-These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by the University of South Florida (USF) Ocean Circulation Lab, facilitated in part by the FDEP and TBEP.
-Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
-Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
-
-In an immediate response to the wastewater discharge, the Ocean Circulation Lab at the University of South Florida, College of Marine Science simulated the Piney Point effluent evolution using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
+Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued over the following months.
+These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.
+The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
 Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of effluent release and subsequent Tampa Bay distribution.
@@
 
-Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
-To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
-Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
-Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
-For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
-Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure \@ref(fig:map)a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
-These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
-
-Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
-Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
-Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
-Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay.
-Data for *K.
-brevis* were also obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
-Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
-Petersberg) impacted by the event.
-Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
-
-Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.
-Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
-Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.
-Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
-Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
-Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
-Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure \@ref(fig:map)a) depending on the analysis described below.
+Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
+Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+Details of the sampling methods are provided in supplement.
+Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
 
@@
 Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
-This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].
+This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].
 For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
@@
 
-A general narrative of 2021 events in Tampa Bay following release of wastewater from Piney Point is shown in Figure \@ref(fig:timeline).
+A general narrative of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline).
 After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgam)b).
@@
 brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.
+Conditions were relatively stable in August and September compared to the months prior.
 A quantitative description of these events follows.
@@
 By August, Pinellas County and the city of St.
-Petersburg removed over 1800 tons of dead fish near public and private shoreline areas (K.
+Petersburg removed over 1600 metric tons of dead fish near public and private shoreline areas (K.
 Hammer Levy, Pinellas County, pers.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1800 tons of dead fish from shoreline areas.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
 The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
@@
 Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
-An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 3630 gallons, [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 285 million gallons discharged from Piney Point.
-Moreover, the estimated nutrient load of 205 tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
+An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters discharged from Piney Point.
+Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
 From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years.
@@
 Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
-However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07] and it is not unreasonable to expect the conditions observed in Tampa Bay to be related to nutrients from Piney Point.
+However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
 
@@
 These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
-Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red time bloom in 2018.
+Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red tide bloom in 2018.
 Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern.
@@
 Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+
+Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K.
+brevis* bloom observed in Tampa Bay this year is difficult in the absence of more quantitative results or mechanistic tools to support understanding.
+Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay.
+Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].
+Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, also created conditions that were favorable for growth of *K.
+brevis*.
+However, the results suggest a likely scenario that residual nutrients from the Piney Point discharge, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K.
+brevis* to the concentrations observed in July [also see @Medina20].
+Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
+Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
+Although Piney Point did not cause red tide, the events of 2021 likely created conditions leading to the extreme bloom concentrations in July, which likely would not have been observed if nutrients from Piney Point were not discharged to Tampa Bay.
 
@@
 
-Compared to past events, the 2021 event differed from previous releases to the bay in that discharge was direct to lower Tampa Bay instead of through channels, rivers or other back-bay habitats (e.g., Alafia River and Bishop Harbor).
-The comparison to past years provided a stark contrast of the effect of discharge location that is specific to Tampa Bay, but can inform understanding of pollutant dispersal and its impacts on different habitats that varies spatially and throughout the water column.
-The Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] was used to simulate Eulerian, passive trace concentration [@Chen03;@Chen08] and expected dilution of nutrients from the point of discharge.
-Although these results were initially used to inform monitoring locations, the model suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
-Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
-This is due to the relatively long residence time of the water in Tampa Bay [@Weisberg06;@Zhu15].
-Although model simulations suggested that nutrient dilution occurred rapidly on the order of 10- to 100-fold in a few days, concentrations were considered biologically relevant given the initial concentrations of labile inorganic nitrogen forms from stack effluent (Table \@ref(tab:stktab)) relative to background concentrations for the bay (Table \@ref(tab:wqsumtab)).
-
 In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.
@@
 Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
+Although the characteristics of water in a gypstack can vary based on site specific criteria, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
 Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.
@@
 Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
@@
 Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
-The final column shows the percentage of total observations that were below detection limits (including Secchi on bottom).
+The final column shows the percentage of total observations that were outside of detection, defined as minimum laboratory detection limits for all parameters and values on the bottom for Secchi observations.
 Medians denoted by "-" could not be calculated due to insufficient values above detection.'

### 2021-10-25 supplement.Rmd
gam fig shade historical lines
DT comments included

@@
 Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
-In 1997, 54 million gallons of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
+In 1997, 204 million liters of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
 
@@
 Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 13 inches of rain, causing over 10 million gallons of wastewater to be released into Bishop Harbor, with an estimated 15.4 tons of nitrogen (pers.
+Tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
 comm.
@@
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 248 million gallons of wastewater from Piney Point were barged 120 miles offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
+Around the same time, 939 million liters of wastewater from Piney Point were barged 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
 Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
@@
 This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
-In 2011, HRK agreed to the storage of 1.5 million cubic yards of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
+In 2011, HRK agreed to the storage of 1.1 million cubic meters of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
 This material was added to an existing gypstack at Piney Point.
-Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 169 million gallons of dredged saltwater slurry and 3.5 tons of nitrogen to receiving waters leading to Bishop Harbor.
+Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 640 million liters of dredged saltwater slurry and 3.2 metric tons of nitrogen to receiving waters leading to Bishop Harbor.
 The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
 HRK maintains ownership and responsibility of the site to present day with oversight by FDEP.
+
+# Sampling methods
+
+Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
+To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
+Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
+Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
+For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
+This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
+Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
+Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
+As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
+These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
+
+Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
+Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
+Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
+Data for *K.
+brevis* were obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
+Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
+Petersberg) impacted by the event.
+Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
+
+Seagrass and macroalgae transect samples were collected approximately biweekly at locations around Piney Point from April to early October 2021.
+Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
+Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.
+Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
+Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
+Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
+Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.
 
@@
 Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure 1a.'}

### 2021-10-26 manu-draft.Rmd
update k brevis figs and text with current data from CL

@@
 brevis* from April to July to bloom concentrations exceeding 10,000 cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1995 to present in Figure \@ref(fig:redtide)a shows the range of cell concentrations sampled in middle and lower Tampa Bay, with only a handful of years having median cell concentrations greater than 10,000 cells/L, notably 2005 and 2018.
+The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
 Median cell concentrations for most years were well below 1,000 cells/L.
-The highest concentration of 17.6 million cells/L was observed in 2021, whereas only 2006 had a maximum observed concentration above 10 million cells/L [@Flaherty11].
+The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
 Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
-Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical baseline values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
+Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
 This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K.

### 2021-10-26 R/figures.R

@@ -25,7 +25,8 @@ box::use(
   vegan = vegan[decostand, metaMDS], 
   FactoMineR = FactoMineR[PCA], 
   ggnewscale[new_scale_fill, new_scale_color], 
-  grid[unit]
+  grid[unit], 
+  readxl[read_excel]
 )
 
 data(rsallpts)
@@ -38,7 +39,6 @@ data(rsstatloc)
 data(rstrndat)
 data(parms)
 data(rstrnpts)
-data(kbrdat)
 data(winddat)
 data(raindat)
 data(hydrodat)
@@ -47,6 +47,17 @@ data(rsphydat)
 data(bstransect)
 
 source(here('R/funcs.R'))
+
+# k brevis data from CL
+habdat <- read_excel(
+    here('data-raw/Tampa Bay Karenia brevis to 2021.10.15 within coordinate box_modified.xlsx'),
+    sheet = 'subsetData_originalSegments'
+  ) %>% 
+  filter(Segment %in% c('MTB','LTB')) %>%
+  select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
+  mutate(
+    date = ymd(date),
+    val = val / 1e5)
 
 # map ---------------------------------------------------------------------
 
@@ -583,26 +594,6 @@ fishdat <- read.csv(here('data-raw/FishKillResultReport.csv')) %>%
   ) %>% 
   filter(date < as.Date('2021-10-01'))
 
-# k brevis data from CL
-habdat <- read.csv(here('data-raw/KB_LowMid_1995-2021.csv')) %>% 
-  filter(Segment %in% c('MTB','LTB')) %>%
-  select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
-  mutate(date = mdy(date))
-
-# add k brevis data from HABSOS that are collected after last date in habdat
-kbrdat <- kbrdat %>% 
-  .[tbseg[tbseg$bay_segment %in% c('MTB', 'LTB'), ], ] %>% 
-  filter(var == 'kb') %>% 
-  filter(date > max(habdat$date)) %>% 
-  mutate(
-    lat = st_coordinates(.)[,2],
-    lng = st_coordinates(.)[,1]
-  ) %>% 
-  st_set_geometry(NULL) %>% 
-  select(date, val, lat, lng)
-
-habdat <- bind_rows(habdat, kbrdat)
-
[…]

### 2021-10-26 R/funcs.R

@@ -264,7 +264,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
     geom_line(aes(y = val, linetype = '2006 - 2020 Baseline modelled (darker more recent)', color = yr), size = 0.5) +
     scale_color_gradientn(colors = colyrs, guide = F) + 
     new_scale_color() + 
-    geom_line(data = prds2, aes(y = val, color = area, linetype = '2021 forecasted by area'), size = 2) + 
+    geom_line(data = prds2, aes(y = val, color = area, linetype = '2021 forecasted +/- 95% CI (colored by area)'), size = 2) + 
     geom_ribbon(data = prds2, aes(ymin = lov, ymax = hiv, fill = area), alpha = 0.3, color = NA) +
     geom_point(data = obs, aes(x = date, y = val, shape = '2021 samples'), color = 'black', alpha = 0.8, size = 0.5, position = position_jitter()) +
     scale_x_date(date_breaks = 'month', date_labels = '%b %d', expand = c(0, 0)) +
@@ -273,7 +273,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
     scale_fill_manual(values = cols, guide = 'none') +
     scale_linetype_manual(values = c(1, 1)) + 
     guides(
-      linetype = guide_legend(override.aes = list(size = c(0.5, 2), color = c('grey', 'black')))
+      linetype = guide_legend(override.aes = list(size = c(0.5, 2), color = c('grey', cols[1])))
     ) +
     labs(
       color = 'Year', 

### 2021-10-26 manu-draft.Rmd
revisions from CA comments

@@
 
+Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.
 From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
@@
 An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Cyanobacteria blooms (*Dapis* spp.) were observed beginning in May, with biomass peaking in June at 48% frequency occurrence for some locations.
-Blooms of *Karenia brevis* were first observed in May and continued through July within Tampa Bay proper.
-Reported fish kills tracked bloom concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
-Combined, these observations indicate abnormal conditions in Tampa Bay following release of wastewater from Piney Point, which is supported by comparison to the decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years.
+Cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.
+Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
+By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release of wastewater from Piney Point are abnormal relative to historic conditions in Tampa Bay.
 
@@
 
-Ecosystem management paradigms for estuaries of the Gulf Coast of Florida, USA are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.
+Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
+Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
+The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
+Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
+The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
+Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain wastewater on site.
+There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
+Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
+
+Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
+Ecosystem management paradigms for estuaries are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.
 The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
@@
 Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
-
-Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
-Because point-source inputs of nutrient loads from wastewater treatment plants and industrial sources into Tampa Bay have been greatly reduced, non-point source loads from wastewater sources and stormwater runoff are now estimated to dominate external nutrient loads to the bay, particularly during the rainy season from June to September [@Janicki08;@Janicki17].
-Atmospheric deposition of nutrients from fossil fuel-based power production and automobile traffic further contribute about one-quarter of the total nitrogen inputs directly to the bay's surface [@Poor13].
-Climate change stressors, such as sea level rise, changing rainfall patterns, and temperature alterations, may further perturb ecosystem dynamics and assimilative capacity by reducing system resilience to nutrient inputs [@Sherwood14;@Burke17].
 The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.
-
-Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
-Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
-Impurities, contaminants, and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
-The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
-There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
-Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
@@
 Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
+The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement.
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
 Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).
-Decreasing holding capacity of the ponds from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.
-Discharges of wastewater from the stacks occurred in the early 2000s and 2011 to nearby Bishop Harbor connected to Tampa Bay.
-Those discharges resulted in spatially-restricted, ecosystem responses [@Switzer11;@Garrett11].
+Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
+Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.
+Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].
 
-Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) when water quality samples with a similar conductivity as the wastewater were detected at onsite seepage interceptor drains.
-In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.
-At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dedging operations was being held in the failing gypstack.
-Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay (Table \@ref(tab:stktab)), particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
-Due to public safety and property concerns over catastrophic failure of the holding walls, an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) was issued by FDEP on March 29th for HRK to begin release of wastewater from the stack into Tampa Bay to reduce physical strain on the stacks.
-Unlike past discharges from the site, HRK was authorized to release untreated wastewater through siphon lines established during dredging operations in 2011 and that discharged through Berth 12 at Port Manatee.
-This was done under the assumption that backwater habitats (e.g., Bishop Harbor) may be spared the impacts of additional effluent discharges from the site, as was observed in prior events.
-From March 30th to April 9th, approximately 814 million liters (215 million gallons) of wastewater were released to lower Tampa Bay.
+Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
+In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
+At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.
+Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay, particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.
 Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
@@
 A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
-We focus primarily on the perspective of the Tampa Bay Estuary Program in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
+We focus primarily on the perspective of the Tampa Bay Estuary Program (TBEP) in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
+The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.
 
@@
 
-Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+These areas (Areas 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals).
 Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
@@
 
-## Timeline of events from April 2021
+## Timeline of 2021 biological response events
 
-A general narrative of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge stopped on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April (Area 1, Figure \@ref(fig:wqgam)b).
-Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum chlorophyll concentration of 265 ug/L, although median concentrations for each week in April were less than 10 ug/L.
+A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline).
+After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.
+Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April chlorophyll concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L.
 The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.
 On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10k cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
+brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
 Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
@@
 
-From April through September 2021, `r nwq` samples were collected for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
+Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).
+Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay.
+Similar differences for total phosphorus, total nitrogen, and chlorophyll-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.
+These contrasts provided a context for the monitoring data collected in 2021.
+
+From April through September 2021, `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
 Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
 The percentage of observations outside of the normal range varied by location and parameter.
-For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
+For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
 Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.
@@
 Red macroalgae were present in high abundances from April to May.
-*Dapis sp.* was first observed on May 24th and was present at all of the sample locations on June 4th and 15th.
+*Dapis sp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
 *Dapis sp.* persisted through June and July, but not was observed after July 20th.
@@
 
-## Nutrient cycling
+## Potential nutrient cycling
 

### 2021-10-26 supplement.Rmd
revisions from CA comments

@@
 Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
-Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds likely contributed to discharge events to surficial and ground waters.
+Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.
 Tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
@@
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
-These areas of interest were delineated based on proximity to Piney Point, the main shipping channel in the bay, inflow boundaries (e.g., Little Manatee River), location of the Skyway Bridge at the mouth of Tampa Bay, and major segments of Tampa Bay used by TBEP for assessing programmatic goals.
 
@@
 Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+This rapid survey design was created by the "Eyes on Seagrass" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
 Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.

### 2021-10-26 R/figures.R

@@ -48,16 +48,18 @@ data(bstransect)
 
 source(here('R/funcs.R'))
 
-# k brevis data from CL
+# k brevis data from CL (not created in piney-point repo)
 habdat <- read_excel(
-    here('data-raw/Tampa Bay Karenia brevis to 2021.10.15 within coordinate box_modified.xlsx'),
-    sheet = 'subsetData_originalSegments'
+    here('data-raw/TB_Karenia_brevis_to_2021-10-15_coord_box_mod_TBsegments_cl.xlsx'),
+    sheet = 'subsetData_originalSegments', 
+    na = 'NA'
   ) %>% 
-  filter(Segment %in% c('MTB','LTB')) %>%
+  filter(BaySegment %in% c('MTB','LTB')) %>%
   select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
   mutate(
     date = ymd(date),
-    val = val / 1e5)
+    val = val / 1e5
+  )
 
 # map ---------------------------------------------------------------------
 

### 2021-10-29 manu-draft.Rmd
CL comments addressed

@@
 data(trnsum)
+data(tnanndat)
 
@@
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.
+For censored data (i.e., above or below detection), secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
 
@@
 brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
-Also during May, *Dapis spp.* (cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
+Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
 *Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
@@
 Concentrations of *K.
-brevis* in middle and lower Tampa Bay peaked in mid-July, with bloom conditions not observed in the bay after July.
+brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.
 Conditions were relatively stable in August and September compared to the months prior.
@@
   filter(`Water quality variable` == 'Secchi (m)') %>% 
-  pull(below) %>%  
+  pull(`% below`) %>%  
   sum
@@
   filter(`Water quality variable` != 'Secchi (m)') %>% 
-  pull(above) %>%  
+  pull(`% above`) %>%  
   sum
-outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 1)
+outrng <- round(100 * (sum(outrng1, outrng2) / nwq), 0)
 
@@
   filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
-  select(Area, above) %>% 
+  select(Area, `% above`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'TN (mg/L)') %>% 
-  select(Area, above) %>% 
+  select(Area, `% above`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'Secchi (m)') %>% 
-  select(Area, below) %>% 
+  select(Area, `% below`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'DO (% sat.)') %>% 
-  select(Area, above) %>% 
+  select(Area, `% above`) %>% 
   deframe %>% 
   round(0)
+ammsum <- rswqdat
 ```
@@
 Physical parameters (salinity, temperature) were generally within range over the study period.
-Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1, similar to the discharge measurements in Table \@ref(tab:stktab).
+Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1.
+Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab).
+Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
 Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
@@
   mutate(
-    fo = round(100 * fo, 1)
+    fo = round(100 * fo, 0)
   ) %>% 
@@
   mutate(
-    foest = round(100 * foest, 1), 
+    foest = round(100 * foest, 0), 
     mo = month(date)
@@
   group_by(taxa) %>% 
-  summarise(foest = round(mean(foest), 1)) %>% 
+  summarise(foest = round(mean(foest), 0)) %>% 
   deframe
@@
   group_by(city) %>% 
-  summarise(cnt = sum(cnt), .groups = 'drop') %>% 
-  mutate(per = round(100 * cnt / sum(cnt), 1))
+  summarise(cnt = sum(cnt), .groups = 'drop')
 wkstpcnt <- fishwk %>% 
@@
 The increase in *K.
-brevis* from April to July to bloom concentrations exceeding 10,000 cells/L was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+brevis* from April to July to bloom concentrations exceeding 10,000 cells/L (based on FWC breakpoint) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
 The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
@@
 Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*).
+This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms ((e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the total nitrogen in the initial discharge.
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.
@@
 
+```{r}
+# tnanndat from https://github.com/tbep-tech/load-estimates
+tnld <- tnanndat %>% 
+  filter(bay_segment == 'Lower Tampa Bay') %>% 
+  filter(year > 2005 & year < 2021) %>% 
+  group_by(year) %>% 
+  summarise(
+    tn_load = sum(tn_load)
+  ) %>% 
+  pull(tn_load) %>% 
+  mean %>% 
+  `/`(1.102) %>% 
+  round(0)
+```
+
 The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
@@
 Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
+External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
 From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years.
@@
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
-From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south.
-These broader trends suggest regional drivers are affecting seagrass communities (e.g., variation in precipitation), yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
-Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent red tide bloom in 2018.
+From 2018 to 2020, seagrass coverage declined by 16\% in Tampa Bay, with similar losses observed in Sarasota Bay (18\%), Lemon Bay (12\%), and Charlotte Harbor (23\%) to the south (Southwest Florida Water Management District, unpublished results).
+These broader trends suggest regional drivers are affecting seagrass communities [e.g., variation in precipitation, @Tomasko20], yet local issues specific to individual bays also pose challenges to managing water quality and subtidal habitats.
+Recent seagrass losses in Sarasota Bay may be linked to decreased light availability from a persistent *K.
+brevis* bloom in 2018.
 Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern.
@@
 Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
-Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
+Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
 Given the observed *K.
@@
 Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].
-The first spill event in 2005 occurred following failure of the retaining walls after a heavy rain event and the second event in 2012 occurred after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
+Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
 Massive fish kills were observed and likely related to low pH of the effluent.
@@
 However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-The weight of evidence that 2021 is an anomaly relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.
+The weight of evidence that 2021 is an anomaly for both nitrogen loading and water quality response relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling and fate of nitrogen from Piney Point.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
+Long-term fate of nitrogen from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
 There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.

### 2021-10-29 supplement.Rmd
CL comments addressed

@@
 To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
-Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) that ensures similar standards and protocols are followed in the collection of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
+Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
 Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
-For this paper, we focus on parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
+Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).
+In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts depending on monitoring agency.
+Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.
+
+For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
+This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
 Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
@@
 
-Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples enumerating major taxa by cell concentrations and qualitative presence/absence samples.
+Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples from light microscopy that enumerated major taxa by cell concentrations and qualitative presence/absence samples.
 Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
-Evaluation of HABs data included specific focus on the red tide organism *Karenia brevis* that can occur in the bay depending on salinity and temperature conditions during the growing season.
+Evaluation herein of HABs data included specific focus on the red tide organism *Karenia brevis*.
 Data for *K.
@@
 Petersberg) impacted by the event.
-Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
+Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx) that provides reports received by FWC via the state's Marine Fish Kill Hotline.
 
@@
 Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
-This rapid survey design was created by the "Eyes on Seagrass" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
+This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
 Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.

### 2021-10-29 R/tables.R

@@ -174,20 +174,20 @@ totab2 <- totab %>%
   left_join(totab1, by = c('area', 'lbs')) %>% 
   mutate(area = ifelse(duplicated(area), '', area)) %>% 
   mutate(
-    `in range` = round(100 * `in range` / cnt, 1),
-    `above` = round(100 * `above` / cnt, 1),
-    `below` = round(100 * `below` / cnt, 1), 
-    `Below detection` = round(100 * nondetect / cnt, 1)
+    `% in range` = round(100 * `in range` / cnt, 0),
+    `% above` = round(100 * `above` / cnt, 0),
+    `% below` = round(100 * `below` / cnt, 0), 
+    `% below detection` = round(100 * nondetect / cnt, 0)
   ) %>% 
   select(
     Area = area, 
     `Water quality variable` = lbs, 
     `Med. (Min., Max.)` = sumv, 
     `N obs.` = cnt, 
-    `in range`, 
-    above, 
-    below, 
-    `Below detection`
+    `% in range`, 
+    `% above`, 
+    `% below`, 
+    `% below detection`
   )
 
 wqsumtab <- totab2

### 2021-10-30 manu-draft.Rmd
added MM comments

@@
 Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
-Phlips (phlips@ufl.edu), Mark Rains (Mark.Rains@floridadep.gov), Gary E.
+Phlips (phlips@ufl.edu), Gary E.
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T.
@@
 Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
-These areas (Areas 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals).
+These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals).
 Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
@@
 
-Statistical assessments were evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
-Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.
-Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a.
-Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.
+Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
+Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the five month period.
+Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a.
+Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-For censored data (i.e., above or below detection), secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
+For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
 
@@
 The R statistical programming language (v4.0.2) was used for all analyses [@RCT20].
-Partner data uploaded or entered manually as Google spreadsheets were imported into R using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages.
-The suite of R packages available in the tidyverse [@Wickham19] were used to wrangle the data into an appropriate format for analysis.
-The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data for Tampa Bay, specifically the EPC water quality data and seagrass transect database.
+We imported data using the googlesheets4 [@Bryan20] and googledrive [@DAgostino20] R packages and used tidyverse [@Wickham19] packages to format data for analysis.
+The tbeptools R package [@Beck21] was used to import and summarize long-term monitoring data (EPC water quality data and seagrass transect data).
 The NADA R package [@Lee20] was used for analysis of censored data.
 All spatial analyses were done using the simple features (sf) R package [@Pebesma18].
-The mgcv R package [@Wood17] was used to create the the GAMs for water quality parameters.
+The mgcv R package [@Wood17] was used to create the GAMs for water quality parameters.
 
@@
 After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.
-Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April chlorophyll concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L.
+Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April Chl-a concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L.
 The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.
@@
 Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay.
-Similar differences for total phosphorus, total nitrogen, and chlorophyll-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.
+Similar differences for total phosphorus, TN, and chl-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.
 These contrasts provided a context for the monitoring data collected in 2021.
 
-From April through September 2021, `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, total nitrogen, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
-Of these samples, `r outrng`\% were outside of the normal range defined by the long-term monthly monitoring data for the baseline period from 2006 to 2020 (below for Secchi depth, above for all others).
-The percentage of observations outside of the normal range varied by location and parameter.
-For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for areas 2 and 3, respectively.
-Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in area 2 and `r tnabv[['3']]`\% in area 3.
-Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in areas 2 and 3.
-Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in area 1, `r dosabv[['2']]`\% in area 2).
-Physical parameters (salinity, temperature) were generally within range over the study period.
-Inorganic nitrogen (ammonia, nitrate/nitrite) was generally within range, although initial time series showed much higher concentrations for ammonia in April near area 1.
+Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month.
+A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
+The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.
+For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 and 3, respectively.
+Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.
+Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.
+Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).
+Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
 Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab).
@@
 
-Total nitrogen, chl-a, and secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).
-For area 1, total nitrogen and chlorophyll concentrations were frequently above normal ranges during April.
+Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).
+For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.
 Concentrations decreased slightly until June and July when values increased again above the seasonal expectation.
-Many secchi observations in area 1 were lower than normal in April and July.
-Observations in areas 2 and 3 were more often within the normal seasonal range, with some exceptions for total nitrogen and chl-a in area 3 in  April, May, and July.
-Statistical comparisons between months for total nitrogen, chl-a, and secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
-Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant for total nitrogen, chl-a, and secchi depth for area 1 and for total nitrogen and chl-a for areas 2 and 3 (Table \@ref(tab:wqcmptab)).
-Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on area and parameter.
+Many Secchi observations in Area 1 were lower than normal in April and July.
+Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.
+Statistical comparisons between months for TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)).
+Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on Area and parameter.
 Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
@@
 Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.
-No transects were sampled in area 2 to the north of Piney Point and no transects were sampled past September in Area 1.
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1.
 Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).
-Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in area 1, `r fomoarea3apr['Red']`\% in area 3).
-Cyanobacteria frequency occurrence peaked in June, with greater coverage in area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
-Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in area 1, `r fomoarea3oct['Green']`\% in October in area 3).
+Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3).
+Cyanobacteria frequency occurrence peaked in June, with greater coverage in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
+Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).
 Brown macroalgae was only observed at one transect.
@@
 Turtle grass (*T.
-testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall area 3), compared to shoal grass (*H.
-wrightii*, `r fomoarea1ave['Halodule wrightii']`\% area 1, `r fomoarea3ave['Halodule wrightii']`\% area 3) and manatee grass (*S.
-filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% area 1, `r fomoarea3ave['Syringodium filiforme']`\% area 3).
-Slight changes in frequency occurrence in area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
+testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall Area 3), compared to shoal grass (*H.
+wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S.
+filiforme*, `r fomoarea1ave['Syringodium filiforme']`\% Area 1, `r fomoarea3ave['Syringodium filiforme']`\% Area 3).
+Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
 Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).
@@
 The increase in *K.
-brevis* from April to July to bloom concentrations exceeding 10,000 cells/L (based on FWC breakpoint) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
 The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
@@
 The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
-Seasonally in 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+In 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
@@
 The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point.
-Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chlorophyll-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms ((e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the total nitrogen in the initial discharge.
-Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of total nitrogen and chlorophyll.
+Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
+This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms ((e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
+Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
 However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
-During peak macroalgae growth, total nitrogen and chlorophyll-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
+During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
 brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).
@@
 These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for total nitrogen, chlorophyll-a, and secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
+The observed water quality conditions throughout this period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
 Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
@@
 
-Several of the results support known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
-Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chlorophyll-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
+Several of the results are consistent with known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
+Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
 Temporally, these changes were observed at different times and for different species of phytoplankton.
-The initial increase in chlorophyll-a was first associated with a diatom bloom in April and then by *K.
+The initial increase in chl-a was first associated with a diatom bloom in April and then by *K.
 brevis* in June and July.
@@
 Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
-Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where total nitrogen that included organic sources were at high concentrations through April and again peaking in July.
+Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.
 Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
@@
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the wastewater, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the discharge, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
 Long-term fate of nitrogen from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
-There may likely never be a definitive link, but this is not necessary to support future actions that ensure that the continued threats of these facilities to the health of coastal resources are mitigated through collaborative efforts by local, regional, and state partners.
+Local, regional, and state partners should continue to pursue management and policy actions actions that can mitigate the continued threats of these facilities to the health of coastal resources.
 These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
@@
 
-```{r wqmap, fig.cap = 'Sampled water quality data for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
-Values outside of the normal range (above for total nitrogen and chlorophyll, below for secchi) are outlined in black and those in normal range are outlined in light grey.
+```{r wqmap, fig.cap = 'Sampled water quality data for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
 Color ramps and point sizes show relative values (reversed for Secchi).
 Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location (Figure \\@ref(fig:map)a).
-Values below detection limits (or secchi on bottom) are not shown.'}
+Values below detection limits (or Secchi on bottom) are not shown.'}
 knitr::include_graphics(here('figs/wqmap.jpeg'))
@@
 
-```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters) by area based on historal seasonal models.
+```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters) by area based on historal seasonal models.
 Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
@@
 
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) area 1 and (b) area 3 (Figure \\@ref(fig:map)a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) Area 1 and (b) Area 3 (Figure \\@ref(fig:map)a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
 Estimates are grouped by sample months in 2021.
@@
 ```{r wqcmptab}
-cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
+cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and Secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
 Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.

### 2021-10-30 R/figures.R

@@ -57,8 +57,7 @@ habdat <- read_excel(
   filter(BaySegment %in% c('MTB','LTB')) %>%
   select(date = Sample_Dat, val = Karenia_br, lat = Latitude, lng = Longitude) %>% 
   mutate(
-    date = ymd(date),
-    val = val / 1e5
+    date = ymd(date)
   )
 
 # map ---------------------------------------------------------------------
@@ -633,7 +632,7 @@ p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
   ) +
   labs(
     x = 'Year',
-    y = 'Cells (100k / L)',
+    y = 'Cells / L',
     title = expression(paste('(a) ', italic('K. brevis'), ' Apr - Sep concentrations by year, middle/lower Tampa Bay'))
   )
 
@@ -676,7 +675,7 @@ p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
   ) +
   labs(
     x= 'Week of',
-    y = 'Cells (100k / L)',
+    y = 'Cells / L',
     title = expression(paste('(b) ', italic('K. brevis'), ' concentrations in 2021 by week, middle/lower Tampa Bay'))
   )
 
@@ -987,8 +986,7 @@ physum <- rsphydat %>%
   ) %>% 
   ungroup() %>% 
   mutate(
-    week = floor_date(date, unit = 'week'), 
-    val = val / 1e5
+    week = floor_date(date, unit = 'week')
   ) %>% 
   filter(species == 'Diatoms') %>% 
   select(week, species, val)
@@ -1022,7 +1020,7 @@ p3 <- ggplot(toplo1, aes(x = week, y = medv)) +
   geom_point() + 
   # geom_line() + 
   geom_errorbar(aes(ymin = lov, ymax = hiv), width = 0) +
-  scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
+  scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = F))) +
   scale_x_date(breaks = brks, date_labels = '%b %d', limits = range(brks), expand = c(0.01, 0.01)) +
   theme_minimal() +
   theme(
@@ -1032,7 +1030,7 @@ p3 <- ggplot(toplo1, aes(x = week, y = medv)) +
   ) +
   labs(
     x = NULL,
-    y = 'Cells (100k / L)',
+    y = 'Cells / L',
[…]

### 2021-10-30 R/funcs.R

@@ -97,9 +97,10 @@ wqplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet =
   
   p1 <- ggplot() + 
     geom_rect(data = bswqtmp, aes(xmin = datestr, xmax = dateend, ymin = minv, ymax = maxv, group = mo, fill = 'Monthly baseline (mean +/- 1 sd)'), alpha = 0.2) +
-    geom_boxplot(data = rswqtmp, aes(x = date, y = val, group = date), fill = bxcls$fillcl, outlier.colour = NA, lwd = 0.5, alpha = 0.8, show.legend = F) + 
-    geom_jitter(data = rswqtmp, aes(x = date, y = val, group = date), alpha = 0.4, size = 0.5) + 
+    # geom_boxplot(data = rswqtmp, aes(x = date, y = val, group = date), fill = bxcls$fillcl, outlier.colour = NA, lwd = 0.5, alpha = 0.8, show.legend = F) + 
+    geom_jitter(data = rswqtmp, aes(x = date, y = val, group = date, color = area), alpha = 0.7, size = 0.5, width = 2) + 
     scale_fill_manual(NULL, values = 'blue') +
+    scale_color_manual(values = cols, guide = F) + 
     scale_linetype_manual(values = 'dashed') + 
     facet_grid(area ~ ., scales = 'free_y') + 
     scale_x_date(breaks = unique(rswqtmp$date), date_labels = '%b %d', expand = c(0.05, 0.05)) +

### 2021-11-02 manu-draft.Rmd
AA edits added

@@
 An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
-An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site, with chlorophyll concentrations exceeding 50 ug/L.
-Cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.
+An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site.
+Cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.
 Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
@@
 
-*Key words*: algae, nitrogen, phosphate mining, seagrass, Tampa Bay, wastewater, water quality
+*Key words*: eutrophication, macroalgae, nitrogen, phosphate mining, seagrass, Tampa Bay
 
@@
 Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of waste precipitate, or phosphogypsum (CaSO$_4\cdot$ H$_2$O).
-Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds [@Burnett01].
+The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).
+Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds capable of storing millions of liters of wastewater [@Burnett01].
 The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
@@
 Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
-Ecosystem management paradigms for estuaries are based primarily on the control of nutrient pollutants from atmospheric, stormwater and wastewater sources.
-The effects of nitrogen from dominant external source inputs are well understood as a limiting nutrient for the growth of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are a primary endpoint for assessing the impacts of nutrient pollution on water quality based on established relationships between nitrogen, phytoplankton growth, water clarity, and light requirements for seagrass species observed in nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
+Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.
+Nitrogen inputs from external sources are well understood as drivers of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
+Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
 Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
@@
 
-The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].
+The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].
 By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
+Currently, seventeen phosphogypsum stacks ([Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
 The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement.
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Departmental of Environmental Protection (FDEP).
+Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Department of Environmental Protection (FDEP).
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
@@
 
-Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued over the following months.
+Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September.
 These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.
@@
 
-Monitoring agencies and local partners that collected data included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
+Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
 Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
@@
 Details of the sampling methods are provided in supplement.
-Additional sampling for contaminants (e.g., heavy metals), invertebrates, and nekton was also conducted, but they are not reported here in anticipation of future monitoring events.
+Additional samples for contaminants (e.g., heavy metals), invertebrates, and nekton were also collected, but they are not reported here in anticipation of future monitoring events.
 
@@
 This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
-In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
+In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference (see https://shiny.tbep.org/piney-point/ for a map of the matches).
 
@@
 Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.
-For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners.
-For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
+For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates.
+For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
 
@@
 On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although observations were limited to lower Tampa Bay.
+brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although bloom conditions were limited to lower Tampa Bay.
 Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
-*Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats below the water column at these locations.
+*Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats at these locations.
 By June 27th, fish kill reports attributed to red tide increased with *K.
@@
 The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.
-For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to teh discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 and 3, respectively.
+For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively.
 Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.
@@
 Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).
-Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were generally within range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
-Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab).
+Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
+Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
 Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
@@
 Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)).
-Results of multiple comparison tests that evaluated differences between pairs of months generally showed that April/May were different from June/July depending on Area and parameter.
+Further analsyis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter.
 Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
@@
 Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
-Brown macroalgae in the genus *Feldmannia* were only observed at `r foall['Brown']`\% frequency occurrence.
+Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence.
 For seagrasses, turtle grass (*T.
@@
 
-Temporal progression of macroalgae and seagrass varied across the months, although a typical pattern observed at many of the transects is shown in Figure \@ref(fig:trnex).
+A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex).
 Transect S3T6 is located less than one kilometer to the north of Port Manatee.
-Macroalgal abundances changed over the course of sampling similar to many of the other transects sampled during the study.
+Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.
 Red macroalgae were present in high abundances from April to May.
-*Dapis sp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
-*Dapis sp.* persisted through June and July, but not was observed after July 20th.
+*Dapis spp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
+*Dapis spp.* persisted through June and July, but not was observed after July 20th.
 Green macroalgae taxa were first observed in July, although at generally low abundances.
 Red macroalgae were the dominant taxa by the end of September.
-Overall abundance of seagrass did not change dramatically from April 22nd through September.
+Overall abundance of seagrass did not change from April 22nd through September.
 The site is dominated by manatee grass that was observed at nearly all of the sample points along the transect at varying coverages.
@@
 brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in middle and lower Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
+The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
 Median cell concentrations for most years were well below 1,000 cells/L.
 The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
-In 2021, bloom concentrations were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+Bloom concentrations in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
@@
 
-Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations (Figure \@ref(fig:redtide)e).
+Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations during June and July 2021 (Figure \@ref(fig:redtide)e).
 In total, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa.
@@
 brevis* cell concentrations (Figure \@ref(fig:redtide)b).
-Notably, the first to last week of fish kill reports covered only one and half months, whereas red tide in the bay was observed for nearly four months.
+Notably, all of the fish kill reports occured within a 1.5 month period, whereas red tide in the bay was observed for nearly four months.
 Increased fish kill reports in early July may coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f), which moved dead fish closer to populated nearshore areas, as noted earlier.
@@
 
-The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide a weight of evidence that demonstrates an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
@@
 The observed water quality conditions throughout this period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
-Although these out-of-range results had specific temporal and spatial patterns, they were widespread in lower and middle Tampa Bay.
+These conditions were widespread in lower and middle Tampa Bay.
 
@@
 
-Several of the results are consistent with known responses of water quality to nutrient loads for shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
+Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
 Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
@@
 This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and seasonally occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in *Ulva spp.* may reflect a natural phenology in Tampa Bay.
-The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis sp.*) in May and June.
-Unusually high abundances of *Dapis sp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.
+The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay.
+The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.
+Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
 Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge.
@@
 brevis* bloom in 2018.
-Although the 2021 bloom in Tampa Bay was short-lived, potential effects on seagrasses remain a concern.
+Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern.
 Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
@@
 Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, restrictions for catch and release only for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
+brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
 For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
@@
 brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
+Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.
 
@@
 Although the characteristics of water in a gypstack can vary based on site specific criteria, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
-Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that can significantly affect coastal resources.
+Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.
 These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.
@@
 Long-term fate of nitrogen from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
-Local, regional, and state partners should continue to pursue management and policy actions actions that can mitigate the continued threats of these facilities to the health of coastal resources.
+Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
 These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
@@
 Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+Funding was provided to the University of Florida from the Ocean Conservancy and the National Science Foundation.
 The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
@@
 
-```{r timeline, fig.cap = 'Graphical timeline of events from the discharge of wastewater effluent at Piney Point starting March 30th, 2021 through the end of July with the gradual decline of red tide in Tampa Bay.'}
-knitr::include_graphics(here('figs/timeline.jpeg'))
+```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the discharge of wastewater from Piney Point.
+Inset image shows blooms of the macroalgae *Dapis spp.*.'}
+knitr::include_graphics(here('figs/timeline.jpg'))
 ```
@@
 Sample dates in 2021 are shown in rows with transect meter results shown in columns.
-Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis sp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
+Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis spp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
 Abundances are Braun-Blanquet coverage estimates.'}
@@
 
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) Area 1 and (b) Area 3 (Figure \\@ref(fig:map)a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) Area 1 and (b) Area 3 (see map Figure \\@ref(fig:map)a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
 Estimates are grouped by sample months in 2021.

### 2021-11-02 supplement.Rmd
added MM comments
AA edits added

@@
 Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.
-Tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
+For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
 comm.
@@
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-From November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+In other event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 939 million liters of wastewater from Piney Point were barged 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
+Around the same time, 939 million liters of wastewater from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
 Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
@@
 Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
-To the extent possible, data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
+Data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
 Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
@@
 Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
-Overall, sample effort was variable given agency resources at the time of the discharge event and over the next few months.
+Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
@@
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along fifty meter transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point.
+Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
 This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
@@
 
-```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (Figure 1a) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
+```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (see map in Figure 1a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
 Estimates are grouped by sample months in 2021.

### 2021-11-02 R/figures.R

@@ -538,11 +538,17 @@ p <- ggplot(toplo, aes(x = date, y = foest)) +
   facet_grid(typ ~ area) +
   theme_minimal(base_size = 14) + 
   scale_y_continuous(limits = c(0, 1))+
-  scale_fill_manual(values = cols) +
+  scale_fill_manual(
+    values = cols,
+    labels = c('Red', 'Green', 'Brown', 'Cyanobacteria', expression(italic('Thalassia testudinum')), 
+               expression(italic('Halodule wrightii')), expression(italic('Syringodium filiforme')),
+               'Total')
+  ) +
   labs(
     y = 'Freq. occurrence'
   ) +
   theme(
+    legend.text.align = 0,
     legend.position = 'top', 
     legend.title = element_blank(),
     strip.background = element_blank(), 
@@ -1381,12 +1387,18 @@ p <- ggplot(toplo, aes(x = date, y = bbave)) +
   facet_grid(typ ~ area, scales = 'free_y') +
   theme_minimal(base_size = 14) + 
   coord_cartesian(ylim = c(0, NA)) +
-  scale_fill_manual(values = cols) +
+  scale_fill_manual(
+    values = cols,
+    labels = c('Red', 'Green', 'Brown', 'Cyanobacteria', expression(italic('Thalassia testudinum')), 
+               expression(italic('Halodule wrightii')), expression(italic('Syringodium filiforme')),
+               'Total')
+    ) +
   labs(
     y = 'Freq. occurrence'
   ) +
   theme(
     legend.position = 'top', 
+    legend.text.align = 0,
     legend.title = element_blank(),
     strip.background = element_blank(), 
     strip.text = element_text(size = 14), 
@@ -1481,11 +1493,15 @@ lns <- physum %>%
 p1 <- ggplot() +
   geom_area(data = physum, aes(x = date, y = foest, fill = species), stat = 'identity', color = 'lightgrey', alpha = 0.7) +
   geom_segment(data = lns, aes(x = date, xend = date, y = 0, yend = ymax), color = 'grey') +
-  scale_fill_manual(values = c('#66c2a5', '#fc8d62')) +
+  scale_fill_manual(
+    values = c('#66c2a5', '#fc8d62'), 
+    labels = c('Diatoms', expression(italic('K. brevis')))
+    ) +
[…]

### 2021-11-02 R/funcs.R

@@ -423,6 +423,7 @@ show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12, rev 
     ggplot2::scale_x_continuous(breaks = savxlms, labels = savxlbs) +
     ggplot2::coord_cartesian(xlim = xlms) +
     ggplot2::theme(
+      legend.text = ggplot2::element_text(face = 'italic'), 
       panel.grid.major.y = ggplot2::element_blank(),
       panel.grid.minor.y = ggplot2::element_blank(),
       panel.grid.minor.x = ggplot2::element_blank(),
@@ -490,7 +491,7 @@ show_rstransect <- function(savdat, mcrdat, savsel, mcrsel, base_size = 12, rev 
     ungroup() %>%
     select(date, location) %>%
     unique()
-  
+
   pb <- ggplot2::ggplot(toplo1b, ggplot2::aes(y = date, x = location)) +
     ggplot2::geom_point(data = toplo2b, colour = 'black', alpha = 1, size = 2) +
     ggplot2::geom_point(inherit.aes = F, aes(colour = 'Empty sample'), x = NA, y = NA) +

### 2021-11-03 manu-draft.Rmd
edits from DC

@@
 Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).
-Contaminants and radionuclides exist in phosphogypsum, making it commercially invaluable and the resulting waste is typically stored on-site in large earthen stacks (gypstacks) or holding ponds capable of storing millions of liters of wastewater [@Burnett01].
-The stacks are usually near distribution centers where fertilizer is shipped elsewhere, such as port facilities close to coastal resources or population centers [@Beck18b].
-Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain wastewater on site.
-There are obvious environmental and human health risks associated with these stacks, primarily through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks.
-Examples exist worldwide demonstrating the potential harm of these facilities on the environment [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
+The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O), which is typically stored on-site in large earthen stacks (gypstacks) capable of storing millions of liters of wastewater.
+Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].
+In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b].
+Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
+Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
@@
 Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.
-Nitrogen inputs from external sources are well understood as drivers of algal blooms that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
+Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
 Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
@@
 By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks ([Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
+Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
 The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement.
 The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
-Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, HRK Holdings, LLC (hereafter HRK), with oversight by the Florida Department of Environmental Protection (FDEP).
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
@@
 
-Recently, HRK suspected leakages from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
-In response, FDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
+Recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
+In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
 At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.
@@
 The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
+We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
 A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
-We focus primarily on the perspective of the Tampa Bay Estuary Program (TBEP) in its role in coordinating monitoring and evaluating short-term impacts, particularly in the context of seagrass resources and long-term management goals that leverage existing partnerships among local resource management institutions.
 The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.
@@
 Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
-Petersburg, TBEP, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
 Details of the sampling methods are provided in supplement.
-Additional samples for contaminants (e.g., heavy metals), invertebrates, and nekton were also collected, but they are not reported here in anticipation of future monitoring events.
 
@@
 Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
-These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segment used by TBEP for assessing programmatic goals).
+These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals).
 Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
@@
 An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.
-For water quality variables, long-term monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.
+For water quality variables, monthly averages from long-term monitoring data were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.
 Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
@@
 Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).
-Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
+Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
 Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
@@
 wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
-The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17].
-At the baywide scale, shoal grass is the dominant species, whereas turtle grass is more common in euhaline waters closer to the Gulf.
+The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf.
+There is no historical macroalgae record for Tampa Bay that is comparable to the resolution of the 2021 samples.
 
@@
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
-From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication and primary production in 2021 were comparatively different than recent prior years.
+From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
 These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
@@
 brevis* in June and July.
+Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
+brevis* bloom.
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
@@
 
-Fisheries resources were undoubtedly impacted by *K.
+Fisheries resources were impacted by *K.
 brevis* blooms in June and July (Figure \@ref(fig:redtide)).
@@
 
-In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining are a serious threat to environmental resources.
+In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.
 In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.
@@
 Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Although the characteristics of water in a gypstack can vary based on site specific criteria, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
+Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the discharge stopped (see supplement).
+The fate of the phosphorus in Tampa Bay is unknown.
+Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
 Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.
@@
 Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the discharge, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
-Long-term fate of nitrogen from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
+Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
 Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
-These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, population growth, habitat loss, episodic weather events, and recurring pollutant sources to the bay.
+These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources to the bay.
 
@@
 Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Funding was provided to the University of Florida from the Ocean Conservancy and the National Science Foundation.
+Funding was provided to the University of Florida from the Ocean Conservancy (award 10027188) and the National Science Foundation (project ID 2130675).
 The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
@@
 
-```{r wqmap, fig.cap = 'Sampled water quality data for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+```{r wqmap, fig.cap = 'Water quality data (all observations) for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
 Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from March to July 2021 in response to wastewater discharge from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from April to September 2021 in response to wastewater discharge from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.

### 2021-11-03 supplement.Rmd
edits from DC

@@
 Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
-In 1997, 204 million liters of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes and impacting 153 hectares of wetland habitat.
+In 1997, 204 million liters of phosphate mining process water from the Mulberry plant spilled into the Alafia River, the second largest tributary to Tampa Bay, killing 1.3 million fishes [@DiPinto01].
 
-The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to FDEP.
+The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to the Florida Department of Environmental Protection (FDEP).
 Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
@@
 
-HRK acquired Piney Point in August 2006 through an administrative agreement with FDEP.
+Piney Point was acquired by HRK Holdings, LLC in August 2006 through an administrative agreement with FDEP, where oversight was still maintained by FDEP.
 This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
@@
 Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
-Discrete water quality samples were taken primarily from surface grabs by boat and processed by the respective laboratories of each participating agency.
+Discrete water quality samples were taken primarily from surface grabs by boat and processed for relevant parameters (see below) by the respective laboratories of each participating agency.
 Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).
@@
 This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
-Seagrasses and macroalgae were identified and abundances were estimated using Braun-Blanquet cover-abundance estimates within a 0.25 m$^2$ quadrat at 10m distances along each transect.
+Seagrasses and macroalgae were identified and Braun-Blanquet abundances were estimated within a 0.25 m$^2$ quadrat at 10m distances along each transect.
 Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
 Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
-Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria) based on expected responses to nutrient pollution.
+Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria).
 Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.

### 2021-11-03 R/figures.R

@@ -160,7 +160,8 @@ tomap <- rsallpts %>%
     type = case_when(
       grepl('\\,', type) ~ 'mixed monitoring', 
       T ~ type
-    )
+    ), 
+    type = factor(type, levels = c('water quality', 'algae', 'mixed monitoring', 'seagrass and macroalgae'))
   ) %>% 
   .[areas, ]
 

### 2021-11-04 manu-draft.Rmd
NW comments added

@@
 ---
-title: "Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
+title: "Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
 output: 
@@
 Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.
-From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater from the Piney Point facility were released into Tampa Bay (Florida, USA).
+From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).
 An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
@@
 Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
-By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release of wastewater from Piney Point are abnormal relative to historic conditions in Tampa Bay.
+By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.
 
@@
 Recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
-In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release wastewater from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
+In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
 At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.
@@
 Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September.
-These data were collected through a coordinated effort under the guidance of an effluent plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.
+These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.
 The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
-Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of effluent release and subsequent Tampa Bay distribution.
+Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution.
 The modeled plume evolution web product (Y.
@@
   filter(`Water quality variable` == 'Secchi (m)') %>% 
-  pull(`% below`) %>%  
+  pull(`% Below`) %>%  
   sum
@@
   filter(`Water quality variable` != 'Secchi (m)') %>% 
-  pull(`% above`) %>%  
+  pull(`% Above`) %>%  
   sum
@@
   filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
-  select(Area, `% above`) %>% 
+  select(Area, `% Above`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'TN (mg/L)') %>% 
-  select(Area, `% above`) %>% 
+  select(Area, `% Above`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'Secchi (m)') %>% 
-  select(Area, `% below`) %>% 
+  select(Area, `% Below`) %>% 
   deframe %>% 
@@
   filter(`Water quality variable` == 'DO (% sat.)') %>% 
-  select(Area, `% above`) %>% 
+  select(Area, `% Above`) %>% 
   deframe %>% 
@@
 
-In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen effluent directly into lower Tampa Bay.
+In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.
 Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
@@
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
-From both a monitoring perspective and the nature of the effluent directly discharged into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
+From both a monitoring perspective and the nature of the discharge into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
 These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
@@
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
-Concerns of the negative impacts of nutrients from the wastewater release focused primarily on the high concentrations of ammonia in the effluent (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
+Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
 Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.
@@
 brevis* to the concentrations observed in July [also see @Medina20].
-Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the effluent was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
+Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the discharge was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
 Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
@@
 Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
-Massive fish kills were observed and likely related to low pH of the effluent.
-Unlike Piney Point, inorganic nitrogen concentrations of the effluent were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
+Massive fish kills were observed and likely related to low pH of the discharge.
+Unlike Piney Point, inorganic nitrogen concentrations of the discharge were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
 Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the discharge stopped (see supplement).
@@
 
-```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to wastewater discharge from Piney Point.
+```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to discharge from Piney Point.
 Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
@@
 
-```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the discharge of wastewater from Piney Point.
+```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the discharge from Piney Point.
 Inset image shows blooms of the macroalgae *Dapis spp.*.'}
@@
 
-```{r wqmap, fig.cap = 'Water quality data (all observations) for April to July 2021 in response to wastewater discarge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+```{r wqmap, fig.cap = 'Water quality data (all observations) for April to September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
 Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
 Color ramps and point sizes show relative values (reversed for Secchi).
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location (Figure \\@ref(fig:map)a).
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at the nearest long-term monitoring site to each sample location (Figure \\@ref(fig:map)a).
 Values below detection limits (or Secchi on bottom) are not shown.'}
@@
 End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
-Missing values were not measured in the effluent or stack water.'
+Missing values were not measured in the discharge or stack water.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from April to September 2021 in response to wastewater discharge from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from April to September 2021 following the discharge from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.

### 2021-11-04 supplement.Rmd
NW comments added

@@
 ---
-title: "Supplement to Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida"
+title: "Supplement to Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
 output: 
@@
 Port operations were primarily for export of phosphate production by the plant.
-Numerous environmental issues were observed in these early years, including suspected wastewater contamination in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
+Numerous environmental issues were observed in these early years, including suspected surface water contamination from Piney Point in nearby Bishop Harbor, groundwater contamination from industrial solvents, and air pollution from plant emissions [@Henderson04].
 Ownership of the facility was transferred to different companies over the course of operation and in 1993 the plant was acquired by Mulberry Phosphates, Inc., which also owned a mining facility in Mulberry, Florida to the north.
@@
 The Mulberry corporation filed for bankruptcy in 2001, transferring regulatory oversight of the Piney Point facility to the Florida Department of Environmental Protection (FDEP).
-Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of wastewater on-site to minimize environmental impacts.
+Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of water on-site to minimize environmental impacts.
 Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.
-For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of wastewater to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
+For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of discharge from Piney Point to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
 comm.
@@
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
-Around the same time, 939 million liters of wastewater from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
+Around the same time, 939 million liters of water from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
 Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
@@
 This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of wastewater contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
+Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
 Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.
@@
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of wastewater on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
+Because of the time-sensitive nature of the potential impacts of discharge from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
 This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 in response to wastewater discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 in response to wastewater discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.

### 2021-11-04 R/tables.R

@@ -105,7 +105,7 @@ tab <- full_join(stkraw, bssum, by = 'var') %>%
     `Water quality variable` = lbs, 
     `2019 stack value` = stkval, 
     `2021 end-of-pipe value` = effval,
-    `Bay median (min, max)` = sumv
+    `2006 - 20202 Bay median (min, max)` = sumv
   ) 
 
 stktab <- tab
@@ -140,7 +140,7 @@ totab <- rswqdat %>%
   st_intersection(ppsegbf) %>% 
   filter(!qual %in% 'S') %>%  # remove secchi on bottom
   mutate(
-    qual = grepl('U', qual) # censored data
+    qual = grepl('^U$|^S$', qual) # censored data
   ) %>% 
   st_set_geometry(NULL)
 
@@ -158,9 +158,12 @@ totab2 <- totab %>%
   left_join(sigs, by = 'lbs') %>% 
   summarise(
     nondetect = sum(qual),
-    medv = round(median(cenfit(val, censored = qual), na.rm = T), unique(sigdig)),
-    minv = round(min(val, na.rm = T), unique(sigdig)), 
-    maxv = round(max(val, na.rm = T), unique(sigdig)), 
+    medv = case_when(
+      grepl('Secchi', unique(lbs)) ~ round(median(val, na.rm = T), unique(sigdig)),
+      T ~ round(median(cenfit(val, censored = qual), na.rm = T), unique(sigdig)),
+    ), 
+    minv = round(min(val, na.rm = T), unique(sigdig)),
+    maxv = round(max(val, na.rm = T), unique(sigdig)),
     cnt = n(), 
     .groups = 'drop'
   ) %>% 
@@ -174,20 +177,20 @@ totab2 <- totab %>%
   left_join(totab1, by = c('area', 'lbs')) %>% 
   mutate(area = ifelse(duplicated(area), '', area)) %>% 
   mutate(
-    `% in range` = round(100 * `in range` / cnt, 0),
-    `% above` = round(100 * `above` / cnt, 0),
-    `% below` = round(100 * `below` / cnt, 0), 
-    `% below detection` = round(100 * nondetect / cnt, 0)
+    `% In range` = round(100 * `in range` / cnt, 0),
+    `% Above` = round(100 * `above` / cnt, 0),
+    `% Below` = round(100 * `below` / cnt, 0), 
+    `% Outside detection` = round(100 * nondetect / cnt, 0)
   ) %>% 
   select(
[…]

### 2021-11-04 R/tables.R
save file

@@ -138,7 +138,7 @@ totab <- rswqdat %>%
   filter(!station %in% nonbay) %>% 
   select(station, date, lbs, val, inrng, qual) %>% 
   st_intersection(ppsegbf) %>% 
-  filter(!qual %in% 'S') %>%  # remove secchi on bottom
+  # filter(!qual %in% 'S') %>%  # remove secchi on bottom
   mutate(
     qual = grepl('^U$|^S$', qual) # censored data
   ) %>% 

### 2021-11-08 manu-draft.Rmd
edit to funding acknowledgements
EM and EP edits added

@@
 ---
-title: "Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
+title: "Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
 output: 
@@
 An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
-An immediate phytoplankton response (non-harmful diatom) was observed in samples closest to the discharge site.
-Cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms in Tampa Bay proper in July.
+In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
+Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms that peaked in July.
 Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
@@
 This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
-In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference (see https://shiny.tbep.org/piney-point/ for a map of the matches).
+In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
+The average distance from a monitoring location in 2021 to the long-term sites was 1.6 km (see https://shiny.tbep.org/piney-point/ for a map of the matches).
 
@@
 
-A general summary of 2021 events in Tampa Bay following discharge rom Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge ceased on April 9th, an initial phytoplankton response was observed near Piney Point with concentrations peaking around mid-April.
-Taxa from the Bacillariophyta phylum (diatoms) were dominant in April, with a maximum April Chl-a concentration of 50.5 ug/L, although median concentrations for each week in April were less than 10 ug/L.
-The initial diatom bloom did not persist past April and was partially dispersed by the passage of a strong weather front across the Tampa Bay region on April 10.
-On April 20th, *K.
-brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay and reached bloom concentrations (>10,000 cells/L) by May 23rd, although bloom conditions were limited to lower Tampa Bay.
-Also during May, *Dapis spp.* (filamentous cyanobacteria macroalgae) were observed at high abundances in Anna Maria Sound and near Port Manatee.
-*Dapis* were observed in large floating mats on the surface and covering benthic and seagrass habitats at these locations.
+A general summary of 2021 events in Tampa Bay following discharge from Piney Point is shown in Figure \@ref(fig:timeline).
+After the discharge ceased on April 9th, a peak in median chl-a concentration was observed near Piney Point (Area 1, Figure \@ref(fig:map)a) mid-April, with peak individual sample values in excess of 50 ug/L.
+Median concentrations for each week in April were less than 10 ug/L.
+The discharge phytoplankton biomass was comprised of over 99% of a spherical nanoplanktonic chlorophyte, with density of 337 x 10$^6$ cells/L.
+The phytoplankton communities near the discharge area in April were generally dominated by diatoms (Bacillariophyta).
+The initial diatom bloom did not persist past April.
+On April 20th, the toxic red tide species *Karenia brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay.
+By May/June, bloom levels of *K.
+brevis* were observed, with peak concentrations in excess of 1 x 10$^6$ cells/L in lower Tampa Bay.
+Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis spp.*) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
+High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface.
 By June 27th, fish kill reports attributed to red tide increased with *K.
@@
 brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.
-Conditions were relatively stable in August and September compared to the months prior.
+Conditions were relatively stable in August and September compared to months prior.
 A quantitative description of these events follows.
@@
 Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
-Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point, Anna Maria Sound, and the northern mouth of Tampa Bay (Figure \@ref(fig:wqmap)).
+Spatial variation among the parameters showed that values were generally above the normal range (or below for Secchi depth) for many locations near Piney Point (Area 1), Anna Maria Sound (Area 3), and the northern mouth of Tampa Bay (Area 3, Figure \@ref(fig:wqmap)).
 
@@
 Macroalgae observed along the transects varied in coverage, with red macroalgae groups having the highest frequency occurrence of `r foall['Red']`\%.
-Common taxa in the red group included *Gracilaria spp.* and *Acanthophora sp.*.
+Common taxa in the red group included genera *Gracilaria* and *Acanthophora*.
 Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
-Common taxa in the green group included *Ulva spp.* and *Caulerpa spp.*, whereas *Dapis spp.* was the only cyanobacteria macroalgae observed.
+Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genera *Dapis*.
 Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence.
@@
 Red macroalgae were present in high abundances from April to May.
-*Dapis spp.* was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
-*Dapis spp.* persisted through June and July, but not was observed after July 20th.
+Benthic filamentous macroalgal cyanobacteria (*Dapis spp.*) was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
+Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.
 Green macroalgae taxa were first observed in July, although at generally low abundances.
@@
 Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1.
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from the TBCOM simulations.
 Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).
@@
 
+Concentrations of the red tide species *K.
+brevis* in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
 brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a handful of years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
+The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
 Median cell concentrations for most years were well below 1,000 cells/L.
 The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
-Bloom concentrations in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
@@
 These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
-The observed water quality conditions throughout this period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
+The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
 These conditions were widespread in lower and middle Tampa Bay.
@@
 Temporally, these changes were observed at different times and for different species of phytoplankton.
-The initial increase in chl-a was first associated with a diatom bloom in April and then by *K.
-brevis* in June and July.
+The initial increase in chl-a was first associated with a diatom bloom in April.
+In June, a bloom of the red tide species *K.
+brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay through July.
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
@@
 
-Macroalgae trends across the study period were much more dramatic than the subtle differences observed in the seagrass community.
-This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay.
-The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.
-Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
-Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge.
-Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
-However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
+There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread through a wide region of lower and middle Tampa Bay.
+Fisheries resources were impacted by *K.
+brevis* blooms in June and July (Figure \@ref(fig:redtide)).
+Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
+Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
+Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
+Given the observed *K.
+brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
+For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
+Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
+brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
+Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.
 
-For seagrasses, both major bloom events produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study.
+For seagrasses, both major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+
+Macroalgae trends across the study period were much more dramatic than the minimal changes observed in the seagrass community.
+This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.
+The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay.
+The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.
+Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
+Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge.
+Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
+However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
 
@@
 
-Fisheries resources were impacted by *K.
-brevis* blooms in June and July (Figure \@ref(fig:redtide)).
-Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
-Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
-Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
-Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
-For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
-Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
-brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.
-
-Additional monitoring and analysis is also required to fully understand the long-term impacts to bay resources beyond water quality.
+Additional monitoring and analysis are also required to fully understand the long-term impacts to bay resources beyond water quality.
 For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
 Comparison of the April samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay will be available in 2022 for both April and September 2021 benthic samples.
+Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay for both April and September 2021 benthic samples are not yet available at the time of writing.
 Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
@@
 However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-The weight of evidence that 2021 is an anomaly for both nitrogen loading and water quality response relative to recent prior years provides justification that our interpretation is within the limits of what would be expected by more mechanistic descriptions.
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
@@
 Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Funding was provided to the University of Florida from the Ocean Conservancy (award 10027188) and the National Science Foundation (project ID 2130675).
+Funding was provided to the University of Florida from the Ocean Conservancy and the National Science Foundation (project ID 2130675).
 The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
@@
 Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.
-Points are offset slightly for readability.')}
+Points are offset slightly for readability.
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from model simulations.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))

### 2021-11-08 supplement.Rmd
EM and EP edits added

@@
 ---
-title: "Supplement to Initial estuarine response to the nutrient dense Piney Point discharge into Tampa Bay, Florida"
+title: "Supplement to Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
 output: 
@@
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of discharge from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified as a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
+Because of the time-sensitive nature of the potential impacts of discharge from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified using a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
 This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for April to July 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March to July 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for April to September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for late March to September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) area 1 and (b) area 3 (see map in Figure 1a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
+```{r trnabu-supp, fig.cap = 'Abundance estimates (+/- 95% confidence) for (a) Area 1 and (b) Area 3 (see map in Figure 1a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = 38) near Piney Point.
 Estimates are grouped by sample months in 2021.
-Points are offset slightly for readability.'}
+Points are offset slightly for readability.
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from model simulations.'}
 knitr::include_graphics(here('figs/trnabu-supp.jpeg'))

### 2021-11-08 R/tables.R

@@ -105,7 +105,7 @@ tab <- full_join(stkraw, bssum, by = 'var') %>%
     `Water quality variable` = lbs, 
     `2019 stack value` = stkval, 
     `2021 end-of-pipe value` = effval,
-    `2006 - 20202 Bay median (min, max)` = sumv
+    `2006 - 2020 lower Tampa Bay median (min, max)` = sumv
   ) 
 
 stktab <- tab

### 2021-11-10 manu-draft.Rmd
personal edits after coauth revise

@@
 Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W.
-Chin (diwchin@gmail.com), Emma E.
-Dontis (edontis@co.pinellas.fl.us), Jayne Gardiner (jgardiner@ncf.edu), Jessica Lewis (jlewis@tbep.org), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
+Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
 Phlips (phlips@ufl.edu), Gary E.
@@
 
+# seagrass transect counts
+ntrnscnt <- length(unique(rstrndat$station))
+ntrns <- english(ntrnscnt)
+
 # # extract bib entries from online
@@
 In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
-Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by *Karenia brevis* blooms that peaked in July.
-Reported fish kills tracked red tide concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
+Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.
+Reported fish kills tracked *K.
+brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
 By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.
@@
 Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O), which is typically stored on-site in large earthen stacks (gypstacks) capable of storing millions of liters of wastewater.
+The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).
+This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
 Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].
 In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b].
-Many of these gypstacks no longer supporting active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
+Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
 Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
-Ongoing threats and challenges to protecting water quality of Gulf Coast estuaries persist despite recent environmental recovery.
+Ongoing threats and challenges to protecting water quality of northern Gulf of Mexico estuaries persist despite recent environmental recovery.
 Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.
@@
 The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement.
-The Piney Point facility located in Palmetto, Florida is a large, remnant phosphogypsum stack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [@Henderson04].
+The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
@@
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater in April, 2021.
+This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.
 The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
@@
 Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September.
-These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science Ocean Circulation Lab.
+These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.
 The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
@@
 Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
-Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling (laboratory processing of discrete samples and *in situ* measurements), phytoplankton cell counts, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
+In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (ug/L), pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%).
+Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point.
+Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples.
+Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
 Details of the sampling methods are provided in supplement.
@@
 These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals).
-Observations at each monitoring station were averaged for each month across years from 2006 to 2020.
-This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares (38,000 acres) [@Greening14;@Sherwood17].
-For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point.
-This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
+Observations at each long-term monitoring station were averaged for each month across years from 2006 to 2020.
+This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 15,378 hectares [38,000 acres, @Greening14;@Sherwood17].
+For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest water quality monitoring station that was sampled in response to Piney Point.
+This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions near the mouth of Tampa Bay are not the same as those in the middle of the bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July).
 In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.
@@
 The historical monitoring data were also used to model an expected seasonal pattern for water quality parameters from April to October in 2021.
-This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).
+This was done by estimating smoothed annual and seasonal splines with Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).
 GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
@@
 Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
-Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the five month period.
+Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.
 Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a.
@@
 Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
-When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.
-These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
+Methods to accommodate measured concentrations of water quality variables that were below detection were used that included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
 
@@
 The percentage of observations outside of the normal range (mean +/- 1 standard deviation from long-term data) varied by location and parameter.
-For chl-a, `r chlabv[['1']]`\% of the observations were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively.
+For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively.
 Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.
@@
 For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.
-Concentrations decreased slightly until June and July when values increased again above the seasonal expectation.
+Concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with the increase in *K.
+brevis* concentrations.
 Many Secchi observations in Area 1 were lower than normal in April and July.
 Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.
-Statistical comparisons between months for TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
+Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
 Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)).
-Further analsyis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter.
-Observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
-brevis* abundances.
+Further analysis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter, such that observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.
+brevis* abundances by mid-summer.
 
@@
 ```{r}
-ntrnscnt <- length(unique(rstrndat$station))
-ntrns <- english(ntrnscnt)
 visdat <- rstrndat %>% 
@@
   )
+fomobrn <- fomo %>% 
+  filter(taxa == 'Brown') %>% 
+  filter(foest > 0) %>% 
+  pull(foest)
 fomoarea1 <- fomo %>% 
@@
 Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genera *Dapis*.
-Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at `r foall['Brown']`\% frequency occurrence.
-For seagrasses, turtle grass (*T.
-testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
-Manatee grass (*S.
-filiforme*) and shoal grass (*H.
-wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
+Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at one transect in April (`r fomobrn`\% frequency occurrence).
+For seagrasses, turtle grass (*Thalassia testudinum*) was the dominant species with frequency occurrence of `r foall['Thalassia testudinum']`\% across all locations and sample dates.
+Manatee grass (*Syringodium filiforme*) and shoal grass (*Halodule wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
 The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf.
@@
 
-Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends across all transects.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from the TBCOM simulations.
+Monthly summaries in frequency occurrence by area (Figure \@ref(fig:trnfrq)) provided an indication of macroalgae and seagrass trends in 2021 across all transects.
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from the TBCOM simulations.
 Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).
 Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3).
-Cyanobacteria frequency occurrence peaked in June, with greater coverage in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
+Reductions in red macroalgae frequency occurrence were observed in June when cyanobacteria frequency occurrence peaked, with greater coverage of cyanobacteria in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
 Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).
@@
 Turtle grass (*T.
-testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall Area 3), compared to shoal grass (*H.
+testudinum*) occurred in higher frequency occurrence in both areas (`r fomoarea1ave['Thalassia testudinum']`\% overall in Area 1, `r fomoarea3ave['Thalassia testudinum']`\% overall in Area 3), compared to shoal grass (*H.
 wrightii*, `r fomoarea1ave['Halodule wrightii']`\% Area 1, `r fomoarea3ave['Halodule wrightii']`\% Area 3) and manatee grass (*S.
@@
 Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
-Statistical analyses with multiple comparison tests confirmed the general trends described above (see supplement).
+Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (see supplement).
+Tests using Braun Blanquet abundances also produced similar results (see supplement).
 
@@
 Concentrations of the red tide species *K.
-brevis* in 2021 were not observed until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+brevis* in 2021 were not observed in Tampa Bay until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
 brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1953 to present in Figure \@ref(fig:redtide)a shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
+The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
 Median cell concentrations for most years were well below 1,000 cells/L.
 The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
-Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (Figure \@ref(fig:redtide)c, d).
-This likely contributed to elevated salinity values in lower and middle Tampa Bay that created conditions favorable for *K.
+Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
+This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K.
 brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.
@@
 brevis* cell concentrations (Figure \@ref(fig:redtide)b).
-Notably, all of the fish kill reports occured within a 1.5 month period, whereas red tide in the bay was observed for nearly four months.
-Increased fish kill reports in early July may coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f), which moved dead fish closer to populated nearshore areas, as noted earlier.
+Notably, all of the fish kill reports occurred within a 1.5 month period during peak *K.
+brevis* cell concentrations, whereas red tide in the bay was observed for nearly four months.
+Increased fish kill reports in early July also coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f, pre-, post-storm wind roses), which moved dead fish closer to populated nearshore areas, as noted earlier.
 By August, Pinellas County and the city of St.
@@
 
-The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April to September, starting with the influx of inorganic nitrogen from Piney Point.
+The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point.
 Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms ((e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
+This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of red tide in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K.
+brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
 The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
@@
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
-brevis* bloom.
+brevis* bloom (see supplement).
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
@@
 Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
-Although Piney Point did not cause red tide, the events of 2021 likely created conditions leading to the extreme bloom concentrations in July, which likely would not have been observed if nutrients from Piney Point were not discharged to Tampa Bay.
+Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
 
 Additional monitoring and analysis are also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106, see https://shiny.tbep.org/piney-point], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
 Comparison of the April samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
@@
 Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
-These efforts are critical to managing the health of Tampa Bay given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources to the bay.
+These efforts are critical to managing the health of coastal resources of the Gulf of Mexico given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
 
@@
 
-```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March to July 2021 by monitoring data type (b) in response to discharge from Piney Point.
+```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March through September 2021 by monitoring data type (b) in response to discharge from Piney Point.
 Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
@@
 ```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the discharge from Piney Point.
-Inset image shows blooms of the macroalgae *Dapis spp.*.'}
+Inset image shows blooms of cyanobacteria macroalgae (*Dapis spp.*).'}
 knitr::include_graphics(here('figs/timeline.jpg'))
@@
 
-```{r wqmap, fig.cap = 'Water quality data (all observations) for April to September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
 Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
@@
 Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, with darker lines for more recent years.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
@@
 ```{r trnex, fig.cap = 'Results for (a) macroalgae and (b) seagrass rapid response transect surveys at a site (S3T6, -82.55866 W longitude, 27.64483 N latitude) near Piney Point.
-Sample dates in 2021 are shown in rows with transect meter results shown in columns.
+Sample dates in 2021 are shown in rows with transect meter results shown in columns (0m nearshore, 50m offshore).
 Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis spp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
@@
 
-```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) Area 1 and (b) Area 3 (see map Figure \\@ref(fig:map)a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ') near Piney Point.
+```{r trnfrq, fig.cap = paste0('Frequency occurrence estimates for (a) Area 1 and (b) Area 3 (see map Figure \\@ref(fig:map)a for locations) for macroalgae (top) and seagrass (bottom) rapid response transect surveys across all transects (n = ', ntrnscnt, ').
 Estimates are grouped by sample months in 2021.
@@
 Points are offset slightly for readability.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from model simulations.')}
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from model simulations.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))
@@
 
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (100k cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after discharge and tropical storm Elsa.
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after Piney Point discharge and tropical storm Elsa.
 *K.
 brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
-brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for all of Tampa Bay from USGS gaged sites, and wind rose plots are from St.
+brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for summed hydrologic loads of major tributaries to Tampa Bay from USGS gaged sites, and wind rose plots are from wind measurements at St.
 Petersburg, Florida.
@@
 End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
-Missing values were not measured in the discharge or stack water.'
+Missing values were not measured in the stack water or discharge.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from April to September 2021 following the discharge from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from April through September 2021 following the discharge from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.
@@
 Total observations (N obs.) and the percentage of observations in range, above, or below normal ranges are also shown.
-Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2021 for values collected at the nearest long-term monitoring site to each sample location.
+Normal ranges are defined as within +/-1 standard deviation of the mean for the month of observation from 2006 to 2020 for values collected at the nearest long-term monitoring site to each sample location.
 The final column shows the percentage of total observations that were outside of detection, defined as minimum laboratory detection limits for all parameters and values on the bottom for Secchi observations.
@@
 column) were used to evaluate pairwise monthly concentrations for each water quality variable in each area.
-Rows that share a letter within each area and water quality variable combination have concentrations that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+Rows that share letters within each area and water quality variable combination have concentrations that are not significantly different between month pairs.
+All statistical tests were performed on the seasonally-corrected water quality values that were based on observations with the long-term monthly median subtracted (observed medians are shown for comparison).
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'

### 2021-11-10 supplement.Rmd
personal edits after coauth revise

@@
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-In other event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+Another event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
@@
 
-Piney Point was acquired by HRK Holdings, LLC in August 2006 through an administrative agreement with FDEP, where oversight was still maintained by FDEP.
+Piney Point was acquired by HRK Holdings, LLC in August 2006 through an administrative agreement with FDEP, where oversight was still maintained by the latter.
 This agreement transferred responsibility of the site to HRK with the intention that any future uses must protect and be compatible with the integrity of stack closure and long-term care.
 In 2011, HRK agreed to the storage of 1.1 million cubic meters of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing gypstack at Piney Point.
+This material was added to an existing gypstack at Piney Point (NGS-S, the point of discharge for 2021).
 Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 640 million liters of dredged saltwater slurry and 3.2 metric tons of nitrogen to receiving waters leading to Bishop Harbor.
@@
 This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (^o C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
+Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
 Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.
@@
 Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria).
-Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis described below.
+Seagrasses and macroalgae abundances were converted to frequency of occurrence estimates (i.e., number of locations present divided by total locations sampled) at the transect scale or within major areas (Figure 1a) depending on the analysis.
 
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March to July 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for late March to September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) total phosphorus (mg/L), (c) orthophosphate (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, with darker lines for more recent years.
+Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure 1a.'}
@@
 Points are offset slightly for readability.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given dispersal patterns of the discharge from model simulations.'}
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from model simulations.'}
 knitr::include_graphics(here('figs/trnabu-supp.jpeg'))
@@
 Rows that share a letter within each area and macroalgae group combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+Data are not seasonally-corrected due to limited long-term seasonal observations.
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
@@
 Rows that share a letter within each area and seagrass species combination have frequency occurrences that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+Data are not seasonally-corrected due to limited long-term seasonal observations.
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
@@
 Rows that share a letter within each area and macroalgae group combination have abundances that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+Data are not seasonally-corrected due to limited long-term seasonal observations.
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'
@@
 Rows that share a letter within each area and seagrass species combination have abundances that are not significantly different between month pairs.
-Probability values were adjusted for the pairwise comparisons using the Bonferroni method in @Holm79.
+Data are not seasonally-corrected due to limited long-term seasonal observations.
 ** p < 0.005, * p < 0.05, blank is not significant at $\\alpha$ = 0.05.'

### 2021-11-10 R/figures.R

@@ -1128,7 +1128,7 @@ p4 <- ggplot(wtsest, aes(x = week, y = medv)) +
   ) + 
   labs(
     y = expression(paste('kg / ', m^2)), 
-    x = 'Week of', 
+    x = NULL, 
     title = '(d) Cyanobacteria macroalgae abundances'
   )
 
@@ -1150,7 +1150,7 @@ p5 <- ggplot(toplo2, aes(x = week, y = medv)) +
     panel.grid.major.x = element_blank()
   ) +
   labs(
-    x = NULL,
+    x = 'Week of',
     y = 'Cells / L',
     title = expression(paste('(e) ', italic('K. brevis'), ' cell concentrations'))
   )
@@ -1287,7 +1287,7 @@ mcrcol <- c('tomato1', 'lightgreen', 'burlywood3', 'lightblue')
 names(mcrcol) <- mcrsel
 mcrcol <- mcrcol[mcrsel]
 cols <- c(mcrcol, savcol)
-cols <- c(cols, Total = 'white')
+# cols <- c(cols, Total = 'white')
 
 # segments
 areas <- ppseg %>% 
@@ -1391,11 +1391,11 @@ p <- ggplot(toplo, aes(x = date, y = bbave)) +
   scale_fill_manual(
     values = cols,
     labels = c('Red', 'Green', 'Brown', 'Cyanobacteria', expression(italic('Thalassia testudinum')), 
-               expression(italic('Halodule wrightii')), expression(italic('Syringodium filiforme')),
-               'Total')
+               expression(italic('Halodule wrightii')), expression(italic('Syringodium filiforme')))#,
+               #'Total')
     ) +
   labs(
-    y = 'Freq. occurrence'
+    y = 'Braun-Blanquet abundance'
   ) +
   theme(
     legend.position = 'top', 

### 2021-11-10 R/tables.R

@@ -471,6 +471,7 @@ savests <- ests %>%
   group_by(area) %>%  
   mutate(
     area = ifelse(duplicated(area), '', area), 
+    taxa = paste0('*', taxa, '*'), 
     taxa = ifelse(duplicated(taxa), '', as.character(taxa)), 
     krusk = ifelse(duplicated(krusk), '', krusk),
     pval = ifelse(duplicated(pval), '', pval)
@@ -635,6 +636,7 @@ savests <- ests %>%
   group_by(area) %>%  
   mutate(
     area = ifelse(duplicated(area), '', area), 
+    taxa = paste0('*', taxa, '*'), 
     taxa = ifelse(duplicated(taxa), '', as.character(taxa)), 
     krusk = ifelse(duplicated(krusk), '', krusk),
     pval = ifelse(duplicated(pval), '', pval)

### 2021-11-11 manu-draft.Rmd
Results sentence
Grammatical nit-picking on plurality.

@@
 
-Total nitrogen, chl-a, and Secchi depth followed a temporal progression in 2021 that was distinct from the long-term seasonal trend estimated from historical data (Figure \@ref(fig:wqgam)).
+Total nitrogen, chl-a, and Secchi depth followed temporal progressions in 2021 that were distinct from long-term seasonal trends estimated from historical data (Figure \@ref(fig:wqgam)).
 For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.

### 2021-11-11 manu-draft.Rmd
Red tide concentration timeline tweak
Suggested edit, as stated above.

@@
 
-Concentrations of the red tide species *K.
-brevis* in 2021 were not observed in Tampa Bay until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+Bloom concentrations of the red tide species *K.
+brevis* in 2021 were first observed in Tampa Bay the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.

### 2021-11-11 manu-draft.Rmd
Bloom descriptions
Some more suggested tweaks for describing red tide 'bloom' concentrations according to FWRI's scale.

--- 2021-11-11 manu-draft.Rmd
@@
 
-Concentrations of the red tide species *K.
-brevis* in 2021 were not observed in Tampa Bay until the week of May 23, with concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+'Medium' bloom concentrations of the red tide species *K.
+brevis* in 2021 were first observed in Tampa Bay the week of May 23, with 'high' bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
-brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum, 'low' bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
 The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.

### 2021-11-11 manu-draft.Rmd
Making the inflow statements a bit clearer to the year 2021
As stated above.

@@
 The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
-Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay were below historical values (1995 - 2020) in the months prior to the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
+Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
 This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K.
-brevis* growth, in addition to the elevated nutrient concentrations from the Piney Point discharge.
+brevis* growth in 2021, in addition to the elevated nutrient concentrations from the Piney Point discharge.
 

### 2021-11-11 manu-draft.Rmd
Discussion tweaks
Mostly focused on changing context to 'release' impacts.

@@
 
-The observed conditions in Tampa Bay in 2021 following the discharge from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+The observed conditions in Tampa Bay in 2021 following unanticipated releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the discharge at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K.
 brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
@@
 In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.
-Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous discharges to Tampa Bay.
-An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters discharged from Piney Point.
+Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay.
+An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point.
 Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
-From both a monitoring perspective and the nature of the discharge into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
+From both a monitoring perspective and the nature of the unanticipated release into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
 These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
@@
 The initial increase in chl-a was first associated with a diatom bloom in April.
-In June, a bloom of the red tide species *K.
-brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay through July.
+In April, the red tide species *K.
+brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay from June through July.
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
@@
 Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
-Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaking in July.
+Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaked in July.
 Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
 
-There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread through a wide region of lower and middle Tampa Bay.
+There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread throughout lower and middle Tampa Bay.
 Fisheries resources were impacted by *K.
@@
 Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) have been extended through the fall.
-For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to the subembayment.
+brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall 2021.
+For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
 Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
 brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-Prior blooms were more localized than the 2021 bloom, potentially allowing fishes to migrate from harmful conditions.
+Prior blooms were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.
 
-For seagrasses, both major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the study.
+For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the initial study period.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Acute stressors from short-term events, such as discharge from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
 
@@
 The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.
-Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the discharge site.
-Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those as a potential response to the Piney Point discharge.
+Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site.
+Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those in potential response to the Piney Point release.
 Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
@@
 brevis*.
-However, the results suggest a likely scenario that residual nutrients from the Piney Point discharge, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K.
+However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K.
 brevis* to the concentrations observed in July [also see @Medina20].
-Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the discharge was widespread throughout the bay and persisted for many months after the discharge ceased at Port Manatee.
+Daily simulation results from the Tampa Bay Coastal Ocean Model [@Chen18;@Chen19] suggested that the plume was widespread throughout the bay and persisted for many months after the release ceased at Port Manatee.
 Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
@@
 In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.
-In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from gypstack release.
+In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.
 Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].
 Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
-Massive fish kills were observed and likely related to low pH of the discharge.
-Unlike Piney Point, inorganic nitrogen concentrations of the discharge were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the discharge stopped (see supplement).
+Massive fish kills were observed and likely related to low pH of the water released.
+Unlike Piney Point, inorganic nitrogen concentrations of the release were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
+Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).
 The fate of the phosphorus in Tampa Bay is unknown.
@@
 Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the discharge, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
 Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.

### 2021-11-11 manu-draft.Rmd
Tweaks to Intro
Changing text to 'release', and updating some descriptors of potential gypstack impacts.

@@
 An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
-In the period immediately after the discharge, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
+In the period immediately after the release, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
 Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.
@@
 brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
-By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the discharge from Piney Point are abnormal relative to historic conditions in Tampa Bay.
+By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release from Piney Point are abnormal relative to historic conditions in Tampa Bay.
 
@@
 Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
-Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled discharge to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
+Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled releases to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
@@
 By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no comprehensive, long-term plan for closure or disposal of waste.
-The potential environmental impacts of these facilities are considered a critical threat that could halt or reverse the decades of prior ecosystem improvement.
+Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.
+Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments.
 The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
-Discharges from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.
-Those discharges resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].
+Unanticipated releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.
+Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].
 
@@
 We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
-A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency discharges from the Piney Point facility.
+A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.
 The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.
@@
 
-Monitoring of the natural resources of Tampa Bay in response to the discharge from Piney Point began in April, 2021 and continued for six months through September.
-These data were collected through a coordinated effort under the guidance of a discharge plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.
-The evolution of discharge from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic discharge rates.
+Monitoring of the natural resources of Tampa Bay in response to the release from Piney Point began in April, 2021 and continued for six months through September.
+These data were collected through a coordinated effort under the guidance of a plume simulation by a numerical circulation model run by the Ocean Circulation Lab at the University of South Florida (USF), College of Marine Science.
+The plume evolution from Piney Point was simulated using the Tampa Bay Coastal Ocean Model (TBCOM) nowcast/forecast system [@Chen18;@Chen19], with an embedded tracer module that included realistic release rates.
 Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution.

### 2021-11-11 manu-draft.Rmd
Title update
Discharge to release

@@
 ---
-title: "Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
+title: "Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida"
 output: 

### 2021-11-17 manu-draft.Rmd
ready to go back to co-authors

@@
 This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.
-The goal is to describe the results of monitoring data of surface waters collected in response to the discharge event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
+The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
 We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
@@
 
-## Monitoring response to the emergency discharge
+## Monitoring response to the emergency release
 
@@
 
-'Medium' bloom concentrations of the red tide species *K.
-brevis* in 2021 were first observed in Tampa Bay the week of May 23, with 'high' bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+Medium bloom concentrations of the red tide species *K.
+brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b, low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L).
 The increase in *K.
-brevis* from April to July to concentrations exceeding 10,000 cells/L (minimum, 'low' bloom concentration based on FWC breakpoints) was an anomaly in 2021 that is not regularly observed in Tampa Bay.
+brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.
 The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
@@
 The initial increase in chl-a was first associated with a diatom bloom in April.
-In April, the red tide species *K.
-brevis* was introduced to Tampa Bay from the Gulf of Mexico, and persisted at high densities in the Bay from June through July.
+The red tide species *K.
+brevis* was also first introduced to Tampa Bay from the Gulf of Mexico in April, but was not observed at high densities in the Bay until June and July.
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
@@
 Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall 2021.
+brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall of 2021.
 For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
@@
 Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
-These efforts are critical to managing the health of coastal resources of the Gulf of Mexico given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
+These efforts are critical to managing Gulf of Mexico ecosystems given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
 
@@
 
-```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March through September 2021 by monitoring data type (b) in response to discharge from Piney Point.
+```{r map, fig.cap = 'Areas of interest and long-term monitoring stations (a) for evaluating status and trends in response-based monitoring data and sample locations from March through September 2021 by monitoring data type (b) in response to release from Piney Point.
 Data types include algae sampling, seagrass and macroalgae, water quality (field-based and laboratory samples), and mixed monitoring (algae, seagrass and macroalgae, water quality).
@@
 
-```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the discharge from Piney Point.
+```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the release from Piney Point.
 Inset image shows blooms of cyanobacteria macroalgae (*Dapis spp.*).'}
@@
 
-```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
 Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
@@
 ```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters) by area based on historal seasonal models.
-Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Forecasts from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
 Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
@@
 Points are offset slightly for readability.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from model simulations.')}
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the plume from model simulations.')}
 knitr::include_graphics(here('figs/trnfrq.jpeg'))
@@
 
-```{r redtide, fig.cap = '*Karenia brevis* concentrations (cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after Piney Point discharge and tropical storm Elsa.
+```{r redtide, fig.cap = '*Karenia brevis* concentrations (cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after Piney Point release and tropical storm Elsa.
 *K.
@@
 End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
-Missing values were not measured in the stack water or discharge.'
+Missing values were not measured in the stack water or release water.'
 knitr::kable(stktab, booktabs = T, caption = cap.val)
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from April through September 2021 following the discharge from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from April through September 2021 following the release water from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.

### 2021-11-17 supplement.Rmd
ready to go back to co-authors

@@
 ---
-title: "Supplement to Initial estuarine response to the nutrient-rich Piney Point discharge into Tampa Bay, Florida"
+title: "Supplement to Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida"
 output: 
@@
 Although phosphate production no longer occurred at the site, focus over the next twenty years centered on containment and treatment of water on-site to minimize environmental impacts.
-Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to discharge events to surficial and ground waters.
-For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing over 38 million liters of discharge from Piney Point to be released into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
+Despite these efforts, reduced holding capacities and degraded physical integrity of the holding ponds, including degradation of the pond liner, likely contributed to the releases to surficial and ground waters.
+For example, tropical storm Gabrielle in 2001 produced 33 centimeters of rain, causing release of over 38 million liters of water from Piney Point into Bishop Harbor, with an estimated 14 metric tons of nitrogen (pers.
 comm.
@@
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-Another event lasting from November 2003 to October 2004, treated process water from Piney Point was discharged to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+Another event lasting from November 2003 to October 2004, treated process water from Piney Point was released to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
 Around the same time, 939 million liters of water from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
-Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of discharge to local areas.
+Efforts for onsite treatment were also increased during this period to increase pH, remove heavy metals, and reduce nutrient concentrations to minimize impacts of unanticipated release to local areas.
 
@@
 In 2011, HRK agreed to the storage of 1.1 million cubic meters of dredged material and seawater from Port Manatee to improve shipping capacity at the port (i.e., Berth 12 construction).
-This material was added to an existing gypstack at Piney Point (NGS-S, the point of discharge for 2021).
-Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency discharge that released 640 million liters of dredged saltwater slurry and 3.2 metric tons of nitrogen to receiving waters leading to Bishop Harbor.
+This material was added to an existing gypstack at Piney Point (NGS-S, the point of release for 2021).
+Placement of the dredged material was suspected in compromising the liner integrity which led to an emergency release of 640 million liters of dredged saltwater slurry and 3.2 metric tons of nitrogen to receiving waters leading to Bishop Harbor.
 The dredging and deposit of slurry at Piney Point continued following structural fortifications to the holding stacks to ensure integrity with additional loadings.
@@
 Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
-Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations at the time of the discharge event and weather restrictions on boating.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the discharge site as indicated by the TBCOM tracer model simulated plume patterns.
+Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations during the event and weather restrictions on boating.
+As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.
 
@@
 brevis* were obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
-Because of the increased occurrence of red tide in July following the emergency discharge, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
+Because of the increased occurrence of red tide in July following the emergency release, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
 Petersberg) impacted by the event.
@@
 Each year, the TBEP coordinates inter-agency sampling among regional partners at over sixty fixed locations throughout the bay [@Sherwood17].
-Because of the time-sensitive nature of the potential impacts of discharge from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified using a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
+Because of the time-sensitive nature of the potential impacts of pollutants from Piney Point on seagrasses near Piney Point, the sampling protocol used at the routine monitoring locations was modified using a "rapid survey" design to sample seagrasses and macroalgae along 50 m transects at several of the long-term monitoring sites, as well as new locations selected along the shore and small subembayments (e.g., Bishop Harbor) to provide a more comprehensive coverage of the seagrass community near Piney Point (Figure 1b).
 This rapid survey design was created by the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" citizen science group working in Charlotte Harbor, Florida, coordinated in part by the University of Florida, Institute of Food and Agricultural Sciences extension program and Florida Sea Grant.
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the discharge from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the discharge from Piney Point for (a) total ammonia nitogen (mg/L), (b) total phosphorus (mg/L), (c) orthophosphate (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
+```{r wqtrndsadd-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total ammonia nitogen (mg/L), (b) total phosphorus (mg/L), (c) orthophosphate (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 ```{r wqgamadd-supp, fig.cap = 'Forecasted 2021 (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
-Forecasts from the historical models for dates during and after the Piney Point discharge are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Forecasts from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
 Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
@@
 Points are offset slightly for readability.
-No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from model simulations.'}
+No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the plume from model simulations.'}
 knitr::include_graphics(here('figs/trnabu-supp.jpeg'))

### 2021-11-19 manu-draft.Rmd
Corrected SS email
Minor edits
Just a quick look at the abstract, will continue later.

I have a concern about the abstract in that it doesn't discuss the responses of seagrass or macro-algae, which didn't seem to be abnormal (seagrass) or too abnormal (macro). As written the abstract states that everything had abnormal results.

I'll try to make the change wherever I see it, but 'spp.' (as in *Dapis* spp.) should not be italicized.

@@
 Phlips (phlips@ufl.edu), Gary E.
-Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolar@tbep.org), Edward T.
+Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolaro@tbep.org), Edward T.
 Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
@@
 From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen load estimates in a matter of days.
+An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen loads in a matter of days.
 In the period immediately after the release, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
-Filamentous cyanobacteria blooms (*Dapis spp.*) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.
+Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.
 Reported fish kills tracked *K.
 brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
-By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses observed after the release from Piney Point are abnormal relative to historic conditions in Tampa Bay.
+By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions in Tampa Bay.
 

### 2021-12-01 manu-draft.Rmd
spp and sp not italicized

@@
 brevis* were observed, with peak concentrations in excess of 1 x 10$^6$ cells/L in lower Tampa Bay.
-Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis spp.*) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
+Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
 High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface.
@@
 Red macroalgae were present in high abundances from April to May.
-Benthic filamentous macroalgal cyanobacteria (*Dapis spp.*) was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
+Benthic filamentous macroalgal cyanobacteria (*Dapis* spp.) was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
 Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.
@@
 Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis sp.* and *Skeletonema sp.*, Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
+This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis* sp.
+and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
@@
 Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
-In particular, increasing abundance in recent years of the green algae *Caulerpa sp.* has been observed at long-term transects that were previously dominated by seagrass.
+In particular, increasing abundance in recent years of the green algae *Caulerpa* sp.
+has been observed at long-term transects that were previously dominated by seagrass.
 These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
@@
 This was expected given both the documented changes from past releases from Piney Point [@Switzer11] and the more rapid response of macroalgae to changing water quality conditions relative to seagrasses [@Valiela97].
-In Tampa Bay, red macroalgae groups (e.g., *Gracilaria spp.*, *Acanthophora sp.*) are more common than green macroalgae (e.g., *Ulva spp.*, *Caulerpa spp.*) and occur earlier in the growing season.
-The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva spp.* may reflect a natural phenology in Tampa Bay.
-The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis spp.*) in May and June.
-Unusually high abundances of *Dapis spp.* were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site.
+In Tampa Bay, red macroalgae groups (e.g., *Gracilaria* spp., *Acanthophora* sp.) are more common than green macroalgae (e.g., *Ulva* spp., *Caulerpa* spp.) and occur earlier in the growing season.
+The dominance of the red groups early in the summer followed by an increase in the green alga *Ulva* spp.
+may reflect a natural phenology in Tampa Bay.
+The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis* spp.) in May and June.
+Unusually high abundances of *Dapis* spp.
+were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site.
 Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those in potential response to the Piney Point release.
@@
 ```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the release from Piney Point.
-Inset image shows blooms of cyanobacteria macroalgae (*Dapis spp.*).'}
+Inset image shows blooms of cyanobacteria macroalgae (*Dapis* spp.).'}
 knitr::include_graphics(here('figs/timeline.jpg'))
@@
 Sample dates in 2021 are shown in rows with transect meter results shown in columns (0m nearshore, 50m offshore).
-Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis spp.* (cyanobacteria) peaking in June and green macroalgae (*Ulva spp.*) increasing in July.
+Results show dominance of manatee grass (*Syringodium filiforme*) and red macroalgae groups, with abundances of *Dapis* spp.
+(cyanobacteria) peaking in June and green macroalgae (*Ulva* spp.) increasing in July.
 Abundances are Braun-Blanquet coverage estimates.'}
@@
 Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurence estimates across all dates.
-Diatom concentrations are based on combined cell counts from *Asterionellopsis sp.* and *Skeletonema sp*.'}
+Diatom concentrations are based on combined cell counts from *Asterionellopsis* sp.
+and *Skeletonema sp*.'}
 knitr::include_graphics('figs/nutrientflow.jpeg')

### 2021-12-01 supplement.Rmd
spp and sp not italicized

@@
 Another event lasting from November 2003 to October 2004, treated process water from Piney Point was released to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
-@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva spp.* and *Gracilaria spp.* were observed as a potential indication of nutrient eutrophication.
+@Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva* spp.
+and *Gracilaria* spp.
+were observed as a potential indication of nutrient eutrophication.
 Around the same time, 939 million liters of water from Piney Point were barged and released 193 kilometers offshore to the Gulf of Mexico to reduce strain on the holding capacity of storage ponds [@Hu03].
@@
 Dominant seagrass species in the bay include *Halodule wrightii*, *Syringodium filiforme*, and *Thalassia testudinum*.
-Other seagrass species (i.e., *Halophila spp.*, *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
+Other seagrass species (i.e., *Halophila* spp., *Ruppia maritima*) were also observed but were present at much lower abundances and were not evaluated herein.
 Macroalgae taxa were aggregated by major group (i.e., red, green, and cyanobacteria).

### 2021-12-01 manu-draft.Rmd
update figs and tabs with package updates

@@
 Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
-Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Area 1 and for TN and chl-a for Areas 2 and 3 (Table \@ref(tab:wqcmptab)).
+Kruskal-Wallis tests that assessed if at least one of the months had significantly different observations for each parameter were significant (*p* < 0.05) for TN, chl-a, and Secchi depth for Areas 1 and 3 and for TN and chl-a for Area 2 (Table \@ref(tab:wqcmptab)).
 Further analysis with multiple comparison tests generally showed that April/May were different from June/July depending on Area and parameter, such that observations in the later months were generally higher (or lower for Secchi) corresponding to increasing *K.

### 2021-12-01 R/figures.R

@@ -74,8 +74,7 @@ areas <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>%
   mutate(
     area = factor(area)
   ) %>% 
@@ -86,9 +85,7 @@ areas <- ppseg %>%
     acres = set_units(acres, 'acres'), 
     acres = as.numeric(acres)
   ) %>% 
-  dplyr::filter(acres > 1e4) %>% 
-  st_buffer(dist = set_units(0.0005, degree)) %>% 
-  st_buffer(dist = set_units(-0.0005, degree)) 
+  dplyr::filter(acres > 1e4)
 
 cols <- c("#E16A86", "#50A315", "#009ADE")
 names(cols) <- levels(areas$area)
@@ -378,8 +375,7 @@ ppsegbf <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )
@@ -458,8 +454,7 @@ areas <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )
@@ -1171,8 +1166,7 @@ ppsegbf <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )
@@ -1205,8 +1199,7 @@ ppsegbf <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
[…]

### 2021-12-01 R/funcs.R

@@ -141,7 +141,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
   nonbay <- c('BH01', 'P Port 2', 'P Port 3', 'PM Out', '20120409-01', 'PPC41', 'P Port 4', 'PMB01', 'NGS-S Pond')
   mindt <- ymd('2021-03-28')
   dtrng <- ymd(c('2006-01-01', '2021-10-01'))
-  
+
   # monitoring data
   rswqtmp <- rswqdat %>% 
     filter(var %in% vr) %>% 
@@ -232,7 +232,7 @@ gamplo_fun <- function(rswqdat, bswqdat, ppseg, vr, cols, logtr = TRUE, rmfacet 
             ),
             xvals = ymd(paste('2021', month(date), day(date), sep = '-'))
           )
-        
+
         return(out)
         
       })

### 2021-12-01 R/tables.R

@@ -119,8 +119,7 @@ ppsegbf <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )
@@ -205,8 +204,7 @@ ppsegbf <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>%  
   mutate(
     area = factor(area)
   )
@@ -361,8 +359,7 @@ areas <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )
@@ -527,8 +524,7 @@ areas <- ppseg %>%
   rename(area = Name) %>% 
   group_by(area) %>% 
   summarise() %>% 
-  st_buffer(dist = set_units(0.0001, degree)) %>% 
-  st_buffer(dist = set_units(-0.0001, degree)) %>% 
+  st_make_valid() %>% 
   mutate(
     area = factor(area)
   )

### 2021-12-01 manu-draft.Rmd
abstract shortening, adding seagrass statement

@@
 
-Aging infrastructure and climate change stressors acting on wastewater containment facilities are global concerns for the environmental protection of surface waters.
 From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 186 metric tons of total nitrogen were exported to Lower Tampa Bay, exceeding typical annual external nitrogen loads in a matter of days.
-In the period immediately after the release, elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
-Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis* that peaked in July.
+An estimated 186 metric tons of total nitrogen were exported to the Bay, exceeding typical annual external nitrogen loads in a matter of days.
+Elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
+Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis*.
 Reported fish kills tracked *K.
-brevis* concentrations, prompting local cleanup efforts to remove over 1600 metric tons of dead fish.
-By comparing these observations to decades of baseline environmental monitoring data for the region and typical seasonal conditions experienced in recent years, we conclude that all of these biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions in Tampa Bay.
+brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish.
+Seagrass resources showed minimal changes over the study period.
+By comparing these results to decades of baseline environmental monitoring data, we conclude that many of the biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions.
 

### 2021-12-01 R/figures.R
nelson et al. cite added

@@ -623,8 +623,8 @@ toplo <- habdat %>%
 
 # plot
 p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
-  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + 
-  scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
+  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
+  scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE)), limits = c(1000, NA)) +
   scale_x_discrete(breaks = seq(1950, 2025, by = 5)) + 
   theme_minimal() +
   theme(
@@ -634,7 +634,7 @@ p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
   ) +
   labs(
     x = 'Year',
-    y = 'Cells / L',
+    y = 'Cells / L (log-scale)',
     title = expression(paste('(a) ', italic('K. brevis'), ' Apr - Sep concentrations by year, middle/lower Tampa Bay'))
   )
 
@@ -666,8 +666,8 @@ p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
   #   aes(ymin = y0, lower = y25, middle = y50, upper = y75, ymax = y100),
   #   stat = "identity", width = 0.75, fill = '#00806E'
   # ) +
-  # scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
-  scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
+  scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE)), limits = c(1000, NA)) +
+  # scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
   # geom_hline(aes(yintercept = 1e5, color = 'Bloom\nconcentration')) +
   theme_minimal() +
   theme(
@@ -677,7 +677,7 @@ p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
   ) +
   labs(
     x= 'Week of',
-    y = 'Cells / L',
+    y = 'Cells / L  (log-scale)',
     title = expression(paste('(b) ', italic('K. brevis'), ' concentrations in 2021 by week, middle/lower Tampa Bay'))
   )
 

### 2021-12-01 manu-draft.Rmd
nelson et al. cite added
Rycyk citation, DAP processing, bloom statement moved

@@
 At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.
-Water quality parameters of NGS-S measured in 2019 were well above baseline conditions typical of surface waters in Tampa Bay, particularly for total phosphorus (160 mg/L) and total nitrogen (230 mg/L).
+Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.
+Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well above baseline conditions typical of surface waters in Tampa Bay.
 From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.
@@
 Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples.
+Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
 Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
@@
 Medium bloom concentrations of the red tide species *K.
-brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b, low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L).
+brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
@@
 brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-Prior blooms were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.
+Prior blooms in Tampa Bay were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.
+In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].
 
@@
 Massive fish kills were observed and likely related to low pH of the water released.
-Unlike Piney Point, inorganic nitrogen concentrations of the release were low and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
+Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
 Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).
@@
 Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.
-These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities.
+These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].
 

### 2021-12-01 manu-draft.Rmd
supplement os materials placeholder added

@@
 The mgcv R package [@Wood17] was used to create the GAMs for water quality parameters.
+
+All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity.
+Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository.
+Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data together through an interactive, online application.
+Links and details are provided in supplement.
 

### 2021-12-01 supplement.Rmd
supplement os materials placeholder added

@@
 ```
+
+# Data, source materials, and dashboard
+
+All datasets used in this study are provided online in an open access data archive hosted on the Knowledge Network for Biocomplexity as part of the DataOne federated network.
+This includes full metadata documentation.
+The repository is available and citable at:
+
+Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository.
+This includes a source RMarkdown file [@Xie20] used for the manuscript text.
+The repository is available at https://github.com/tbep-tech/piney-point-manu.
+
+The Piney Point Environmental Monitoring Dashboard can be used to view all data together through an interactive, online application.
+It provides a synthesis of data to assess baseline conditions prior to April 2021 and to evaluate changing water quality conditions as new data become available.
+The dashboard is available at https://shiny.tbep.org and citable at @Beck21c.
 

### 2021-12-02 manu-draft.Rmd
redtide fig colored by seg and breakpoints shown

@@
 *K.
-brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), Fish kill reports are from Florida Fish and Wildlife Conservation Commission, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
+brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), low/medium/high cell concentrations in a/b are from Florida Fish and Wildlife Conservation Commission (FWC), Fish kill reports are from FWC, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
 brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for summed hydrologic loads of major tributaries to Tampa Bay from USGS gaged sites, and wind rose plots are from wind measurements at St.

### 2021-12-02 R/figures.R

@@ -613,33 +613,47 @@ weeklv <- seq.Date(from = as.Date('2021-01-01'), to = Sys.Date(), by = 'days') %
 
 # habdat p1
 toplo <- habdat %>%
+  st_as_sf(coords = c('lng', 'lat'), crs = 4326) %>% 
+  st_intersection(tbseg[tbseg$bay_segment %in% c('MTB', 'LTB'), ]) %>% 
+  st_set_geometry(NULL) %>% 
   filter(date < as.Date('2021-10-01')) %>% 
   filter(month(date) > 3 & month(date) < 10) %>% 
   mutate(
     yr = year(date),
-    yr = factor(yr, levels = seq(min(yr), max(yr)))
+    yr = factor(yr, levels = seq(min(yr), max(yr))), 
+    long_name = factor(long_name, levels = c('Lower Tampa Bay', 'Middle Tampa Bay'))
   ) %>% 
-  complete(yr)
+  complete(yr, long_name)
 
 # plot
-p1 <- ggplot(toplo, aes(x = yr, y = 1 + val)) +
-  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
+p1 <- ggplot(toplo, aes(x = yr, y = 1 + val, group = long_name)) +
+  geom_hline(aes(yintercept = 1e4), linetype = 'dotted') +
+  geom_hline(aes(yintercept = 1e5), linetype = 'dotted') +
+  geom_hline(aes(yintercept = 1e6), linetype = 'dotted') +
+  geom_text(aes(y = 1e4, x = 1), label = 'Low', vjust = 0, hjust = 0, size = 3, nudge_y = 0.1) +
+  geom_text(aes(y = 1e5, x = 1 ), label = 'Medium', vjust = 0, hjust = 0, size = 3, nudge_y = 0.1) +
+  geom_text(aes(y = 1e6, x = 1 ), label = 'High', vjust = 0, hjust = 0, size = 3, nudge_y = 0.1) +
+  geom_point(position = position_jitter(width = 0.1), alpha = 0.9, aes(color = long_name)) +
   scale_y_log10(labels = function(x) as.numeric(format(x, scientific = FALSE)), limits = c(1000, NA)) +
   scale_x_discrete(breaks = seq(1950, 2025, by = 5)) + 
+  scale_color_brewer('Location', palette = 'Pastel1') + 
   theme_minimal() +
   theme(
     axis.text.x = element_text(angle = 45, size = 8, hjust = 1),
[…]

### 2021-12-03 manu-draft.Rmd
edit author list, add KA, correct NW email

@@
 Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W.
-Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
+Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Katherine A.
+Hubbard (katherine.hubbard@myfwc.com), Yonggang Liu (yliu@usf.edu),  Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
 Phlips (phlips@ufl.edu), Gary E.
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolaro@tbep.org), Edward T.
-Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.org), Joe Whalen (jwhalen@tbep.org)"
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.gov), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue

### 2021-12-06 manu-draft.Rmd
nw second revision done

@@
 For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates.
-For censored data (i.e., above or below detection), Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis, whereas all other non-detects that were below laboratory standards of detection were evaluated with methods described below.
+Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis.
+Observations for other parameters that were below laboratory standards of detection were evaluated with methods described below.
 
@@
 Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
-Ammonia concentrations were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
+Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
 Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
@@
 This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K.
-brevis* growth in 2021, in addition to the elevated nutrient concentrations from the Piney Point discharge.
+brevis* growth in 2021 (Figure S2f, S3f), in addition to the elevated nutrient concentrations from the Piney Point discharge.
 
@@
 This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis* sp.
-and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were a majority percentage of the TN in the initial discharge.
+and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
@@
 The co-occurring decline in macroalgae and increase in *K.
-brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point were likely present (as noted by modelling efforts).
+brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).
 Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
@@
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
-From both a monitoring perspective and the nature of the unanticipated release into Tampa Bay, indicators of eutrophication from the 2021 water quality data were comparatively different than recent prior years.
+As anticipated, the 2021 results showed that indicators of eutrophication from the water quality data were comparatively different than recent prior years.
 These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
@@
 
-For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were minimal over the initial study period.
+For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were not observed over the initial study period.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).
-The fate of the phosphorus in Tampa Bay is unknown.
+The fate of the phosphorus in Tampa Bay is less understood than that of nitrogen.
 Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
@@
 ```{r stktab}
-cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and end-of-pipe samples from April 2021 for relevant water quality variables.
+cap.val <- 'Measured concentrations from the phosphogypsum stack (NGS-S) at Piney Point from a 2019 sample and samples from April 2021 for relevant water quality variables.
 Values are compared to normal annual medians (min, max) for concentrations in lower Tampa Bay.
 Normal medians are based on data for a baseline period from 2006 to 2020 from long-term monitoring stations in lower Tampa Bay (Figure \\@ref(fig:map)a).
-End-of-pipe concentrations were taken as averages from two samples on April 6th and 13th at the outflow at Port Manatee.
+The 2021 samples are from the NGS-S stack on April 13th and directly from the outflow site at Port Manatee on April 6th.
 Missing values were not measured in the stack water or release water.'

### 2021-12-06 supplement.Rmd
nw second revision done

@@
 
-Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP in agreement with USEPA standards and those of the inter-agency partners.
-Data quality objectives followed guidelines outlined in the TBEP Data Quality Management Plan [@tbep1620].
+Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.
+Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620].
 Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
@@
 This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
-Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen, mg/L).
-Overall, sample frequency was approximately biweekly over the subsequent months, with some variability given agency resource limitations during the event and weather restrictions on boating.
+Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
+Overall, sample frequency was approximately biweekly over the subsequent months with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.

### 2021-12-06 R/tables.R

@@ -62,7 +62,8 @@ effraw <- tibble(
   var = c('no23', 'nh34', 'tn', 'tp', 'orthop', 'dosat', 'ph', 'chla'),
   lbs = c("Nitrate/Nitrite (mg/L)", "NH3, NH4+ (mg/L)", "TN (mg/L)", 
           "TP (mg/L)", "Ortho-P (mg/L)", "DO (% sat.)", "pH", "Chl-a (ug/L)"),
-  effval = c(0.292, 210, 220, mean(c(140, 161)), mean(c(140, 155)), NA, NA, 105)
+  effval = c(0.004, 210, 220, 140, 140, NA, NA, NA), 
+  stkval2 = c(0.292, NA, NA, 161, 155, NA, NA, 105)
 )
 
 bssum <- bswqdat %>% 
@@ -99,12 +100,14 @@ tab <- full_join(stkraw, bssum, by = 'var') %>%
   full_join(effraw, by = c('var', 'lbs')) %>% 
   mutate(
     stkval = ifelse(is.na(stkval), '-', as.character(stkval)),
-    effval = ifelse(is.na(effval), '-', as.character(effval))
+    effval = ifelse(is.na(effval), '-', as.character(effval)), 
+    stkval2 = ifelse(is.na(stkval2), '-', as.character(stkval2)), 
   ) %>% 
   select(
     `Water quality variable` = lbs, 
-    `2019 stack value` = stkval, 
-    `2021 end-of-pipe value` = effval,
+    `2019 stack value` = stkval,
+    `2021 stack value` = stkval2,
+    `2021 pipe value` = effval,
     `2006 - 2020 lower Tampa Bay median (min, max)` = sumv
   ) 
 

### 2021-12-06 manu-draft.Rmd
cl and kh second revision done

@@
 
-From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into Tampa Bay (Florida, USA).
-An estimated 186 metric tons of total nitrogen were exported to the Bay, exceeding typical annual external nitrogen loads in a matter of days.
+From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA).
+This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days.
 Elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
-Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by blooms of the red tide organism *Karenia brevis*.
+Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*.
 Reported fish kills tracked *K.
@@
 In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b].
-Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change related stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
+Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
 Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled releases to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
@@
 Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Historical gains in seagrass coverage in southwest Florida estuaries (e.g., Tampa Bay, Sarasota Bay, Charlotte Harbor) have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
+Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth and persistence, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
+Historical gains in seagrass coverage in southwest Florida estuaries have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
 Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
-The three contiguous National Estuary Programs of southwest Florida and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.
+The three contiguous National Estuary Programs of southwest Florida (Tampa Bay, Sarasota Bay, Charlotte Harbor) and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.
 
@@
 
-Recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
+More recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
 In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
-At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations was being held in the failing gypstack.
+At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack.
 Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.
@@
 
-This paper provides an initial assessment of environmental conditions in Tampa Bay over six months following the recent release of legacy phosphate mining wastewater from March 30th to April 9th, 2021.
+This paper provides an initial assessment of environmental conditions in Tampa Bay over the six month period after the release of legacy phosphate mining wastewater from the Piney Point facility in 2021.
 The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
 We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
-A timeline of events in 2021 is provided, which is supported by the quantitative results from 2021 response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.
+A timeline of events is provided, which is supported by the quantitative results from 2021 routine and response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.
 The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.
@@
 
-Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
+Monitoring agencies and local partners that collected data using standardized protocols included FDEP, Environmental Protection Commission (EPC) of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute (FWRI) of the Florida Fish and Wildlife Conservation Commission (FWC), City of St.
 Petersburg, Tampa Bay Estuary Program (TBEP), Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (ug/L), pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%).
-Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point.
-Phytoplankton samples included a mix of quantitative (cells/l) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples.
+In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (psu), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
+Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
+
+Phytoplankton samples included a mix of quantitative (cells/L) and qualitative (presence/absence) samples for major taxa at similar frequency and spatial distribution as the water quality samples.
+Harmful Algal Bloom (HAB) data for *Karenia brevis* were obtained from event-based monitoring samples from the [FWC-FWRI HAB Monitoring Database](https://myfwc.com/research/redtide/monitoring/database/).
 Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
+Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
 Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
-Details of the sampling methods are provided in supplement.
+Finally, precipitation data were from Tampa International Airport, inflow estimates to Tampa Bay were basd on summed hydrologic loads of major tributaries from US Geological Survey gaged sites, and wind data were from Albert Whitted Airfield at St.
+Petersburg, Florida.
+Additional details of the sampling methods and data sources are provided in supplement.
 
@@
 
-Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
-These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing programmatic goals).
+Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/, Hillsborough County collected monthly, Manatee County collected quarterly) were used to establish baseline conditions for major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above.
+These areas (Area 1: closest to Piney Point; Area 2: north of Piney Point; Area 3: south of Piney Point including northern Sarasota Bay) were identified based on anticipated impacts from expected plume patterns following the TBCOM simulations and other prominent bay boundaries relative to Piney Point (i.e., the main shipping channel in the bay, inflow boundaries, location of the Skyway Bridge at the mouth of Tampa Bay, and major bay segments used by TBEP for assessing annual water quality targets).
 Observations at each long-term monitoring station were averaged for each month across years from 2006 to 2020.
@@
 Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository.
-Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data together through an interactive, online application.
+Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application.
 Links and details are provided in supplement.
@@
 A general summary of 2021 events in Tampa Bay following discharge from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge ceased on April 9th, a peak in median chl-a concentration was observed near Piney Point (Area 1, Figure \@ref(fig:map)a) mid-April, with peak individual sample values in excess of 50 ug/L.
-Median concentrations for each week in April were less than 10 ug/L.
-The discharge phytoplankton biomass was comprised of over 99% of a spherical nanoplanktonic chlorophyte, with density of 337 x 10$^6$ cells/L.
-The phytoplankton communities near the discharge area in April were generally dominated by diatoms (Bacillariophyta).
+After the discharge ceased on April 9th, a peak in median chl-a concentration was observed near Piney Point (Area 1, Figure \@ref(fig:map)a) in mid-April, with peak individual sample values in excess of 50 $\mu$g/L.
+Median concentrations for each week in April were less than 10 $\mu$g/L.
+The discharge phytoplankton assemblage was comprised of over 99% of a spherical nanoplanktonic chlorophyte (3.37 x 10$^8$ cells/L).
+The phytoplankton communities near the discharge area in April were generally dominated by diatoms.
 The initial diatom bloom did not persist past April.
-On April 20th, the toxic red tide species *Karenia brevis* was first observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay.
+On April 20th, the HAB species *Karenia brevis* was observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay; this first Tampa Bay influx was related to an ongoing coastal bloom.
 By May/June, bloom levels of *K.
-brevis* were observed, with peak concentrations in excess of 1 x 10$^6$ cells/L in lower Tampa Bay.
+brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with peak concentrations in excess of 1 x 10$^6$ cells/L.
 Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
 High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface.
-By June 27th, fish kill reports attributed to red tide increased with *K.
-brevis* cell concentrations in lower and middle Tampa Bay.
+By June 27th, fish kill reports attributed to *K.
+brevis* increased as cellular abundance climbed in lower and middle Tampa Bay.
 The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.
-This shift contributed to an increase in fish kill reports by moving dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St.
-Petersburg and Tampa.
+This shift in winds likely disturbed the water column and altered the spatial distribution of *K.
+brevis* in the bay.
+Strong southeasterly winds also likely moved dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St.
+Petersburg and Tampa, contributing to an increase in fish kill reports.
 Concentrations of *K.
@@
 chlabv <- wqsumfl %>% 
-  filter(`Water quality variable` == 'Chl-a (ug/L)') %>% 
+  filter(`Water quality variable` == 'Chl-a ($\\mu$g/L)') %>% 
   select(Area, `% Above`) %>% 
@@
 Notable differences were also observed for dissolved oxygen (e.g., `r dosabv[['1']]`\% were above in Area 1, `r dosabv[['2']]`\% in Area 2).
-Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal range, although initial time series showed much higher concentrations for ammonia in April near Area 1.
+Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal ranges, although initial time series showed much higher concentrations for ammonia in April near Area 1.
 Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
@@
 
-Medium bloom concentrations of the red tide species *K.
-brevis* in 2021 were first observed in Tampa Bay the week of May 23, with high bloom concentrations peaking by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+Bloom concentrations of the red tide species *K.
+brevis* in 2021 were first observed in Tampa Bay the week of May 23, with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
 brevis* from April to July was an anomaly in 2021 that is not regularly observed in Tampa Bay.
-The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10,000 cells/L, notably 1963, 1971, 2005, 2018, and 2021.
+The historical record from 1953 to present (Figure \@ref(fig:redtide)a) shows cell concentrations sampled in Tampa Bay between April and September, with only a few years having cell concentrations greater than 10$^5$ cells/L, notably 1963, 1971, 2005, 2018, and 2021.
 Median cell concentrations for most years were well below 1,000 cells/L.
-The highest concentration of 20 million cells/L was observed in 1971, whereas the second highest concentration of 17.6 million cells/L was observed in 2021.
+The two highest concentrations were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.
 Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
@@
 
-Fish kill reports attributed to red tide at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations during June and July 2021 (Figure \@ref(fig:redtide)e).
+Fish kill reports attributed to $K.
+brevis$ at the cities of Tampa and Saint Petersburg, FL closely tracked cell concentrations during June and July 2021 (Figure \@ref(fig:redtide)e).
 In total, `r wkstpcnt` reports were made in Saint Petersburg and `r wktmpcnt` in Tampa.
@@
 brevis* cell concentrations (Figure \@ref(fig:redtide)b).
-Notably, all of the fish kill reports occurred within a 1.5 month period during peak *K.
-brevis* cell concentrations, whereas red tide in the bay was observed for nearly four months.
-Increased fish kill reports in early July also coincided with a shift in winds from Tropical Storm Elsa (Figure \@ref(fig:redtide)f, pre-, post-storm wind roses), which moved dead fish closer to populated nearshore areas, as noted earlier.
+Notably, all of the fish kill reports occurred within a 1.5 month period when *K.
+brevis* cell concentrations were consistently above the medium threshold (10$^4$ cells/L).
+The center of Tropical Storm Elsa (Figure \@ref(fig:redtide)f, pre-, post-storm wind roses) also passed through the bay area on July 5th, causing a shift in winds that likely altered the location of *K.
+brevis* cells and dead fish in the bay.
+It is important to note that medium cell concentrations and fish kill reports were observed both before and after storm passage.
 By August, Pinellas County and the city of St.
@@
 Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
-This initial peak in water quality parameters suggested a unique response of the phytoplankton community as an increase in diatoms (e.g., centric species, also *Asterionellopsis* sp.
-and *Skeletonema* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
+The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
-However, macroalgal biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
+However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
 During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
@@
 The co-occurring decline in macroalgae and increase in *K.
-brevis* suggests a release of nutrients from the former that stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).
+brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).
 Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of Cyanobacteria in Anna Maria Sound and near Port Manatee, 3) bloom concentrations of the ride tide organism *K.
+These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K.
 brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
@@
 Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
-Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included organic sources were at high concentrations through April and again peaked in July.
+Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included particulate and dissolved organic sources were at high concentrations through April and again peaked in July.
 Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
 
-There were also concerns that such releases may have contributed to the persistence and intensity of the red tide event that spread throughout lower and middle Tampa Bay.
+There were also concerns that such releases may have contributed to the persistence and intensity of *K.
+brevis* that spread throughout lower and middle Tampa Bay.
 Fisheries resources were impacted by *K.
@@
 For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
-Fish community structure and species composition did not differ compared to a pre-impact period, although bloom concentrations of HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
+Fish community structure and species composition did not differ compared to a pre-impact period, although HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
 brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-Prior blooms in Tampa Bay were more localized in comparison to the 2021 bloom event, potentially allowing fishes to migrate from harmful conditions.
+Prior blooms in Tampa Bay were more localized and *K.
+brevis* was at lower abundances in comparison to the 2021 bloom event, potentially mitigating exposure of fishes to related harmful conditions.
 In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].
@@
 Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern.
-Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from the annual transect monitoring results for Tampa Bay.
+Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.
 In particular, increasing abundance in recent years of the green algae *Caulerpa* sp.
@@
 may reflect a natural phenology in Tampa Bay.
-The most notable change in the macroalgae community was a high abundance of cyanobacteria (i.e., *Dapis* spp.) in May and June.
+The most notable change in the macroalgal community in 2021 was a high abundance of filamentous cyanobacteria (i.e., *Dapis* spp.) in May and June.
 Unusually high abundances of *Dapis* spp.
 were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site.
-Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal changes and those in potential response to the Piney Point release.
-Cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
+Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release.
+Cyanobacteria macroalgae has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
 However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
@@
 Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay for both April and September 2021 benthic samples are not yet available at the time of writing.
-Finally, effects of changing environmental conditions and red tide on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
-Twenty preliminary manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
+Finally, effects of changing environmental conditions and *K.
+brevis* on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
+Twenty manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
 This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
@@
 
-```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters).
+```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters).
 Values outside of the normal range (above for total nitrogen and chlorophyll-a, below for Secchi) are outlined in black and those in normal range are outlined in light grey.
@@
 
-```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) Secchi disk depth (meters) by area based on historal seasonal models.
+```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters) by area based on historical seasonal models.
 Forecasts from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
@@
 Estimates are grouped by sample months in 2021.
-Frequency occurrences are absolute for each taxa based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.
+Frequency occurrences are absolute for each taxon based on presence/absence, whereas the total frequency occurrence applies to any taxa observed on each transect.
 Points are offset slightly for readability.
@@
 ```{r redtide, fig.cap = '*Karenia brevis* concentrations (cells/L) (a) by year and (b) by week in 2021, (c) cumulative precipitation in 2021 compared to past years, (d) cumulative inflow in 2021 compared to past years, (e) fish kill reports in 2021, and (f) wind rose plots for 2021 with notable breaks before/after Piney Point release and tropical storm Elsa.
-*K.
-brevis* cell counts are from NOAA Harmful Algal BloomS Observing System (HABSOS, https://www.ncei.noaa.gov/maps/habsos), low/medium/high cell concentrations in a/b are from Florida Fish and Wildlife Conservation Commission (FWC), Fish kill reports are from FWC, Fish and Wildlife Research Insitute Fish Kill Database, attributed to *K.
-brevis* (https://public.myfwc.com/FWRI/FishKillReport/), precipitation data are from Tampa International Airport, inflow estimates are for summed hydrologic loads of major tributaries to Tampa Bay from USGS gaged sites, and wind rose plots are from wind measurements at St.
-Petersburg, Florida.
 Wind roses show relative counts of six minute observations in directional (30 degree bins, north is vertical) and speed (m/s) categories.'}
@@
 brevis* peaking in early July.
-Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurence estimates across all dates.
+Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurrence estimates across all dates.
 Diatom concentrations are based on combined cell counts from *Asterionellopsis* sp.
@@
 cap.val <- 'Comparison of total nitrogen, chlorophyll-a, and Secchi depth by areas of interest (Figure \\@ref(fig:map)a) and month.
-Overall signifance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Overall significance of differences of concentrations between months for each water quality variable and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
 Multiple comparisons with Mann-Whitney U tests (Comp.

### 2021-12-06 supplement.Rmd
cl and kh second revision done

@@
 For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a (ug/L) concentrations.
+This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a ($\mu$g/L) concentrations.
 Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
@@
 Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples from light microscopy that enumerated major taxa by cell concentrations and qualitative presence/absence samples.
-Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms (Bacillariophyta and other centric taxa), as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
+Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms, as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
 Evaluation herein of HABs data included specific focus on the red tide organism *Karenia brevis*.
-Data for *K.
-brevis* were obtained from event-based monitoring samples collected by FWC and available from a joint FWC-USF short-term red tide tracking product (http://ocgweb.marine.usf.edu/hab_tracking/wfcom_hab.html) and from the Harmful Algal BloomS Observing System ([HABSOS](https://habsos.noaa.gov/)).
 Because of the increased occurrence of red tide in July following the emergency release, fish kill reports from FWC were also evaluated in relation to key municipalities (Tampa, St.
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a (ug/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 cap.val <- 'Comparison of macroalgae frequency occurrence by areas of interest (Figure 1a) and month.
-Overall signifance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Overall significance of differences of frequency occurrence between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
 Multiple comparisons with Mann-Whitney U tests (Comp.
@@
 cap.val <- 'Comparison of seagrass species frequency occurrence by areas of interest (Figure 1a) and month.
-Overall signifance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Overall significance of differences of frequency occurrence between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
 Multiple comparisons with Mann-Whitney U tests (Comp.
@@
 cap.val <- 'Comparison of macroalgae Braun-Blanquet abundances by areas of interest (Figure 1a) and month.
-Overall signifance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Overall significance of differences of abundances between months for macroalgae groups and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
 Multiple comparisons with Mann-Whitney U tests (Comp.
@@
 cap.val <- 'Comparison of seagrass species Braun-Blanquet abundances by areas of interest (Figure 1a) and month.
-Overall signifance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
+Overall significance of differences of abundances between months for seagrass species and area combination are shown with Chi-squared statistics based on Kruskall-Wallis rank sum tests.
 Multiple comparisons with Mann-Whitney U tests (Comp.

### 2021-12-06 R/figures.R

@@ -387,7 +387,7 @@ datin <- rswqdat %>%
   filter(!(var == 'secchi' & val >= 9.5)) # outlier secchi
 
 p1 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = T, rmfacet = T, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
-p2 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = T, rmfacet = T, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
+p2 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = T, rmfacet = T, ttl = '(b) Chlorophyll-a', ylb = expression(paste(mu, 'g/L (log-scale)')))
 p3 <- gamplo_fun(datin, bswqdat, ppsegbf, vr = 'secchi', cols, logtr = F, rmfacet = F, ttl = '(c) Secchi', ylb = 'meters')
 
 p <- (p1 + p2 + p3 + plot_layout(ncol = 3)) / wrap_elements(grid::textGrob('Day of year', gp = gpar(fontsize=14))) + 
@@ -1204,7 +1204,7 @@ datin <- rswqdat %>%
   filter(!(var == 'secchi' & val >= 9.5)) # outlier secchi
 
 p1 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'tn', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(a) Total Nitrogen', ylb = 'mg/L (log-scale)')
-p2 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = 'ug/L (log-scale)')
+p2 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'chla', cols, logtr = TRUE, rmfacet = TRUE, ttl = '(b) Chlorophyll-a', ylb = expression(paste(mu, 'g/L (log-scale)')))
 p3 <- wqplo_fun(datin, bswqdat, ppsegbf, vr = 'secchi', cols, logtr = FALSE, ttl = '(c) Secchi', ylb = 'meters')
 
[…]

### 2021-12-06 R/tables.R

@@ -52,7 +52,7 @@ stkraw <- read.csv(here('data-raw/stack_samples.csv')) %>%
   bind_rows(
     tibble(
       var = 'chla', 
-      lbs = c('Chl-a (ug/L)'), 
+      lbs = c('Chl-a ($\\mu$g/L)'), 
       stkval = NA
     )
   )
@@ -61,7 +61,7 @@ stkraw <- read.csv(here('data-raw/stack_samples.csv')) %>%
 effraw <- tibble(
   var = c('no23', 'nh34', 'tn', 'tp', 'orthop', 'dosat', 'ph', 'chla'),
   lbs = c("Nitrate/Nitrite (mg/L)", "NH3, NH4+ (mg/L)", "TN (mg/L)", 
-          "TP (mg/L)", "Ortho-P (mg/L)", "DO (% sat.)", "pH", "Chl-a (ug/L)"),
+          "TP (mg/L)", "Ortho-P (mg/L)", "DO (% sat.)", "pH", "Chl-a ($\\mu$g/L)"),
   effval = c(0.004, 210, 220, 140, 140, NA, NA, NA), 
   stkval2 = c(0.292, NA, NA, 161, 155, NA, NA, 105)
 )
@@ -182,7 +182,11 @@ totab2 <- totab %>%
     `% In range` = round(100 * `in range` / cnt, 0),
     `% Above` = round(100 * `above` / cnt, 0),
     `% Below` = round(100 * `below` / cnt, 0), 
-    `% Outside detection` = round(100 * nondetect / cnt, 0)
+    `% Outside detection` = round(100 * nondetect / cnt, 0), 
+    lbs = case_when(
+      lbs == 'Chl-a (ug/L)' ~ 'Chl-a ($\\mu$g/L)', 
+      T ~ lbs
+    )
   ) %>% 
   select(
     Area = area, 
@@ -324,7 +328,7 @@ cmps <- rswqsub %>%
   unnest('ests') %>% 
   ungroup() %>% 
   mutate(
-    lbs = factor(lbs, levels = c('TN (mg/L)', 'Chl-a (ug/L)', 'Secchi (m)'))
+    lbs = factor(lbs, levels = c('TN (mg/L)', 'Chl-a (ug/L)', 'Secchi (m)'), labels = c('TN (mg/L)', 'Chl-a ($\\mu$g/L)', 'Secchi (m)'))
   ) %>% 
   arrange(area, lbs, mo) %>% 
   group_by(area) %>%  

### 2021-12-06 manu-draft.Rmd
DC second edits partially done

@@
 Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading effects of nitrogen, phytoplankton growth and persistence, water clarity, light limitation, and epiphytic growth that can eliminate seagrass from nearshore environments [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
+Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading negative effects of nitrogen, phytoplankton growth and persistence, water clarity, light limitation, and epiphyte loading on seagrass growth and survival [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
 Historical gains in seagrass coverage in southwest Florida estuaries have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
@@
 Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments.
-The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility, @Henderson04].
+The Piney Point facility located in Palmetto, Florida is a large, remnant gypstack with three holding ponds located 3 kilometers from the shore of Tampa Bay and near two Florida Aquatic Preserves [see supplement for a history of the facility; @Henderson04].
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
@@
 In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
-At that time, approximately 1.8 billion liters of mix legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack.
+At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack.
 Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.
-Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well above baseline conditions typical of surface waters in Tampa Bay.
+Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well outside normal ranges typical of surface waters in Tampa Bay.
 From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.
@@
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (psu), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
+In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
 Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
@@
 GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
-The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].
+The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods as @Murphy19].
 The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
@@
 
-Statistical assessments were evaluated only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
+Statistical assessments were conducted only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
 Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.
@@
 Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
-Methods to accommodate measured concentrations of water quality variables that were below detection were used that included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
+Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
 
@@
 
-Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).
+Water quality conditions in the northern gypstack measured in 2019 and measured directly at the point of discharge in 2021 showed concentrations that were generally much higher for key water quality parameters as compared to baseline conditions in Tampa Bay (Table \@ref(tab:stktab)).
 Notably, total ammonia nitrogen was measured at 210 mg/L at Piney Point and in the discharge, compared to a long-term median of 0.02 mg/L in lower Tampa Bay.
@@
 Common taxa in the red group included genera *Gracilaria* and *Acanthophora*.
-Green and cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
-Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genera *Dapis*.
+Green macroalgae and filamentous cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
+Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genus *Dapis*.
 Brown macroalgae (primarily in the genus *Feldmannia*) were only observed at one transect in April (`r fomobrn`\% frequency occurrence).
@@
 The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf.
-There is no historical macroalgae record for Tampa Bay that is comparable to the resolution of the 2021 samples.
+There is no historical macroalgae record for Tampa Bay that is comparable to the spatial and temporal resolution of the 2021 samples.
 
@@
 Red macroalgae were present in high abundances from April to May.
-Benthic filamentous macroalgal cyanobacteria (*Dapis* spp.) was first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
+Benthic filamentous cyanobacteria (*Dapis* spp.) mats were first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
 Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.
@@
 Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (see supplement).
-Tests using Braun Blanquet abundances also produced similar results (see supplement).
+Tests using Braun Blanquet cover estimates also produced similar results (see supplement).
 
@@
 The water quality conditions observed during the study period, particularly for TN, chl-a, and Secchi depth, were outside of normal seasonal ranges for many of the observations (Figures \@ref(fig:wqmap), Table \@ref(tab:wqsumtab)).
-These conditions were widespread in lower and middle Tampa Bay.
-
-In addition to the changing environmental conditions observed during the summer, the Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.
+The Piney Point event also represented an anomalous volume and load of labile nitrogen released directly into lower Tampa Bay.
 Spill events [reported to FDEP](https://prodenv.dep.state.fl.us/DepPNP/reports/viewIncidentDetails) (e.g., industrial spills, service line failures, sanitary sewer overflows) provide additional context for Piney Point relative to other potential anomalous releases to Tampa Bay.
-An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point.
+An assessment of over 800 reports to FDEP for the Tampa Bay watershed over the last five years showed spill volumes for these events are small (median volume 13.7 thousand liters [TBEP unpublished analysis](https://tbep-tech.github.io/piney-point-analysis/spills)) compared to the 814 million liters released from Piney Point.
 Moreover, the estimated nutrient load of 186 metric tons of nitrogen to Tampa Bay from Piney Point over the ten day period, exceeded current annual estimates of all external loading sources into lower Tampa Bay [@tbep0417].
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
-As anticipated, the 2021 results showed that indicators of eutrophication from the water quality data were comparatively different than recent prior years.
-These events were interpreted in the context of broad goals of the Tampa Bay Estuary Program and its partners to collectively manage ecosystem health following a nutrient management paradigm with seagrasses as both an indicator of environmental quality and an asset for biological resources in Tampa Bay.
 
-Several of the results of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
-Regarding the nutrient management paradigm for Tampa Bay, the relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
+Several of the water quality responses are consistent with observations of nutrient loading in other shallow Gulf Coast estuaries [@Doering06;@Caffrey14;@Greening14].
+The relationship between nutrients, chl-a, and water transparency followed expectations of reduced water quality with increased nutrient loads.
 Temporally, these changes were observed at different times and for different species of phytoplankton.
@@
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
-Concerns of the negative impacts of nutrients focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
-Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included particulate and dissolved organic sources were at high concentrations through April and again peaked in July.
+Management concerns of the negative impacts of nutrients on water quality focused primarily on the high concentrations of ammonia in the discharge (Table \@ref(tab:stktab)), which can be utilized rapidly by many phytoplankton taxa [@Bates76;@Domingues11].
+Low concentrations of ammonia after April may be explained by quick uptake by the initial diatom bloom, where TN that included particulate and dissolved organic sources was at high concentrations through April and again peaked in July.
 Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
 
-There were also concerns that such releases may have contributed to the persistence and intensity of *K.
-brevis* that spread throughout lower and middle Tampa Bay.
-Fisheries resources were impacted by *K.
-brevis* blooms in June and July (Figure \@ref(fig:redtide)).
-Quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
+There were also concerns that the release from Piney may have contributed to the persistence and intensity of *K.
+brevis*, having negative effects on fisheries resources in June and July (Figure \@ref(fig:redtide)).
+In addition to fish kill reports, quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
 Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.

### 2021-12-06 supplement.Rmd
DC second edits partially done

@@
 This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a ($\mu$g/L) concentrations.
-Samples for pH, salinity (psu), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
+Samples for pH, salinity (ppt), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
 Overall, sample frequency was approximately biweekly over the subsequent months with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.

### 2021-12-07 manu-draft.Rmd
added knb links

@@
 The mgcv R package [@Wood17] was used to create the GAMs for water quality parameters.
-
-All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity.
+All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity [@Beck21d].
 Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository.
-Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application.
+Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application [@Beck21c].
 Links and details are provided in supplement.

### 2021-12-07 supplement.Rmd
added knb links

@@
 This includes full metadata documentation.
-The repository is available and citable at:
+The repository is available at https://knb.ecoinformatics.org/view/doi:10.5063/F1959G05 and citable as @Beck21d.
 
@@
 The Piney Point Environmental Monitoring Dashboard can be used to view all data together through an interactive, online application.
-It provides a synthesis of data to assess baseline conditions prior to April 2021 and to evaluate changing water quality conditions as new data become available.
-The dashboard is available at https://shiny.tbep.org and citable at @Beck21c.
+It provides a synthesis of data to assess baseline conditions prior to April 2021 and all 2021 monitoring data collected in response to the Piney Point event.
+The dashboard is available at https://shiny.tbep.org and citable as @Beck21c.
 

### 2021-12-08 manu-draft.Rmd
DC second revision done

@@
 brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with peak concentrations in excess of 1 x 10$^6$ cells/L.
-Also during May/June, high abundances of benthic filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
+Also during May/June, high abundances of filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
 High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface.
@@
 Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal ranges, although initial time series showed much higher concentrations for ammonia in April near Area 1.
-Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (see supplement, similar to the discharge measurements in Table \@ref(tab:stktab)).
+Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (Figures S2, S3), similar to the discharge measurements in Table \@ref(tab:stktab)).
 Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
@@
 Common taxa in the red group included genera *Gracilaria* and *Acanthophora*.
-Green macroalgae and filamentous cyanobacteria macroalgae were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
+Green macroalgae and filamentous cyanobacteria were less common, with frequency occurrences of `r foall['Green']`\% and `r foall['Cyanobacteria']`\%.
 Common taxa in the green group included genera *Ulva* and *Caulerpa*, whereas cyanobacteria biomass was dominated by the benthic filamentous genus *Dapis*.
@@
 Manatee grass (*Syringodium filiforme*) and shoal grass (*Halodule wrightii*) had similar coverage across all transects, with frequency occurrences of `r foall['Syringodium filiforme']`\% and `r foall['Halodule wrightii']`\%, respectively.
-The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17], with turtle grass being the dominant species in more euhaline waters closer to the Gulf.
+The frequency occurrences of seagrasses near Piney Point were similar to the long-term record of seagrass transect data available for Tampa Bay [@Sherwood17, also see https://shiny.tbep.org/seagrasstransect-dash], with turtle grass being the dominant species in more euhaline waters closer to the Gulf.
 There is no historical macroalgae record for Tampa Bay that is comparable to the spatial and temporal resolution of the 2021 samples.
@@
 Red macroalgae were present in high abundances from April to May.
-Benthic filamentous cyanobacteria (*Dapis* spp.) mats were first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
+Filamentous cyanobacteria (*Dapis* spp.) mats were first observed on May 24th and was present at all of the sample locations along this transect on June 4th and 15th.
 Filamentous cyanobacteria persisted through June and July, but was not observed in abundance after July 20th.
@@
 Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
-Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (see supplement).
-Tests using Braun Blanquet cover estimates also produced similar results (see supplement).
+Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (Tables S1, S2).
+Tests using Braun Blanquet cover estimates also produced similar results (Tables S3, S4).
 
@@
 brevis* cells and dead fish in the bay.
-It is important to note that medium cell concentrations and fish kill reports were observed both before and after storm passage.
+It is important to note that high cell concentrations (>10$^6$ cells/L) were observed in middle Tampa Bay (Figure \@ref(fig:redtide)b) and fish kills were reported both before and after storm passage (Figure \@ref(fig:redtide)e).
 By August, Pinellas County and the city of St.
@@
 Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
-These distinct temporal periods were readily identified through an ordination plot (see supplement) for the observed data in Figure \@ref(fig:nutrientflow).
+These distinct temporal periods were readily identified through an ordination plot (Figure S7) for the observed data in Figure \@ref(fig:nutrientflow).
 
@@
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
-brevis* bloom (see supplement).
+brevis* bloom (Figures S2, S3).
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.
@@
 brevis* bloom in 2018.
-Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern.
+Although the 2021 red tide in Tampa Bay was short-lived, potential long-term effects on seagrasses remain a concern [e.g., alteration of sediment geochemistry, @Eldridge04].
 Ecosystem shifts from seagrass to macroalgae dominated communities are also a concern, both in 2021 and as observed at some locations in recent years from annual transect monitoring results for Tampa Bay.
@@
 These changes may be indicative of broader ecosystem shifts concurrent with alteration of nutrient loads or system resilience at the expense of seagrass communities [@Lloret05;@Stafford06].
-Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges in the context of broader issues affecting ecosystem health.
+Acute stressors from short-term events, such as unanticipated releases from Piney Point, create additional and often preventable challenges to managing seagrass health.
 
@@
 The most notable change in the macroalgal community in 2021 was a high abundance of filamentous cyanobacteria (i.e., *Dapis* spp.) in May and June.
-Unusually high abundances of *Dapis* spp.
-were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site.
+High abundances of *Dapis* spp.
+were observed in Anna Maria Sound near the mouth of Tampa Bay and near Port Manatee at the release site, which is uncommon at these locations.
 Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release.
-Cyanobacteria macroalgae has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
+Filamentous cyanobacteria has been observed during routine annual transect monitoring in Tampa Bay and it has previously been documented in public reports to the Florida Department of Environmental Protection.
 However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
@@
 Additional monitoring and analysis are also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments sampled in April near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106, see https://shiny.tbep.org/piney-point], possibly reflecting the high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
-Comparison of the April samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Sediment contaminant results for a suite of metals and chemicals as well as percent silt/clay for both April and September 2021 benthic samples are not yet available at the time of writing.
+For benthic communities, sediments sampled in April and September near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106, see https://shiny.tbep.org/piney-point], possibly reflecting the generally high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
+Comparison of the April and September samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([TBEP unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
+Differences between the April and September samples were not observed.
 Finally, effects of changing environmental conditions and *K.
-brevis* on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources within the broader distribution of these species.
+brevis* on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources.
 Twenty manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
-This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that is likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
+This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that are likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
 
-In the broader context of mining impacts to surface waters, the results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.
-In addition to the previous studies and reports for Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.
+In the broader context of mining impacts to surface waters, these results reinforce the understanding that legacy pollutants from phosphate mining can negatively affect environmental resources.
+In addition to Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.
 Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].
@@
 Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (see supplement).
+Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (Figures S2, S3).
 The fate of the phosphorus in Tampa Bay is less understood than that of nitrogen.
@@
 ```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the release from Piney Point.
-Inset image shows blooms of cyanobacteria macroalgae (*Dapis* spp.).'}
+Inset image shows blooms of filamentous cyanobacteria (*Dapis* spp.).'}
 knitr::include_graphics(here('figs/timeline.jpg'))
@@
 
-```{r wqgam, fig.cap = 'Forecasted 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters) by area based on historical seasonal models.
-Forecasts from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
+```{r wqgam, fig.cap = 'Expected 2021 (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters) by area based on historical seasonal models.
+Predictions (expected values) from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Expected values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure \\@ref(fig:map)a.'}
@@
 
-```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) cyanobacteria macroalgal abundances, and (e) *Karenia brevis* cell concentrations.
+```{r nutrientflow, fig.cap = 'Weekly summarized observations (medians, 2.5th to 97.5th percentiles) across all sampled locations for (a) total nitrogen concentrations, (b) chlorophyll-a concentrations, (c) diatom cell concentrations, (d) filamentous cyanobacteria abundances, and (e) *Karenia brevis* cell concentrations.
 Values are summarized for all samples within each week.
-The values suggest nutrient cycling between water column phytoplankton in the initial April diatom bloom, then to cyanobacteria macroalgae in May to June, and then to *K.
+The values suggest nutrient cycling between water column phytoplankton in the initial April diatom bloom, then to filamentous cyanobacteria in May to June, and then to *K.
 brevis* peaking in early July.
-Quantitative cell counts for diatoms are missing for several weeks, but see supplementary figures for frequency occurrence estimates across all dates.
+Quantitative cell counts for diatoms are missing for several weeks, but see Figure S6 for frequency occurrence estimates across all dates.
 Diatom concentrations are based on combined cell counts from *Asterionellopsis* sp.
@@
 ```{r wqsumtab}
-cap.val <- 'Summary of water quality variables collected in Tampa Bay from April through September 2021 following the release water from Piney Point.
+cap.val <- 'Summary of water quality variables collected in Tampa Bay from April through September 2021 following the release of water from Piney Point.
 Variables are grouped by major areas of interest for evaluating status and trends shown in Figure \\@ref(fig:map)a.

### 2021-12-08 supplement.Rmd
DC second revision done

@@
 Species of phytoplankton associated with harmful algal blooms were observed around this time [@Garrett11].
-Another event lasting from November 2003 to October 2004, treated process water from Piney Point was released to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
+During another event lasting from November 2003 to October 2004, treated process water from Piney Point was released to Bishop Harbor to further reduce the likelihood of an uncontrolled spill.
 @Switzer11 reported minimal impacts to nekton communities, although macroalgal blooms of *Ulva* spp.
@@
 Discrete water quality samples were taken primarily from surface grabs by boat and processed for relevant parameters (see below) by the respective laboratories of each participating agency.
-Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, secchi depth, etc.).
-In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts depending on monitoring agency.
+Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).
+In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.
 Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.
@@
 Samples for pH, salinity (ppt), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
-Overall, sample frequency was approximately biweekly over the subsequent months with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.
+Overall, sample frequency was approximately biweekly over the months following the Piney Point release, with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.
 As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.
@@
 Phytoplankton samples were also collected by multiple partners and included a mix of quantitative samples from light microscopy that enumerated major taxa by cell concentrations and qualitative presence/absence samples.
-Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms, as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs), as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
+Taxa were aggregated into major groups of interest for Tampa Bay, with a focus on diatoms as common primary producers observed throughout the growing season, and species associated with harmful algal blooms (HABs) as a potentially adverse outcome of these species outcompeting others in response to nutrient inputs from Piney Point.
 Evaluation herein of HABs data included specific focus on the red tide organism *Karenia brevis*.
@@
 
-```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) secchi disk depth (meters).
+```{r wqtrnds-supp, fig.cap = 'Sampled water quality data by week for late March through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters).
 Observations are aggregated by week and within assessment areas shown in Figure 1a.
@@
 
-```{r wqgamadd-supp, fig.cap = 'Forecasted 2021 (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
-Forecasts from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
-Forecasted values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
+```{r wqgamadd-supp, fig.cap = 'Expected 2021 (a) total ammonia nitogen (mg/L), (b) orthophosphate (mg/L), (c) total phosphorus (mg/L), (d) dissolved oxygen saturation (%), (e) turbidity (NTU), and (f) salinity (ppt) by area based on historal seasonal models.
+Predictions (expected values) from the historical models for dates during and after the Piney Point release are shown in thick lines (+/- 95% confidence), with observed samples overlaid on the plots to emphasize deviation of 2021 data from historical seasonal estimates.
+Expected values are based on Generalized Additive Models fit to historical baseline data from 2006 to early 2021, where historical predictions are shown as thin grey lines, with darker lines for more recent years.
 Results are grouped by assessment areas shown in Figure 1a.'}
@@
 Sample dates are noted by vertical grey lines in each plot.
-Diatoms are based on presence/absense of *Asterionellopsis sp*.
+Diatoms are based on presence/absence of *Asterionellopsis sp*.
 and *Skeletonema sp*.
@@
 
-```{r nutrientfloword-supp, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and cyanobacteria macroalgal abundances.
+```{r nutrientfloword-supp, fig.cap = 'Ordination results comparing weekly summarized observations across all sampled locations for total nitrogen concentrations, chlorophyll-a concentrations, diatom cell concentrations, *Karenia brevis* cell concentrations, and filamentous cyanobacteria abundances.
 Ordination results are from non-metric multi-dimensional scaling performed on the 97.5th percentile values of observations in each week for each parameter.

### 2021-12-08 R/figures.R

@@ -1056,7 +1056,7 @@ p3 <- ggplot(toplo1, aes(x = week, y = medv)) +
   labs(
     x = NULL,
     y = 'Cells / L',
-    title = '(c) Diatom cell concentrations'
+    title = '(c) Diatom cell concentration'
   )
 
 # macroalgal weight summarized
@@ -1147,7 +1147,7 @@ p4 <- ggplot(wtsest, aes(x = week, y = medv)) +
   labs(
     y = expression(paste('kg / ', m^2)), 
     x = NULL, 
-    title = '(d) Cyanobacteria macroalgae abundances'
+    title = '(d) Filamentous cyanobacteria abundance'
   )
 
 # k brevis plot
@@ -1170,7 +1170,7 @@ p5 <- ggplot(toplo2, aes(x = week, y = medv)) +
   labs(
     x = 'Week of',
     y = 'Cells / L',
-    title = expression(paste('(e) ', italic('K. brevis'), ' cell concentrations'))
+    title = expression(paste('(e) ', italic('K. brevis'), ' cell concentration'))
   )
 
 # combine all

### 2021-12-08 manu-draft.Rmd
abstract shorten again, added highlights

@@
 
-From March 30th to April 9th, 2021, 814 million liters of mixed legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA).
+From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA).
 This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days.
-Elevated levels of phytoplankton (non-harmful diatoms) were observed in the nearshore regions of the lower Bay.
+Elevated levels of phytoplankton (non-harmful diatoms) were first observed in April in the lower Bay.
 Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*.
@@
 brevis* concentrations, prompting cleanup of over 1600 metric tons of dead fish.
-Seagrass resources showed minimal changes over the study period.
-By comparing these results to decades of baseline environmental monitoring data, we conclude that many of the biological responses, observed after the release from Piney Point, are abnormal relative to historic conditions.
+Seagrasses had minimal changes over the study period.
+By comparing these results to baseline environmental monitoring data, we conclude that many of the biological responses observed after the release from Piney Point are abnormal relative to historic conditions.
 

### 2021-12-10 manu-draft.Rmd
added pers comm to RW

@@
 Chen, Y.
-Sun, personal communication) served as the principal guidance for coordinating the data collection during the event.
+Sun, University of South Florida, pers.
+comm.
+Apr.
+2021) served as the principal guidance for coordinating the data collection during the event.
 
@@
 However, these communities can respond rapidly to external nutrient inputs [@Albert05;@Ahern07], often exhibiting lagged responses with characteristic growth/decay periods similar to observations herein [@Estrella13], and it is not unreasonable to expect these trends to be related to nutrients from Piney Point.
+Although long-term seasonal data are unavailable for comparison, anecdotal reports suggested that the observed biomass in 2021 was very unusual (R.
+Woithe, Environmental Science Associates, pers.
+comm.
+Dec.
+2021).
 

### 2021-12-14 manu-draft.Rmd
removed dep from co-auth, add graph abs

@@
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolaro@tbep.org), Edward T.
-Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Nijole Wellendorf (nijole.wellendorf@floridadep.gov), Joe Whalen (jwhalen@tbep.org)"
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue

### 2022-01-26 manu-draft.Rmd
fixing author names

@@
 Raulerson (graulerson@tbep.org), Sheila Scolaro (sscolaro@tbep.org), Edward T.
-Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Bob Weisberg (weisberg@usf.edu), Joe Whalen (jwhalen@tbep.org)"
+Sherwood (esherwood@tbep.org), David Tomasko (dave@sarasotabay.org), Robert H.
+Weisberg (weisberg@usf.edu), Joe Whalen (jwhalen@tbep.org)"
 urlcolor: blue

### 2022-01-26 revision/review_response.Rmd
reviewer response doc created

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files('revision-response.Rmd')
```

We sincerely thank the reviewers for providing comments on our manuscript.  We provie a point-by-point response to these comments below. In general, most of the revisions address concerns raised by all three reviewers, including 1) improvement in writing style for clarity and brevity, and 2) additional information on sampling methods.  We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.

## Reviewer 1

The authors (a very long list of authors) have described the sequence of events that followed a phosphate mining spill. The paper is long and tedious- and in the end cannot conclude any strong relationships with red tide occurrence. I wish I could be more positive about this paper, but it has the feeling that authors were rushing to get something in press, rather than taking the time to write a well crafted manuscript. There are some very good writers on the list of authors- so I am a bit surprised that they all agreed to this final version.

graphic abstract- much too complicated; fonts too tiny, too much text

highlights- the term Piney Point is meaningless except for locals

abstract- poorly written

Introduction- needs a complete rewrite- it takes more than 3 pages to get to the point that there was a spill. The general background is interesting but you need to grab the readers attention first.

p. 5- inadequate explanation for why the focus on N. Why not both?

Methods-
the first paragraph on monitoring is all about modeling, not monitoring

p. 7- how were the Karenia data quantified...they didn't just appear in a database.

p. 7 - eyes on seagrass is meaningless (except for locals)

stats section- poorly written, confusing, wordy

Results- rambling, disorganized and non-quantitative. Paragraphs don't make sense (try using topic sentences and structuring concepts accordingly)

Discussion- I was left wondering what I learned. No conclusions drawn

In all, it seemed like an interesting exercise to describe all the patterns, but this does not make a quality manuscript that will stand the test of time.

## Reviewer 2

Beck and colleagues report on the short- and mid-term effect of a very large nutrient-rich release from Piney Point. The authors focus on a few water quality indicators, but also include data for several additional responses (e.g., seagrass coverage). The results are largely what you would expect, initial rapid response of phytoplankton with cascading secondary responses.

First, a few of the things that make this a worthwhile contribution to MPB. The authors should be commended for their quick response to data collection and the large number of data streams that were brought to the study. It really is quite impressive. Further, compared to citizen monitoring, I expect the data is high quality and consistent. Second, the writing was quite good; there were some sections where flow or organization could be improved (beginning of results; last section of discussion); however, these considerations are very minor.

Second, a few things that could be improved or warrant some critique. What is the rationale for the lack of continuous line numbers? Journal suggestion? Reviewing is so much easier when line numbers line up with the text and continue throughout the document. I realize this is not something that should be fixed now, but it is bothersome. That said, my one major critique of the manuscript, one that is difficult to reconcile with the available data, is the speculative nature of last section of the Results, titled Potential nutrient cycling. This is clearly a very important section and, in my opinion, warrants additional consideration. The title of this section is correct, yes; thus, I would like to see 1) this section moved to the discussion, 2) the MAJOR lines of evidence supporting this thesis outlined and discussed consecutively in one section, 3) major alternatives entertained and addressed. The authors might also think about what it would take to make the connections more explicit, if possible (stable isotopes, modeling, etc.).

MINOR COMMENTS/SPECIFIC SUGGESTIONS

P3/L56: Recommend removing "in other countries". When I read that passage it seems to suggest a broader critique, one that is more political. That may not have been your intention, but that was how I read it.

P4/L21-24: I am not sure if "unanticipated releases" captures the spirt in which these releases occurred. In fact, the title of the FWC report was "Response of estuarine nekton to the regulated discharge of treated phosphate-production process water". This regulated discharge seemed to occur after extensive consultation with the EPA.

P4/L51: worth indicating that these concentrations, especially for TP, are several orders of magnitude greater than typical surface waters?

P4/L56: deliberately "released"

P5/L4: I see the use of the term lower Tampa Bay; how does that compare to Figure 1, the areas of interest? Ultimately, lower Tampa Bay is used extensively, but I don't see it defined. Also, is this region ecologically based or defined by currents?

P5/L22-27: I think a citation may be warranted here; while predictable, numerous studies ultimately supported the statement that nitrogen is limiting in Tampa Bay.

P5/L29: interesting that discharge was to Port Manatee and not Bishop Harbor like the release discussed previously.

P6/L24-39: that is an impressive and extensive list of agencies and partners; any potential drawbacks to so many data streams?

P6/L54-65: is it routine to take water samples directly from the surface? Ultimately, were the authors worried about any potential stratification, particularly of Karenia brevis.

P7/L32-37: why get data from Tampa Airport and the Airfield in St. Petersburg? There is an NWS location not far from Piney Point.

P9/L27-29: not sure what to make of the aggregating data to weeks or months; isn't there a way to maintain the actual sampling regime?

P9/L49: This is a bit confusing; is this comparing "between" pre and post?

P11/L12-14: as personal preference, I see no reason to lead off the results with a general statement about the location of a figure summary. Instead, the authors might reference the figure when they describe, in general, that a bloom was observed ~2 weeks after the release began, red tide was observed within x weeks, the bloom peaks, and conditions become stable on xx.

P11/L20-: this first part of the results has a choppy structure

P11/L31-34: the authors indicate the Tampa Bay influx was related to an ongoing coastal bloom; is this indicating the red tide bloom was ongoing prior to the Piney Point discharge? Also, not clear what "related" means in this context. This important statement may require additional context.

P16/L26: not sure I see the rationale for analyzing the data using Braun Blanquet estimates, which, in my impression, was developed as a rapid assessment

P17/Potential nutrient cycling: Unfortunately, much of the latter portions of this section are speculative and better presented in the discussion. Further, it would have been nice to see something a bit more concrete linking the nutrient shift from photoplankton to macroalgae and release of macroalgae nutrients to K. brevis; stable isotopes perhaps; the nitrogen sources could have been highly enriched

P18/L18: or a shift in the relative availability, N:P

## Reviewer 4

This is an interesting study in that it utilizes a quasi-Lagrangian approach to quantify the influence of a nutrient-rich waste plume on an estuarine ecosystem and is the first that I am aware of that documents the impacts of the Piney Point episode. Nonetheless, there are some points that require clarification to give the reader confidence in the findings. In general, more details are needed on several aspects of the methodology, and I provide specific details below on this as well as other issues.

Page 1, Line 38 - "Elevated levels of phytoplankton" - needs to be clearer… is this referring to biomass or abundance?

Pages 6, Lines 4-22 - The sampling regime was largely dictated by output of a model of plume evolution. It would be useful to see more details on how the model is calibrated, and if/how its output is cross-verified. In other words, are field observations of T/S compared with model output to evaluate its performance in terms of accuracy of plume location? This seems like critical information for assuring that the sampling regime actually followed the plume.

Page 6, Lines 46-54 - More information is needed on laboratory procedures. In particular, with many different entities running samples and presumably utilizing different analytical approaches for each variable, it is important to know how comparable the data is between programs.

Page 7, Line 22 - Were the breakpoints for K. brevis, or total phytoplankton abundance?

Page 7, Line 37 - What about from ungauged flows?

Page 11, Line 30 - Need to put Anna Maria Sound on the map. Likewise for Port Manatee.

Page 12-14, Water quality trends section - Very little attention is given to the role of the hurricane that hit the area in early July. Would be interesting to hear more about its effects on water quality in the study area.

Page 14, Macroalgae and seagrass trends section - This appears to be one of the weaker components of the project, although it is outside of my expertise. My biggest concern is that it is not clear to me how useful "% occurrence" is. As far as I can tell, this is not equivalent to biomass or areal coverage. Perhaps I am mistaken? Regardless, it would be useful to see a better justification for this metric. Also, how did the 2021 data compare to historic spatial-temporal trends?

Page 17, Lines 17-20 - How did the 2021 fish kills compare to historical kill spatial-temporal trends?

Page 17, Lines 37-40 - Could there have been a role for low D.O. in the post-storm fish kills?

Discussion - In general, it would be useful if the team would spend some time establishing the prevalence of N-limitation in the system, either based on their own data or previous studies. The interpretation of plume impacts is predicated on the system being N-limited, so that needs to be established up front.

Page 20, Line 22 - Ammonium, not ammonia. Or was pH such that ammonia would be the dominant form?

Page 20, Line 39 - Might be worth having a separate paragraph that talks about the timing/duration of the K. brevis bloom in 2021 compared to historic conditions.

Page 20 (last paragraph)-21 (first paragraph) - Without showing methods and/or data, I feel that it is inappropriate to introduce the nekton abundance/composition here.

Page 21, Lines 32-47 - Much of this text on seagrass seems speculative, and again, unclear about use of % occurrence as a metric.

Page 24, Lines 17-52 - Speculative without showing methods of data collection or the actual data.

### 2022-01-26 R/figures.R
graph abs revised, bay segment boundary to map, figs to revision folder

@@ -66,7 +66,7 @@ habdat <- read_excel(
 pineypoint <- tibble(
   lon = -82.52469352586753, 
   lat = 27.629819505234703
-  ) %>% 
+) %>% 
   st_as_sf(coords = c('lon', 'lat'), crs = 4326)
 
 # segments
@@ -165,6 +165,8 @@ tomap <- rsallpts %>%
 p2a <- ggmap(bsmap1_transparent) +
   geom_sf(data = tomap, aes(color = type), inherit.aes = F) +
   geom_sf(data = pineypoint, fill = 'black', pch = 24, color = 'black', size = 3, inherit.aes= F) + 
+  geom_sf(data = tbseglines[1, ], aes(fill = 'middle/lower Tampa\nBay boundary'), color = 'black', inherit.aes = F) + 
+  # scale_fill_manual(values = 'grey') +
   theme_bw() + 
   theme(
     panel.grid = element_blank(), 
@@ -178,6 +180,7 @@ p2a <- ggmap(bsmap1_transparent) +
     legend.position = 'right'
   ) +
   labs(
+    fill = NULL,
     color = 'Response monitoring',
     title = '(b) Sample locations'
   ) +
@@ -208,9 +211,9 @@ lbs2 <- tibble(
 
 p2b <- ggplot() + 
   geom_sf(data = states, fill = 'grey', colour = 'grey') +
-  geom_sf(data = insetbb, fill = NA, color = 'blue', size = 1.5) +
-  geom_sf_text(data = lbs1, aes(label = label), size = 5) + 
-  geom_sf_text(data = lbs2, aes(label = label), size = 5, angle = -65) + 
+  geom_sf(data = insetbb, fill = NA, color = 'blue', size = 1.25) +
+  geom_sf_text(data = lbs1, aes(label = label), size = 3.75) + 
+  geom_sf_text(data = lbs2, aes(label = label), size = 4, angle = -65) + 
   coord_sf(ylim = insetylim, xlim = insetxlim) +
   theme_void() +
   theme( 
@@ -219,26 +222,26 @@ p2b <- ggplot() +
   ) 
 
 p2 <- p2a + 
-  inset(ggplotGrob(p2b), xmin = -82.33, xmax = -82.02, ymin = 27.345, ymax = 27.57)
+  inset(ggplotGrob(p2b), xmin = -82.33, xmax = -82.1, ymin = 27.34, ymax = 27.5)
 
[…]

### 2022-01-26 revision/review_response.Rmd
graph abs revised, bay segment boundary to map, figs to revision folder
update map with anna maria and port manatee locations

@@
 
+Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Please see the details below.
+
 graphic abstract- much too complicated; fonts too tiny, too much text
+
+* __Response__: The graphical abstract was simplified and the font size increased.
 
@@
 
+* __Response__: The boundary between middle and lower Tampa Bay was added to Figure 1.
+These boundaries are used by the Tampa Bay Estuary Program to track annual attainment of water quality targets under a Reasonable Assurance plan for TMDL reporting to the Florida Department of Environmental Protection.
+We do not define the rationale for the boundaries in the main text because the regulatory delineations of the bay are not of interest to a broader audience, but we use the delineation as it is useful for our discussion of trends in the paper.
+Our area delineations (areas 1 - 3) were meant to provide a more relevant grouping for the paper.
+
 P5/L22-27: I think a citation may be warranted here; while predictable, numerous studies ultimately supported the statement that nitrogen is limiting in Tampa Bay.
@@
 
+* __Response__: Locations were added to the map.
+
 Page 12-14, Water quality trends section - Very little attention is given to the role of the hurricane that hit the area in early July.

### 2022-01-26 manu-draft.Rmd
update map with anna maria and port manatee locations

@@
 Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
-Finally, precipitation data were from Tampa International Airport, inflow estimates to Tampa Bay were basd on summed hydrologic loads of major tributaries from US Geological Survey gaged sites, and wind data were from Albert Whitted Airfield at St.
+Finally, precipitation data were from Tampa International Airport, inflow estimates to Tampa Bay were based on summed hydrologic loads of major tributaries from US Geological Survey gaged sites, and wind data were from Albert Whitted Airfield at St.
 Petersburg, Florida.

### 2022-01-26 R/figures.R

@@ -69,6 +69,20 @@ pineypoint <- tibble(
 ) %>% 
   st_as_sf(coords = c('lon', 'lat'), crs = 4326)
 
+# anna maria sound loc
+annamaria <- tibble(
+  lon = -82.69431, 
+  lat = 27.48293
+) %>% 
+  st_as_sf(coords = c('lon', 'lat'), crs = 4326)
+
+# port manatee
+portmanatee <- tibble(
+  lon = -82.56343, 
+  lat = 27.63378
+) %>% 
+  st_as_sf(coords = c('lon', 'lat'), crs = 4326)
+
 # segments
 areas <- ppseg %>% 
   rename(area = Name) %>% 
@@ -131,9 +145,24 @@ p1 <- ggmap(bsmap1_transparent) +
   new_scale_fill() + 
   geom_sf(data = bsstatloc, aes(fill = 'Long-term monitoring', shape = 'Long-term monitoring'), color = 'black', size = 2, inherit.aes= F) +
   geom_sf(data = pineypoint, aes(fill = 'Piney Point', shape = 'Piney Point'), color = 'black', size = 3, inherit.aes= F) + 
+  geom_sf(data = portmanatee, aes(fill = 'Port Manatee', shape = 'Port Manatee'), color = 'black', size = 2.5, inherit.aes= F) + 
+  geom_sf(data = annamaria, aes(fill = 'Anna Maria Sound', shape = 'Anna Maria Sound'), color = 'black', size = 2.5, inherit.aes= F) + 
   geom_sf_text(data = areas, aes(label = area), color = 'black', inherit.aes = F, alpha = 0.8, size = 6) +
-  scale_fill_manual('test', values = c('white', 'black')) +
-  scale_shape_manual('test', values= c(21, 24)) +
+  scale_fill_manual('test', 
+                    values = list(
+                      `Long-term monitoring` = 'white', 
+                      `Piney Point`  = 'black', 
+                      `Port Manatee` = 'black', 
+                      `Anna Maria Sound` = 'black')
+  ) +
+  scale_shape_manual('test', 
+                     values = list(
+                       `Long-term monitoring` = 21, 
+                       `Piney Point`  = 24,
+                       `Port Manatee` = 23, 
+                       `Anna Maria Sound` = 22
+                     ) 
[…]

### 2022-01-27 revision/review_response.Rmd
better rain data, link about inflow estimates

@@
 
+* __Response__: We have updated our weather data to include wind and precipitation from Albert Whitted Field in St.
+Petersburg and have excluded the precipitation data from Tampa International Airport.
+Our initial rationale for including Tampa was a longer period of record.
+However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data with the historical observations.
+Also, the NWS location is in Ruskin, Florida and is farther from the bay proper than the location in St.
+Petersburg.
+
 P9/L27-29: not sure what to make of the aggregating data to weeks or months; isn't there a way to maintain the actual sampling regime?
@@
 
+* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012.
+This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs in to the bay.
+We have provided a citation to this report in the paper.
+
+     Janicki Environmental, Inc., 2012. Development of a screening level tool for estimating annual hydrologic loadings to Tampa Bay (No. 05-12). Tampa Bay Estuary Program, St. Petersburg, Florida.
+
 Page 11, Line 30 - Need to put Anna Maria Sound on the map. Likewise for Port Manatee.

### 2022-01-27 manu-draft.Rmd
better rain data, link about inflow estimates

--- 2022-01-26 manu-draft.Rmd
@@
 Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
-Finally, precipitation data were from Tampa International Airport, inflow estimates to Tampa Bay were basd on summed hydrologic loads of major tributaries from US Geological Survey gaged sites, and wind data were from Albert Whitted Airfield at St.
-Petersburg, Florida.
+Finally, precipitation and wind data were from Albert Whitted Airfield at St.
+Petersburg, Florida and inflow estimates to Tampa Bay were based on summed hydrologic loads of major tributaries from US Geological Survey gaged sites [similar to @Janicki12].
 Additional details of the sampling methods and data sources are provided in supplement.
@@
 The two highest concentrations were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.
-Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (1995 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
+Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (2006 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
 This likely contributed to elevated salinity in lower and middle Tampa Bay that created conditions favorable for *K.

### 2022-01-27 R/figures.R

@@ -764,12 +764,13 @@ p2 <- ggplot(toplo, aes(x = week, y =  1 + val)) +
 
 # precip 
 toplo <- raindat %>% 
-  filter(grepl('^Tampa', station)) %>% 
+  filter(grepl('^St\\.\\sPete', station)) %>% 
   mutate(
     doy = yday(date), 
     yr = year(date), 
     mo = month(date)
   ) %>% 
+  filter(yr >= 2006) %>% 
   group_by(yr) %>% 
   arrange(yr, doy) %>% 
   mutate(precip_cm = cumsum(precip_cm)) %>% 
@@ -778,7 +779,7 @@ toplo <- raindat %>%
   mutate(
     flvl = case_when(
       yr == 2021 ~ '2021', 
-      T ~ '1995 - 2020'
+      T ~ '2006 - 2020'
     ), 
     xvals = ymd(paste('2021', month(date), day(date), sep = '-'))
   )
@@ -789,12 +790,12 @@ toplo2 <- toplo %>%
   filter(yr < 2021) %>% 
   group_by(xvals) %>% 
   summarise(
-    lov = quantile(precip_cm, 0.25, na.rm = T),
-    hiv = quantile(precip_cm, 0.75, na.rm = T),
+    lov = quantile(precip_cm, 0.05, na.rm = T),
+    hiv = quantile(precip_cm, 0.95, na.rm = T),
   ) %>% 
   ungroup()
 p3 <- ggplot() + 
-  geom_ribbon(data = toplo2, aes(x = xvals, ymin = lov, ymax = hiv, fill = '1995 - 2020\n25th - 75th %tile'), alpha = 0.7) +
+  geom_ribbon(data = toplo2, aes(x = xvals, ymin = lov, ymax = hiv, fill = '2006 - 2020\n5th - 95th %tile'), alpha = 0.7) +
   geom_line(data = toplo1, aes(x = xvals, y = precip_cm, color = '2021'), size = 1.2) +
   scale_color_manual(values = '#00806E') +
   scale_fill_manual(values = 'grey') + 
@@ -821,6 +822,7 @@ toplo <- hydrodat %>%
     yr = year(date), 
     mo = month(date)
   ) %>% 
+  filter(yr >= 2006) %>% 
   group_by(yr) %>% 
   arrange(yr, doy) %>% 
   mutate(
@@ -832,7 +834,7 @@ toplo <- hydrodat %>%
   mutate(
[…]

### 2022-01-27 revision/review_response.Rmd
getting through reviewer 4 comments

@@
 link-citations: true
+bibliography: ../refs.bib
 ---
@@
 
-Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Please see the details below.
+* __Response__: Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Please see the details below.
 
@@
 
+* __Response__: The first highlight was changed to "186 metric tons of total nitrogen from wastewater were added to Tampa Bay".
+
 abstract- poorly written
@@
 
+* __Response__: Please see our response to a similar comment from reviewer 4.
+Many studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated that Tampa Bay is nitrogen-limited.
+Phosphorus trends are also provided in the supplement.
+
 Methods-
 the first paragraph on monitoring is all about modeling, not monitoring
+
+* __Response__: We have added a "Modeling" sub-section at the beginning of the methods.
 
@@
 p. 7 - eyes on seagrass is meaningless (except for locals)
+
+* __Response__: We have removed this from the text.
 
@@
 I realize this is not something that should be fixed now, but it is bothersome.
+
+* __Response__: Unfortunately, the line numbers are generated automatically by the journal submission system and we have no control over how they are presented.
+We appreciate your use of both page and line numbers in your comments to reference text locations.
+
 That said, my one major critique of the manuscript, one that is difficult to reconcile with the available data, is the speculative nature of last section of the Results, titled Potential nutrient cycling.
@@
 
+* __Response__: Seagrass transect monitoring in Tampa Bay has occurred annually since the mid-1990s using similar methods as in this paper [i.e., @Sherwood17].
+The primary difference for the routine monitoring is the transect length, which is typically much longer than the 50m used in the 2021 rapid design.
+This shorter distance was chosen to allow for quicker sampling times and greater coverage in response to the rapidly changing conditions in 2021.
+Thus, we consider the rapid design as providing data that are of similar quality to the long-term data, with notable exceptions indicated in the mansucript (i.e., no seasonal estimates).
+Our rationale for evaluating Braun Blanquet coverage was to verify the results obtained evaluating frequency occurrence estimates.
+Line xxx was modified to make this clear:
+
+     "Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4)."
+
 P17/Potential nutrient cycling: Unfortunately, much of the latter portions of this section are speculative and better presented in the discussion.
@@
 
+* __Response__: Thank you for your constructive comments.  Please see the detailed responses below.
+
 Page 1, Line 38 - "Elevated levels of phytoplankton" - needs to be clearer… is this referring to biomass or abundance?
+
+*__Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".
+This was based on chlorohpyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.
+The word limit for the absract prevents these details, but they are presented in text (i.e., line xxx and xxx).
 
@@
 
+* __Response__: TBCOM was previously tested for veracity against in situ observations [e.g., @Chen18].
+A prior version upon which TBCOM was constructed was also veracity tested and used in several studies [Zhu et al., 2015[a](https://doi.org/10.1007/s12237-014-9815-4),[b](https://doi.org/10.1007/s12237-014-9793-6),[c](https://doi.org/10.1016/j.csr.2015.07.001), and @Chen19] More systematic model/data comparisons were also reported in a recent PhD dissertation (J.
+Chen, Univ.
+of South Florida, 2021).
+Thus, prior to this application, TBCOM was found to perform well in simulating the circulation in the Tampa Bay, and TBCOM has provided publicly available, daily nowcast/forecasts (http://ocgweb.marine.usf.edu/~tbm/index.html) for the past four years.
+
+     More detailed model description or analysis is beyond the scope this work.
+A dedicated manuscript is being prepared to document the TBCOM tracer model as a rapid response to the Piney Point event.
+Preliminary results can be found in CERF and AGU presentations:
+
+     Liu, Y., Weisberg, R.H., Zheng, L., Sun, Y., Chen, J.
+(2021), Nowcast/Forecast of the Tampa Bay, Piney Point Effluent Plume: A Rapid Response, Abstract (OS35B-1036) presented at AGU Fall Meeting, New Orleans, Louisiana, December 2021.
+
+     Liu, Y., Weisberg, R.H., Zheng, L., Sun, Y., Chen, J.
+(2021), Nowcast/Forecast of the Tampa Bay, Piney Point Effluent Plume: A Rapid Response, Coastal and Estuarine Research Federation (CERF) 2021 Virtual Meeting, November 2021.
+
+     Lines xxx were edited to include some of the above details: "The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
+Preliminary model results were reported in @Liu21."
+
 Page 6, Lines 46-54 - More information is needed on laboratory procedures.
@@
 
+* __Response__: Our initial submission included supplement material that described the sample collection and processing in more detail.
+We have moved that content from the supplement to the main text, beginning on lines xxx:
+
+     "Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
+Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.
+Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620].
+Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of split samples between laboratories to check precision of measured values.
+Samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).
+In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.
+Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP."
+
 Page 7, Line 22 - Were the breakpoints for K. brevis, or total phytoplankton abundance?
+
+* __Response__: These were applied to K. brevis.  The sentence was revised: "Bloom sizes for HAB species were described qualitatively as low/medium/high concentrations..."
 
@@
 
-* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012.
+* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012 [@Janicki12].
 This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs in to the bay.
 We have provided a citation to this report in the paper.
-
-     Janicki Environmental, Inc., 2012. Development of a screening level tool for estimating annual hydrologic loadings to Tampa Bay (No. 05-12). Tampa Bay Estuary Program, St. Petersburg, Florida.
 
@@
 
+* __Response__: As noted in the red tide section and discussion, Tropical Storm Elsa was considered a confounding factor when interpreting drivers of the red tide event.
+It is challenging to disentangle an isolated effect of the storm on the red tide and the same could be said for the broader suite of parameters that were used to assess water quality.
+For example, a clear chlorophyll spike in early July coincided with the red tide peak, which also coincided with passage of the storm.
+A lack of continuous monitoring in the bay also prevented a more comprehensive assessment of potential storm effects.
+However, water quality conditions were changing rapidly from the red tide, preceding the arrival of the storm by several days, and we consider the water quality conditions at the time of passage to be driven primarily by the *K.
+brevis* bloom in the bay.
+We have added some additional text describing these interpretations in response to your comment below about dissolved oxygen.
+
 Page 14, Macroalgae and seagrass trends section - This appears to be one of the weaker components of the project, although it is outside of my expertise.
@@
 
+* __Response__:
+
 Page 17, Lines 17-20 - How did the 2021 fish kills compare to historical kill spatial-temporal trends?
 
+* __Response__: We were also interested in this question and had included a long-term trend plot in an earlier draft, with results as early as 1995.
+However, this was not included due to differences in reporting methods that have changed over time.
+For example, online reporting iculties in comparing the long-term trend to current results
+
 Page 17, Lines 37-40 - Could there have been a role for low D.O. in the post-storm fish kills?
+
+* __Response__: We included some text describing this issue in the discussion.
+
+     "Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen.
+@Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.
+Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality."
 
@@
 
+* __Response__: We agree that this nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus.
+The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.
+We include some additional text and citations in the final paragraph of the introduction as further justification:
+
+     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
+As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."
+
+     We also note that phosphorus trends are included in the supplement.
+
 Page 20, Line 22 - Ammonium, not ammonia. Or was pH such that ammonia would be the dominant form?
+
+* __Response__:
 
@@
 
+* __Response__:
+
 Page 20 (last paragraph)-21 (first paragraph) - Without showing methods and/or data, I feel that it is inappropriate to introduce the nekton abundance/composition here.
+
+* __Response__: Agreed, much of this content was removed.
 
@@
 
+* __Response__:
+
 Page 24, Lines 17-52 - Speculative without showing methods of data collection or the actual data.
+
+* __Response__:
+
+# References

### 2022-01-27 manu-draft.Rmd
getting through reviewer 4 comments

@@
 This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days.
-Elevated levels of phytoplankton (non-harmful diatoms) were first observed in April in the lower Bay.
+An initial phytoplankton bloom (non-harmful diatoms) was first observed in April in the lower Bay.
 Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*.
@@
 The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
-We focus on nitrogen inputs as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
+Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
+As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
 A timeline of events is provided, which is supported by the quantitative results from 2021 routine and response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.
@@
 
-## Monitoring response to the emergency release
+## Simulation modeling
 
@@
 Normalized tracer distributions were automatically updated each day, providing 1-day hindcasts and 3.5-day forecasts throughout the period of discharge and subsequent Tampa Bay distribution.
-The modeled plume evolution web product (Y.
-Liu, R.H.
-Weisberg, J.
-Chen, Y.
-Sun, University of South Florida, pers.
-comm.
-Apr.
-2021) served as the principal guidance for coordinating the data collection during the event.
+The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
+Preliminary model results were reported in @Liu21.
+
+## Monitoring response to the emergency release
 
@@
 Monitoring efforts focused on a suite of parameters expected to respond to increased nutrient loads into the bay, including water quality sampling, phytoplankton identification, and seagrass and macroalgae transect surveys (Figure \@ref(fig:map)).
-In short, water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
+
+Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
 Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
+Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.
+Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620].
+Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of split samples between laboratories to check precision of measured values.
+Samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).
+In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.
+Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.
 
@@
 Harmful Algal Bloom (HAB) data for *Karenia brevis* were obtained from event-based monitoring samples from the [FWC-FWRI HAB Monitoring Database](https://myfwc.com/research/redtide/monitoring/database/).
-Phytoplankton blooms were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
+Bloom sizes for HAB species were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
 Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
-Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design following the "[Eyes on Seagrass](http://blogs.ifas.ufl.edu/news/2019/03/19/citizen-scientists-needed-to-survey-seagrass-seaweed-in-charlotte-harbor/)" method developed by a local citizen science group in cooperation with academic and federal partners.
+Seagrass and macroalgae sampling occurred approximately biweekly at `r ntrns` transects using a modified rapid assessment design, where species were identified and enumerated using Braun-Blanquet abundances in a 0.25 m$^2$ quadrat at 10m distances along each 50m transect (see supplement).
 Finally, precipitation and wind data were from Albert Whitted Airfield at St.
@@
 Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (Tables S1, S2).
-Tests using Braun Blanquet cover estimates also produced similar results (Tables S3, S4).
+Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4).
 
@@
 brevis*, having negative effects on fisheries resources in June and July (Figure \@ref(fig:redtide)).
-In addition to fish kill reports, quantitative data on changes in nekton abundance and diversity in Tampa Bay in 2021 are forthcoming.
-Routine sampling of fisheries occurs monthly in Tampa Bay and a long-term record back to 1998 provides detailed information for the major bay segments.
-Results from the Tampa Bay Nekton Index showed a decline in fisheries resources following a significant red tide event that persisted for several months in lower Tampa Bay in 2005 [@Flaherty11;@Schrandt21].
-Given the observed *K.
-brevis* concentrations in 2021 and the magnitude of fish kills, mandates for catch and release for popular sportfishes (*Sciaenops ocellatus*, *Cynoscion nebulosus*, and *Centropomus undecimalis*) were extended through the fall of 2021.
+Fisheries resources in Tampa Bay have previously been negatively affected by red tide [e.g., in 2005, @Flaherty11;@Schrandt21].
 For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
@@
 In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].
+Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen.
+@Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.
+Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality.
 

### 2022-01-27 supplement.Rmd
getting through reviewer 4 comments

@@
 # Sampling methods
-
-Established laboratory and field sample protocols for all survey methods were based on an [Interagency Monitoring Project Plan](https://drive.google.com/drive/u/0/folders/1oBGvjdve-Gpo4Kn3Ovn8a8-yVoP25eec) maintained by the TBEP and those of the inter-agency partners.
-Data quality objectives followed guidelines outlined in the USEPA-approved TBEP Data Quality Management Plan [@tbep1620].
-Many of the local partners also participate in the Southwest Florida [Regional Ambient Monitoring Program](https://tbep.org/our-work/boards-committees/technical-advisory-committee/#ramp) (RAMP) that ensures similar standards and protocols are followed in the collection and processing of monitoring data, including routine cross-reference of samples between laboratories to check precision of measured values.
-Discrete water quality samples were taken primarily from surface grabs by boat and processed for relevant parameters (see below) by the respective laboratories of each participating agency.
-Surface samples requiring laboratory analysis (e.g., nutrient assays) were obtained primarily from bottle collection at the surface, whereas in situ measurements were available for many parameters (e.g., dissolved oxygen, Secchi depth, etc.).
-In situ measurements were collected using common monitoring equipment, such as YSI sondes or Seabird CTD casts, depending on monitoring agency.
-Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.
-
-For data analysis in this paper, we focus on water quality parameters related to the Tampa Bay nutrient management paradigm and the expected phytoplankton response from a dense, inorganic nitrogen plume entering the bay.
-This included statistical evaluation of total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), and chlorophyll-a ($\mu$g/L) concentrations.
-Samples for pH, salinity (ppt), temperature (C), and dissolved oxygen saturation (\%) were also evaluated given the role these parameters can have as indicators of surface water contamination (pH), physical drivers of primary production (salinity, temperature), and indicators of primary production and respiration (dissolved oxygen).
-Overall, sample frequency was approximately biweekly over the months following the Piney Point release, with greatest effort during the first three months and with some variability given agency resource limitations during the event and weather restrictions on boating.
-As appropriate, water quality data were aggregated at the weekly scale and by major areas of interest (Figure 1a) given the hypothesized impacts adjacent to the release site as indicated by the TBCOM tracer model simulated plume patterns.
 

### 2022-01-28 revision/review_response.Rmd
rev4 comments addressed

@@
 
-*__Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".
+* __Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".
 This was based on chlorohpyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.
-The word limit for the absract prevents these details, but they are presented in text (i.e., line xxx and xxx).
+The word limit for the abstract prevents these details, but they are presented in text (i.e., line xxx and xxx).
 
@@
 * __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012 [@Janicki12].
-This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs in to the bay.
+This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs to the bay.
 We have provided a citation to this report in the paper.
@@
 
-* __Response__: As noted in the red tide section and discussion, Tropical Storm Elsa was considered a confounding factor when interpreting drivers of the red tide event.
+* __Response__: As noted in the red tide section and discussion, Tropical Storm Elsa was a confounding factor when interpreting drivers of the red tide event.
 It is challenging to disentangle an isolated effect of the storm on the red tide and the same could be said for the broader suite of parameters that were used to assess water quality.
@@
 brevis* bloom in the bay.
-We have added some additional text describing these interpretations in response to your comment below about dissolved oxygen.
+We have added some additional text describing these interpretations in response to your comment below about dissolved oxygen and fish kills.
 
@@
 
-* __Response__:
+* __Response__: Frequency occurrence is a routine reporting metric used by the Tampa Bay Estuary Program to summarize the long-term annual transect monitoring [see @Johansson16;@Sherwood17].
+Additional text on line xxx was added to make this clear:
+
+     "Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17]."
+
+     We realize that additional metrics could have been used to summarize the data and we presented trend assessments using the Braun-Blanquet abundance estimates as confirmation for the results from the frequency occurrence analyses (line xxx).
+However, this does not provide an assessment of biomass or coverage.
+Weight data were collected for select macroalgae samples during the 2021 surveys and a preliminary analysis (see the Seagrass -> Biomass estimates tab [here](https://tbep-tech.github.io/piney-point-analysis/trend)) suggested similar trends as the frequnecy occurrence results.
+Because we did not have comparable weight data for seagrasses, the analysis was not included.
+It is also worth mentioning that coverage estimates are obtained every two years for Tampa Bay as an annual snapshot (noted on line xxx in the discussion), but these are not of sufficient temporal resolution to support the analyses in this paper.
+
+     There were limitations to our analysis that were noted in the original draft, primarily the lack of long-term historical data on seasonal trends for seagrasses and macroalgae.
+For example, line xxx in the original discussion: "Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release."
 
@@
 
-* __Response__: We were also interested in this question and had included a long-term trend plot in an earlier draft, with results as early as 1995.
+* __Response__: We were also interested in this question and had included a long-term trend plot in an earlier draft, with results as early as 1995 (see [here](https://github.com/tbep-tech/piney-point-manu/blob/c45049af6885e1e7d29aeb414d161a0ce4bdeadb/figs/redtide.jpeg)).
 However, this was not included due to differences in reporting methods that have changed over time.
-For example, online reporting iculties in comparing the long-term trend to current results
+Online reporting became an option in recent years and we considered the results incomparable to the earlier data that were based on phone reporting methods.
 
@@
 
-* __Response__: We agree that this nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus.
+* __Response__: We agree that the nutrient limitation needs to be clearly stated as the basis for focusing primarily on nitrogen and not phosphorus.
 The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.
@@
 
-     We also note that phosphorus trends are included in the supplement.
+     Phosphorus summary plots are also included in the supplement.
 
@@
 
-* __Response__:
+* __Response__: This is in reference to total ammonia nitrogen, as was provided by the water quality sampling.
+We have made a note on line xxx in the methods that any mention of ammonia refers to total ammonia nitrogen.
 
@@
 
-* __Response__:
+* __Response__: Previous content in the discussion below this paragraph provided some historical context on red tide events in the region.
+Specifically, starting on line xxx: "Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf..."
+
+     We have added some text below the existing pararaph to provide more context: "Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
+Severity in estuaries is typically less than Gulf waters as *K.
+brevis* is limited to higher salinity."
 
@@
 
-* __Response__:
+* __Response__: We have replaced the beginning of this paragraph with more definitive text:
+
+     "Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein].
+The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses."
+
+     Also, please see our response above about frequency occurrence.
 
@@
 
-* __Response__:
+* __Response__: The paragraph was removed for brevity.
 

### 2022-01-28 manu-draft.Rmd
rev4 comments addressed

@@
 
-Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
+Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L, hereafter referred to as ammonia), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
 Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
@@
 
+Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17].
 Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
@@
 
-For seagrasses, major bloom events in 2021 produced unfavourable water quality conditions, although changes in frequency occurrence of seagrasses were not observed over the initial study period.
+Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein].
+The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].
-Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, also created conditions that were favorable for growth of *K.
-brevis*.
+Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
+Severity in estuaries is typically less than Gulf waters as *K.
+brevis* is limited to higher salinity.
+Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, created conditions that were favorable for growth of *K.
+brevis* in Tampa Bay.
 However, the results suggest a likely scenario that residual nutrients from the Piney Point release, or indirectly through nutrients made available from the growth and decomposition of other primary producers (e.g., diatoms, macroalgae) stimulated by inputs from Piney Point, were sufficiently available to allow growth of *K.
@@
 Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
-
-Additional monitoring and analysis are also required to fully understand the long-term impacts to bay resources beyond water quality.
-For benthic communities, sediments sampled in April and September near Port Manatee and surrounding waters suggested a mix of conditions dominated by "intermediate" and "healthy" benthic index scores [@tbep0106, see https://shiny.tbep.org/piney-point], possibly reflecting the generally high spatial variability of macroinvertebrate communities in coastal habitats [@tbep0620;@Gillett21].
-Comparison of the April and September samples to historical conditions suggested relatively consistent benthic invertebrate community structure from 1993 to present ([TBEP unpublished results](https://tbep-tech.github.io/piney-point-analysis/benthic)).
-Differences between the April and September samples were not observed.
-Finally, effects of changing environmental conditions and *K.
-brevis* on marine mammals (e.g., cetaceans, sirenians) was also a concern given their use of bay resources.
-Twenty manatee (*Trichechus manatus latirostris*) mortalities were [reported](https://myfwc.com/media/25649/2021preliminaryredtide.pdf) in the red tide boundary of the impacted counties of Tampa Bay through August 2021.
-This is of particular concern given the recent [unusual mortality events](https://myfwc.com/research/manatee/rescue-mortality-response/ume/) for Florida manatees that are likely linked to seagrass losses on the east coast of Florida (e.g., Indian River Lagoon) and current seagrass losses for southwest Florida.
 

### 2022-01-28 revision/review_response.Rmd
getting through reviewer 2

@@
 We sincerely thank the reviewers for providing comments on our manuscript.
-We provie a point-by-point response to these comments below.
+We provide a point-by-point response to these comments below.
 In general, most of the revisions address concerns raised by all three reviewers, including 1) improvement in writing style for clarity and brevity, and 2) additional information on sampling methods.
@@
 
+* __Response__: We really appreciate your comments.  Please see the detailed responses below.
+
 First, a few of the things that make this a worthwhile contribution to MPB.
@@
 Second, the writing was quite good; there were some sections where flow or organization could be improved (beginning of results; last section of discussion); however, these considerations are very minor.
+
+* __Response__: Thank you for pointing out the merits of this paper and providing suggestions where the writing can be improved.  We have edited the sections you point out for brevity and clarity.
 
@@
 
+* __Response__: Thank you for this comment.
+We agree this section is important as it relates the timeline of events in a plausible manner, although it is definitely speculative without more quantitative results linking the pieces.
+As such, we have moved the content to the discussion and amended to text to include a more nuanced description of the trends, including what information may be needed to support the narrative.
+This includes information from the final paragraph of the discussion, which was removed.
+The new section is included from lines xxx to xxx.
+
 MINOR COMMENTS/SPECIFIC SUGGESTIONS
@@
 
+* __Response__: Our point was to suggest that mining products in Florida are generally exported and have little benefit for the local communities, aside from localized economic benefits.
+However, this may be a bit editorialized and was removed.
+
 P4/L21-24: I am not sure if "unanticipated releases" captures the spirt in which these releases occurred.
@@
 
+* __Response__: Agreed, we have removed "unanticipated".
+We agree that the releases often occur after consultation or approval from regulatory bodies, but they are generally unanticipated in that site management has not planned for these events, nor the potential impacts on the environment.
+
 P4/L51: worth indicating that these concentrations, especially for TP, are several orders of magnitude greater than typical surface waters?
 
+* __Response__: The sentence was edited: "Water quality of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay."
+
 P4/L56: deliberately "released"
+
+* __Response__: Changed to "released under FDEP authorization".
 
@@
 
+* __Response__: Agreed, we have revised this statement to include additional citations and explanation:
+
+     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
+As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."
+
 P5/L29: interesting that discharge was to Port Manatee and not Bishop Harbor like the release discussed previously.
+
+* __Response__: Yes, we believe the intent was to minimize impacts to sensitive back-bay habitats that were observed in previous releases, although we have no information to verify this claim and have not included it in the text.
 
@@
 
+* __Response__: Data quality assurance and control was a primary concern with multiple partners collecting information.
+We have included additional information about the data collection procedures and how the Tampa Bay Estuary Program worked to facilitate quality of the data.
+Please see the additions starting on line xxx.
+
 P6/L54-65: is it routine to take water samples directly from the surface? Ultimately, were the authors worried about any potential stratification, particularly of Karenia brevis.
+
+* __Response__:  Most of the water quality and phytoplankton samples were surface collections following standard sampling methods used by our partners (see the text addition about sampling).
+Tampa Bay also has a very shallow average depth of just over 3m and is not strongly stratified for most of the year.
+So, we feel confident that the majority of samples used in our analyses were generally representative of conditions in the water column.
+Anecdotally, benthic fishes were observed during the fish kills (e.g., eels, rays), so it is likely that *K.
+brevis* was present throughout the water column, although we have very little quantitative data to describe the vertical distribution.
 
@@
 
+* __Response__:
+
 P9/L49: This is a bit confusing; is this comparing "between" pre and post?
+
+* __Response__:
 
@@
 
+* __Response__:
+
 P11/L20-: this first part of the results has a choppy structure
+
+* __Response__:
 
@@
 This important statement may require additional context.
+
+* __Response__:
 
@@
 
+* __Response__: This section was moved to the discussion, please see the response to your general comment above.
+
 P18/L18: or a shift in the relative availability, N:P
+
+* __Response__:
 
@@
 The second paragraph in the introduction was meant to provide this justification, but this was clearly insufficient as noted in the review.
-We include some additional text and citations in the final paragraph of the introduction as further justification:
-
-     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
-As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."
-
-     Phosphorus summary plots are also included in the supplement.
+We include some additional text and citations in the final paragraph of the introduction as further justification - see the response to the similar critique from reviewer 2.
+Also note that phosphorus summary plots were previously included in the supplement.
 

### 2022-01-28 manu-draft.Rmd
getting through reviewer 2

@@
 
-The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production in other countries [@Henderson04].
+The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer to support agricultural production [@Henderson04].
 By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
@@
 Holding capacity of the ponds has decreased over time from seasonal rain events, tropical storms, and storage of dredging material from nearby Port Manatee.
-Unanticipated releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.
+Releases from the stacks occurred in the early 2000s and in 2011 to nearby Bishop Harbor connected to Tampa Bay.
 Those releases resulted in spatially-restricted, ecosystem responses including localized harmful algal blooms and increased macroalgal abundance [@Switzer11;@Garrett11].
@@
 Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.
-Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) well outside normal ranges typical of surface waters in Tampa Bay.
-From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay.
+Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay.
+From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released under FDEP authorization to lower Tampa Bay.
 Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
@@
 
-## Potential nutrient cycling
-
-The above results can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point.
-Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
-The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
-Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
-However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
-During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
-brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).
-The co-occurring decline in macroalgae and increase in *K.
-brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter, although residual water column nutrients from Piney Point may have also been present (as suggested by modelling efforts).
-Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
-These distinct temporal periods were readily identified through an ordination plot (Figure S7) for the observed data in Figure \@ref(fig:nutrientflow).
-
 # Discussion
@@
 
-The observed conditions in Tampa Bay in 2021 following unanticipated releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
+The observed conditions in Tampa Bay in 2021 following releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
@@
 External nitrogen loads to lower Tampa Bay averaged `r tnld` metric tons per year for the baseline period of 2006 to 2020 (https://tbep-tech.github.io/load-estimates/).
+
+## Potential nutrient cycling
+
+The events of 2021 can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point.
+Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
+The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
+These results were evidenced by taxonomic enumeration of phytoplankton samples collected near Port Manatee.
+Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
+However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
+During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
+brevis* concentrations peaked (Figure \@ref(fig:nutrientflow)e).
+The co-occurring decline in macroalgae and increase in *K.
+brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter.
+Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
+
+Our quantitative results provide some evidence to support the progression of events outlined above as a flow of nutrients over time.
+The distinct temporal progression can be readily identified through an ordination plot (Figure S7) for the observed data in Figure \@ref(fig:nutrientflow).
+Weekly summaries of the data are clearly separated in the ordination into monthly groups where different communities were dominant and is partially explained by orientation of the water quality vectors relative to cyanobacteria, diatoms, and *K.
+brevis*.
+For example, total nitrogen and chlorophyll are strongly aligned with the *K.
+brevis* axis as nutrients were likely available in organic form during the peak of the red tide event.
+However, this simple analysis only demonstrates an association in the observed data and cannot be verified without additional information.
+Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
+Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
+Additional confounding variables can also obscure the association between water quality and community changes.
+Bay conditions preceding the 2021 events, as well as the passage of tropical storm Elsa, could obscure these associations (described below).
+
+## Additional interpretation of impacts
 
@@
 These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].
-
-Limitations of our analyses are also important to note to inform future event-based monitoring and additional research.
-All of the analyses are correlative based on associations between the measured water quality observations, macroalgae, and seagrass results and may not represent explicit cause and effect mechanisms.
-However, the interpretations are supported by previous research on drivers of primary production and eutrophication of coastal waters.
-Additional data to support these results could include explicit load-based estimates for all sources entering the bay through 2021 and these estimates are forthcoming.
-Laboratory-based methods, such as isotopic analyses of nutrient signatures found in biological tissues (e.g., macroalgae) compared to those from the release, could provide a more comprehensive description of the recycling of nitrogen from Piney Point.
-Long-term fate of nutrients from Piney Point is uncertain and continued monitoring can further support understanding of ecosystem response in Tampa Bay beyond the initial results in 2021.
-Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
-These efforts are critical to managing Gulf of Mexico ecosystems given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
 

### 2022-01-30 revision/review_response.Rmd
reviewer 2 comments addressed

@@
 brevis* was present throughout the water column, although we have very little quantitative data to describe the vertical distribution.
+Please see the addition of information on *K.
+brevis* sampling on line xxx provided in response to reviewer 1.
 
@@
 Petersburg and have excluded the precipitation data from Tampa International Airport.
-Our initial rationale for including Tampa was a longer period of record.
-However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data with the historical observations.
+Our initial rationale for using data from Tampa was a longer period of record.
+However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data with historical observations.
 Also, the NWS location is in Ruskin, Florida and is farther from the bay proper than the location in St.
@@
 
-* __Response__:
+* __Response__: The weekly aggregation was only done to allow plotting and comparison between variables on a common scale (e.g., Figure 7 and 8).
+None of the statistical tests used data with fixed sample dates.
+We removed the text because the statement is inaccurate and somewhat misleading.
 
@@
 
-* __Response__:
+* __Response__: No, these tests were meant to compare all observations post-release to help identify periods of time when observations differed, e.g., when chlorophyll was at its maximum.
+The text was revised for clarity.
+
+     "Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point."
 
@@
 
-* __Response__:
+* __Response__: The first section of the results describing the timeline was removed and text was placed in the appropriate sections that followed.
+The timeline is now referenced in the first paragraph of the discussion to provide a synthesis of the main results that were described.
 
@@
 
-* __Response__:
+* __Response__: Please see the response to the previous comment.
 
@@
 
-* __Response__:
+* __Response__: Yes, red tide was observed in the Gulf of Mexico prior to the occurrence in Tampa Bay.  Content in the red tide section was edited to make this clear (line xxx):
+
+     "This first Tampa Bay influx originated from an ongoing coastal bloom in the Gulf of Mexico, as is common when red tide is observed in the bay [@Steidinger72;@Flaherty11]."
+
+     Blooms in Tampa Bay originate from *K.
+brevis* that is transported in from the Gulf of Mexico.
+An important distinction that we make in this paper is that, although blooms in Tampa Bay are not uncommon, the severity of the 2021 bloom was very likely related to the favorable conditions in the bay prior to July.
+These favorable conditons were related to 1) nutrient availability from Piney Point, and 2) high salinity conditions from low rainfall.
+The passage of tropical storm Elsa was also a confounding factor.
+It is nearly impossible to identify a single factor causing the observed conditions, but we describe the considerations in the discussion (lines xxx to xxx).
 
@@
 
-* __Response__:
+* __Response__: Added text "...or changing availability of nutrient ratios."
 

### 2022-01-30 manu-draft.Rmd
reviewer 2 comments addressed

@@
 Statistical assessments were conducted only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
-Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred on different days during the six month period.
 Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a.
 Variables with log-normal distributions were log$_10$-transformed (i.e., nutrients, chl-a) prior to analysis.
-For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection dates.
+Only the water quality data from FDEP were used for statistical analysis given the consistency of sample location and collection dates.
 Secchi observations that were visually identified on the bottom (`r sumb` of `r totb` observations in the FDEP data) were removed from analysis.
@@
 
-Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17].
-Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point.
 Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
@@
 Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist.
+Frequency occurrence estimates were used to evaluate macroalgae and seagrasses as a standard metric used in previous analyses in Tampa Bay [@Johansson16;@Sherwood17].
 Methods used to accommodate measured concentrations of water quality variables that were below detection included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
@@
 # Results
-
-## Timeline of 2021 biological response events
-
-A general summary of 2021 events in Tampa Bay following discharge from Piney Point is shown in Figure \@ref(fig:timeline).
-After the discharge ceased on April 9th, a peak in median chl-a concentration was observed near Piney Point (Area 1, Figure \@ref(fig:map)a) in mid-April, with peak individual sample values in excess of 50 $\mu$g/L.
-Median concentrations for each week in April were less than 10 $\mu$g/L.
-The discharge phytoplankton assemblage was comprised of over 99% of a spherical nanoplanktonic chlorophyte (3.37 x 10$^8$ cells/L).
-The phytoplankton communities near the discharge area in April were generally dominated by diatoms.
-The initial diatom bloom did not persist past April.
-On April 20th, the HAB species *Karenia brevis* was observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay; this first Tampa Bay influx was related to an ongoing coastal bloom.
-By May/June, bloom levels of *K.
-brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with peak concentrations in excess of 1 x 10$^6$ cells/L.
-Also during May/June, high abundances of filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1).
-High levels of cyanobacteria coverage on benthic and seagrass habitats were observed, in addition to large floating mats on the surface.
-By June 27th, fish kill reports attributed to *K.
-brevis* increased as cellular abundance climbed in lower and middle Tampa Bay.
-The center of tropical storm Elsa passed to the west of Tampa Bay on July 5th, causing a shift in prevailing winds from the southeast.
-This shift in winds likely disturbed the water column and altered the spatial distribution of *K.
-brevis* in the bay.
-Strong southeasterly winds also likely moved dead fish closer to heavily populated areas of Tampa Bay, specifically near the cities of St.
-Petersburg and Tampa, contributing to an increase in fish kill reports.
-Concentrations of *K.
-brevis* in middle and lower Tampa Bay peaked in early to mid July, with bloom conditions not observed in the bay after July.
-Conditions were relatively stable in August and September compared to months prior.
-A quantitative description of these events follows.
 
@@
 For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.
+Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, altough median concentrations for each week in April were less than 10 $\mu$g/L.
+The initial chlorophyll peak was associated with a localized phytoplankton bloom generally dominated by diatoms.
+The initial diatom bloom did not persist past April.
 Concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with the increase in *K.
@@
 Reductions in red macroalgae frequency occurrence were observed in June when cyanobacteria frequency occurrence peaked, with greater coverage of cyanobacteria in Area 3 (`r fomoarea3jun['Cyanobacteria']`\%) compared to Area 1 (`r fomoarea1jun['Cyanobacteria']`\%).
+Notable blooms of the filamentous cyanobacteria (*Dapis* spp.) were observed in Anna Maria Sound (Area 3) and near Port Manatee (Area 1) (Figure \@ref(fig:map)), typically observed covering benthic and seagrass habitats, in addition to large floating mats on the surface.
 Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).
@@
 
-Bloom concentrations of the red tide species *K.
-brevis* in 2021 were first observed in Tampa Bay the week of May 23, with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+On April 20th, the HAB species *Karenia brevis* was observed near Anna Maria Sound at the southern edge of the mouth of Tampa Bay.
+This first Tampa Bay influx likely originated from an ongoing coastal bloom in the Gulf of Mexico, as is common when red tide is observed in the bay [@Steidinger72;@Flaherty11].
+By May 23, bloom concentrations of *K.
+brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
@@
 brevis* cell concentrations were consistently above the medium threshold (10$^4$ cells/L).
-The center of Tropical Storm Elsa (Figure \@ref(fig:redtide)f, pre-, post-storm wind roses) also passed through the bay area on July 5th, causing a shift in winds that likely altered the location of *K.
-brevis* cells and dead fish in the bay.
+The center of Tropical Storm Elsa (Figure \@ref(fig:redtide)f, pre-, post-storm wind roses) also passed through the bay area on July 5th, causing a shift in winds that likely disturbed the water column and altered the spatial distribution of *K.
+brevis* in the bay.
+Strong southeasterly winds also likely moved dead fish closer to heavily populated areas of Tampa Bay, specifically near St.
+Petersburg and Tampa, contributing to an increase in fish kill reports.
 It is important to note that high cell concentrations (>10$^6$ cells/L) were observed in middle Tampa Bay (Figure \@ref(fig:redtide)b) and fish kills were reported both before and after storm passage (Figure \@ref(fig:redtide)e).
-By August, Pinellas County and the city of St.
-Petersburg removed over 1600 metric tons of dead fish near public and private shoreline areas (K.
+By August, cleanup efforts removed over 1600 metric tons of dead fish near public and private shoreline areas (K.
 Hammer Levy, Pinellas County, pers.
@@
 Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K.
+These anomalous events (Figure \@ref(fig:timeline)) included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K.
 brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
@@
 Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
-However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities.
+However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities or changing availability of nutrient ratios.
 During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
@@
 Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf [@Steidinger75;@Weisberg14;@Weisberg19;@Liu16] and occasionally occurring at bloom concentrations in lower and middle Tampa Bay.
-Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger71], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].
+Although bloom concentrations in 2021 were extreme, historical blooms have been observed in Tampa Bay with notable events occurring in 1971 [@Steidinger72], 2005 [@Flaherty11], and recently in 2018 [@Skripnikov21].
 Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
@@
 
-```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the release from Piney Point.
-Inset image shows blooms of filamentous cyanobacteria (*Dapis* spp.).'}
-knitr::include_graphics(here('figs/timeline.jpg'))
-```
-
 ```{r wqmap, fig.cap = 'Water quality data (raw observations) for April through September 2021 following the release from Piney Point for (a) total nitrogen (mg/L), (b) chlorophyll-a ($\\mu$g/L), and (c) Secchi disk depth (meters).
@@
 knitr::include_graphics('figs/redtide.jpeg')
+```
+
+```{r timeline, fig.cap = 'Graphical timeline of events in Tampa Bay from March 30th through September 2021 following the release from Piney Point.
+Inset image shows blooms of filamentous cyanobacteria (*Dapis* spp.).'}
+knitr::include_graphics(here('figs/timeline.jpg'))
 ```

### 2022-01-30 revision/review_response.Rmd
adding responses to reviewer 1

@@
 
-* __Response__: Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Please see the details below.
+* __Response__: Thank you for your comments.
+We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted.
+Note that many of the edits in response to the more specific comments from the other reviewers are used to address some of the comments below.
 
@@
 
+* __Response__: Please see the responses to reviewer 2.
+
 Results- rambling, disorganized and non-quantitative. Paragraphs don't make sense (try using topic sentences and structuring concepts accordingly)
+
+* __Response__: Please see the responses to reviewer 2, particularly regarding the first section of the results.
 
@@
 
+* __Response__: Please see the responses to reviewer 2, particularly regarding the edits to the discussion sections.
+
 In all, it seemed like an interesting exercise to describe all the patterns, but this does not make a quality manuscript that will stand the test of time.
+
+* __Response__: We are confident that the edits made in response to your comments and those from the other reviewers have made this a more compelling and valuable contribution.
 

### 2022-01-31 revision/review_response.Rmd
revisions done, need to proof and format for resubmission

--- 2022-01-30 revision/review_response.Rmd
@@
 We provide a point-by-point response to these comments below.
-In general, most of the revisions address concerns raised by all three reviewers, including 1) improvement in writing style for clarity and brevity, and 2) additional information on sampling methods.
+In general, most of the revisions address concerns raised by all three reviewers, including 1) numerouis edits to improve the writing for clarity and brevity, and 2) additional information on sampling methods.
 We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.
@@
 
-* __Response__: Thank you for your comments. We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted. Please see the details below.
+* __Response__: Thank you for your comments.
+We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted.
+Note that many of the edits in response to the specific comments from the other reviewers are used to address some of the comments below.
 
@@
 
+* __Response__: Abstract was edited.
+
 Introduction- needs a complete rewrite- it takes more than 3 pages to get to the point that there was a spill. The general background is interesting but you need to grab the readers attention first.
+
+* __Response__: Introduction was shortened and edits were made to improve clarity.
 
@@
 
+* __Response__: Please see the responses to reviewer 2, particularly regarding the edits to the discussion sections.
+
 In all, it seemed like an interesting exercise to describe all the patterns, but this does not make a quality manuscript that will stand the test of time.
+
+* __Response__: We are confident that the edits made in response to your comments and those from the other reviewers have made this manuscript a more compelling and valuable contribution.
 
@@
 * __Response__: Thank you for this comment.
-We agree this section is important as it relates the timeline of events in a plausible manner, although it is definitely speculative without more quantitative results linking the pieces.
+We agree this section is important, although it is definitely speculative without more quantitative results that more definitively link the pieces.
 As such, we have moved the content to the discussion and amended to text to include a more nuanced description of the trends, including what information may be needed to support the narrative.
-This includes information from the final paragraph of the discussion, which was removed.
+This includes information from the final paragraph of the discussion, which was shortened and combined with the prior paragraph.
 The new section is included from lines xxx to xxx.

### 2022-01-31 manu-draft.Rmd
revisions done, need to proof and format for resubmission

@@
 
-From March 30th to April 9th, 2021, 814 million liters of legacy phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA).
+Legacy mining facilities pose significant risks to aquatic resources.
+From March 30th to April 9th, 2021, 814 million liters of phosphate mining wastewater and marine dredge water from the Piney Point facility were released into lower Tampa Bay (Florida, USA).
 This resulted in an estimated addition of 186 metric tons of total nitrogen, exceeding typical annual external nitrogen load estimates to lower Tampa Bay in a matter of days.
-An initial phytoplankton bloom (non-harmful diatoms) was first observed in April in the lower Bay.
+An initial phytoplankton bloom (non-harmful diatoms) was first observed in April.
 Filamentous cyanobacteria blooms (*Dapis* spp.) peaked in June, followed by a bloom of the red tide organism *Karenia brevis*.
@@
 Seagrasses had minimal changes over the study period.
-By comparing these results to baseline environmental monitoring data, we conclude that many of the biological responses observed after the release from Piney Point are abnormal relative to historic conditions.
+By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to external nitrogen loads into Tampa Bay.
 
@@
 Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11]. 
-Phosphate fertilizer is produced through the "wet process" reaction to create phosphoric acid by treating mined phosphate rock with sulfuric acid [@Burnett01;@Perez16].
-The process generates large amounts of waste, creating approximately one unit of phosphoric acid per five units of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O).
-This waste is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
+The production of phosphate fertilizer generates large amounts of phosphogpysum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
 Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].
-In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than those of natural surface waters [@Burnett01;@Beck18b].
+In addition to elevated phosphorus concentrations, other nutrients, contaminants, and radionuclides may be present at values much higher than natural surface waters [@Burnett01;@Beck18b].
 Many of these gypstacks no longer support active mining and aging infrastructure combined with climate change and seasonal stressors (e.g., heavy precipitation events) have reduced the capacity of these facilities to maintain water on site.
-Environmental and human health risks associated with these stacks can occur through controlled or uncontrolled releases to surface waters or groundwater contamination through leaching from unlined or poorly maintained stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
-
-Ongoing threats and challenges to protecting water quality of northern Gulf of Mexico estuaries persist despite recent environmental recovery.
-Successful restoration has been accomplished through ecosystem management paradigms that are based primarily on the control of nutrient pollutants from atmospheric deposition, stormwater, and wastewater.
-Nitrogen inputs from external sources are well understood as drivers of algal blooms in coastal environments that can degrade water quality, having a negative effect on inter- and subtidal habitats [@Nixon95;@Howarth06;@Parker12;@Greening14].
-Seagrasses in particular are valuable habitat-defining species that provide many ecosystem services, but are especially vulnerable to the impacts of nutrient pollution on water quality because of cascading negative effects of nitrogen, phytoplankton growth and persistence, water clarity, light limitation, and epiphyte loading on seagrass growth and survival [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
-Historical gains in seagrass coverage in southwest Florida estuaries have been achieved through public-private partnerships and consensus-based approaches to science applications that seek to limit the total nutrient loads delivered to surface waters [@Janicki96;@Greening16;@Tomasko18;@Tomasko20].
-Together, these efforts have resulted in the long-term recovery of Tampa Bay and other Gulf Coast estuaries through a reduction in external nitrogen loads, improvements in water clarity, and baywide expansion of seagrass coverage to benchmark targets established for the region [@Greening14;@Sherwood17].
-The three contiguous National Estuary Programs of southwest Florida (Tampa Bay, Sarasota Bay, Charlotte Harbor) and their partners have been instrumental in coordinating efforts to address legacy pollutants and current threats to the long-term protection of estuarine resources.
+Numerous studies have documented the environmental and human health risks associated with these stacks [@Tayibi09;@Sanders13;@elzrelli15;@Perez16;@Beck18b].
 
@@
 By 2001, an estimated 36 million metric tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].
-Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed.
 Effective management and final closure of these facilities are imperative to reduce threats to prior ecosystem recovery efforts and investments.
@@
 
-More recently, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
-In response, the Florida Department of Environmental Protection (FDEP) authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 29th, 2021 to release water from the southern gypstack directly into lower Tampa Bay to prevent catastrophic failure of the facility.
+In March 2021, leakages were detected from a tear in the plastic liner of the southern holding pond (NGS-S) at Piney Point.
 At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack.
-Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of both phosphorus and nitrogen.
+Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of nitrogen, in addition to phosphorus.
 Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay.
-From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released under FDEP authorization to lower Tampa Bay.
+From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay following an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) authorized by the Florida Department of Environmental Protection (FDEP).
 Over this ten day period, an estimated 186 metric tons (205 tons) of nitrogen were delivered to the bay, exceeding contemporary annual estimates of external nutrient loads to lower Tampa Bay in a matter of days [@tbep0417].
@@
 As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
+Our analysis evaluated datasets that are descriptive of the vulnerability of seagrasses to nutrient pollution though cascading negative effects of nitrogen, phytoplankton growth and persistence, and water clarity on seagrass growth and survival [@Dixon95;@Kenworthy96;@Greening06;@Beck18a].
 A timeline of events is provided, which is supported by the quantitative results from 2021 routine and response-based monitoring of conditions in and around Port Manatee, FL -- the focal point of emergency releases from the Piney Point facility.
-The results from this paper provide examples of anticipated short-term responses to acute nutrient loadings from legacy mining facilities in the broader context of historical conditions that may influence system response to these events.
+The results from this paper provide an unprecedented chronology of short-term estuarine response to acute nutrient loadings from legacy mining facilities, where context would not have been possible without the long-term monitoring datasets available for the region.
 
@@
 Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
-These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point.
+These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point, e.g., were July conditions significantly different from April.
 Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
@@
 All datasets used in this study are available from an open access data archive hosted on the Knowledge Network for Biocomplexity [@Beck21d].
-Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a GitHub repository.
+Materials for reproducing the analyses, figures, tables, and other content in this paper are provided in a [GitHub repository](https://github.com/tbep-tech/piney-point-manu).
 Finally, the Piney Point Environmental Monitoring Dashboard can be used to view all data included in this paper through an interactive, online application [@Beck21c].
@@
 Similar differences for total phosphorus, TN, and chl-a were observed when comparing stack conditions with those of the ambient conditions in Tampa Bay.
-These contrasts provided a context for the monitoring data collected in 2021.
 
-Samples collected in the bay between April through September 2021 indicated that water quality conditions were unusual and outside of normal values expected for each month.
+Samples collected in the bay between April through September 2021 indicated that water quality conditions were outside of normal values expected for each month.
 A total of `r nwq` samples were collected and analyzed for chl-a, dissolved oxygen, TN, total phosphorus, total ammonia nitrogen, nitrate/nitrite, pH, salinity, Secchi depth, and temperature (Table \@ref(tab:wqsumtab)).
@@
 Physical parameters (salinity, temperature) and inorganic nitrogen (ammonia, nitrate/nitrite) were more often in normal ranges, although initial time series showed much higher concentrations for ammonia in April near Area 1.
-Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L, about three orders of magnitude above baseline (Figures S2, S3), similar to the discharge measurements in Table \@ref(tab:stktab)).
+Ammonia concentrations near the point of discharge were observed in excess of 10 mg/L in April, about three orders of magnitude above baseline (Figures S2, S3), similar to the discharge measurements in Table \@ref(tab:stktab).
 Inorganic nitrogen did not persist at high concentrations past April as it was likely rapidly utilized by phytoplankton (see below).
@@
 The initial diatom bloom did not persist past April.
-Concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with the increase in *K.
-brevis* concentrations.
+Chlorophyll concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with an increase in *K.
+brevis* concentrations to bloom levels.
 Many Secchi observations in Area 1 were lower than normal in April and July.
 Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in  April, May, and July.
+
 Statistical comparisons between months for seasonally-corrected observations of TN, chl-a, and Secchi depth (Table \@ref(tab:wqcmptab)) supported the results in Figure \@ref(fig:wqgam).
@@
 
-A typical pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex).
-Transect S3T6 is located less than one kilometer to the north of Port Manatee.
+A typical temporal pattern for macroalgae and seagrass observed at many of the transects is shown in Figure \@ref(fig:trnex), using transect S3T6 near Port Manatee as an example.
 Macroalgal abundances changed over the course of sampling similar to the remainder of transects sampled during the study.
@@
 No transects were sampled in Area 2 to the north of Piney Point and no transects were sampled past September in Area 1 given allocated sampling effort following projected dispersal patterns of the discharge from the TBCOM simulations.
-Macroalgal dominance varied across the months, similar to the example in Figure \@ref(fig:trnex).
 Red macroalgae was the dominant group across all months and areas, with the highest frequency occurrences observed in April (`r fomoarea1apr['Red']`\% in Area 1, `r fomoarea3apr['Red']`\% in Area 3).
@@
 Green macroalgae had the second lowest frequency occurrence, although it increased slightly by the end of the study period (`r fomoarea1sep['Green']`\% in September in Area 1, `r fomoarea3oct['Green']`\% in October in Area 3).
-Brown macroalgae was only observed at one transect.
 For seagrass, both areas had generally stable total frequency occurrence.
@@
 Slight changes in frequency occurrence in Area 3 were observed for all species starting in July, with a slight reduction in frequency occurrence of turtle grass and an increase in shoal grass and manatee grass.
-Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed only for macroalgae (Tables S1, S2).
+Statistical analyses with multiple comparison tests confirmed the general trends described above, with significant changes observed over time only for macroalgae (Tables S1, S2).
 Tests using Braun Blanquet cover estimates confirmed the results from the frequency occurrence estimates (Tables S3, S4).
@@
 By May 23, bloom concentrations of *K.
-brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th, after which concentrations declined (Figure \@ref(fig:redtide)b).
+brevis* were observed in lower Tampa Bay (lower/middle bay boundary Figure \@ref(fig:map)b), with concentrations peaking (10$^6$ to 10$^7$ cells/L) by the week of July 4th in middle Tampa Bay, after which concentrations declined (Figure \@ref(fig:redtide)b).
 The increase in *K.
@@
 Median cell concentrations for most years were well below 1,000 cells/L.
-The two highest concentrations were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.
+The two highest concentrations in the long-term record were observed in 1971 (20 million cells/L) and 2021 (17.6 million cells/L), both being over an order of magnitude above the high category.
 Cumulative rainfall and associated inflow from the main rivers entering Tampa Bay in 2021 were below historical values (2006 - 2020) in the months preceding the highest bloom concentrations (i.e., January to June, Figure \@ref(fig:redtide)c, d).
@@
 These results were evidenced by taxonomic enumeration of phytoplankton samples collected near Port Manatee.
-Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a.
-However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities or changing availability of nutrient ratios.
+Water quality indicators improved slightly following the decrease in diatoms in late April, as noted by relatively lower concentrations of TN and chl-a as the bloom dispersed.
+However, filamentous cyanobacteria biomass increased after the initial diatom bloom and peaked in June (Figure \@ref(fig:nutrientflow)d), suggesting a shift of nutrients from phytoplankton to drift macroalgae communities or changing availability of nutrient ratios creating favorable conditions for macroalgae growth [@Valiela97;@Cohen06].
 During peak macroalgae growth, TN and chl-a concentrations remained relatively low as nutrients were likely retained in macroalgae, until late June and early July when *K.
@@
 The co-occurring decline in macroalgae and increase in *K.
-brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter.
+brevis* suggests a release of nutrients from the former that could have stimulated growth of the latter, although residual nutrients from the initial release from Piney Point were likely still available [@Liu21].
 Finally, conditions were relatively stable in August and September with relatively improved water quality conditions and no dominant algal blooms.
@@
 Variation in observed concentrations of nutrients is complex given that high concentrations may suggest availability to support phytoplankton growth, whereas low concentrations may imply cycling of available nitrogen in organic forms already utilized by different taxa, including macroalgae [@Valiela97;@Cohen06].
-
-There were also concerns that the release from Piney may have contributed to the persistence and intensity of *K.
-brevis*, having negative effects on fisheries resources in June and July (Figure \@ref(fig:redtide)).
-Fisheries resources in Tampa Bay have previously been negatively affected by red tide [e.g., in 2005, @Flaherty11;@Schrandt21].
-For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
-Fish community structure and species composition did not differ compared to a pre-impact period, although HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
-brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
-Prior blooms in Tampa Bay were more localized and *K.
-brevis* was at lower abundances in comparison to the 2021 bloom event, potentially mitigating exposure of fishes to related harmful conditions.
-In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].
-Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen.
-@Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.
-Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality.
 
@@
 
+There were also concerns that the release from Piney may have contributed to the persistence and intensity of *K.
+brevis*, having negative effects on fisheries resources in June and July (Figure \@ref(fig:redtide)).
+Fisheries resources in Tampa Bay have previously been negatively affected by red tide [e.g., in 2005, @Flaherty11;@Schrandt21].
+For past Piney Point events, @Switzer11 evaluated nekton communities in Bishop Harbor from November 2003 to October 2004 following discharge to this subembayment.
+Fish community structure and species composition did not differ compared to a pre-impact period, although HAB species (*Prorocentrum minimum*, *Heterosigma akashiwo*), including *K.
+brevis* and diatoms, were observed in Bishop Harbor during this time [@Garrett11].
+Prior blooms in Tampa Bay were more localized and *K.
+brevis* was at lower abundances in comparison to the 2021 bloom event, potentially mitigating exposure of fishes to related harmful conditions.
+In Sarasota Bay to the south, fish activity measured by passive acoustic methods was significantly lower during a 2018 red tide event as compared to pre-bloom levels [@Rycyk20].
+Water quality conditions before and after passage of tropical storm Elsa may have also contributed to fish kills by reducing bottom-water dissolved oxygen.
+@Stevens06 documented impacts of a category 4 storm on fish resources in the Charlotte Harbor estuary, although tropical storm Elsa was much smaller and fish kills were documented prior to and after arrival of the storm.
+Lack of continuous monitoring data for bottom waters in Tampa Bay prevents a more detailed assessment of impacts of the storm on water quality.
+
 Establishing causal linkages between the nutrient inputs from Piney Point and the severity of the *K.
@@
 In addition to Tampa Bay [@Garrett11;@Switzer11], other Gulf Coast estuaries have been affected by pollutants from unanticipated gypstack releases.
-Grand Bay is a 7500 hectare protected area in southern Mississippi that has been exposed to phosphorus-rich and highly acidic water from a defunct gypstack [@Dillon15;@Beck18b].
-Two spills have occurred in Grand Bay, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the gypstack was exceeded again with heavy rainfall.
-Massive fish kills were observed and likely related to low pH of the water released.
-Unlike Piney Point, inorganic nitrogen concentrations of the release were low due to a different fertilizer production method and concerns of the long-term impacts focused primarily on heavy loads of orthophosphate [@Dillon15].
-Phosphate loads to Tampa Bay from Piney Point were similar in magnitude to the nitrogen loads, although concentrations were within normal baseline ranges within a month after the release stopped (Figures S2, S3).
-The fate of the phosphorus in Tampa Bay is less understood than that of nitrogen.
-Regardless, the historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
+For example, two spills have occurred in Grand Bay, Mississippi, the first in 2005 following failure of the retaining walls after a heavy rain event and the second in 2012 after passage of Hurricane Isaac when the holding capacity of the local gypstack was exceeded again with heavy rainfall [@Dillon15;@Beck18b].
+The historical context of Grand Bay is similar to Piney Point and other international examples, e.g., Huelva estuary in Spain [@Perez10;@Perez16].
 Legacy wastewater from fertilizer production has been poorly maintained at some facilities and long-term plans are insufficient to safely dispose of remnant pollutants that pose a risk of significant impacts to coastal resources that increases over time.
 These are not isolated examples and enhanced regulatory oversight is needed to safely and effectively close these types of facilities [@Nelson21].
+Local, regional, and state partners should continue to pursue management and policy actions that can mitigate the continued threats of these facilities to the health of coastal resources.
+These efforts are critical to managing Gulf of Mexico ecosystems given past successes and the need to address ongoing threats of climate change, human population growth, habitat loss, severe weather events, and recurring pollutant sources.
 

### 2022-02-01 revision/review_response.Rmd
revisions proofed

@@
 ```{r echo = F, cache = F, eval = F}
-spelling::spell_check_files('revision-response.Rmd')
+spelling::spell_check_files(here::here('revision/review_response.Rmd'))
 ```
@@
 We provide a point-by-point response to these comments below.
-In general, most of the revisions address concerns raised by all three reviewers, including 1) numerouis edits to improve the writing for clarity and brevity, and 2) additional information on sampling methods.
-We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.
+In general, most of the revisions address concerns raised by all three reviewers, including 1) numerous edits to improve the writing for clarity and brevity, and 2) additional information on sampling methods.
+We have also addressed concerns of Reviewer 2 about potential nutrient cycling by adding and expanding a section in the discussion and concerns of Reviewer 4 regarding additional detail on the simulation modelling.
+
+All page and line numbers in the reviewer comments refer to the original submission.  Page and line numbers noted in our response refer to the revised submission.
 
@@
 We have made revisions to improve the writing, by reducing length in many locations and making additional edits where noted.
-Note that many of the edits in response to the specific comments from the other reviewers are used to address some of the comments below.
+Many of the edits in response to the specific comments from the other reviewers are used to address some of the comments below.
 
@@
 p. 7- how were the Karenia data quantified...they didn't just appear in a database.
+
+* __Response__: Text was added: "HAB sampling typically occurs in response to bloom events or fish kills with extensive quality control of cell counts conducted by FWC-FWRI [additional details in @Stumpf22].
+HAB data were restricted to Tampa Bay boundaries and over 90\% of the samples were collected within one meter of the surface."
 
@@
 * __Response__: Thank you for this comment.
-We agree this section is important, although it is definitely speculative without more quantitative results that more definitively link the pieces.
-As such, we have moved the content to the discussion and amended to text to include a more nuanced description of the trends, including what information may be needed to support the narrative.
+We agree this section is important, although it is definitely speculative without more quantitative results that definitively link the pieces.
+As such, we have moved the content to the discussion and expanded to text to include a more nuanced description of the trends, including what information may be needed to support the narrative.
 This includes information from the final paragraph of the discussion, which was shortened and combined with the prior paragraph.
@@
 
-* __Response__: Our point was to suggest that mining products in Florida are generally exported and have little benefit for the local communities, aside from localized economic benefits.
+* __Response__: Our point was to suggest that mining products in Florida are generally exported and have little benefit for the local communities.
 However, this may be a bit editorialized and was removed.
@@
 * __Response__: Agreed, we have removed "unanticipated".
-We agree that the releases often occur after consultation or approval from regulatory bodies, but they are generally unanticipated in that site management has not planned for these events, nor the potential impacts on the environment.
+We agree that the releases often occur after consultation or approval from regulatory bodies, but they are generally unanticipated in that long-term site management has not planned for these events, nor the potential impacts on the environment.
 
@@
 
-* __Response__: Changed to "released under FDEP authorization".
+* __Response__: Changed to "From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay following an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) authorized by the Florida Department of Environmental Protection (FDEP)."  A similar, redundant sentence earlier in the paragraph was removed.
 
@@
 
-     "Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
+     "Numerous studies, as well as the successful nutrient management paradigm, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
 As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production."
@@
 Our initial rationale for using data from Tampa was a longer period of record.
-However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data with historical observations.
+However, we have revised the historical baseline in the plot to include only years 2006 to 2020, consistent with our comparison of the water quality data to historical observations.
 Also, the NWS location is in Ruskin, Florida and is farther from the bay proper than the location in St.
@@
 
-* __Response__: The weekly aggregation was only done to allow plotting and comparison between variables on a common scale (e.g., Figure 7 and 8).
+* __Response__: The weekly aggregation was only done to allow plotting and comparison between variables on a common scale (e.g., Figure 6 and 8 in the revised text).
 None of the statistical tests used data with fixed sample dates.
@@
 
-     "Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
-These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point."
+     "Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure 1a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
+These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point, e.g., were July conditions significantly different from April?"
 
@@
 An important distinction that we make in this paper is that, although blooms in Tampa Bay are not uncommon, the severity of the 2021 bloom was very likely related to the favorable conditions in the bay prior to July.
-These favorable conditons were related to 1) nutrient availability from Piney Point, and 2) high salinity conditions from low rainfall.
+These favorable conditions were described in the paper as 1) nutrient availability from Piney Point, and 2) high salinity conditions from low rainfall.
 The passage of tropical storm Elsa was also a confounding factor.
@@
 * __Response__: Seagrass transect monitoring in Tampa Bay has occurred annually since the mid-1990s using similar methods as in this paper [i.e., @Sherwood17].
-The primary difference for the routine monitoring is the transect length, which is typically much longer than the 50m used in the 2021 rapid design.
+The primary difference for the annual monitoring is the transect length, which is typically much longer than the 50m used in the 2021 rapid design.
 This shorter distance was chosen to allow for quicker sampling times and greater coverage in response to the rapidly changing conditions in 2021.
-Thus, we consider the rapid design as providing data that are of similar quality to the long-term data, with notable exceptions indicated in the mansucript (i.e., no seasonal estimates).
+Thus, the rapid design provides data that are similar quality to the long-term data, with notable exceptions indicated in the manuscript (i.e., no seasonal estimates).
 Our rationale for evaluating Braun Blanquet coverage was to verify the results obtained evaluating frequency occurrence estimates.
@@
 
-* __Response__: Added text "...or changing availability of nutrient ratios."
+* __Response__: Added text "...or changing availability of nutrient ratios creating favorable conditions for macroalgae growth [@Valiela97;@Cohen06]."
 
@@
 * __Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".
-This was based on chlorohpyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.
+This was based on chlorophyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.
 The word limit for the abstract prevents these details, but they are presented in text (i.e., line xxx and xxx).
@@
      Lines xxx were edited to include some of the above details: "The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
-Preliminary model results were reported in @Liu21."
+Preliminary model results for Piney Point are reported in @Liu21 and previous model veracity testing was described in @Chen18 and @Chen19 (and references therein)"
 
@@
 
-* __Response__: Our methods for estimating inflows into Tampa Bay are based on methods that are similarly presented in a technical report published by the Tampa Bay Estuary Program in 2012 [@Janicki12].
+* __Response__: Our methods for estimating inflows into Tampa Bay are based on those similarly presented in a technical report published by the Tampa Bay Estuary Program [@Janicki12].
 This report provides the basis for compliance assessment of our TMDL for reporting to the Florida Department of Environmental Protection and represents our best estimate at hydrologic load inputs to the bay.
@@
 However, this does not provide an assessment of biomass or coverage.
-Weight data were collected for select macroalgae samples during the 2021 surveys and a preliminary analysis (see the Seagrass -> Biomass estimates tab [here](https://tbep-tech.github.io/piney-point-analysis/trend)) suggested similar trends as the frequnecy occurrence results.
+Weight data were collected for select macroalgae samples during the 2021 surveys and a preliminary analysis (see the Seagrass -> Biomass estimates tab [here](https://tbep-tech.github.io/piney-point-analysis/trend)) suggested similar trends as the frequency occurrence results.
 Because we did not have comparable weight data for seagrasses, the analysis was not included.
@@
 However, this was not included due to differences in reporting methods that have changed over time.
-Online reporting became an option in recent years and we considered the results incomparable to the earlier data that were based on phone reporting methods.
+Online reporting became available in recent years and we considered the results incomparable to the earlier data that were based on phone reporting methods.
 
@@
 
-     We have added some text below the existing pararaph to provide more context: "Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
+     We have added some text below the existing paragraph to provide more context: "Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
 Severity in estuaries is typically less than Gulf waters as *K.

### 2022-02-01 manu-draft.Rmd
revisions proofed

@@
 ```{r echo = F, cache = F, eval = F}
-spelling::spell_check_files('manu_draft.Rmd')
+spelling::spell_check_files(here::here('manu-draft.Rmd'))
 ```
@@
 Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
-The production of phosphate fertilizer generates large amounts of phosphogpysum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
+The production of phosphate fertilizer generates large amounts of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
 Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].
@@
 The goal is to describe the results of monitoring data of surface waters collected in response to the event to assess relative deviation of current conditions from long-term, seasonal records of water quality, phytoplankton, and seagrass/macroalgae datasets available for the region.
-Numerous studies, as well as the successful nutrient management paradigm for Tampa Bay, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
+Numerous studies, as well as the successful nutrient management paradigm, have demonstrated nitrogen-limitation in Tampa Bay and the system is generally considered phosphorus enriched [@Wang99;@Greening06;@Greening14].
 As such, we focus on nitrogen in our analyses as the identified limiting nutrient for Tampa Bay and its potential to create water quality conditions unfavorable for seagrass growth due to enhanced algal production.
@@
 The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
-Preliminary model results were reported in @Liu21.
+Preliminary model results for Piney Point are reported in @Liu21 and previous model veracity testing was described in @Chen18 and @Chen19 (and references therein).
 
@@
 Harmful Algal Bloom (HAB) data for *Karenia brevis* were obtained from event-based monitoring samples from the [FWC-FWRI HAB Monitoring Database](https://myfwc.com/research/redtide/monitoring/database/).
-Bloom sizes for HAB species were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
+HAB sampling typically occurs in response to bloom events or fish kills with extensive quality control of cell counts conducted by FWC-FWRI [additional details in @Stumpf22].
+HAB data were restricted to Tampa Bay boundaries and over 90\% of the samples were collected within one meter of the surface.
+Bloom sizes for *K.
+brevis* were described qualitatively as low/medium/high concentrations based on [FWC breakpoints](https://myfwc.com/research/redtide/statewide/) at 10,000/100,000/1,000,000 cells/L.
 Fish kill reports were obtained from the FWC [online database](https://public.myfwc.com/fwri/FishKillReport/searchresults.aspx).
@@
 GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year.
-The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods as @Murphy19].
+The continuous year value was modeled with a thin plate regression spline and the day of year value was modeled with a cyclic spline [following similar methods as @Murphy19].
 The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.
@@
 Differences in observations between months for April to September for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13].
-These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point, e.g., were July conditions significantly different from April.
+These tests were used to statistically characterize the temporal progression of changes in the bay following release from Piney Point, e.g., were July conditions significantly different from April?
 Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons.
@@
 For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.
-Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, altough median concentrations for each week in April were less than 10 $\mu$g/L.
+Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, although median concentrations for each week in April were less than 10 $\mu$g/L.
 The initial chlorophyll peak was associated with a localized phytoplankton bloom generally dominated by diatoms.
@@
 There is no historical macroalgae record for Tampa Bay that is comparable to the spatial and temporal resolution of the 2021 samples.
+Only annual historical data are available for seagrasses, with no seasonal data comparable to the results herein.
 

### 2022-02-02 manu-draft.Rmd
Tampa Bay Watch to acknowledgments

@@
 We thank the administrative staff, field crews, and laboratory staff from the Florida Department of Environmental Protection, Environmental Protection Commission of Hillsborough County, Parks and Natural Resources Department of Manatee County, Pinellas County Division of Environmental Management, Fish and Wildlife Research Institute of the Florida Fish and Wildlife Conservation Commission, City of St.
-Petersburg, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
+Petersburg, Tampa Bay Watch, Sarasota Bay Estuary Program, Environmental Science Associates, University of South Florida, University of Florida, and New College of Florida.
 Funding was provided to the University of Florida from the Ocean Conservancy and the National Science Foundation (project ID 2130675).

### 2022-02-04 manu-draft.Rmd
Minor edits - will continue on Monday
Trying to find a good estimate of total holding capacity of stacks, 2020 DEP presentation to RPC give a total of 1.3 billion gallons capacity for all of PP, so that's about 4.9 billion liters... - https://drive.google.com/drive/u/0/folders/1NC47EqfqivSDJ8GP_0H6AbUv3GO5chw_ - slide 16, p 8

@@
 Seagrasses had minimal changes over the study period.
-By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to external nitrogen loads into Tampa Bay.
+By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads into Tampa Bay.
 
@@
 Wastewater byproducts from mining are a global threat to the quality of surface and groundwater resources [@Tayibi09;@Hudson11].
-The production of phosphate fertilizer generates large amounts of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding millions of liters of process water.
+The production of phosphate fertilizer generates large amounts of phosphogypsum waste (CaSO$_4\cdot$ H$_2$O) that is typically stored on-site in large earthen stacks (gypstacks) capable of holding hundreds of millions of liters of process water.
 Water quality in gypstacks can vary depending on processing method used at the mining facility, background geological characteristics of the region, and on-site practices for managing stormwater or other activities that can introduce additional materials to the holding ponds [@Henderson04;@Perez10].

### 2022-02-07 R/figures.R
reproc figs to correct order for revision folder

@@ -270,10 +270,6 @@ jpeg(here('revision/Fig1.jpeg'), height = 4.2, width = 9, family = 'serif', unit
 print(pout)
 dev.off()
 
-# timeline copy -----------------------------------------------------------
-
-file.copy(here('figs/timeline.jpg'), here('revision/Fig2.jpg'))
-
 # nutrients, chloropyll, secchi map ---------------------------------------
 
 # nonbay stations
@@ -411,7 +407,7 @@ jpeg(here('figs/wqmap.jpeg'), height = 7, width = 12, units = 'in', res = 500, f
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig3.jpeg'), height = 7, width = 12, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision/Fig2.jpeg'), height = 7, width = 12, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -445,7 +441,7 @@ jpeg(here('figs/wqgam.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, 
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig4.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision/Fig3.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -484,7 +480,7 @@ jpeg(here('figs/trnex.jpeg'), height = 7, width = 8, units = 'in', res = 500, fa
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig5.jpeg'), height = 7, width = 8, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision/Fig4.jpeg'), height = 7, width = 8, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -616,7 +612,7 @@ jpeg(here('figs/trnfrq.jpeg'), height = 6, width = 9, units = 'in', res = 500, f
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig6.jpeg'), height = 6, width = 9, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision/Fig5.jpeg'), height = 6, width = 9, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
[…]

### 2022-02-09 revision/review_response.Rmd
formatting done for revision resubmission to MPB

@@
 
-All page and line numbers in the reviewer comments refer to the original submission.  Page and line numbers noted in our response refer to the revised submission.
+All page and line numbers in the reviewer comments refer to the original submission.  Page numbers noted in our response refer to the revised submission.
 
@@
 This includes information from the final paragraph of the discussion, which was shortened and combined with the prior paragraph.
-The new section is included from lines xxx to xxx.
+The new section begins on page 18.
 
@@
 We have included additional information about the data collection procedures and how the Tampa Bay Estuary Program worked to facilitate quality of the data.
-Please see the additions starting on line xxx.
+Please see the additions starting on page 6.
 
@@
 Please see the addition of information on *K.
-brevis* sampling on line xxx provided in response to reviewer 1.
+brevis* sampling at the top of page 7 in response to reviewer 1.
 
@@
 
-* __Response__: Yes, red tide was observed in the Gulf of Mexico prior to the occurrence in Tampa Bay.  Content in the red tide section was edited to make this clear (line xxx):
+* __Response__: Yes, red tide was observed in the Gulf of Mexico prior to the occurrence in Tampa Bay.  Content in the red tide section was edited to make this clear (page 15):
 
@@
 The passage of tropical storm Elsa was also a confounding factor.
-It is nearly impossible to identify a single factor causing the observed conditions, but we describe the considerations in the discussion (lines xxx to xxx).
+It is nearly impossible to identify a single factor causing the observed conditions, but we describe the considerations in the discussion (page 23).
 
@@
 Our rationale for evaluating Braun Blanquet coverage was to verify the results obtained evaluating frequency occurrence estimates.
-Line xxx was modified to make this clear:
+The text was modified to make this clear (page 15):
 
@@
 * __Response__: This line was changed to "An initial phytoplankton bloom (non-harmful diatoms) was first observed...".
-This was based on chlorophyll measurements in the vicinity of Port Manataee and cell concentrations enumerated to taxa from laboratory analysis.
-The word limit for the abstract prevents these details, but they are presented in text (i.e., line xxx and xxx).
+This was based on chlorophyll measurements in the vicinity of Port Manatee and cell concentrations enumerated to taxa from laboratory analysis.
+The word limit for the abstract prevents these details, but they are presented in text (page 18).
 
@@
 
-     Lines xxx were edited to include some of the above details: "The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
+     The main text was edited to include some of the above details (page 5): "The modeled plume evolution web product (http://ocgweb.marine.usf.edu/~liu/Tracer/) served as the principal guidance for coordinating the data collection during the event.
 Preliminary model results for Piney Point are reported in @Liu21 and previous model veracity testing was described in @Chen18 and @Chen19 (and references therein)"
@@
 * __Response__: Our initial submission included supplement material that described the sample collection and processing in more detail.
-We have moved that content from the supplement to the main text, beginning on lines xxx:
+We have moved that content from the supplement to the main text, beginning on page 6:
 
@@
 * __Response__: Frequency occurrence is a routine reporting metric used by the Tampa Bay Estuary Program to summarize the long-term annual transect monitoring [see @Johansson16;@Sherwood17].
-Additional text on line xxx was added to make this clear:
+Additional text on page 10 was added to make this clear:
 
@@
 
-     We realize that additional metrics could have been used to summarize the data and we presented trend assessments using the Braun-Blanquet abundance estimates as confirmation for the results from the frequency occurrence analyses (line xxx).
+     We realize that additional metrics could have been used to summarize the data and we presented trend assessments using the Braun-Blanquet abundance estimates as confirmation for the results from the frequency occurrence analyses (Tables S3 and S4).
 However, this does not provide an assessment of biomass or coverage.
@@
 Because we did not have comparable weight data for seagrasses, the analysis was not included.
-It is also worth mentioning that coverage estimates are obtained every two years for Tampa Bay as an annual snapshot (noted on line xxx in the discussion), but these are not of sufficient temporal resolution to support the analyses in this paper.
+It is also worth mentioning that coverage estimates are obtained every two years for Tampa Bay as an annual snapshot (noted at the top of page 21 in the discussion), but these are not of sufficient temporal resolution to support the analyses in this paper.
 
      There were limitations to our analysis that were noted in the original draft, primarily the lack of long-term historical data on seasonal trends for seagrasses and macroalgae.
-For example, line xxx in the original discussion: "Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release."
+For example, page 22,line 52 in the original discussion: "Long-term monitoring data describing normal seasonal variation in macroalgae are unavailable and we cannot distinguish between seasonal and interannual changes and those in potential response to the Piney Point release."
 
@@
 * __Response__: This is in reference to total ammonia nitrogen, as was provided by the water quality sampling.
-We have made a note on line xxx in the methods that any mention of ammonia refers to total ammonia nitrogen.
+We have made a note in the methods section (top of page 6) that any mention of ammonia refers to total ammonia nitrogen.
 
@@
 * __Response__: Previous content in the discussion below this paragraph provided some historical context on red tide events in the region.
-Specifically, starting on line xxx: "Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf..."
+Specifically, on page 23: "Occurrence of this species has historically been spatially distinct, with blooms originating in subsurface water offshore on the West Florida Shelf..."
 

### 2022-02-09 manu-draft.Rmd
formatting done for revision resubmission to MPB
revision submitted to MPB

@@
 Seagrasses had minimal changes over the study period.
-By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads into Tampa Bay.
+By comparing these results to baseline environmental monitoring data, we demonstrate adverse water quality changes in response to abnormally high and rapidly delivered nitrogen loads.
 
@@
 
-## Data Analysis
+## Data analysis
 
@@
 Peaks in dissolved oxygen saturation were also observed as an indicator of elevated phytoplankton production [@Kemp80], particularly in July with the peak *K.
-brevis* bloom (Figures S2, S3).
+brevis* bloom (Figures S2d, S3d).
 Of note is that inorganic species of nitrogen, mainly ammonia, were only present at high concentrations in early April.

### 2022-03-09 revision2/review_response.Rmd
starting second revision

---
output: 
  bookdown::word_document2:
    reference_docx: response_styles.docx
    number_sections: false
always_allow_html: true
urlcolor: blue
link-citations: true
bibliography: ../refs.bib
---

```{r setup, echo = F, warning = F, message = F, results = 'hide'}
# figure path, chunk options
knitr::opts_chunk$set(fig.path = 'figs/', warning = F, message = F, echo = F, cache = F, dev.args = list(family = 'serif'), dpi = 300, warning = F)
```

```{r echo = F, cache = F, eval = F}
spelling::spell_check_files(here::here('revision2/review_response.Rmd'))
```

We sincerely thank the reviewers again for providing comments on our manuscript.  As before, we provide a point-by-point response to these comments below.

All page and line numbers in the reviewer comments refer to the revised manuscript from the previous submission.

## Reviewer 1

This revised version is significantly improved compared to the first submission. However, there are still some issues that need addressing. It remains an overall descriptive study with little or no mechanistic understanding of why the bloom happened- but perhaps that is OK. The writing is better.

Title- no one other than local people will know what a Piney Point is. Needs revision to something along the lines of: Initial estuarine response to a spill of phosphate mining waste: time course of a red tide following  the Piney Point spill in Tampa Bay, USA

The graphics abstract needs a lot of work. It has far too many words and too much detail. Such an abstract is intended as a tease, not as a way to condense every detail of the paper. The little blurred icons are not helpful- I can't even tell what some of the blobs are supposed to represent. No one needs dates and every detail for this presentation. It should not be a miniaturized poster. Simplify, simplify. Key points: the spill and the bloom followed.

Figure 8d looks like K brevis occurred on only 1 day or 2-- odd.

The paper lays out a lot of background and will be helpful to follow up papers.

## Reviewer 2

This is a nice paper that addresses the effects of a high visibility/publicity wastewater spill on an iconic estuary (Tampa bay). Most of my comments were addressed, but there are still a few issues that need additional attention.

1. The authors didn't address my question of how the plume model was verified, other than to point to a couple of papers. I read those papers and don't fully understand them as I am not a physical oceanographer, but as far as I could tell, neither paper clearly stated how verification was made using field observations. Given that this paper is likely to be read mostly by water quality experts and not physical oceanographers, I feel strongly that the authors need to include a brief (plain language… not physical oceanographer-speak) description of modeled plume verification, especially considering that the field sampling and interpretation of results is so strongly dependent on accurate quantification of plume evolution.

2. The authors speculate that nutrients were likely a driver of the abnormally high K. brevis abundances during the study. Is there any way to determine if the abundances observed in Tampa Bay were a result of growth vs. a physical concentration mechanism? This seems like it would be a good first step for giving the reader some confidence that nutrients from Piney Point played a role. For example, if you know the initial abundance coming into Tampa Bay via the inlet, can the plume model be used to estimate transit time to whatever location/time point maximum abundances were observed, and from that, can you use known growth rates to determine if those abundances are logical based on growth alone vs physical concentration?

3. The authors argue in several locations that elevated salinities are conducive to K. brevis growth. Need references to support this.

4. The discussion is awkwardly laid out, e.g., the first paragraph of "Additional interpretation of impacts" would seemingly belong in the previous section on "Nutrient cycling".

5. In the discussion section on seagrass impacts, there is a sentence that is confusing and possibly contradictory: "The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses."

# References

### 2022-03-09 R/figures.R

@@ -266,7 +266,7 @@ jpeg(here('figs/map.jpeg'), height = 4.2, width = 9, family = 'serif', units = '
 print(pout)
 dev.off()
 
-jpeg(here('revision/Fig1.jpeg'), height = 4.2, width = 9, family = 'serif', units = 'in', res = 500)
+jpeg(here('revision2/Fig1.jpeg'), height = 4.2, width = 9, family = 'serif', units = 'in', res = 500)
 print(pout)
 dev.off()
 
@@ -407,7 +407,7 @@ jpeg(here('figs/wqmap.jpeg'), height = 7, width = 12, units = 'in', res = 500, f
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig2.jpeg'), height = 7, width = 12, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision2/Fig2.jpeg'), height = 7, width = 12, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -441,7 +441,7 @@ jpeg(here('figs/wqgam.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, 
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig3.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision2/Fig3.jpeg'), height = 6, width = 8.5, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -480,7 +480,7 @@ jpeg(here('figs/trnex.jpeg'), height = 7, width = 8, units = 'in', res = 500, fa
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig4.jpeg'), height = 7, width = 8, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision2/Fig4.jpeg'), height = 7, width = 8, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
@@ -612,7 +612,7 @@ jpeg(here('figs/trnfrq.jpeg'), height = 6, width = 9, units = 'in', res = 500, f
 print(p)
 dev.off()
 
-jpeg(here('revision/Fig5.jpeg'), height = 6, width = 9, units = 'in', res = 500, family = 'serif')
+jpeg(here('revision2/Fig5.jpeg'), height = 6, width = 9, units = 'in', res = 500, family = 'serif')
 print(p)
 dev.off()
 
[…]

### 2022-03-09 revision2/review_response.Rmd
review one revision done

@@
 
-We sincerely thank the reviewers again for providing comments on our manuscript.  As before, we provide a point-by-point response to these comments below.
-
+We sincerely thank the reviewers again for providing comments on our manuscript.
+As before, we provide a point-by-point response to these comments below.
 All page and line numbers in the reviewer comments refer to the revised manuscript from the previous submission.
@@
 
+* __Response__: Thank you for your comments and we are pleased to hear that the draft has improved. We have addressed the remaining concerns below.
+
 Title- no one other than local people will know what a Piney Point is.
 Needs revision to something along the lines of: Initial estuarine response to a spill of phosphate mining waste: time course of a red tide following  the Piney Point spill in Tampa Bay, USA
+
+* __Response__: Agreed, the title has been changed to "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida".
+However, we have aded "Piney Point" to the key words to assist with article indexing.
 
@@
 
+* __Response__: The graphical abstract has again been revised for simplicity.  We have enlarged the icons, removed unnecessary icons, and simplified the text.
+
 Figure 8d looks like K brevis occurred on only 1 day or 2-- odd.
+
+* __Response__: Each point on the x-axis is one week.
+However, the upper limit of the y-axis was truncated to better emphasize the date ranges within which *K.
+brevis* was observed.
+This is also noted in the figure caption.
 

### 2022-03-09 manu-draft.Rmd
review one revision done

@@
 ---
-title: "Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida"
+title: "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida"
 output: 
@@
 
-*Key words*: eutrophication, macroalgae, nitrogen, phosphate mining, seagrass, Tampa Bay
+*Key words*: macroalgae, nitrogen, phosphate mining, Piney Point, seagrass, Tampa Bay
 
@@
 brevis* peaking in early July.
+The upper limit of the y-axis on (e) is truncated to emphasize trends.
 Quantitative cell counts for diatoms are missing for several weeks, but see Figure S6 for frequency occurrence estimates across all dates.

### 2022-03-09 supplement.Rmd
rev4 comments addressed
reproc figs to correct order for revision folder
review one revision done

@@
 ---
-title: "Supplement to Initial estuarine response to the nutrient-rich Piney Point release into Tampa Bay, Florida"
+title: "Supplement to Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida"
 output: 
@@
 Diatoms are based on presence/absence of *Asterionellopsis sp*.
-and *Skeletonema sp*.
-Note that frequency occurrence is not a precise measure of cell concentrations.'}
+and *Skeletonema sp*.'}
 knitr::include_graphics('figs/phymcrfoest-supp.jpeg')

### 2022-03-09 R/figures.R

@@ -1226,6 +1226,7 @@ p5 <- ggplot(toplo2, aes(x = week, y = medv)) +
   scale_y_continuous(labels = function(x) as.numeric(format(x, scientific = FALSE))) +
   scale_x_date(breaks = brks, date_labels = '%b %d', limits = range(brks), expand = c(0.01, 0.01)) +
   # geom_hline(aes(yintercept = 1e5, color = 'Bloom\nconcentration')) +
+  coord_cartesian(ylim = c(0, 5e6)) +
   theme_minimal() +
   theme(
     axis.text.x = element_text(angle = 45, size = 8, hjust = 1),

### 2022-03-09 revision2/review_response.Rmd
working on reviewer two comments

@@
 
+     * __Response__: We have added citations to the discussion, second to last paragraph: "Severe *K.
+brevis* blooms are rarer in estuaries because high abundances are most common at higher salinities typical of coastal or oceanic waters [@Steidinger98;@Villac20]."
+
 4. The discussion is awkwardly laid out, e.g., the first paragraph of "Additional interpretation of impacts" would seemingly belong in the previous section on "Nutrient cycling".
@@
 
+     * __Response__: This sentence was revised as "Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chlorophyll concentrations, low secchi observations), although the conditions likely did not persist long enough to impact seagrasses."
+
 # References

### 2022-03-09 manu-draft.Rmd
working on reviewer two comments

@@
 Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein].
-The results observed in 2021 suggested water quality conditions were not supportive of seagrass growth, although changes were not observed and the conditions likely did not persist long enough to impact seagrasses.
+Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chlorophyll concentrations, low secchi observations), although the conditions likely did not persist long enough to impact seagrasses.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 Seasonal persistence in Gulf waters in southwest Florida can vary between years, with some blooms lasting as short as a few weeks, while others have been present for longer than a year [the 2018 bloom lasted sixteen months, @Skripnikov21].
-Severity in estuaries is typically less than Gulf waters as *K.
-brevis* is limited to higher salinity.
+Severe *K.
+brevis* blooms are rarer in estuaries because high abundances are most common at higher salinities typical of coastal or oceanic waters [@Steidinger98;@Villac20].
 Contributing factors in 2021, such as low rainfall preceding the bloom and varying wind patterns, created conditions that were favorable for growth of *K.

### 2022-03-12 revision2/review_response.Rmd
reviewer two revision almost done

@@
 link-citations: true
+csl: ../marine-pollution-bulletin.csl
 bibliography: ../refs.bib
@@
 
+* __Response__: Thank you again for your comments. We have made efforts to address your final concerns as noted below.
+
 1.
@@
 
+     * __Response__: These are all valid points when considering the potential mechanisms driving the observed red tide in 2021.
+However, without additional data on timing and locations of cell concentrations and substantial modelling efforts, we cannot fully address this question.
+We do note in several locations in the discussion that the bloom did originate in the Gulf of Mexico, yet the availability of nutrients from Piney Point very likely increased the severity of the bloom in July.
+We provide some additional citations in the discussion to make this point more clearly [@Medina20;@Medina22].
+
 3. The authors argue in several locations that elevated salinities are conducive to K. brevis growth. Need references to support this.
@@
 
+     * __Response__: This paragraph was moved to the preceding section.
+
 5.

### 2022-03-13 revision2/review_response.Rmd
revisions for reviewer 2 done

@@
 
+     * __Response__: The Tampa Bay Coastal Ocean Model was previously tested for sea level and velocity field veracity under extreme conditions, e.g., Hurricane Irma [@Chen18] and under more regular variations in tide, wind and river forcing by @Zhu15 and @Chen19.
+Model results closely followed observations over the time scale associated with tides, wind forcing and the longer term averaged net estuarine circulation.
+It was with this backdrop that we felt justified in adding a tracer component to examine how the Piney Point effluent would spread throughout Tampa Bay.
+Two additional tests were then performed.
+The first was to demonstrate that the model faithfully conserved the amount of tracer.
+Thus we plotted what was known to be released and compared this with the integrated amount of tracer within the Tampa Bay Coastal Ocean model simulation.
+Both were identical.
+Next we compared patterns of modeled tracer distribution with ocean color observations by satellite remote sensing.
+Again, the agreements were very good.
+In essence, if a model is known to faithfully reproduce the three-dimensional circulation features, then it should perform similarly in tracking a passive scalar because the scalar is transported and spread by the circulation.
+A dedicated paper of the modeling work (including model/data comparisons) is still in preparation.
+More details will be reported in that paper.
+Preliminary results were reported at a recent conference (see Weisberg 2022, [BASIS7](https://tbep.org/estuary/basis/basis7/)).
+
 2.

### 2022-03-13 revision2/review_response.Rmd
final formatting done before resubmission

@@
 
-We sincerely thank the reviewers again for providing comments on our manuscript.
-As before, we provide a point-by-point response to these comments below.
-All page and line numbers in the reviewer comments refer to the revised manuscript from the previous submission.
+We sincerely thank the reviewers again for providing comments on our manuscript.  As before, we provide a point-by-point response to these comments below.
 
@@
 * __Response__: Agreed, the title has been changed to "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida".
-However, we have aded "Piney Point" to the key words to assist with article indexing.
+However, we have added "Piney Point" to the key words to assist with article indexing.
 
@@
 The first was to demonstrate that the model faithfully conserved the amount of tracer.
-Thus we plotted what was known to be released and compared this with the integrated amount of tracer within the Tampa Bay Coastal Ocean model simulation.
+Thus, we plotted what was known to be released and compared this with the integrated amount of tracer within the Tampa Bay Coastal Ocean model simulation.
 Both were identical.
-Next we compared patterns of modeled tracer distribution with ocean color observations by satellite remote sensing.
+Next, we compared patterns of modeled tracer distribution with ocean color observations by satellite remote sensing.
 Again, the agreements were very good.

### 2022-03-14 manu-draft.Rmd
reviewer two revision almost done
CH added as co-author, CH comments added

@@
 Burke (mburke@tbep.org), Jing Chen (jchen15@usf.edu), Diana W.
-Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Katherine A.
+Chin (diwchin@gmail.com), Jayne Gardiner (jgardiner@ncf.edu), Chuanmin Hu (huc@usf.edu), Katherine A.
 Hubbard (katherine.hubbard@myfwc.com), Yonggang Liu (yliu@usf.edu), Cary Lopez (cary.lopez@myfwc.com), Miles Medina (miles.medina@ufl.edu), Elise Morrison (elise.morrison@essie.ufl.edu), Edward J.
@@
 At that time, approximately 1.8 billion liters of mixed legacy phosphate mining wastewater and seawater from port dredging operations were being held in the failing gypstack.
-Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of nitrogen, in addition to phosphorus.
-Water quality parameters of NGS-S measured in 2019 showed total phosphorus (160 mg/L) and total nitrogen (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay.
+Piney Point historically produced Diammonium Phosphate ((NH$_4$)$_2$HPO$_4$) and the remnant stackwater has very high concentrations of total nitrogen (TN), in addition to total phosphorus (TP).
+Water quality parameters of NGS-S measured in 2019 showed TP (160 mg/L) and TN (230 mg/L) were approximately three orders of magnitude higher than typical concentrations in Tampa Bay.
 From March 30th to April 9th, approximately 814 million liters (215 million gallons) of stack water were released to lower Tampa Bay following an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) authorized by the Florida Department of Environmental Protection (FDEP).
@@
 
-Water quality parameters included discrete, laboratory-processed and *in situ* samples for total nitrogen (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L, hereafter referred to as ammonia), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), total phosphorus (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a ($\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
+Water quality parameters included discrete, laboratory-processed and *in situ* samples for TN (mg/L), total ammonia nitrogen (NH$_3$ + NH$_4^+$, mg/L, hereafter referred to as ammonia), nitrate/nitrite (NO$_3^-$ + NO$_2^-$, mg/L), TP (mg/L), orthophosphate (PO$_4^{3-}$, mg/L), chlorophyll-a (chl-a, $\mu$g/L), pH, salinity (ppt), temperature ($^\circ$C), and dissolved oxygen saturation (\%).
 Most samples were surface collections by boat, with sample frequency approximately biweekly for locations around Piney Point, although effort varied by monitoring group and was more consistent during the first three months after the release.
@@
 Laboratory methods used to process samples were based on accepted procedures promoted through the Southwest Florida RAMP.
+Additionally, the Sentinel-3 satellites were used to derive chl-a maps, which were subsequently calibrated using field-measured chl-a in surface waters.
 
@@
 
-Statistical assessments were conducted only on total nitrogen (TN), chlorophyll-a (chl-a), and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
+Statistical assessments were conducted only on TN, chl-a, and Secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems.
 Spatial comparisons were based primarily on the three areas identified in Figure \@ref(fig:map)a.
@@
 For chl-a, `r chlabv[['1']]`\% of the observations from April through September were above the normal range for Area 1 located closest to the discharge point, whereas only `r chlabv[[2]]`\% and `r chlabv[[3]]`\% were above for Areas 2 (to the north) and 3 (to the south), respectively.
-Total nitrogen concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.
+TN concentrations were above the normal range for `r tnabv[['1']]`\% of observations in Area 1, whereas concentrations were above for `r tnabv[['2']]`\% of observations in Area 2 and `r tnabv[['3']]`\% in Area 3.
 Secchi observations were below the normal range for `r secblw[['1']]`\% of observations in Area 1 and for `r secblw[['2']]`\% and `r secblw[['3']]`\% of observations in Areas 2 and 3.
@@
 
-Total nitrogen, chl-a, and Secchi depth followed temporal progressions in 2021 that were distinct from long-term seasonal trends estimated from historical data (Figure \@ref(fig:wqgam)).
+TN, chl-a, and Secchi depth followed temporal progressions in 2021 that were distinct from long-term seasonal trends estimated from historical data (Figure \@ref(fig:wqgam)).
 For Area 1, TN and chl-a concentrations were frequently above normal ranges during April.
-Chlorophyll-a concentrations were observed in excess of 50 $\mu$g/L, although median concentrations for each week in April were less than 10 $\mu$g/L.
-The initial chlorophyll peak was associated with a localized phytoplankton bloom generally dominated by diatoms.
+Chl-a concentrations were observed in excess of 50 $\mu$g/L, although median concentrations for each week in April were less than 10 $\mu$g/L.
+The initial chl-a peak was associated with a localized phytoplankton bloom generally dominated by diatoms.
 The initial diatom bloom did not persist past April.
-Chlorophyll concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with an increase in *K.
+Chl-a concentrations decreased slightly until June and July when values increased again above the seasonal expectation, coincident with an increase in *K.
 brevis* concentrations to bloom levels.
@@
 Observations in Areas 2 and 3 were more often within the normal seasonal range, with some exceptions for TN and chl-a in Area 3 in April, May, and July.
+These field-based observations were in line with remotely-estimated chl-a using satellite observations.
+These observations showed an initial bloom on April 5, which peaked on April 9 with a bloom area of about 25 km$^2$ (about 10 km alongshore and 2.5 km cross-shore) in Area 1 of Figure \@ref(fig:map)a, with chl-a ranging between 5 and 40 $\mu$g/L.
+The bloom disappeared on April 12 but reappeared on April 15 at the same location, then disappeared after April 22.
+Notably, similar blooms at this location were not observed from satellite in the month of April since Sentinel-3 satellite data became available in 2016.
+Clearly, the bloom was induced by the wastewater discharge, but localized and also short lived.
 
@@
 The observed conditions in Tampa Bay in 2021 following releases from Piney Point provide multiples lines of evidence for an adverse environmental response to a large pulse of inorganic nitrogen into the system.
-Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay.
-These anomalous events (Figure \@ref(fig:timeline)) included 1) a large diatom bloom in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K.
+Collectively, these observations show that conditions in 2021 were anomalous when compared to long-term monitoring data for Tampa Bay, although some of the anomalies may not be related to the Piney Point release.
+These anomalous events (Figure \@ref(fig:timeline)) included 1) a large diatom bloom (~ 25 km$^2$, chl-a between 5 and 40 $\mu$g/L) in April in the vicinity of the release at Port Manatee, 2) high abundance of filamentous cyanobacteria in Anna Maria Sound and near Port Manatee, 3) medium to high bloom concentrations of the ride tide organism *K.
 brevis* in lower and middle Tampa Bay from June through July, and 4) high incidence of fish kill reports prompting local governments to remove over 1600 metric tons of dead fish from shoreline areas.
@@
 The events of 2021 can be considered together to develop a narrative of the temporal shift of nutrient pools between ecosystem components of the bay from April through September, starting with the influx of inorganic nitrogen from Piney Point.
-Total nitrogen concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
+TN concentrations first peaked in April (Figure \@ref(fig:nutrientflow)a), as did chl-a concentrations (Figure \@ref(fig:nutrientflow)b).
 The initial peak in water quality parameters suggested a rapid response of the phytoplankton community as an increase in diatoms (e.g., centric species, such as *Skeletonema* sp., and also *Asterionellopsis* sp., Figure \@ref(fig:nutrientflow)c) that can readily utilize inorganic forms of nitrogen that were present in the initial discharge [@Bates76;@Domingues11].
@@
 brevis*.
-For example, total nitrogen and chlorophyll are strongly aligned with the *K.
+For example, TN and chl-a are strongly aligned with the *K.
 brevis* axis as nutrients were likely available in organic form during the peak of the red tide event.
@@
 Bay conditions preceding the 2021 events, as well as the passage of tropical storm Elsa, could obscure these associations (described below).
-
-## Additional interpretation of impacts
 
@@
 
+## Additional interpretation of impacts
+
 Previous research for Tampa Bay has identified water quality conditions that are likely to promote seagrass growth [@Greening06;@Greening14, and references therein].
-Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chlorophyll concentrations, low secchi observations), although the conditions likely did not persist long enough to impact seagrasses.
+Water quality results in 2021 suggested that conditions may have been light-limiting for seagrass growth (e.g., high chl-a concentrations, low Secchi observations), although the conditions likely did not persist long enough to impact seagrasses.
 The long-term effects of the Piney Point discharge on the seagrass community remains uncertain.
@@
 Plume dispersal also suggested that both open-water and back-bay habitats were exposed to nutrient concentrations sufficient to stimulate phytoplankton production.
-Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
+Although Piney Point did not cause red tide (i.e., it originates in the Gulf of Mexico), the events of 2021 may have created conditions in Tampa Bay conducive for the extreme bloom concentrations observed in July.
+Similarly, recent studies have highlighted the role of anthropogenic forcing in increasing bloom intensity in southwest Florida [@Medina20;@Medina22].
 
@@
 The progress achieved in restoring the Tampa Bay ecosystem over recent decades would not be possible without the collaborative partnerships fostered in the region.
-Our partners’ willingness to adapt and implement innovative monitoring and management actions in response to Piney Point and the ever evolving challenges threatening Tampa Bay is greatly appreciated.
+Our partners’ willingness to adapt and implement innovative monitoring and management actions in response to Piney Point and the ever-evolving challenges threatening Tampa Bay is greatly appreciated.
 

### 2022-03-17 revision2/review_response.Rmd
additional response to reviewer 2
ES comments added to revision

@@
 
-* __Response__: Agreed, the title has been changed to "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida".
+* __Response__: Agreed, the title has been changed to "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility adjacent to Tampa Bay, Florida".
 However, we have added "Piney Point" to the key words to assist with article indexing.
@@
 More details will be reported in that paper.
-Preliminary results were reported at a recent conference (see Weisberg 2022, [BASIS7](https://tbep.org/estuary/basis/basis7/)).
+Preliminary results were reported at recent conferences:
+
+      Weisberg, R.H. (2022). Tampa Bay Coastal Ocean Model Applications, Abstract presented at the Bay Area Scientific Information Symposium, St. Petersburg, Florida, March 2022.
+
+      Liu, Y., Weisberg, R.H., Zheng, L., Sun, Y., Chen, J.
+(2021), Nowcast/Forecast of the Tampa Bay, Piney Point Effluent Plume: A Rapid Response, Abstract (OS35B-1036) presented at AGU Fall Meeting, New Orleans, Louisiana, December 2021.
 
@@
 We provide some additional citations in the discussion to make this point more clearly [@Medina20;@Medina22].
+Additional research will also be published in the near future that will describe the results of stable isotope analyses as a follow up to this paper, which is already noted in the discussion.
 

### 2022-03-17 manu-draft.Rmd
ES comments added to revision

@@
 ---
-title: "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida"
+title: "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility adjacent to Tampa Bay, Florida"
 output: 

### 2022-03-17 supplement.Rmd
CH added as co-author, CH comments added
ES comments added to revision

@@
 ---
-title: "Supplement to Initial estuarine response to inorganic nutrient inputs from a legacy mining facility to Tampa Bay, Florida"
+title: "Initial estuarine response to inorganic nutrient inputs from a legacy mining facility adjacent to Tampa Bay, Florida"
 output: 

### 2022-03-17 revision2/review_response.Rmd
second revision resubmitted

@@
 
+      Also note that we have included additional results in the text that describe satellite-based estimates of chlorophyll concentration.
+The results were consistent with model simulation results in tracking the initial phytoplankon bloom at the point of discharge (see bottom of page 12, third paragraph of initial water quality results).
+
 2.
